### 2015-02-01 augur/notes/freq_est.tex

\documentclass[11pt]{article}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage[english]{babel}
\usepackage[utf8x]{inputenc}
\usepackage{amsmath,amsfonts,amssymb,eucal,eurosym,textcomp}
\usepackage{color}
\usepackage{graphicx}
\usepackage[numbers]{natbib}
%\usepackage{lineno}
%\linenumbers

\title{Estimation of frequency trajectories}

\begin{document}
\maketitle
What we got is a series of dated observations that either belong to a particular clade or not
\begin{equation}
    (t_i, c_i) \; \mathrm{for}\; i\in 1,\ldots,n \; \mathrm{and}\; c_i\in[0,1],\; t_i = date
\end{equation}
the frequency of the clade in question is a continuous time trajectory $x(t)$. Given the 
trajectory, the likelihood of observing the data is
\begin{equation}
    P(\{(t_i, c_i)\}|x(t)) \sim \prod_{i=1}^n x(t_i)^{c_i}(1-x(t_i))^{1-c_i}
\end{equation}
We want to estimate $x(t)$, which changes over time due to selection and chance. This could 
modeled as 
\begin{equation}
    \dot{x} = x(1-x) (s-\bar{s}) + \sqrt{2Dx(1-x)}\eta + \mu(2x-1)
\end{equation}
where $\eta$ is a white noise Wiener process and $D$ is the diffusion constant (what other 
people would call the inverse effective population size) and the last term could be 
the mutational influx. The probability of a particular trajectory 
\begin{equation}
    \log P(x(t) | \{(t_i, c_i)\} ) \sim \sum_{i=1}^n c_i\log(x(t_i)) + (1-c_i)\log(1-x(t_i))
     - \int dt \frac{(\dot{x}-x(1-x)(s-\bar{s}))^2}{4Dx(1-x)}
\end{equation}
One conveniently discretizes the time integral to 
\begin{equation}
    \sum_j \Delta t \frac{(x_{j+1}-x_j-x_j(1-x_j)(s-\bar{s}))^2}{4Dx_j(1-x_j)} 
\end{equation}
while interpreting the stochastic ODE as Ito.
An alternative variable would be $z = x/(1-x)$ (such that $x = z/(1+z)$ and 
$1-x = \frac{1}{1+z}$). Ignoring the determinant for now, the posterior becomes
\begin{equation}
\begin{split}
    \log P(x(t) | \{(t_i, c_i)\} ) & \sim \sum_{i=1}^n c_i\log(z(t_i)/(1+z(t_i))) + (1-c_i)\log(1/(1+z(t_i)))\\
    &+\sum_j \Delta t (1+z_j)^2\frac{(z_{j+1}-z_j-z_j(s-\bar{s}))^2}{4Dz_j}     \\
    &=\sum_{i=1}^n c_i\log(z(t_i)) - \log(1+z(t_i))\\
    &+\sum_j \Delta t (1+z_j)^2\frac{(z_{j+1}-z_j-z_j(s-\bar{s}))^2}{4Dz_j} 
\end{split}
\end{equation}
The mix of exponential growth and diffusion means that this isn't very simple, but it 
is not too bad after all. Taking derivatives with respect to $z_k$, we find
\begin{equation}
\begin{split}
    &\sum_{i=1 | t_i \in \Delta t_k}^n \left[\frac{c_i}{z_k} - \frac{1}{1+z_k}\right] +\Delta t (1+z_k)\frac{(z_{k+1}-z_k-z_k(s-\bar{s}))^2}{2Dz_k}\\
    &-\Delta t (1+z_k)^2\frac{(z_{k+1}-z_k-z_k(s-\bar{s}))^2}{4Dz_k^2} -\Delta t (1+z_k)^2\frac{(z_{k+1}-z_k-z_k(s-\bar{s}))(s-\bar{s})}{2Dz_k}  
\end{split}
\end{equation}
Signs are mostly wrong. But I think we can solve this for the most likely trajectory -- at least approximately.

\end{document}

### 2015-02-02 augur/notes/freq_est.tex

@@
     \log P(x(t) | \{(t_i, c_i)\} ) & \sim \sum_{i=1}^n c_i\log(z(t_i)/(1+z(t_i))) + (1-c_i)\log(1/(1+z(t_i)))\\
-    &+\sum_j \Delta t (1+z_j)^2\frac{(z_{j+1}-z_j-z_j(s-\bar{s}))^2}{4Dz_j}     \\
+    &-\sum_j \Delta t (1+z_j)^2\frac{(z_{j+1}-z_j(1+s-\bar{s}))^2}{4Dz_j}     \\
     &=\sum_{i=1}^n c_i\log(z(t_i)) - \log(1+z(t_i))\\
-    &+\sum_j \Delta t (1+z_j)^2\frac{(z_{j+1}-z_j-z_j(s-\bar{s}))^2}{4Dz_j} 
+    &-\sum_j \Delta t (1+z_j)^2\frac{(z_{j+1}-z_j(1+s-\bar{s}))^2}{4Dz_j} 
 \end{split}
@@
 \begin{split}
-    &\sum_{i=1 | t_i \in \Delta t_k}^n \left[\frac{c_i}{z_k} - \frac{1}{1+z_k}\right] +\Delta t (1+z_k)\frac{(z_{k+1}-z_k-z_k(s-\bar{s}))^2}{2Dz_k}\\
-    &-\Delta t (1+z_k)^2\frac{(z_{k+1}-z_k-z_k(s-\bar{s}))^2}{4Dz_k^2} -\Delta t (1+z_k)^2\frac{(z_{k+1}-z_k-z_k(s-\bar{s}))(s-\bar{s})}{2Dz_k}  
+    &\frac{\partial}{\partial z_k}\log P(x(t) | \{(t_i, c_i)\} )  = 0 = 
+            \sum_{i=1 | t_i \in \Delta t_k}^n \left[\frac{c_i}{z_k} - \frac{1}{1+z_k}\right] \\
+           &-\Delta t (1+z_k)\frac{(z_{k+1}-z_k(1+s-\bar{s}))^2}{2Dz_k}
+            +\Delta t (1+z_k)^2\frac{(z_{k+1}-z_k(1+s-\bar{s}))^2}{4Dz_k^2}\\
+           &-\Delta t (1+z_k)^2\frac{(z_{k+1}-z_k(1+s-\bar{s}))(1+s-\bar{s})}{2Dz_k}  
+           -\Delta t (1+z_{k-1})^2\frac{(z_{k}-z_{k-1}(1+s-\bar{s}))}{2Dz_{k-1}}
 \end{split}
 \end{equation}
-Signs are mostly wrong. But I think we can solve this for the most likely trajectory -- at least approximately.
+We need to solve this for all $z_k$ simultaneously to obtain the most likely trajectory. 
+Regorganizing slightly and ignoring higher order terms, we have
+\begin{equation}
+\begin{split}
+     0  \approx  &4D \sum_{i=1 | t_i \in \Delta t_k}^n \left[\frac{c_i}{z_k} - \frac{1}{1+z_k}\right] \\
+           &-\Delta t \left[2z_k(1+z_k)+(1+z_k)^2\right](z_{k+1}/z_k-(1+s-\bar{s}))^2\\
+           &-\Delta t 2(1+z_k)^2 (z_{k+1}/z_k+z_{k}/z_{k-1}-2(1+s-\bar{s}))  
+\end{split}
+\end{equation}
+More promising, however, might be straight numerical minimization of the log-likelihood. 
+We could in principle fit $s$ for individual clades while calculating $\bar{s}$ as an average
+\begin{equation}
+    \bar{s}(t_i) = \sum_j x_j(t_i)s_j
+\end{equation}
+where $x_j$ is the frequency of clade $j$. What is not obvious to me at this point is how to
+consist the fitness of the subclades with that of the parent clade.
 

### 2015-03-19 manuscript/nextflu.tex
added manuscript files

\documentclass{bioinfo}
\copyrightyear{2015}
\pubyear{2015}
\usepackage[english]{babel}
\usepackage{amssymb,amsfonts,amsmath}
\usepackage{hyperref}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{natbib}

\newcommand{\EQ}[1]{Eq.~(\ref{eq:#1})}
\newcommand{\EQS}[2]{Eqs.~(\ref{eq:#1}) and (\ref{eq:#2})}
\newcommand{\FIG}[1]{Fig.~\ref{fig:#1}}
\newcommand{\TAB}[1]{Tab.~\ref{tab:#1}}
\newcommand{\REF}[1]{ref.~\citep{#1}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{document}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title[Tracking of seasonal influenza H3N2 virus evolution]{nextflu: Real-time tracking of seasonal influenza H3N2 virus evolution in humans}
\author{Trevor Bedford$^{1}$ and Richard~A.~Neher$^{2}$} % I don't care about the order
\address{$^{2}$Max Planck Institute for Developmental Biology, 72076 T\"ubingen,
Germany}
\history{Received on XXXXX; revised on XXXXX; accepted on XXXXX}

\editor{Associate Editor: XXXXXXX}

%\date{\today}
\maketitle
\bibliographystyle{natbib}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
\section{Motivation:}
\section{Results:}
\section{Availability:} \url{http://nextflu.org} and \url{https://github.com/blab/nextflu}.
\end{abstract}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Introduction}

\begin{itemize}
	\item Motivation
	\item data sources
	\item pipe line and design
	\item frequency estimation details
	\item genotype and trait coloring
	\item generalizability
\end{itemize}

\begin{figure}[bhtp]
	\begin{center}
%\includegraphics[width=0.99\columnwidth]{figures/}
\caption[labelInTOC]{.}
\label{fig:gspace}
\end{center}
\end{figure}

\paragraph{Funding\textcolon}This work is supported by the ERC though
Stg-260686.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\bibliography{bib}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\end{document}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2015-03-19 manuscript/nextflu.tex
took a first pass at the ms

@@
 \maketitle
-\bibliographystyle{natbib}
+\bibliographystyle{plain}
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{abstract}
-\section{Motivation:}
-\section{Results:}
+\begin{abstract} \section{Motivation:} Seasonal influenza viruses evolve
+rapidly and change their antigenic properties. This allows the virus
+population to evade immunity in their human hosts and reinfect previously
+infected individuals. Similarly, vaccines against seasonal influenza need to
+be updated frequently to protect against evolving virus population.
+
+\section{Results:} We have developed a processing pipeline and web-browser
+based visualization that allows convenient exploration and analysis of the
+most recent influenza virus sequence data. This web-application is live at
+\url{nextflu.org} and displays a phylogenetic tree that can be decorated with
+several layers of additional information such the viral genotype at specific
+sites, sampling location, and sequence or tree based statistics that have been
+shown to predictive of future virus dynamics.
+
 \section{Availability:} \url{http://nextflu.org} and \url{https://github.com/blab/nextflu}.
@@
 
+nextflu consists of a processing pipeline writtin in python and a javascript
+based visualization called auspice. As input, augur requires a fasta file of
+sequences of the HA segment with fasta labels containing relevant information
+such as the strain name, the sampling date and passage history. In a first
+step, viruses without complete date  or geographic information and viruses
+passaged in eggs are removed. In addition, local outbreaks are filtered by
+keeping only one instance of identical sequences sampled at the same location
+on the same day.  Following filtering, the viruses are subsampled to achieve
+an appoximately even temporal and global distribution (typically 50 viruses
+per month). For our standard display period of 3 years, this typically results
+in \~1800 viruses, which we align us \texttt{mafft} \citep{mafft}.
+
+Once aligned, the set of virus sequences is further cleaned by removing
+insertions relative  to the outgroup and by removing sequences that are either
+much closer or further from the outgroup then expected given their date of
+sampling (filtering viruses that don't follow an approximate molecular clock).
+REASSORTANTS.
+
+From the filtered and cleaned alignment, augur builds a phylogenetic tree using
+\texttt{fasttree} \citep{Price:2009p47657} which is further refined by one hour
+of RAxML optimization \citep{RAxml}. Next, the state of every internal node
+of the tree is inferred using marginal maximum likelihood method and missing 
+sequence is filled with the nearest ancestral sequence. Internal branches 
+without mutations are collapsed into polytomies. The final tree is decorated
+with the attributes to be displayed in the browser, see below.
+
+\begin{figure}[t!]
+	\begin{center}
+	\includegraphics[width=0.99\columnwidth]{figures/tree_screenshot}
+\caption[]{The \texttt{nextflu} website with the user interface on the left and
+the tree on the right.}
+\label{fig:tree}
+\end{center}
+\end{figure}
+
+In addition to the phylogenetic tree, augur estimates the frequency
+trajectories of mutations, genotypes and clades in the tree. Frequency
+trajectories are  represented as a linear interpolation $x(t)$ of values $x_i$
+at time points  $t_i$ spaced one month apart. The pivot frequencies
+corresponding to a feature $\phi$ (e.g. mutation) are estimated by minimizing
+\begin{equation}
+\label{eq:freq}
+	\begin{split}
+	-\log LH(\{x_i\} | \phi)  =& \sum_v I(v,\phi)\log(x(t_v)) + (1-I(v,\phi))\log(1-x(t_v)) \\
+			&+\sum_i \frac{(\Delta x_i - \epsilon\Delta x_{i-1})^2}{2\gamma \delta t_i}
+\end{split}
+\end{equation}
+where $I(v,\phi)=1$ if virus $v$ carries attribute $\phi$ and $I(v,\phi)=0$
+otherwise. $\Delta x_i = x_i-x_{i-1}$, $\Delta t_i = t_i-t_{i-1}$, $\gamma$
+parameterizes the smoothing imposed on the frequency estimate (related to the
+frequency changes by genetic drift) and $\epsilon$ parameterizes our
+expectation of frequency changes in interval $i$ based on the observed change
+in interval $i-1$. $\epsilon=0$ implies that frequency changes are
+uncorrelated from one interval to the next, while $\epsilon=1$ implies that
+the most likely $\Delta x_i$ is $\Delta x_{i-1}$. We typically use $\epsilon =
+0.7$. \EQ{freq} is minimized using SciPy's implementation of a downhill
+simplex algorithm \citep{Oliphant:2007p25672}. To speed up convergence,
+minization is first done using a coarse grid of pivot points which is
+subsequently refined to 12 pivots per year. PUT IN DETAILS ON HOW WE ESTIMATE 
+FREQUENCIES ON THE TREE?
+
+The tree decorated with attributes and clade frequencies is written to file in
+a nested json format for display in the browser by auspice. In addition to the
+tree, json files containing the sequences, frequencies of mutations and
+genotypes, as well as meta information are exported.
+
+The tree is visualized by auspice using d3 \citep{d3}. The user can explore
+the data interactively by selecting viruses from different dates (moving the
+date slider, top right corner in \FIG{tree}) or by coloring the tree by the
+following attributes:
+\begin{itemize}
+	\item epitope mutations: each tip is colored by the number of aminoacid mutations at \~50 
+		  positions relative to the root of the tree. These positions are known to
+		  be frequent targets of antibodies \citep{shih_simultaneous_2007} and have been
+		  suggested to be predictive of future success \citep{luksza_predictive_2014}.
+	\item non-epitope mutations: color by the number of mutations outside the above
+		  mentioned epitope sites. Mutations outside epitopes tend to be damaging and 
+		  anti-correlated with clade expansion \citep{luksza_predictive_2014}.
+    \item receptor binding mutations: color by the number of mutations at 7 positions
+    	  close to the receptor binding site that have been shown to be responsible 
+    	  for major antigenic positions in the past decades \citep{koel_substitutions_2013}.
+    \item local branching index is the exponentially weighted tree length surrounding 
+    	  a node, which is associated with rapid branching and expansion of clades 
+    	  \citep{neher_predicting_2014}. It is recalculated everytime the date slider is
+    	  moved and based only on the highlighted viruses.
+    \item geographic location
+    \item HA1 genotype: entering amino acid positions (separated by a comma) in the 
+    	  text field will color every node according to the genotype at these positions.
+\end{itemize}
+
+The frequency plot below the tree (see \FIG{freq}) displays the frequency trajectory
+of clades in the tree whenever the mouse hovers above the branch defining the clade. 
+Furthermore, trajectories of individual mutations, combinations of two mutations, and 
+predifined clades such as 3c3.a can be plotted by entering them in text field
 \begin{figure}[bhtp]
 	\begin{center}
-%\includegraphics[width=0.99\columnwidth]{figures/}
-\caption[labelInTOC]{.}
-\label{fig:gspace}
+	\includegraphics[width=0.99\columnwidth]{figures/frequencies}
+\caption[]{The frequency diagram on \url{nextflu.org} allows geography specific
+		plotting of frequencies of individual mutations, pairs of mutations, 
+		predefined genotypes, as well as clades in the tree. }
+\label{fig:freq}
 \end{center}
 \end{figure}
+
+\section{Conclusion}
+We built nextflu.org to facilitate the analysis and exploration of seasonal influenza
+sequence data collected by laboratories around to world. By using the most recent data
+and integrating phylogenies with frequency trajectories and sequence or tree based
+predictors of successful clades, we hope that nextflu can inform the choice of  
+strains used in seasonal influenza vaccines.
+
+nextflu was designed to be readily adapted to other rapidly evolving viruses. 
+To that end, the generic classes used for processing the sequence data to include features 
+specific to the virus in question.
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\bibliography{bib}
+\bibliography{nextflu}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2015-04-01 augur/src/H3N2_process.py
streamlined the skipping. now allows multiple skips and works like this: --skip genotypes export added genotype_frequencies step, can be separately skipped unified min_freq threshold used for variable sites and the threshold used for frequency estimation

@@ -264,7 +264,9 @@ class H3N2_process(process, H3N2_filter, H3N2_clean, H3N2_refine):
 		if 'frequencies' in steps:
 			print "--- Estimating frequencies at " + time.strftime("%H:%M:%S") + " ---"
 			self.determine_variable_positions()
-			self.estimate_frequencies()
+			self.estimate_frequencies(tasks = ["mutations", "clades", "tree"])
+			if 'genotype_frequencies' in steps: 
+					self.estimate_frequencies(tasks = ["genotypes"])
 			self.dump()
 		if 'export' in steps:
 			self.temporal_regional_statistics()
@@ -273,7 +275,7 @@ class H3N2_process(process, H3N2_filter, H3N2_clean, H3N2_refine):
 			                       annotations = ['3c3.a', '3c2.a'])
 
 if __name__=="__main__":
-	all_steps = ['filter', 'align', 'clean', 'tree', 'ancestral', 'refine', 'frequencies', 'export']
+	all_steps = ['filter', 'align', 'clean', 'tree', 'ancestral', 'refine', 'frequencies','genotype_frequencies', 'export']
 	parser = argparse.ArgumentParser(description='Process virus sequences, build tree, and prepare of web visualization')
 	parser.add_argument('-y', '--years_back', type = int, default=3, help='number of past years to sample sequences from')
 	parser.add_argument('-v', '--viruses_per_month', type = int, default = 50, help='number of viruses sampled per month')
@@ -283,7 +285,7 @@ if __name__=="__main__":
 	parser.add_argument('--test', default = False, action="store_true",  help ="don't run the pipeline")
 	parser.add_argument('--start', default = 'filter', type = str,  help ="start pipeline at specified step")
 	parser.add_argument('--stop', default = 'export', type=str,  help ="run to end")
-	parser.add_argument('--skip_frequencies', default = False, action="store_true",  help ="don't run frequency estimation")	
+	parser.add_argument('--skip', nargs='+', type = str,  help ="analysis steps to skip")	
 	params = parser.parse_args()
 	lt = time.localtime()
 	num_date = round(lt.tm_year+(lt.tm_yday-1.0)/365.0,2)
@@ -293,8 +295,11 @@ if __name__=="__main__":
 	dt= params.time_interval[1]-params.time_interval[0]
 	params.pivots_per_year = 12.0 if dt<5 else 6.0 if dt<10 else 3.0
[…]

### 2015-04-01 augur/src/bernoulli_frequency.py

@@ -286,8 +286,8 @@ class virus_frequencies(object):
 			for pos2 in relevant_pos[i1+1:]:
 				for ai1, aa1 in enumerate(self.aa_alphabet):
 					for ai2, aa2 in enumerate(self.aa_alphabet):
-						if self.aa_frequencies[ai1,pos1]>0.3*threshold \
-						and self.aa_frequencies[ai2,pos2]>0.3*threshold:
+						if self.aa_frequencies[ai1,pos1]>threshold \
+						and self.aa_frequencies[ai2,pos2]>threshold:
 							gt = [(pos1,aa1),(pos2,aa2)]
 							if self.verbose: print "estimating freq of ", gt
 							freq, (tps, obs) = self.estimate_genotype_frequency(sub_aln, gt, min_observations = 10)

### 2015-04-01 augur/src/process.py

@@ -14,12 +14,14 @@ class process(virus_frequencies):
 	def __init__(self, prefix = 'data/', time_interval = (2012.0, 2015.0), 
 	             auspice_frequency_fname ='../auspice/data/frequencies.json', 
 				 auspice_sequences_fname='../auspice/data/sequences.json', 
-				 auspice_tree_fname='../auspice/data/tree.json', min_freq = 0.01, **kwargs):
+				 auspice_tree_fname='../auspice/data/tree.json', 
+				 min_mutation_frequency = 0.01, min_genotype_frequency = 0.1, **kwargs):
 		self.tree_fname = prefix+'tree.pkl'
 		self.virus_fname = prefix+'virus.pkl'
 		self.frequency_fname = prefix+'frequencies.pkl'
 		self.aa_seq_fname = prefix+'aa_seq.pkl'
-		self.min_freq = min_freq
+		self.min_mutation_frequency = min_mutation_frequency
+		self.min_genotype_frequency = min_genotype_frequency
 		self.time_interval = tuple(time_interval)
 
 		self.auspice_tree_fname = auspice_tree_fname
@@ -245,7 +247,7 @@ class process(virus_frequencies):
 	def determine_variable_positions(self):
 		'''
 		calculates nucleoties_frequencies and aa_frequencies at each position of the alignment
-		also computes consensus sequences and position at which the major allele is at less than 1-min_freq
+		also computes consensus sequences and position at which the major allele is at less than 1-min_mutation_frequency
 		results are stored as
 		self.nucleoties_frequencies
 		self.aa_frequencies
@@ -257,7 +259,7 @@ class process(virus_frequencies):
 		for ni,nuc in enumerate(self.nuc_alphabet):
 			self.nucleoties_frequencies[ni,:]=(aln_array==nuc).mean(axis=0)
 
-		self.variable_nucleotides = np.where(np.max(self.nucleoties_frequencies,axis=0)<1.0-self.min_freq)[0]
+		self.variable_nucleotides = np.where(np.max(self.nucleoties_frequencies,axis=0)<1.0-self.min_mutation_frequency)[0]
 		self.consensus_nucleotides = "".join(np.fromstring(self.nuc_alphabet, 'S1')[np.argmax(self.nucleoties_frequencies,axis=0)])
 
 		if hasattr(self, 'aa_aln'):
[…]

### 2015-04-01 auspice/js/auspice.js
adjusted color scale for delta frequencies to +-1/year, normalized to the time difference between pivots

@@ -158,7 +158,7 @@ function calcDfreq(node, freq_ii){
 		for (var i1=0; i1<node.children.length; i1++) {
 			if (node.children[i1].freq["global"] != "undefined"){
 				var tmp_freq = node.children[i1].freq["global"]
-				node.children[i1].dfreq = 0.5*(tmp_freq[freq_ii] - tmp_freq[freq_ii-1])/(tmp_freq[freq_ii] + tmp_freq[freq_ii-1] + 0.05);
+				node.children[i1].dfreq = 0.5*(tmp_freq[freq_ii] - tmp_freq[freq_ii-1])/(tmp_freq[freq_ii] + tmp_freq[freq_ii-1] + 0.1);
 			}else{
 				node.children[i1].dfreq = node.dfreq;
 			}
@@ -361,8 +361,12 @@ d3.json("data/tree.json", function(error, root) {
 	var nodes = tree.nodes(root),
 		links = tree.links(nodes);
 	var tree_legend;
-
 	var rootNode = nodes[0];
+	if (typeof rootNode.pivots != "undefined"){
+		var dt = rootNode.pivots[1]-rootNode.pivots[0];		
+	}else{
+		var dt = 1.0/12;
+	}
 	var tips = gatherTips(rootNode, []);
 	var internals = gatherInternals(rootNode, []);
 	calcBranchLength(rootNode);
@@ -444,7 +448,7 @@ d3.json("data/tree.json", function(error, root) {
 		.range(colors);
 
 	var dfreqColorScale = d3.scale.linear()
-		.domain([-0.2, -0.15, -0.1,-0.05, 0.0, 0.05,  0.1, 0.15, 0.2])
+		.domain(([-1.0, -0.8, -0.6,-0.4, -0.2, 0, 0.2, 0.4, 0.6, 0.8]).map(function(d){return Math.round(d*dt*100)/100;}))
 		.range(colors);
 
 	var colorScale;

### 2015-04-01 auspice/js/auspice.js
added legend explanation for frequency change

@@ -616,7 +616,10 @@ d3.json("data/tree.json", function(error, root) {
     		}
    			if (colorBy == "genotype") {
     			return "Genotype"
-    		}        			
+    		}
+   			if (colorBy == "dfreq") {
+    			return "Frequency change (per "+Math.round(12*dt)+" month)";
+    		}
     	});
     
 		var tmp_leg = legend.selectAll(".legend")

### 2015-04-01 auspice/js/auspice.js
made the interval over which delta frequency is calculated a variable

@@ -153,12 +153,13 @@ function calcLBI(node, allnodes){
 /**
  * for each node, calculate the derivative of the frequency tranjectory. if none exists, copy parent
 **/
+var dfreq_dn = 2;
 function calcDfreq(node, freq_ii){
 	if (typeof node.children != "undefined") {
 		for (var i1=0; i1<node.children.length; i1++) {
 			if (node.children[i1].freq["global"] != "undefined"){
 				var tmp_freq = node.children[i1].freq["global"]
-				node.children[i1].dfreq = 0.5*(tmp_freq[freq_ii] - tmp_freq[freq_ii-1])/(tmp_freq[freq_ii] + tmp_freq[freq_ii-1] + 0.1);
+				node.children[i1].dfreq = 0.5*(tmp_freq[freq_ii] - tmp_freq[freq_ii-dfreq_dn])/(tmp_freq[freq_ii] + tmp_freq[freq_ii-dfreq_dn] + 0.1);
 			}else{
 				node.children[i1].dfreq = node.dfreq;
 			}
@@ -448,7 +449,7 @@ d3.json("data/tree.json", function(error, root) {
 		.range(colors);
 
 	var dfreqColorScale = d3.scale.linear()
-		.domain(([-1.0, -0.8, -0.6,-0.4, -0.2, 0, 0.2, 0.4, 0.6, 0.8]).map(function(d){return Math.round(d*dt*100)/100;}))
+		.domain(([-1.0, -0.8, -0.6,-0.4, -0.2, 0, 0.2, 0.4, 0.6, 0.8]).map(function(d){return Math.round(d*dt*dfreq_dn*100)/100;}))
 		.range(colors);
 
 	var colorScale;
@@ -618,7 +619,7 @@ d3.json("data/tree.json", function(error, root) {
     			return "Genotype"
     		}
    			if (colorBy == "dfreq") {
-    			return "Frequency change (per "+Math.round(12*dt)+" month)";
+    			return "Frequency change (per "+Math.round(12*dfreq_dn*dt)+" month)";
     		}
     	});
     
@@ -816,7 +817,7 @@ d3.json("data/tree.json", function(error, root) {
 		var num_date = globalDate/1000/3600/24/365.25+1970;	
 		for (var ii=0; ii<rootNode.pivots.length-1; ii++){
 			if (rootNode.pivots[ii]<num_date && rootNode.pivots[ii+1]>=num_date){
-				freq_ii=ii+1;
[…]

### 2015-04-01 auspice/js/auspice.js
changed virus tool tip slightly to improve display

@@ -273,11 +273,12 @@ var virusTooltip = d3.tip()
 		if (typeof d.date != "undefined") {
 			string += "<br>Date: " + d.date;
 		}
-		if (typeof d.isolate_id != "undefined") {
-			string += "<br>Isolate ID: " + d.isolate_id;
+		if ((typeof d.db != "undefined") && (typeof d.accession != "undefined") && (typeof d.db=='GISAID')) {
+			string += "<br>GISAID ID: EPI" + d.accession;
 		}
 		if (typeof d.lab != "undefined") {
-			string += "<br>Orig. lab: " + d.lab;
+			string += "<br>Orig. lab: " + d.lab.substring(0,30);
+			if (d.lab.substring.length>30) string+='...';
 		}
 		if (typeof d.region != "undefined") {
 			string += "<br>Region: " + d.region.replace(/([A-Z])/g, ' $1');

### 2015-04-01 augur/src/H3N2_process.py
pulled the base argument parser and virus_config into process.py. additional arguments can be added at will fixed small mistake in auspice lab display in tool tip added .replace(' ', '') in strain name fixing, since this got cut from the fasta parsing

@@ -2,7 +2,7 @@ import time, argparse,re,os
 from virus_filter import flu_filter
 from virus_clean import virus_clean
 from tree_refine import tree_refine
-from process import process
+from process import process, virus_config
 from Bio import SeqIO
 from Bio.Seq import Seq
 from Bio.Align import MultipleSeqAlignment
@@ -10,36 +10,25 @@ import numpy as np
 from itertools import izip
 
 epitope_mask = np.fromstring("0000000000000000000000000000000000000000000011111011011001010011000100000001001011110011100110101000001100000100000001000110101011111101011010111110001010011111000101011011111111010010001111101110111001010001110011111111000000111110000000101010101110000000000011100100000001011011100000000000001001011000110111111000000000000000000000000000000000000000000000000000000000000000000000000000000000000000000000000000000000000000000000000000000000000000000000000000000000000000000000000000000000000000000000000000000000000000000000000000000000000000000000", dtype='S1')
-	
-virus_config = {
+receptor_binding_sites = [144, 154, 155, 157, 158, 188, 192]
+
+
+virus_config.update({
 	# data source and sequence parsing/cleaning/processing
 	'virus':'H3N2',
 	'alignment_file':'data/gisaid_epiflu_sequence.fasta',
-	'fasta_fields':{0:'strain', 1:'isolate_id', 3:'passage', 5:'date', 7:'lab', 8:"accession"},
 	'outgroup':'A/Beijing/32/1992',
 	#'force_include':'source-data/HI_strains.txt',
 	'force_include_all':False,
 	'max_global':True,   # sample as evenly as possible from different geographic regions 
 	'cds':[48,None], # define the HA1 start i n 0 numbering
-	'n_iqd':3,     # standard deviations from clock
-
-	# frequency estimation parameters
-	'aggregate_regions': [  ("global", None), ("NA", ["NorthAmerica"]), ("EU", ["Europe"]), 
-							("AS", ["China", "SoutheastAsia", "JapanKorea"]), ("OC", ["Oceania"]) ],
-	'frequency_stiffness':10.0,
-	'min_freq':0.01,
 	# define relevant clades in canonical HA1 numbering (+1)
 	'clade_designations': { "3c3.a":[(128,'A'), (142,'G'), (159,'S')],
 						   "3c3":  [(128,'A'), (142,'G'), (159,'F')],
 						   "3c2.a":[(144,'S'), (159,'Y'), (225,'D'), (311,'H'),(489,'N')],
[…]

### 2015-04-01 augur/src/process.py

@@ -9,6 +9,34 @@ from bernoulli_frequency import virus_frequencies
 from tree_util import delimit_newick
 import numpy as np
 
+parser = argparse.ArgumentParser(description='Process virus sequences, build tree, and prepare of web visualization')
+parser.add_argument('-y', '--years_back', type = int, default=3, help='number of past years to sample sequences from')
+parser.add_argument('-v', '--viruses_per_month', type = int, default = 50, help='number of viruses sampled per month')
+parser.add_argument('-r', '--raxml_time_limit', type = float, default = 1.0, help='number of hours raxml is run')
+parser.add_argument('--interval', nargs = '+', type = float, default = None, help='interval from which to pull sequences')
+parser.add_argument('--prefix', type = str, default = 'data/', help='path+prefix of file dumps')
+parser.add_argument('--test', default = False, action="store_true",  help ="don't run the pipeline")
+parser.add_argument('--start', default = 'filter', type = str,  help ="start pipeline at specified step")
+parser.add_argument('--stop', default = 'export', type=str,  help ="run to end")
+parser.add_argument('--skip', nargs='+', type = str,  help ="analysis steps to skip")	
+
+
+virus_config = {
+	'fasta_fields':{0:'strain', 1:'isolate_id', 3:'passage', 5:'date', 7:'lab', 8:"accession"},
+	# frequency estimation parameters
+	'aggregate_regions': [  ("global", None), ("NA", ["NorthAmerica"]), ("EU", ["Europe"]), 
+							("AS", ["China", "SoutheastAsia", "JapanKorea"]), ("OC", ["Oceania"]) ],
+	'frequency_stiffness':10.0,
+	'verbose':2, 
+	'tol':1e-4, #tolerance for frequency optimization
+	'pc':1e-3, #pseudocount for frequencies 
+	'extra_pivots': 6,  # number of pivot point for or after the last observations of a mutations
+	'inertia':0.7,		# fraction of frequency change carry over in the stiffness term
+	'n_iqd':3,     # standard deviations from clock
+}
+
+
+
 class process(virus_frequencies):
 	"""generic template class for processing virus sequences into trees"""
 	def __init__(self, prefix = 'data/', time_interval = (2012.0, 2015.0), 

### 2015-04-01 augur/src/virus_filter.py

@@ -213,7 +213,7 @@ class flu_filter(virus_filter):
 
 	def fix_strain_names(self):
 		for v in self.viruses:
-			v['strain'] = v['strain'].replace('\'','').replace('(','').replace(')','').replace('H3N2','').replace('Human','').replace('human','').replace('//','/')
+			v['strain'] = v['strain'].replace(' ', '').replace('\'','').replace('(','').replace(')','').replace('H3N2','').replace('Human','').replace('human','').replace('//','/')
 
 	def filter_passage(self):
 		self.viruses = filter(lambda v: re.match(r'^E\d+', v.get('passage',''), re.I) == None, self.viruses)

### 2015-04-01 manuscript/nextflu.tex
Add affiliation and revise abstract.

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\title[Tracking of seasonal influenza H3N2 virus evolution]{nextflu: Real-time tracking of seasonal influenza H3N2 virus evolution in humans}
-\author{Trevor Bedford$^{1}$ and Richard~A.~Neher$^{2}$} % I don't care about the order
-\address{$^{2}$Max Planck Institute for Developmental Biology, 72076 T\"ubingen,
-Germany}
+\title[Tracking of seasonal influenza H3N2 virus evolution]{nextflu: Real-time 
+tracking of seasonal influenza H3N2 virus evolution in humans}
+\author{Richard~A.~Neher$^{1}$ and Trevor Bedford$^{2}$}
+\address{$^{1}$Max Planck Institute for Developmental Biology, 72076 T\"ubingen, 
+Germany, and $^{2}$Vaccine and Infectious Disease Division, Fred Hutchinson
+Cancer Research Center, Seattle, WA 98109, USA}
 \history{Received on XXXXX; revised on XXXXX; accepted on XXXXX}
@@
 infected individuals. Similarly, vaccines against seasonal influenza need to
-be updated frequently to protect against evolving virus population.
+be updated frequently to protect against an evolving virus population.
 
@@
 \url{nextflu.org} and displays a phylogenetic tree that can be decorated with
-several layers of additional information such the viral genotype at specific
+additional information such the viral genotype at specific
 sites, sampling location, and sequence or tree based statistics that have been
-shown to predictive of future virus dynamics.
+shown to predictive of future virus dynamics. Additionally, mutation frequency
+trajectories are calculated and displayed.
 

### 2015-04-01 manuscript/nextflu.tex
Switch to one line per sentence.
I'm sorry if this makes you crazy, but one-line-per-sentence really
helps reduce merge conflicts when collaborating via TeX.

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\title[Tracking of seasonal influenza H3N2 virus evolution]{nextflu: Real-time 
-tracking of seasonal influenza H3N2 virus evolution in humans}
+\title[Tracking of seasonal influenza H3N2 virus evolution]{nextflu: Real-time tracking of seasonal influenza H3N2 virus evolution in humans}
 \author{Richard~A.~Neher$^{1}$ and Trevor Bedford$^{2}$}
-\address{$^{1}$Max Planck Institute for Developmental Biology, 72076 T\"ubingen, 
-Germany, and $^{2}$Vaccine and Infectious Disease Division, Fred Hutchinson
-Cancer Research Center, Seattle, WA 98109, USA}
+\address{$^{1}$Max Planck Institute for Developmental Biology, 72076 T\"ubingen, Germany, and $^{2}$Vaccine and Infectious Disease Division, Fred Hutchinson Cancer Research Center, Seattle, WA 98109, USA}
 \history{Received on XXXXX; revised on XXXXX; accepted on XXXXX}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{abstract} \section{Motivation:} Seasonal influenza viruses evolve
-rapidly and change their antigenic properties. This allows the virus
-population to evade immunity in their human hosts and reinfect previously
-infected individuals. Similarly, vaccines against seasonal influenza need to
-be updated frequently to protect against an evolving virus population.
+\begin{abstract} \section{Motivation:} Seasonal influenza viruses evolve rapidly and change their antigenic properties.
+This allows the virus population to evade immunity in their human hosts and reinfect previously infected individuals.
+Similarly, vaccines against seasonal influenza need to be updated frequently to protect against an evolving virus population.
 
-\section{Results:} We have developed a processing pipeline and web-browser
-based visualization that allows convenient exploration and analysis of the
-most recent influenza virus sequence data. This web-application is live at
-\url{nextflu.org} and displays a phylogenetic tree that can be decorated with
-additional information such the viral genotype at specific
-sites, sampling location, and sequence or tree based statistics that have been
-shown to predictive of future virus dynamics. Additionally, mutation frequency
-trajectories are calculated and displayed.
+\section{Results:} We have developed a processing pipeline and web-browser based visualization that allows convenient exploration and analysis of the most recent influenza virus sequence data.
+This web-application is live at \url{nextflu.org} and displays a phylogenetic tree that can be decorated with additional information such the viral genotype at specific sites, sampling location, and sequence or tree based statistics that have been shown to predictive of future virus dynamics.
+Additionally, mutation frequency trajectories are calculated and displayed.
 
@@
 
-nextflu consists of a processing pipeline writtin in python and a javascript
-based visualization called auspice. As input, augur requires a fasta file of
-sequences of the HA segment with fasta labels containing relevant information
-such as the strain name, the sampling date and passage history. In a first
-step, viruses without complete date  or geographic information and viruses
-passaged in eggs are removed. In addition, local outbreaks are filtered by
-keeping only one instance of identical sequences sampled at the same location
-on the same day.  Following filtering, the viruses are subsampled to achieve
-an appoximately even temporal and global distribution (typically 50 viruses
-per month). For our standard display period of 3 years, this typically results
-in \~1800 viruses, which we align us \texttt{mafft} \citep{mafft}.
+nextflu consists of a processing pipeline writtin in python and a javascript based visualization called auspice.
+As input, augur requires a fasta file of sequences of the HA segment with fasta labels containing relevant information such as the strain name, the sampling date and passage history.
+In a first step, viruses without complete date  or geographic information and viruses passaged in eggs are removed.
+In addition, local outbreaks are filtered by keeping only one instance of identical sequences sampled at the same location on the same day.
+Following filtering, the viruses are subsampled to achieve an appoximately even temporal and global distribution (typically 50 viruses per month).
+For our standard display period of 3 years, this typically results in \~1800 viruses, which we align us \texttt{mafft} \citep{mafft}.
 
-Once aligned, the set of virus sequences is further cleaned by removing
-insertions relative  to the outgroup and by removing sequences that are either
-much closer or further from the outgroup then expected given their date of
-sampling (filtering viruses that don't follow an approximate molecular clock).
+Once aligned, the set of virus sequences is further cleaned by removing insertions relative to the outgroup and by removing sequences that are either much closer or further from the outgroup then expected given their date of sampling (filtering viruses that don't follow an approximate molecular clock).
 REASSORTANTS.
 
-From the filtered and cleaned alignment, augur builds a phylogenetic tree using
-\texttt{fasttree} \citep{Price:2009p47657} which is further refined by one hour
-of RAxML optimization \citep{RAxml}. Next, the state of every internal node
-of the tree is inferred using marginal maximum likelihood method and missing 
-sequence is filled with the nearest ancestral sequence. Internal branches 
-without mutations are collapsed into polytomies. The final tree is decorated
-with the attributes to be displayed in the browser, see below.
+From the filtered and cleaned alignment, augur builds a phylogenetic tree using \texttt{fasttree} \citep{Price:2009p47657} which is further refined by one hour of RAxML optimization \citep{RAxml}.
+Next, the state of every internal node of the tree is inferred using marginal maximum likelihood method and missing sequence is filled with the nearest ancestral sequence.
+Internal branches without mutations are collapsed into polytomies.
+The final tree is decorated with the attributes to be displayed in the browser, see below.
 
@@
 
-In addition to the phylogenetic tree, augur estimates the frequency
-trajectories of mutations, genotypes and clades in the tree. Frequency
-trajectories are  represented as a linear interpolation $x(t)$ of values $x_i$
-at time points  $t_i$ spaced one month apart. The pivot frequencies
-corresponding to a feature $\phi$ (e.g. mutation) are estimated by minimizing
+In addition to the phylogenetic tree, augur estimates the frequency trajectories of mutations, genotypes and clades in the tree. 
+Frequency trajectories are represented as a linear interpolation $x(t)$ of values $x_i$ at time points  $t_i$ spaced one month apart. 
+The pivot frequencies corresponding to a feature $\phi$ (e.g. mutation) are estimated by minimizing
 \begin{equation}
@@
 \end{equation}
-where $I(v,\phi)=1$ if virus $v$ carries attribute $\phi$ and $I(v,\phi)=0$
-otherwise. $\Delta x_i = x_i-x_{i-1}$, $\Delta t_i = t_i-t_{i-1}$, $\gamma$
-parameterizes the smoothing imposed on the frequency estimate (related to the
-frequency changes by genetic drift) and $\epsilon$ parameterizes our
-expectation of frequency changes in interval $i$ based on the observed change
-in interval $i-1$. $\epsilon=0$ implies that frequency changes are
-uncorrelated from one interval to the next, while $\epsilon=1$ implies that
-the most likely $\Delta x_i$ is $\Delta x_{i-1}$. We typically use $\epsilon =
-0.7$. \EQ{freq} is minimized using SciPy's implementation of a downhill
-simplex algorithm \citep{Oliphant:2007p25672}. To speed up convergence,
-minization is first done using a coarse grid of pivot points which is
-subsequently refined to 12 pivots per year. PUT IN DETAILS ON HOW WE ESTIMATE 
-FREQUENCIES ON THE TREE?
+where $I(v,\phi)=1$ if virus $v$ carries attribute $\phi$ and $I(v,\phi)=0$ otherwise.
+$\Delta x_i = x_i-x_{i-1}$, $\Delta t_i = t_i-t_{i-1}$, $\gamma$ parameterizes the smoothing imposed on the frequency estimate (related to the frequency changes by genetic drift) and $\epsilon$ parameterizes our expectation of frequency changes in interval $i$ based on the observed change in interval $i-1$.
+$\epsilon=0$ implies that frequency changes are uncorrelated from one interval to the next, while $\epsilon=1$ implies that
+the most likely $\Delta x_i$ is $\Delta x_{i-1}$.
+We typically use $\epsilon = 0.7$.
+\EQ{freq} is minimized using SciPy's implementation of a downhill simplex algorithm \citep{Oliphant:2007p25672}.
+To speed up convergence, minization is first done using a coarse grid of pivot points which is
+subsequently refined to 12 pivots per year.
+PUT IN DETAILS ON HOW WE ESTIMATE FREQUENCIES ON THE TREE?
 
-The tree decorated with attributes and clade frequencies is written to file in
-a nested json format for display in the browser by auspice. In addition to the
-tree, json files containing the sequences, frequencies of mutations and
-genotypes, as well as meta information are exported.
+The tree decorated with attributes and clade frequencies is written to file in a nested json format for display in the browser by auspice.
+In addition to the tree, json files containing the sequences, frequencies of mutations and genotypes, as well as meta information are exported.
 
-The tree is visualized by auspice using d3 \citep{d3}. The user can explore
-the data interactively by selecting viruses from different dates (moving the
-date slider, top right corner in \FIG{tree}) or by coloring the tree by the
-following attributes:
+The tree is visualized by auspice using d3 \citep{d3}.
+The user can explore the data interactively by selecting viruses from different dates (moving the date slider, top right corner in \FIG{tree}) or by coloring the tree by the following attributes:
 \begin{itemize}
-	\item epitope mutations: each tip is colored by the number of aminoacid mutations at \~50 
-		  positions relative to the root of the tree. These positions are known to
-		  be frequent targets of antibodies \citep{shih_simultaneous_2007} and have been
-		  suggested to be predictive of future success \citep{luksza_predictive_2014}.
-	\item non-epitope mutations: color by the number of mutations outside the above
-		  mentioned epitope sites. Mutations outside epitopes tend to be damaging and 
-		  anti-correlated with clade expansion \citep{luksza_predictive_2014}.
-    \item receptor binding mutations: color by the number of mutations at 7 positions
-    	  close to the receptor binding site that have been shown to be responsible 
-    	  for major antigenic positions in the past decades \citep{koel_substitutions_2013}.
-    \item local branching index is the exponentially weighted tree length surrounding 
-    	  a node, which is associated with rapid branching and expansion of clades 
-    	  \citep{neher_predicting_2014}. It is recalculated everytime the date slider is
-    	  moved and based only on the highlighted viruses.
+	\item epitope mutations: each tip is colored by the number of aminoacid mutations at \~50 positions relative to the root of the tree.
+These positions are known to be frequent targets of antibodies \citep{shih_simultaneous_2007} and have been suggested to be predictive of future success \citep{luksza_predictive_2014}.
+	\item non-epitope mutations: color by the number of mutations outside the above mentioned epitope sites.
+	Mutations outside epitopes tend to be damaging and anti-correlated with clade expansion \citep{luksza_predictive_2014}.
+    \item receptor binding mutations: color by the number of mutations at 7 positions close to the receptor binding site that have been shown to be responsible for major antigenic positions in the past decades \citep{koel_substitutions_2013}.
+    \item local branching index is the exponentially weighted tree length surrounding a node, which is associated with rapid branching and expansion of clades \citep{neher_predicting_2014}.
+It is recalculated everytime the date slider is moved and based only on the highlighted viruses.
     \item geographic location
-    \item HA1 genotype: entering amino acid positions (separated by a comma) in the 
-    	  text field will color every node according to the genotype at these positions.
+    \item HA1 genotype: entering amino acid positions (separated by a comma) in the text field will color every node according to the genotype at these positions.
 \end{itemize}
 
-The frequency plot below the tree (see \FIG{freq}) displays the frequency trajectory
-of clades in the tree whenever the mouse hovers above the branch defining the clade. 
-Furthermore, trajectories of individual mutations, combinations of two mutations, and 
-predifined clades such as 3c3.a can be plotted by entering them in text field
+The frequency plot below the tree (see \FIG{freq}) displays the frequency trajectory of clades in the tree whenever the mouse hovers above the branch defining the clade. 
+Furthermore, trajectories of individual mutations, combinations of two mutations, and predifined clades such as 3c3.a can be plotted by entering them in text field.
+
 \begin{figure}[bhtp]
@@
 \section{Conclusion}
-We built nextflu.org to facilitate the analysis and exploration of seasonal influenza
-sequence data collected by laboratories around to world. By using the most recent data
-and integrating phylogenies with frequency trajectories and sequence or tree based
-predictors of successful clades, we hope that nextflu can inform the choice of  
-strains used in seasonal influenza vaccines.
+We built nextflu.org to facilitate the analysis and exploration of seasonal influenza sequence data collected by laboratories around to world.
+By using the most recent data and integrating phylogenies with frequency trajectories and sequence or tree based predictors of successful clades, we hope that nextflu can inform the choice of strains used in seasonal influenza vaccines.
 
 nextflu was designed to be readily adapted to other rapidly evolving viruses.
-To that end, the generic classes used for processing the sequence data to include features 
-specific to the virus in question.
+To that end, the generic classes used for processing the sequence data to include features specific to the virus in question.
 
-\paragraph{Funding\textcolon}This work is supported by the ERC though
-Stg-260686.
+\paragraph{Funding\textcolon}This work is supported by the ERC though Stg-260686.
 

### 2015-04-01 manuscript/nextflu.tex
Revise abstract to fit Application Note style.
See http://www.oxfordjournals.org/our_journals/bioinformatics/for_authors/general.html.

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{abstract} \section{Motivation:} Seasonal influenza viruses evolve rapidly and change their antigenic properties.
+\begin{abstract} \section{Summary:} Seasonal influenza viruses evolve rapidly and change their antigenic properties.
 This allows the virus population to evade immunity in their human hosts and reinfect previously infected individuals.
 Similarly, vaccines against seasonal influenza need to be updated frequently to protect against an evolving virus population.
-
-\section{Results:} We have developed a processing pipeline and web-browser based visualization that allows convenient exploration and analysis of the most recent influenza virus sequence data.
-This web-application is live at \url{nextflu.org} and displays a phylogenetic tree that can be decorated with additional information such the viral genotype at specific sites, sampling location, and sequence or tree based statistics that have been shown to predictive of future virus dynamics.
+We have thus developed a processing pipeline and web-browser based visualization that allows convenient exploration and analysis of the most recent influenza virus sequence data.
+This web-application displays a phylogenetic tree that can be decorated with additional information such the viral genotype at specific sites, sampling location, and sequence or tree based statistics that have been shown to predictive of future virus dynamics.
 Additionally, mutation frequency trajectories are calculated and displayed.
 
-\section{Availability:} \url{http://nextflu.org} and \url{https://github.com/blab/nextflu}.
+\section{Availability and implementation:} Python and Javascript source code is freely available from \url{https://github.com/blab/nextflu}, while the web-application is live at \url{http://nextflu.org}.
+
+\section{Contact:} tbedford@fredhutch.org
+
 \end{abstract}

### 2015-04-01 augur/src/H1N1pdm_process.py
added H1N1pdm vaccine strain to augur and auspice added H1N1pdm receptor binding site mutations and epitope mutations repurposed H3N2_refine from H3N2_process for H1N1 processing added auspice export file names to virus config and a prefix variable for loading in auspice

@@ -1,46 +1,41 @@
-import time, argparse,re,os
+import time, re, os
 from virus_filter import flu_filter
 from virus_clean import virus_clean
 from tree_refine import tree_refine
-from process import process
+from H3N2_process import H3N2_refine as H1N1pdm_refine
+from process import process, virus_config
 from Bio import SeqIO
 from Bio.Seq import Seq
 from Bio.Align import MultipleSeqAlignment
 import numpy as np
 from itertools import izip
 
-epitope_mask = np.fromstring("0000000000000000000000000000000000000000000011111011011001010011000100000001001011110011100110101000001100000100000001000110101011111101011010111110001010011111000101011011111111010010001111101110111001010001110011111111000000111110000000101010101110000000000011100100000001011011100000000000001001011000110111111000000000000000000000000000000000000000000000000000000000000000000000000000000000000000000000000000000000000000000000000000000000000000000000000000000000000000000000000000000000000000000000000000000000000000000000000000000000000000000000", dtype='S1')
-	
-virus_config = {
+epitope_mask = np.array(['1' if pos in [141,142,145,146,172,176,178,179,180,181,183,184,185, #Sa
+										170,173,174,177,206,207,210,211,212,214,216,		 #Sb
+										183,187,191,196,221,225,254,258,288,				 #Ca1
+										154,157,158,159,161,163,238,239,242,243,			 #Ca2
+										87, 88, 90, 91, 92, 95, 96, 98, 99, 100, 132, 139	 #Cb
+									   ]
+						else '0' for pos in xrange(1,1725)])
+
+receptor_binding_sites = [159,169,170,172,173,203,207]
+
+
+virus_config.update({
 	# data source and sequence parsing/cleaning/processing
 	'virus':'H1N1pdm',
 	'alignment_file':'data/H1N1pdm_gisaid_epiflu_sequence.fasta',
-	'fasta_fields':{0:'strain', 1:'accession', 3:'passage', 5:'date' },
 	'outgroup':'A/Swine/Indiana/P12439/00',
[…]

### 2015-04-01 augur/src/process.py

@@ -10,7 +10,7 @@ from tree_util import delimit_newick
 import numpy as np
 
 parser = argparse.ArgumentParser(description='Process virus sequences, build tree, and prepare of web visualization')
-parser.add_argument('-y', '--years_back', type = int, default=3, help='number of past years to sample sequences from')
+parser.add_argument('-y', '--years_back', type = float, default=3, help='number of past years to sample sequences from')
 parser.add_argument('-v', '--viruses_per_month', type = int, default = 50, help='number of viruses sampled per month')
 parser.add_argument('-r', '--raxml_time_limit', type = float, default = 1.0, help='number of hours raxml is run')
 parser.add_argument('--interval', nargs = '+', type = float, default = None, help='interval from which to pull sequences')

### 2015-04-01 auspice/js/auspice.js

@@ -1,3 +1,5 @@
+var file_prefix = 'H1N1pdm_';
+
 function gatherTips(node, tips) {
 	if (typeof node.children != "undefined") {
 		for (var i=0, c=node.children.length; i<c; i++) {
@@ -22,6 +24,7 @@ function gatherInternals(node, internals) {
 
 function getVaccines(tips) {
 	vaccineChoice = {};
+	vaccineChoice['A/California/07/1999'] = "2009-09-25"; //H1N1pdm CHECK
 	vaccineChoice['A/Fujian/411/2002'] = "2003-09-25";
 	vaccineChoice['A/California/7/2004'] = "2005-02-21";
 	vaccineChoice['A/Wisconsin/67/2005'] = "2006-02-21";
@@ -359,7 +362,7 @@ var gt_chart = c3.generate({
 });
 
 
-d3.json("data/tree.json", function(error, root) {
+d3.json("data/"+file_prefix+"tree.json", function(error, root) {
 
 	if (error) return console.warn(error);
 
@@ -1178,7 +1181,7 @@ d3.json("data/tree.json", function(error, root) {
 
 });
 
-d3.json("data/meta.json", function(error, json) {
+d3.json("data/"+file_prefix+"meta.json", function(error, json) {
 	if (error) return console.warn(error);
 	d3.select("#updated").text(json['updated']);
 	commit_id = json['commit'];
@@ -1190,12 +1193,12 @@ d3.json("data/meta.json", function(error, json) {
 
 });
 
-d3.json("data/sequences.json", function(error, json) {
+d3.json("data/"+file_prefix+"sequences.json", function(error, json) {
 	if (error) return console.warn(error);
 	cladeToSeq=json;
 });
 
-d3.json("data/frequencies.json", function(error, json){
+d3.json("data/"+file_prefix+"frequencies.json", function(error, json){
 	console.log(error);
 	var pivots= json["mutations"]["global"]["pivots"].map(function (d) {return Math.round(parseFloat(d)*100)/100;});
 	var ticks = [Math.round(pivots[0])];

### 2015-04-01 augur/src/H1N1pdm_process.py
changed auspice output to prefix+standard

@@ -32,9 +32,7 @@ virus_config.update({
 	'cds':[0,None], # define the HA1 start i n 0 numbering
 	# define relevant clades in canonical HA1 numbering (+1)
 	'clade_designations': {},
-	'auspice_frequency_fname':'../auspice/data/H1N1pdm_frequencies.json',
-	'auspice_sequences_fname':'../auspice/data/H1N1pdm_sequences.json',
-	'auspice_tree_fname':'../auspice/data/H1N1pdm_tree.json',
+	'auspice_prefix':'H1N1pdm_',
 	})
 
 

### 2015-04-01 augur/src/process.py

@@ -40,9 +40,7 @@ virus_config = {
 class process(virus_frequencies):
 	"""generic template class for processing virus sequences into trees"""
 	def __init__(self, prefix = 'data/', time_interval = (2012.0, 2015.0), 
-	             auspice_frequency_fname ='../auspice/data/frequencies.json', 
-				 auspice_sequences_fname='../auspice/data/sequences.json', 
-				 auspice_tree_fname='../auspice/data/tree.json', 
+	             auspice_prefix = '', 
 				 min_mutation_frequency = 0.01, min_genotype_frequency = 0.1, **kwargs):
 		self.tree_fname = prefix+'tree.pkl'
 		self.virus_fname = prefix+'virus.pkl'
@@ -52,9 +50,10 @@ class process(virus_frequencies):
 		self.min_genotype_frequency = min_genotype_frequency
 		self.time_interval = tuple(time_interval)
 
-		self.auspice_tree_fname = auspice_tree_fname
-		self.auspice_sequences_fname = auspice_sequences_fname
-		self.auspice_frequency_fname = auspice_frequency_fname
+		self.auspice_tree_fname = auspice_prefix + 'tree.json'
+		self.auspice_sequences_fname = auspice_prefix + 'sequences.json'
+		self.auspice_frequency_fname = auspice_prefix + 'frequencies.json'
+		self.auspice_meta_fname = auspice_prefix + 'meta.json'
 		self.nuc_alphabet = 'ACGT-N'
 		self.aa_alphabet = 'ACDEFGHIKLMNPQRSTVWY*X'
 		virus_frequencies.__init__(self, **kwargs)
@@ -167,8 +166,7 @@ class process(virus_frequencies):
 			meta["regions"] = self.regions
 			meta["virus_stats"] = [ [str(y)+'-'+str(m)] + [self.date_region_count[(y,m)][reg] for reg in self.regions]
 									for y,m in sorted(self.date_region_count.keys()) ]
-		meta_fname = "../auspice/data/meta.json"
-		write_json(meta, meta_fname, indent=0)
+		write_json(meta, self.auspice_meta_fname, indent=0)
 
 	def align(self):
 		'''

### 2015-04-01 augur/src/process.py
fixed vaccine designation for h1n1pdm,
added auspice_prefix to process.py

caught missing genotype frequency in auspice

@@ -50,10 +50,10 @@ class process(virus_frequencies):
 		self.min_genotype_frequency = min_genotype_frequency
 		self.time_interval = tuple(time_interval)
 
-		self.auspice_tree_fname = auspice_prefix + 'tree.json'
-		self.auspice_sequences_fname = auspice_prefix + 'sequences.json'
-		self.auspice_frequency_fname = auspice_prefix + 'frequencies.json'
-		self.auspice_meta_fname = auspice_prefix + 'meta.json'
+		self.auspice_tree_fname = 		'../auspice/data/' + auspice_prefix + 'tree.json'
+		self.auspice_sequences_fname = 	'../auspice/data/' + auspice_prefix + 'sequences.json'
+		self.auspice_frequency_fname = 	'../auspice/data/' + auspice_prefix + 'frequencies.json'
+		self.auspice_meta_fname = 		'../auspice/data/' + auspice_prefix + 'meta.json'
 		self.nuc_alphabet = 'ACGT-N'
 		self.aa_alphabet = 'ACDEFGHIKLMNPQRSTVWY*X'
 		virus_frequencies.__init__(self, **kwargs)

### 2015-04-01 auspice/js/auspice.js

@@ -24,7 +24,7 @@ function gatherInternals(node, internals) {
 
 function getVaccines(tips) {
 	vaccineChoice = {};
-	vaccineChoice['A/California/07/1999'] = "2009-09-25"; //H1N1pdm CHECK
+	vaccineChoice['A/California/07/2009'] = "2009-09-25"; //H1N1pdm CHECK
 	vaccineChoice['A/Fujian/411/2002'] = "2003-09-25";
 	vaccineChoice['A/California/7/2004'] = "2005-02-21";
 	vaccineChoice['A/Wisconsin/67/2005'] = "2006-02-21";
@@ -1240,7 +1240,7 @@ d3.json("data/"+file_prefix+"frequencies.json", function(error, json){
 	function get_frequencies(region, gt){
 		var freq = [];
 		for (var pi=0; pi<pivots.length; pi++){freq[freq.length]=0;}
-		if (json["genotypes"][region][gt]!=undefined) {
+		if ((typeof json["genotypes"] !="undefined") && (json["genotypes"][region][gt]!=undefined)) {
 			console.log(gt+" found as genotype");
 			for (var pi=0; pi<freq.length; pi++){
 				freq[pi]+=json["genotypes"][region][gt][pi];

### 2015-04-01 augur/src/H3N2_process.py
Fix main so H3N2_process will run without `--skip` specified.

@@ -275,10 +275,11 @@ if __name__=="__main__":
 	dt= params.time_interval[1]-params.time_interval[0]
 	params.pivots_per_year = 12.0 if dt<5 else 6.0 if dt<10 else 3.0
 	steps = all_steps[all_steps.index(params.start):(all_steps.index(params.stop)+1)]
-	for tmp_step in params.skip:
-		if tmp_step in steps:
-			print "skipping",tmp_step
-			steps.remove(tmp_step)
+	if params.skip is not None:
+		for tmp_step in params.skip:
+			if tmp_step in steps:
+				print "skipping",tmp_step
+				steps.remove(tmp_step)
 
 	# add all arguments to virus_config (possibly overriding)
 	virus_config.update(params.__dict__)

### 2015-04-01 augur/src/H1N1pdm_process.py
copied Trevor's fix for missing --skip to H1N1pdm

@@ -159,10 +159,11 @@ if __name__=="__main__":
 	dt= params.time_interval[1]-params.time_interval[0]
 	params.pivots_per_year = 12.0 if dt<5 else 6.0 if dt<10 else 3.0
 	steps = all_steps[all_steps.index(params.start):(all_steps.index(params.stop)+1)]
-	for tmp_step in params.skip:
-		if tmp_step in steps:
-			print "skipping",tmp_step
-			steps.remove(tmp_step)
+	if params.skip is not None:
+		for tmp_step in params.skip:
+			if tmp_step in steps:
+				print "skipping",tmp_step
+				steps.remove(tmp_step)
 
 	# add all arguments to virus_config (possibly overriding)
 	virus_config.update(params.__dict__)

### 2015-04-01 augur/src/bernoulli_frequency.py
fixed problem with frequency estimation when frequency shifts are very rapid. in this example, frequencies went from 0 to 1 very quickly resulting in a
000001000001000000011111111111111111111111111111111111111111111111111

pattern of observations. To speed up freq estimation, we had restricted
the interval used to 6 pivots after the last 0. With stiff prior on frequency
change, the frequency had not reached 1 by that point.

to fix this, a few oversights in argument passing got fixed, frequency estimation
made less stiff, and the number of pivots after the last 0 increased to 12

@@ -204,13 +204,14 @@ class frequency_estimator(object):
 
 class virus_frequencies(object):
 	def __init__(self, time_interval = (2012.0, 2015.1),
-				stiffness = 10.0, pivots_per_year = 12.0, 
+				frequency_stiffness = 10.0, pivots_per_year = 12.0, 
 				clade_designations={}, aggregate_regions = None,
 				extra_pivots = 5, **kwarks):
-		self.stiffness = stiffness
+		self.stiffness = frequency_stiffness
 		self.pivots_per_year = pivots_per_year
 		self.clade_designations=clade_designations
 		self.aggregate_regions = aggregate_regions
+		self.extra_pivots = extra_pivots
 		self.pivots = get_pivots(self.time_interval[0], self.time_interval[1], self.pivots_per_year)
 		self.kwarks = kwarks
 		if not hasattr(self, "frequencies"):
@@ -236,7 +237,7 @@ class virus_frequencies(object):
 		if len(tps)>threshold and np.sum(obs)>min_observations:
 			if self.verbose: 
 				print "# of time points",len(tps), "# observations",sum(obs) 
-			fe = frequency_estimator(zip(tps, obs), pivots=self.pivots,
+			fe = frequency_estimator(zip(tps, obs), pivots=self.pivots, extra_pivots = self.extra_pivots,
 			               stiffness=self.stiffness*float(len(observations))/len(self.viruses), 
 		                   logit=True, **self.kwarks)
 			fe.learn()
@@ -325,23 +326,23 @@ class virus_frequencies(object):
 		ci=0
 		# need to resort, since the clade size order might differs after subsetting to regions
 		children_by_size = sorted(node.child_nodes(), key = lambda x:len(x.tips), reverse=True)
+		if debug: print '###',len(node.tips), frequency_left[-5:]
 		for child in children_by_size[:-1]: # clades are ordered by decreasing size
 			if len(child.tips)<threshold: # skip tiny clades
 				break
 			else:
-				if debug: print child.num_date, len(child.tips)
-
 				obs = np.in1d(tps, all_dates[tip_to_date_index[child.tips]])
 
 				# make n pivots a year, interpolate frequencies
 				# FIXME: adjust stiffness to total number of observations in a more robust manner
 				fe = frequency_estimator(zip(tps, obs), pivots=self.pivots, stiffness=self.stiffness*len(all_dates)/2000.0, 
-										logit=True, **self.kwarks)
[…]

### 2015-04-01 augur/src/process.py

@@ -26,11 +26,11 @@ virus_config = {
 	# frequency estimation parameters
 	'aggregate_regions': [  ("global", None), ("NA", ["NorthAmerica"]), ("EU", ["Europe"]), 
 							("AS", ["China", "SoutheastAsia", "JapanKorea"]), ("OC", ["Oceania"]) ],
-	'frequency_stiffness':10.0,
+	'frequency_stiffness':5.0,
 	'verbose':2, 
 	'tol':1e-4, #tolerance for frequency optimization
 	'pc':1e-3, #pseudocount for frequencies 
-	'extra_pivots': 6,  # number of pivot point for or after the last observations of a mutations
+	'extra_pivots': 12,  # number of pivot point for or after the last observations of a mutations
 	'inertia':0.7,		# fraction of frequency change carry over in the stiffness term
 	'n_iqd':3,     # standard deviations from clock
 }

### 2015-04-01 auspice/js/auspice.js
Switch to acronym rather than abbreviation for originating lab.

@@ -277,8 +277,10 @@ var virusTooltip = d3.tip()
 			string += "<br>GISAID ID: EPI" + d.accession;
 		}
 		if (typeof d.lab != "undefined") {
-			string += "<br>Orig. lab: " + d.lab.substring(0,30);
-			if (d.lab.length>30) string+='...';
+			if (d.lab != "") {
+				var abbr = d.lab.match(/[A-Z]/g).join('')
+				string += "<br>Orig. lab: " + abbr;
+			}
 		}
 		if (typeof d.region != "undefined") {
 			string += "<br>Region: " + d.region.replace(/([A-Z])/g, ' $1');

### 2015-04-02 manuscript/nextflu.tex
Revise text, mostly introduction and augur.

@@
 \maketitle
-\bibliographystyle{plain}
+\bibliographystyle{bioinformatics}
 
@@
 This web-application displays a phylogenetic tree that can be decorated with additional information such the viral genotype at specific sites, sampling location, and sequence or tree based statistics that have been shown to predictive of future virus dynamics.
-Additionally, mutation frequency trajectories are calculated and displayed.
+Additionally, mutation, genotype and clade frequency trajectories are calculated and displayed.
 
@@
 
-\begin{itemize}
-	\item Motivation
-	\item data sources
-	\item pipe line and design
-	\item frequency estimation details
-	\item genotype and trait coloring
-	\item generalizability
-\end{itemize}
-
-nextflu consists of a processing pipeline writtin in python and a javascript based visualization called auspice.
-As input, augur requires a fasta file of sequences of the HA segment with fasta labels containing relevant information such as the strain name, the sampling date and passage history.
-In a first step, viruses without complete date  or geographic information and viruses passaged in eggs are removed.
-In addition, local outbreaks are filtered by keeping only one instance of identical sequences sampled at the same location on the same day.
-Following filtering, the viruses are subsampled to achieve an appoximately even temporal and global distribution (typically 50 viruses per month).
-For our standard display period of 3 years, this typically results in \~1800 viruses, which we align us \texttt{mafft} \citep{mafft}.
-
-Once aligned, the set of virus sequences is further cleaned by removing insertions relative to the outgroup and by removing sequences that are either much closer or further from the outgroup then expected given their date of sampling (filtering viruses that don't follow an approximate molecular clock).
-REASSORTANTS.
-
-From the filtered and cleaned alignment, augur builds a phylogenetic tree using \texttt{fasttree} \citep{Price:2009p47657} which is further refined by one hour of RAxML optimization \citep{RAxml}.
-Next, the state of every internal node of the tree is inferred using marginal maximum likelihood method and missing sequence is filled with the nearest ancestral sequence.
-Internal branches without mutations are collapsed into polytomies.
-The final tree is decorated with the attributes to be displayed in the browser, see below.
+Every year, seasonal influenza infects between 10\% and 20\% of the global population and influenza subtype A/H3N2 is responsible for the bulk of human morbidity and mortality \citep{flufactsheet}.
+Vaccination remains the most effective public health measure to combat seasonal epidemics.
+However, influenza viruses constantly evolve and thereby undergo antigenic drift, allowing drifted viruses to reinfect those in the human population with acquired immunity to previously circulating strains.
+Owing to antigenic drift, the seasonal influenza vaccine needs frequent updating to remain effective.
+In any given year, the particular choice of vaccine strain plays a major role in determining vaccine efficacy and so it is of critical importance to develop tools to analyze the ongoing evolution of the influenza virus population in order to aid vaccine strain selection.
+The program nextflu presents a near real-time display of genetic relationships among influenza H3N2 viruses and allows investigation of currently available sequence data.
 
@@
 	\begin{center}
-	\includegraphics[width=0.99\columnwidth]{figures/tree_screenshot}
+	\includegraphics[width=0.99\columnwidth]{figures/tree}
 \caption[]{The \texttt{nextflu} website with the user interface on the left and
@@
 \end{figure}
+
+In implementation, nextflu consists of a processing pipeline written in Python called \textit{augur} that takes sequence data and infers evolutionary information and a JavaScript-based browser visualization called \textit{auspice} that displays this processed information.
+As input, \textit{augur} requires a FASTA file of sequences with FASTA header labels containing relevant information such as the strain name, the sampling date and passage history.
+For this purpose, influenza H3N2 sequence data for the hemagglutinin (HA) gene is downloaded from the GISAID EpiFlu database \citep{GISAID}, which contains the most up-to-date collection of seasonal influenza viruses.
+The first step in the processing pipeline is to automatically select a subset of representative H3N2 viruses.
+Here, viruses without complete date or geographic information, viruses passaged in eggs and sequences less than 987 bases are removed.
+In addition, local outbreaks are filtered by keeping only one instance of identical sequences sampled at the same location on the same day.
+Following filtering, the viruses are subsampled to achieve a more equitable temporal and geographic distribution.
+For our standard display period of 3 years and 50 viruses per month, this typically results in $\sim$1800 viruses, which we align using MAFFT \citep{katoh_mafft_2013}.
+Once aligned, the set of virus sequences is further cleaned by removing insertions relative to the outgroup to enforce canonical HA site numbering and by removing sequences that show either too much or too little divergence relative to the expectation given data of sampling.
+Additionally, known reassortant clusters are removed, including the triple-reassortant influenza viruses sporadically sampled in humans throughout North America \citep{bastien_human_2010}.
+
+From the filtered and cleaned alignment, \textit{augur} builds a phylogenetic tree using FastTree \citep{price_fasttree_2009}, which is then further refined using RAxML \citep{stamatakis_raxml_2014}.
+Next, the state of every internal node of the tree is inferred using marginal maximum likelihood method and missing sequence is filled with the nearest ancestral sequence.
+Internal branches without mutations are collapsed into polytomies.
+The final tree is decorated with the attributes to be displayed in the browser, see below.
 
@@
 
-The tree is visualized by auspice using d3 \citep{d3}.
+The tree is visualized by auspice using d3 \citep{bostock_d3_2011}.
 The user can explore the data interactively by selecting viruses from different dates (moving the date slider, top right corner in \FIG{tree}) or by coloring the tree by the following attributes:

### 2015-04-03 augur/src/make_all.py
added shebang and extra binary variable in case I run it on the tuebingen cluster.

@@ -1,3 +1,4 @@
+#!/usr/bin/env python
 import subprocess,glob, os
 
 run_pipeline = []
@@ -8,24 +9,26 @@ for flu in ["H3N2", "H1N1pdm", "H1N1", "Vic", "Yam"]:
 		if len(auspice_files)==0 or min([os.path.getmtime(fname) for fname in auspice_files])<max([os.path.getmtime(fname) for fname in input_files]):
 			run_pipeline.append(flu)
 
+#python_binary = '/ebio/ag-neher/share/programs/bin/python'
+python_binary = 'python'
 
 if 'H3N2' in run_pipeline:
-	call = map(str, ['python', 'src/H3N2_process.py', '-v', 30, '-y', 3, '--skip', 'genotype_frequencies', '-r', 1.0, '--prefix', 'data/H3N2_'])
+	call = map(str, [python_binary, 'src/H3N2_process.py', '-v', 50, '-y', 3, '--skip', 'genotype_frequencies', '-r', 1.0, '--prefix', 'data/H3N2_'])
 	print call
 	subprocess.call(call)
 if 'H1N1' in run_pipeline:
-	call = map(str, ['python', 'src/H1N1historical_process.py', '-v', 20, '--interval', 1990, 2010, '--skip', 'genotype_frequencies', '-r', 1.0, '--prefix', 'data/H1N1_'])
+	call = map(str, [python_binary, 'src/H1N1historical_process.py', '-v', 20, '--interval', 1990, 2010, '--skip', 'genotype_frequencies', '-r', 1.0, '--prefix', 'data/H1N1_'])
 	print call
 	subprocess.call(call)
 if 'H1N1pdm' in run_pipeline:
-	call = map(str, ['python', 'src/H1N1pdm_process.py', '-v', 30, '-v', 6, '--skip', 'genotype_frequencies', '-r', 1.0, '--prefix', 'data/H1N1pdm_'])
+	call = map(str, [python_binary, 'src/H1N1pdm_process.py', '-v', 30, '-y', 6, '--skip', 'genotype_frequencies', '-r', 1.0, '--prefix', 'data/H1N1pdm_'])
 	print call
 	subprocess.call(call)
 if 'Vic' in run_pipeline:
-	call = map(str, ['python', 'src/Vic_process.py', '-v', 30, '-v', 6, '--skip', 'genotype_frequencies', '-r', 1.0, '--prefix', 'data/Vic_'])
[…]

### 2015-04-03 auspice/js/auspice.js
changed tooltip to display country rather than region

@@ -266,8 +266,8 @@ var virusTooltip = d3.tip()
 		string +="</div>";
 		string += "<div class=\"smallspacer\"></div>";
 		string += "<div class=\"smallnote\">";
-		if (typeof d.region != "undefined") {
-			string += d.region.replace(/([A-Z])/g, ' $1');
+		if (typeof d.country != "undefined") {
+			string += d.country.replace(/([A-Z])/g, ' $1');
 		}
 		if (typeof d.date != "undefined") {
 			string += ", " + d.date;

### 2015-04-03 augur/src/H1N1pdm_process.py
ammended definition of clades 6c and 6b

@@ -37,8 +37,8 @@ virus_config.update({
 		'4':[(142, 'D'), (151 ,'A'), (200, 'S')],
 		'5':[(104, 'N'), (222, 'K'), (233, 'V'), (266, 'L')],
 		'6':[(202,'T'),  (114, 'N'), (214, 'A')],
-		'6c':[(251,'I'), (114, 'N'), (214, 'A')],
-		'6b':[(180,'Q'), (273,'T'), (214, 'A')],
+		'6c':[(251,'I'), (114, 'N'), (214, 'A'), (300,'E')],
+		'6b':[(180,'Q'), (273, 'T'), (214, 'A'), (300,'E')],
 		'7':[(202,'T'),  (114, 'N'), (214, 'T')],
 		'8':[(203,'T'), (289,'A')],
 		},

### 2015-04-03 auspice/js/Vic.js
added vaccine strains to B-flu js

@@ -2,7 +2,10 @@ var file_prefix = 'Vic_';
 var dfreq_dn = 2;
 
 var	vaccineChoice = {};
-//vaccineChoice['A/California/07/2009'] = "2009-09-25";
+vaccineChoice['B/Shangdong/7/97'] = "1999-09-25";
+vaccineChoice['B/HongKong/330/2001'] = "2002-09-25";
+vaccineChoice['B/Malaysia/2506/2004'] = "2006-09-25";
+vaccineChoice['B/Brisbane/60/2008'] = "2009-09-25";
 
 var epiColorDomain = [0,1,2,3, 4,5,6,7,8,9,10];
 var nonEpiColorDomain = [0,1,2,3,4,5,6,7,8,9,10,11];

### 2015-04-03 auspice/js/Yam.js

@@ -2,7 +2,11 @@ var file_prefix = 'Yam_';
 var dfreq_dn = 2;
 
 var	vaccineChoice = {};
-//vaccineChoice['A/California/07/2009'] = "2009-09-25";
+vaccineChoice['B/Beijing/184/93'] = "1998-11-01";
+vaccineChoice['B/Sichuan/379/99'] = "2001-09-25";
+vaccineChoice['B/Shanghai/361/2002'] = "2004-09-25";
+vaccineChoice['B/Florida/4/2006'] = "2008-09-25";
+vaccineChoice['B/PHUKET/3073/2013'] = "2015-03-25";
 
 var epiColorDomain = [0,1,2,3, 4,5,6,7,8,9,10];
 var nonEpiColorDomain = [0,1,2,3,4,5,6,7,8,9,10,11];

### 2015-04-03 augur/src/Vic_process.py
added another mutation to the definition of Vic/1A

@@ -32,7 +32,7 @@ virus_config.update({
 	'cds':[0,None], # define the HA1 start i n 0 numbering
 	# define relevant clades in canonical HA1 numbering (+1)
 	'clade_designations': {
-		'1A': [(101,'K'), (84, 'L'), (191, 'K')],
+		'1A': [(101,'K'), (84, 'L'), (191, 'K'), (613,'S')],
 		'1B': [(101,'K'), (84, 'P'), (191, 'K')]
 	},
 	'auspice_prefix':'Vic_',

### 2015-04-03 augur/src/bernoulli_frequency.py
record clade names in lower case, check for presence of clade frequency before checking for mutations since names like 1a etc can be parsed as mutations rather then clades

@@ -307,7 +307,7 @@ class virus_frequencies(object):
 			print "estimating frequency of clade", clade_name, clade_gt
 			freq, (tps, obs) = self.estimate_genotype_frequency(sub_aln, [(pos-1, aa) for pos, aa in clade_gt])
 			if freq is not None:
-				clade_frequencies[clade_name] = list(np.round(logit_inv(freq.y),3))
+				clade_frequencies[clade_name.lower()] = list(np.round(logit_inv(freq.y),3))
 		return clade_frequencies
 
 	def estimate_sub_frequencies(self, node, all_dates, tip_to_date_index, threshold=50, region_name="global"):

### 2015-04-03 auspice/js/auspice.js

@@ -1243,7 +1243,13 @@ d3.json("/data/" + file_prefix + "frequencies.json", function(error, json){
 	function get_frequencies(region, gt){
 		var freq = [];
 		for (var pi=0; pi<pivots.length; pi++){freq[freq.length]=0;}
-		if ((typeof json["genotypes"] !="undefined") && (json["genotypes"][region][gt]!=undefined)) {
+		if (json["clades"][region][gt.toLowerCase()]!=undefined) {
+			console.log(gt+" found as clade");
+			for (var pi=0; pi<freq.length; pi++){
+				freq[pi]+=json["clades"][region][gt.toLowerCase()][pi];
+			}
+		}
+		else if ((typeof json["genotypes"] !="undefined") && (json["genotypes"][region][gt]!=undefined)) {
 			console.log(gt+" found as genotype");
 			for (var pi=0; pi<freq.length; pi++){
 				freq[pi]+=json["genotypes"][region][gt][pi];
@@ -1252,11 +1258,6 @@ d3.json("/data/" + file_prefix + "frequencies.json", function(error, json){
 			console.log(gt+" found as mutation");
 			for (var pi=0; pi<freq.length; pi++){
 				freq[pi]+=json["mutations"][region][gt][pi];
-			}
-		}else if (json["clades"][region][gt.toLowerCase()]!=undefined) {
-			console.log(gt+" found as clade");
-			for (var pi=0; pi<freq.length; pi++){
-				freq[pi]+=json["clades"][region][gt.toLowerCase()][pi];
 			}
 		}
 		return freq.map(function (d) {return Math.round(d*100)/100;});

### 2015-04-03 augur/src/make_all.py
silly error in make_all...

@@ -29,6 +29,6 @@ if 'Vic' in run_pipeline:
 	print call
 	subprocess.call(call)
 if 'Yam' in run_pipeline:
-	call = map(str, [python_binary, 'src/Vic_process.py', '-v', 30, '-y', 6, '--skip', 'genotype_frequencies', '-r', 1.0, '--prefix', 'data/Yam_'])
+	call = map(str, [python_binary, 'src/Yam_process.py', '-v', 30, '-y', 6, '--skip', 'genotype_frequencies', '-r', 1.0, '--prefix', 'data/Yam_'])
 	print call
 	subprocess.call(call)

### 2015-04-03 manuscript/nextflu.tex
shortened and streamlined

@@
 \newcommand{\REF}[1]{ref.~\citep{#1}}
+\newcommand{\augur}{\textit{augur}}
+\newcommand{\auspice}{\textit{auspice}}
+\newcommand{\nextflu}{\textit{nextflu}}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\title[Tracking of seasonal influenza H3N2 virus evolution]{nextflu: Real-time tracking of seasonal influenza H3N2 virus evolution in humans}
+\title[Tracking of seasonal influenza virus evolution]{nextflu: Real-time tracking of seasonal influenza virus evolution in humans}
 \author{Richard~A.~Neher$^{1}$ and Trevor Bedford$^{2}$}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{abstract} \section{Summary:} Seasonal influenza viruses evolve rapidly and change their antigenic properties.
-This allows the virus population to evade immunity in their human hosts and reinfect previously infected individuals.
+\begin{abstract} \section{Summary:} Seasonal influenza viruses evolve rapidly, which allows the virus population to evade immunity in their human hosts and reinfect previously infected individuals.
 Similarly, vaccines against seasonal influenza need to be updated frequently to protect against an evolving virus population.
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section*{Introduction}
+%\section*{Introduction}
 
-Every year, seasonal influenza infects between 10\% and 20\% of the global population and influenza subtype A/H3N2 is responsible for the bulk of human morbidity and mortality \citep{flufactsheet}.
+Every year, seasonal influenza infects between 10\% and 20\% of the global population resulting in substantial human morbidity and mortality \citep{flufactsheet}.
 Vaccination remains the most effective public health measure to combat seasonal epidemics.
@@
 In any given year, the particular choice of vaccine strain plays a major role in determining vaccine efficacy and so it is of critical importance to develop tools to analyze the ongoing evolution of the influenza virus population in order to aid vaccine strain selection.
-The program nextflu presents a near real-time display of genetic relationships among influenza H3N2 viruses and allows investigation of currently available sequence data.
+The program \nextflu{} presents a near real-time display of genetic relationships among influenza viruses and allows investigation of currently available sequence data.
 
@@
 	\includegraphics[width=0.99\columnwidth]{figures/tree}
-\caption[]{The \texttt{nextflu} website with the user interface on the left and
+\caption[]{The \nextflu{} website with the user interface on the left and
 the tree on the right.}
@@
 
-In implementation, nextflu consists of a processing pipeline written in Python called \textit{augur} that takes sequence data and infers evolutionary information and a JavaScript-based browser visualization called \textit{auspice} that displays this processed information.
-As input, \textit{augur} requires a FASTA file of sequences with FASTA header labels containing relevant information such as the strain name, the sampling date and passage history.
-For this purpose, influenza H3N2 sequence data for the hemagglutinin (HA) gene is downloaded from the GISAID EpiFlu database \citep{GISAID}, which contains the most up-to-date collection of seasonal influenza viruses.
+In implementation, \nextflu{} consists of a processing pipeline written in Python called \augur{} that analyzes virus sequence data and a JavaScript-based browser visualization called \auspice{} that displays this processed information.
+As input, \augur{} requires a FASTA file of sequences with FASTA header labels containing relevant information such as the strain name, the sampling date and passage history.
+For this purpose, influenza sequence data for the hemagglutinin (HA) gene is downloaded from the GISAID EpiFlu database \citep{GISAID}, which contains the most up-to-date collection of seasonal influenza viruses.
 The first step in the processing pipeline is to automatically select a subset of representative H3N2 viruses.
@@
 
-From the filtered and cleaned alignment, \textit{augur} builds a phylogenetic tree using FastTree \citep{price_fasttree_2009}, which is then further refined using RAxML \citep{stamatakis_raxml_2014}.
+From the filtered and cleaned alignment, \augur{} builds a phylogenetic tree using FastTree \citep{price_fasttree_2009}, which is then further refined using RAxML \citep{stamatakis_raxml_2014}.
 Next, the state of every internal node of the tree is inferred using marginal maximum likelihood method and missing sequence is filled with the nearest ancestral sequence.
@@
 Frequency trajectories are represented as a linear interpolation $x(t)$ of values $x_i$ at time points $t_i$ spaced one month apart. 
-The pivot frequencies corresponding to a feature $\phi$ (e.g. mutation) are estimated by minimizing
+The pivot frequencies corresponding to a feature $\phi$ (e.g. mutation) are estimated by maximizing
 \begin{equation}
@@
 	\begin{split}
-	-\log LH(\{x_i\} | \phi)  =& \sum_v I(v,\phi)\log(x(t_v)) + (1-I(v,\phi))\log(1-x(t_v)) \\
-			&+\sum_i \frac{(\Delta x_i - \epsilon\Delta x_{i-1})^2}{2\gamma \delta t_i}
+	\prod_{v\in V_\phi}x(t_v)\prod_{v\notin V_{\phi}}(1-x(t_v))\times e^{ 
+			-\sum_i \frac{(\Delta x_i - \epsilon\Delta x_{i-1})^2}{2\gamma \delta t_i}
+			}
 \end{split}
 \end{equation}
-where $I(v,\phi)=1$ if virus $v$ carries attribute $\phi$ and $I(v,\phi)=0$ otherwise.
-$\Delta x_i = x_i-x_{i-1}$, $\Delta t_i = t_i-t_{i-1}$, $\gamma$ parameterizes the smoothing imposed on the frequency estimate (related to the frequency changes by genetic drift) and $\epsilon$ parameterizes our expectation of frequency changes in interval $i$ based on the observed change in interval $i-1$.
-$\epsilon=0$ implies that frequency changes are uncorrelated from one interval to the next, while $\epsilon=1$ implies that
-the most likely $\Delta x_i$ is $\Delta x_{i-1}$.
+where $V_{\phi}$ is the set of viruses carring attribute $\phi$, $\Delta x_i = x_i-x_{i-1}$, $\Delta t_i = t_i-t_{i-1}$, $\gamma$ parameterizes the smoothing imposed on the frequency estimate and $\epsilon$ parameterizes our expectation of frequency changes in interval $i$ based on the observed change in interval $i-1$.
+$\epsilon=0$ implies that frequency changes are uncorrelated from one interval to the next, while $\epsilon=1$ implies that the most likely $\Delta x_i$ is $\Delta x_{i-1}$.
 We typically use $\epsilon = 0.7$.
-\EQ{freq} is minimized using SciPy's implementation of a downhill simplex algorithm \citep{Oliphant:2007p25672}.
-To speed up convergence, minization is first done using a coarse grid of pivot points which is
-subsequently refined to 12 pivots per year.
-PUT IN DETAILS ON HOW WE ESTIMATE FREQUENCIES ON THE TREE?
+The logarithm of \EQ{freq} is maximized using SciPy's implementation of a downhill simplex algorithm \citep{Oliphant:2007p25672}.
+To speed up convergence, optimization is first done using a coarse grid of pivot points which is subsequently refined to 12 pivots per year.
+Frequency estimation is done for mutations, particular genotypes, and major clades of the tree.
 
-The tree decorated with attributes and clade frequencies is written to file in a nested json format for display in the browser by auspice.
+The tree decorated with attributes and clade frequencies is written to file in a nested json format for display in the browser by \auspice.
 In addition to the tree, json files containing the sequences, frequencies of mutations and genotypes, as well as meta information are exported.
 
-The tree is visualized by auspice using d3 \citep{bostock_d3_2011}.
-The user can explore the data interactively by selecting viruses from different dates (moving the date slider, top right corner in \FIG{tree}) or by coloring the tree by the following attributes:
+The tree is visualized by \auspice{} using d3 \citep{bostock_d3_2011}.
+The user can explore the data interactively by selecting viruses from different dates (moving the date slider, top right corner in \FIG{tree}) or by coloring the tree by attributes such as:
 \begin{itemize}
-	\item epitope mutations: each tip is colored by the number of aminoacid mutations at \~50 positions relative to the root of the tree.
-These positions are known to be frequent targets of antibodies \citep{shih_simultaneous_2007} and have been suggested to be predictive of future success \citep{luksza_predictive_2014}.
-	\item non-epitope mutations: color by the number of mutations outside the above mentioned epitope sites.
-	Mutations outside epitopes tend to be damaging and anti-correlated with clade expansion \citep{luksza_predictive_2014}.
-    \item receptor binding mutations: color by the number of mutations at 7 positions close to the receptor binding site that have been shown to be responsible for major antigenic positions in the past decades \citep{koel_substitutions_2013}.
+	\item epitope mutations: each tip is colored by the number of aminoacid mutations at antibody epitopes ($\sim 50$ sites) relative to the root of the tree.
+Evolution at epitopes has been suggested to be predictive of future success \citep{luksza_predictive_2014}.
+%	\item non-epitope mutations: 
+%	Mutations outside epitopes tend to be damaging and anti-correlated with clade expansion \citep{luksza_predictive_2014}.
+    \item receptor binding mutations: Mutations at seven positions close to the receptor binding site have been shown to be responsible for major antigenic positions in the past decades \citep{koel_substitutions_2013}.
     \item local branching index is the exponentially weighted tree length surrounding a node, which is associated with rapid branching and expansion of clades \citep{neher_predicting_2014}.
-It is recalculated everytime the date slider is moved and based only on the highlighted viruses.
-    \item geographic location
-    \item HA1 genotype: entering amino acid positions (separated by a comma) in the text field will color every node according to the genotype at these positions.
+ %   \item geographic location
+    \item HA genotype: enter amino acid positions to color every node according to the genotype at these positions.
 \end{itemize}
@@
 The frequency plot below the tree (see \FIG{freq}) displays the frequency trajectory of clades in the tree whenever the mouse hovers above the branch defining the clade. 
-Furthermore, trajectories of individual mutations, combinations of two mutations, and predifined clades such as 3c3.a can be plotted by entering them in text field.
+Furthermore, trajectories of individual mutations, combinations of two mutations, and predefined clades such as \textit{3c3.a} can be plotted by entering them in text field.
 
-\begin{figure}[bhtp]
+\begin{figure}[t]
 	\begin{center}
@@
 
-\section{Conclusion}
-We built nextflu.org to facilitate the analysis and exploration of seasonal influenza sequence data collected by laboratories around to world.
-By using the most recent data and integrating phylogenies with frequency trajectories and sequence or tree based predictors of successful clades, we hope that nextflu can inform the choice of strains used in seasonal influenza vaccines.
-
-nextflu was designed to be readily adapted to other rapidly evolving viruses.
+%\section{Conclusion}
+We built \nextflu{} to facilitate the analysis and exploration of seasonal influenza sequence data collected by laboratories around to world.
+By using the most recent data and integrating phylogenies with frequency trajectories and sequence or tree based predictors of successful clades, we hope that \nextflu{} can inform the choice of strains used in seasonal influenza vaccines.
+\nextflu{} was designed to be readily adapted to other rapidly evolving viruses.
 To that end, the generic classes used for processing the sequence data to include features specific to the virus in question.

### 2015-04-03 auspice/js/auspice.js
adjusted legend text for frequency change

@@ -628,7 +628,12 @@ d3.json("/data/" + file_prefix + "tree.json", function(error, root) {
     			return "Genotype"
     		}
    			if (colorBy == "dfreq") {
-    			return "Frequency change (prev. "+Math.round(12*dfreq_dn*dt)+" month)";
+   				var tmp_nmonth = Math.round(12*dfreq_dn*dt);
+   				var tmp_text = "Freq. change ("+tmp_nmonth+" month";
+   				if (tmp_nmonth>1){
+   					tmp_text+='s';
+   				}
+    			return tmp_text+')';
     		}
     	});
     

### 2015-04-03 auspice/js/H1N1pdm.js
Simple fix for ticks on the frequency plot x axis.

@@ -7,3 +7,5 @@ vaccineChoice['A/California/07/2009'] = "2009-09-25";
 var epiColorDomain = [0,1,2,3, 4,5,6,7,8,9,10];
 var nonEpiColorDomain = [0,1,2,3,4,5,6,7,8,9,10,11];
 var rbsColorDomain = [0,0.5, 1,1.5, 2];
+
+var time_ticks = [2010, 2011, 2012, 2013, 2014, 2015];

### 2015-04-03 auspice/js/H3N2.js

@@ -11,7 +11,8 @@ vaccineChoice['A/Victoria/361/2011'] = "2012-02-21";
 vaccineChoice['A/Texas/50/2012'] = "2013-09-25";
 vaccineChoice['A/Switzerland/9715293/2013'] = "2014-09-25";
 
-
 var epiColorDomain = [4,5,6,7,8,9,10,11,12,13];
 var nonEpiColorDomain = [2,3,4,5,6,7,8,9,10,11];
 var rbsColorDomain = [0,1,2,3,4];
+
+var time_ticks = [2012.5, 2013, 2013.5, 2014, 2014.5, 2015];

### 2015-04-03 auspice/js/Vic.js

@@ -10,3 +10,5 @@ vaccineChoice['B/Brisbane/60/2008'] = "2009-09-25";
 var epiColorDomain = [0,1,2,3, 4,5,6,7,8,9,10];
 var nonEpiColorDomain = [0,1,2,3,4,5,6,7,8,9,10,11];
 var rbsColorDomain = [0,0.5, 1,1.5, 2];
+
+var time_ticks = [2010, 2011, 2012, 2013, 2014, 2015];

### 2015-04-03 auspice/js/Yam.js

@@ -11,3 +11,5 @@ vaccineChoice['B/PHUKET/3073/2013'] = "2015-03-25";
 var epiColorDomain = [0,1,2,3, 4,5,6,7,8,9,10];
 var nonEpiColorDomain = [0,1,2,3,4,5,6,7,8,9,10,11];
 var rbsColorDomain = [0,0.5, 1,1.5, 2];
+
+var time_ticks = [2010, 2011, 2012, 2013, 2014, 2015];

### 2015-04-03 auspice/js/auspice.js

@@ -352,7 +352,7 @@ var gt_chart = c3.generate({
 				position: 'outer-center'	
 			},
 			tick: {
-				//values: [2012,2012.5,2013,2013.5,2014,2014.5,2015],
+				values: time_ticks,
 				outer: false				
 			}
 		}

### 2015-04-07 auspice/js/auspice.js
* export entropy intro frequency Json * add bar graph to auspice * very rough!!

@@ -1324,6 +1324,63 @@ d3.json("/data/" + file_prefix + "frequencies.json", function(error, json){
 		});
 	}
 
+	var ent = [['x'],['']];
+	for (var ii=0;ii<json["entropy"].length;ii+=1){
+		if (Math.round(10000*json["entropy"][ii][1])/10000>0){
+			ent[1].push(Math.round(10000*json["entropy"][ii][1])/10000);
+			ent[0].push(ii+1);
+		}
+	}
+
+	
+	var entropy_chart = c3.generate({
+		bindto: '#entropy',
+		size: {width: width, height: 350},
+		legend: {position: position},
+		axis: {
+			y: {
+				label: {
+					text: 'entropy',
+					position: 'outer-middle'	
+				},
+			},
+			x: {
+				label: {
+					text: 'position',
+					position: 'outer-center'	
+				},
+				tick: {
+					outer: false				
+				}
+			}
+		},			
+		data: {
+			x: 'x',
+			columns: ent,
+			type: "bar",
+			onclick: function (d,i) { 
+						console.log(d);
+		            	if (json["entropy"][d.x-1][2].length>1){
+		            		var tmp = [];
+		            		for (var ii=0;ii<json["entropy"][d.x-1][2].length;ii+=1){
+								tmp.push(["global",d.x+json["entropy"][d.x-1][2][ii]]);
+		            		}
+		            		console.log(tmp);
+		            		make_gt_chart(tmp);
+		            	}
+		            }
+		},
+	    tooltip: {
+	        format: {
+	            title: function (d) { 
+	            	return 'Position ' + d + json["entropy"][d-1][2].join(","); },
+	            value: function (value, ratio, id) {
+	                return "Entropy: "+value;
+	            }
+	        }
+		},
+	});
+	
 	d3.select("#plotfreq")
 		.on("click", function (){
 			gt = parse_gt_string(document.getElementById("gtspec").value);			

### 2015-04-07 augur/src/process.py
* export entropy intro frequency Json * add bar graph to auspice * very rough!!

@@ -8,6 +8,7 @@ import dendropy
 from bernoulli_frequency import virus_frequencies
 from tree_util import delimit_newick
 import numpy as np
+from itertools import izip
 
 parser = argparse.ArgumentParser(description='Process virus sequences, build tree, and prepare of web visualization')
 parser.add_argument('-y', '--years_back', type = float, default=3, help='number of past years to sample sequences from')
@@ -158,6 +159,12 @@ class process(virus_frequencies):
 			
 		# Include genotype frequencies
 		if hasattr(self, 'frequencies'):
+			if not hasattr(self, 'aa_entropy'):
+				self.determine_variable_positions()
+
+			self.frequencies["entropy"] = [ [pos, S, muts] for pos,S,muts in 
+					izip(xrange(self.aa_entropy.shape[0]), self.aa_entropy,self.variable_aa_identities) ]
+
 			write_json(self.frequencies, self.auspice_frequency_fname)
 
 		# Write out metadata
@@ -308,6 +315,10 @@ class process(virus_frequencies):
 
 			self.variable_aa = np.where(np.max(self.aa_frequencies,axis=0)<1.0-self.min_mutation_frequency)[0]
 			self.consensus_aa = "".join(np.fromstring(self.aa_alphabet, 'S1')[np.argmax(self.aa_frequencies,axis=0)])
+			self.aa_entropy = -np.sum(self.aa_frequencies*np.log(np.maximum(1e-10,self.aa_frequencies)), axis=0)
+			self.variable_aa_identities = [ [self.aa_alphabet[ii] for ii in np.where(self.aa_frequencies[:,pos])[0]]
+											for pos in xrange(self.aa_frequencies.shape[1])]
+
 
 	def estimate_frequencies(self, tasks = ['mutations','genotypes', 'clades', 'tree']):
 		if 'mutations' in tasks:

### 2015-04-07 auspice/js/auspice.js
colors the tree by genotype on click on bargraph at a particular position

@@ -327,6 +327,7 @@ var virusTooltip = d3.tip()
 		return string;
 	});
 treeplot.call(virusTooltip);
+var cbg;
 
 var linkTooltip = d3.tip()
 	.direction('e')
@@ -1170,7 +1171,8 @@ d3.json("/data/" + file_prefix + "tree.json", function(error, root) {
 		}
 		tree_legend = makeLegend();
 	}
-	
+
+	cbg = colorByGenotypePosition;	
 	d3.select("#coloring")
 		.style("cursor", "pointer")
 		.on("change", colorByTrait);
@@ -1367,6 +1369,7 @@ d3.json("/data/" + file_prefix + "frequencies.json", function(error, json){
 		            		}
 		            		console.log(tmp);
 		            		make_gt_chart(tmp);
+		            		cbg([d.x-1]);
 		            	}
 		            }
 		},

### 2015-04-07 manuscript/nextflu.tex
Revise manuscript.
I've broken up the frequency trajectory equations in an attempt to
make it more digestible.

@@
 \newcommand{\REF}[1]{ref.~\citep{#1}}
-\newcommand{\augur}{\textit{augur}}
-\newcommand{\auspice}{\textit{auspice}}
-\newcommand{\nextflu}{\textit{nextflu}}
+\newcommand{\augur}{\textbf{augur}}
+\newcommand{\auspice}{\textbf{auspice}}
+\newcommand{\nextflu}{\textbf{nextflu}}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{abstract} \section{Summary:} Seasonal influenza viruses evolve rapidly, which allows the virus population to evade immunity in their human hosts and reinfect previously infected individuals.
+\begin{abstract} \section{Summary:} Seasonal influenza viruses evolve rapidly, allowing them to evade immunity in their human hosts and reinfect previously infected individuals.
 Similarly, vaccines against seasonal influenza need to be updated frequently to protect against an evolving virus population.
-We have thus developed a processing pipeline and web-browser based visualization that allows convenient exploration and analysis of the most recent influenza virus sequence data.
-This web-application displays a phylogenetic tree that can be decorated with additional information such the viral genotype at specific sites, sampling location, and sequence or tree based statistics that have been shown to predictive of future virus dynamics.
+We have thus developed a processing pipeline and browser-based visualization that allows convenient exploration and analysis of the most recent influenza virus sequence data.
+This web-application displays a phylogenetic tree that can be decorated with additional information such the viral genotype at specific sites, sampling location, and derived statistics that have been shown to predictive of future virus dynamics.
 Additionally, mutation, genotype and clade frequency trajectories are calculated and displayed.
@@
 \caption[]{The \nextflu{} website with the user interface on the left and
-the tree on the right.}
+the phylogenetic tree on the right.}
 \label{fig:tree}
@@
 For our standard display period of 3 years and 50 viruses per month, this typically results in $\sim$1800 viruses, which we align using MAFFT \citep{katoh_mafft_2013}.
-Once aligned, the set of virus sequences is further cleaned by removing insertions relative to the outgroup to enforce canonical HA site numbering and by removing sequences that show either too much or too little divergence relative to the expectation given data of sampling.
+Once aligned, the set of virus sequences is further cleaned by removing insertions relative to the outgroup to enforce canonical HA site numbering and by removing sequences that show either too much or too little divergence relative to the expectation given sampling date.
 Additionally, known reassortant clusters are removed, including the triple-reassortant influenza viruses sporadically sampled in humans throughout North America \citep{bastien_human_2010}.
@@
 From the filtered and cleaned alignment, \augur{} builds a phylogenetic tree using FastTree \citep{price_fasttree_2009}, which is then further refined using RAxML \citep{stamatakis_raxml_2014}.
-Next, the state of every internal node of the tree is inferred using marginal maximum likelihood method and missing sequence is filled with the nearest ancestral sequence.
+Next, the state of every internal node of the tree is inferred using a marginal maximum likelihood method and missing sequence data at phylogeny tips is filled with the nearest ancestral sequence at these sites.
 Internal branches without mutations are collapsed into polytomies.
-The final tree is decorated with the attributes to be displayed in the browser, see below.
+The final tree is decorated with the attributes to be displayed in the browser, as described below.
 
-In addition to the phylogenetic tree, augur estimates the frequency trajectories of mutations, genotypes and clades in the tree. 
-Frequency trajectories are represented as a linear interpolation $x(t)$ of values $x_i$ at time points $t_i$ spaced one month apart. 
-The pivot frequencies corresponding to a feature $\phi$ (e.g. mutation) are estimated by maximizing
+In addition to the phylogenetic tree, \augur{} estimates the frequency trajectories of mutations, genotypes and clades in the tree. 
+Frequency trajectories are represented as a linear interpolation $x(t)$ of pivot frequencies $x_i$ at time points $t_i$ spaced one month apart. 
+The frequencies $x_i$ corresponding to a feature $\phi$ (e.g. mutation) are estimated by simultaneously maximizing the Bernoulli observation model 
+\begin{equation}
+	\label{eq:obs}
+	f(\mathbf{x}) = \prod_{v\in V_\phi} x(t_v) \prod_{v\notin V_{\phi}} (1-x(t_v))
+\end{equation}
+where $V_{\phi}$ is the set of viruses carring attribute $\phi$ and $x(t_v)$ is the sampling timepoint of virus $v$, and the process noise model
 \begin{equation}
 	\label{eq:freq}
-	\begin{split}
-	\prod_{v\in V_\phi}x(t_v)\prod_{v\notin V_{\phi}}(1-x(t_v))\times e^{ 
+	g(\mathbf{x}) = \mathrm{exp}\left(
 		-\sum_i \frac{(\Delta x_i - \epsilon\Delta x_{i-1})^2}{2\gamma \delta t_i}
-			}
-\end{split}
+	\right)
 \end{equation}
-where $V_{\phi}$ is the set of viruses carring attribute $\phi$, $\Delta x_i = x_i-x_{i-1}$, $\Delta t_i = t_i-t_{i-1}$, $\gamma$ parameterizes the smoothing imposed on the frequency estimate and $\epsilon$ parameterizes our expectation of frequency changes in interval $i$ based on the observed change in interval $i-1$.
+where $\Delta x_i = x_i-x_{i-1}$, $\Delta t_i = t_i-t_{i-1}$, $\gamma$ parameterizes the smoothing imposed on the frequency estimate and $\epsilon$ parameterizes our expectation of frequency changes in interval $i$ based on the observed change in interval $i-1$.
 $\epsilon=0$ implies that frequency changes are uncorrelated from one interval to the next, while $\epsilon=1$ implies that the most likely $\Delta x_i$ is $\Delta x_{i-1}$.
 We typically use $\epsilon = 0.7$.
-The logarithm of \EQ{freq} is maximized using SciPy's implementation of a downhill simplex algorithm \citep{Oliphant:2007p25672}.
+Here, the objective function $\mathrm{log}(f) + \mathrm{log}(g)$ is maximized using SciPy's implementation of a downhill simplex algorithm \citep{oliphant_python_2007}.
 To speed up convergence, optimization is first done using a coarse grid of pivot points which is subsequently refined to 12 pivots per year.
 Frequency estimation is done for mutations, particular genotypes, and major clades of the tree.
+At the end of the \augur{} pipeline, JSON files are exported containing the annotated phylogenetic tree, sequence data and frequency trajectories.
 
-The tree decorated with attributes and clade frequencies is written to file in a nested json format for display in the browser by \auspice.
-In addition to the tree, json files containing the sequences, frequencies of mutations and genotypes, as well as meta information are exported.
-
-The tree is visualized by \auspice{} using d3 \citep{bostock_d3_2011}.
-The user can explore the data interactively by selecting viruses from different dates (moving the date slider, top right corner in \FIG{tree}) or by coloring the tree by attributes such as:
+These JSON files are then visualized by \auspice{} using D$^3$ \citep{bostock_d3_2011} and the phylogenetic tree displayed (\FIG{tree}).
+The user can explore the data interactively by selecting viruses from different dates or by coloring the tree by attributes such as:
 \begin{itemize}
-	\item epitope mutations: each tip is colored by the number of aminoacid mutations at antibody epitopes ($\sim 50$ sites) relative to the root of the tree.
-Evolution at epitopes has been suggested to be predictive of future success \citep{luksza_predictive_2014}.
+	\item \textit{epitope mutations} at sites generally associated with antibody binding have been suggested to be predictive of future clade success \citep{luksza_predictive_2014},
 %	\item non-epitope mutations: 
 %	Mutations outside epitopes tend to be damaging and anti-correlated with clade expansion \citep{luksza_predictive_2014}.
-    \item receptor binding mutations: Mutations at seven positions close to the receptor binding site have been shown to be responsible for major antigenic positions in the past decades \citep{koel_substitutions_2013}.
-    \item local branching index is the exponentially weighted tree length surrounding a node, which is associated with rapid branching and expansion of clades \citep{neher_predicting_2014}.
+    \item \textit{receptor binding mutations} at seven positions close to the receptor binding site have been shown to be responsible for major antigenic transitions in the past decades \citep{koel_substitutions_2013},
+    \item \textit{local branching index} is the exponentially weighted tree length surrounding a node, which is associated with rapid branching and expansion of clades \citep{neher_predicting_2014},
  %   \item geographic location
-    \item HA genotype: enter amino acid positions to color every node according to the genotype at these positions.
+    \item \textit{HA genotype} directly colors the tree by genotype at specific amino acid positions.
 \end{itemize}
 
-The frequency plot below the tree (see \FIG{freq}) displays the frequency trajectory of clades in the tree whenever the mouse hovers above the branch defining the clade. 
-Furthermore, trajectories of individual mutations, combinations of two mutations, and predefined clades such as \textit{3c3.a} can be plotted by entering them in text field.
+The frequency plot below the tree (\FIG{freq}) displays the frequency trajectory of clades in the tree whenever the mouse hovers above the branch defining the clade. 
+Furthermore, trajectories of individual mutations, combinations of two mutations, and predefined clades such as 3c3.a can be plotted.
 
@@
 	\includegraphics[width=0.99\columnwidth]{figures/frequencies}
-\caption[]{The frequency diagram on \url{nextflu.org} allows geography specific
-		plotting of frequencies of individual mutations, pairs of mutations, 
-		predefined genotypes, as well as clades in the tree. }
+\caption[]{The frequency diagram allows geography-specific plotting of frequencies of individual mutations, pairs of mutations and clades in the tree.}
 \label{fig:freq}
@@
 We built \nextflu{} to facilitate the analysis and exploration of seasonal influenza sequence data collected by laboratories around to world.
-By using the most recent data and integrating phylogenies with frequency trajectories and sequence or tree based predictors of successful clades, we hope that \nextflu{} can inform the choice of strains used in seasonal influenza vaccines.
+By using the most recent data and integrating phylogenies with frequency trajectories and predictors of successful clades, we hope that \nextflu{} can inform the choice of strains used in seasonal influenza vaccines.
 \nextflu{} was designed to be readily adapted to other rapidly evolving viruses.
-To that end, the generic classes used for processing the sequence data to include features specific to the virus in question.
 
-\paragraph{Funding\textcolon}This work is supported by the ERC though Stg-260686.
+\paragraph{Funding\textcolon}This work is supported by the ERC though Stg-260686 and by the NIH through U54 GM111274.
 

### 2015-04-07 manuscript/nextflu.tex
Fix references to conform to Bioinformatics et al. format.

@@
 \begin{figure}[t!]
-	\begin{center}
+	\centering
 	\includegraphics[width=0.99\columnwidth]{figures/tree}
-\caption[]{The \nextflu{} website with the user interface on the left and
-the phylogenetic tree on the right.}
+	\caption[]{The \nextflu{} website with the user interface on the left and the phylogenetic tree on the right.}
 	\label{fig:tree}
-\end{center}
 \end{figure}
@@
 \begin{figure}[t]
-	\begin{center}
+	\centering
 	\includegraphics[width=0.99\columnwidth]{figures/frequencies}
@@
 	\label{fig:freq}
-\end{center}
 \end{figure}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\bibliography{nextflu}
+\begin{thebibliography}{}
+
+\bibitem[Bastien {\em et~al.}, 2010]{bastien_human_2010}
+Bastien,N. {\em et~al.} (2010{\em{}})
+  Human infection with a triple-reassortant swine influenza {A (H1N1)} virus
+  containing the hemagglutinin and neuraminidase genes of seasonal influenza
+  virus.
+\newblock {\em J Infect Dis, } {\bf 201} (8), 1178--1182.
+
+\bibitem[Bogner {\em et~al.}, 2006]{GISAID}
+Bogner,P. {\em et~al.} (2006{\em{}}) A global
+  initiative on sharing avian flu data.
+\newblock {\em Nature, } {\bf 442} (7106), 981--981.
+
+\bibitem[Bostock {\em et~al.}, 2011]{bostock_d3_2011}
+Bostock,M. {\em et~al.} (2011{\em{}}) D$^3$: data-driven
+  documents.
+\newblock {\em IEEE Trans Vis Comput Graphics, } {\bf 17} (12), 2301--2309.
+
+\bibitem[Katoh and Standley, 2013]{katoh_mafft_2013}
+Katoh,K. and Standley,D.M. (2013{\em{}}) {MAFFT} multiple sequence alignment
+  software version 7: improvements in performance and usability.
+\newblock {\em Mol Biol Evol, } {\bf 30}, 772--780.
+
+\bibitem[Koel {\em et~al.}, 2013]{koel_substitutions_2013}
+Koel,B.F. {\em et~al.} (2013{\em{}})
+  Substitutions near the receptor binding site determine major antigenic change
+  during influenza virus evolution.
+\newblock {\em Science, } {\bf 342} (6161), 976--979.
+
+\bibitem[\L{}uksza and L\"a{}ssig, 2014]{luksza_predictive_2014}
+\L{}uksza,M. and L\"a{}ssig,M. (2014{\em{}}) A predictive fitness model for
+  influenza.
+\newblock {\em Nature, } {\bf 507} (7490), 57--61.
+
+\bibitem[Neher {\em et~al.}, 2014]{neher_predicting_2014}
+Neher,R.A. {\em et~al.} (2014{\em{}}) Predicting evolution
+  from the shape of genealogical trees.
+\newblock {\em eLife Sciences, } {\bf 3}, e03568.
+
+\bibitem[Oliphant, 2007]{oliphant_python_2007}
+Oliphant,T. (2007{\em{}}) Python for scientific computing.
+\newblock {\em Comput Sci Eng, } {\bf 9} (3), 10--20.
+
+\bibitem[Price {\em et~al.}, 2009]{price_fasttree_2009}
+Price,M.N. {\em et~al.} (2009{\em{}}) {FastTree:} computing
+  large minimum evolution trees with profiles instead of a distance matrix.
+\newblock {\em Mol Biol Evol, } {\bf 26} (7), 1641--50.
+
+\bibitem[Stamatakis, 2014]{stamatakis_raxml_2014}
+Stamatakis,A. (2014{\em{}}) {RAxML} version 8: a tool for phylogenetic analysis
+  and post-analysis of large phylogenies.
+\newblock {\em Bioinformatics, } {\bf 30} (9), 1312--1313.
+
+\bibitem[{World Health Organization}, 2009]{flufactsheet}
+{World Health Organization} (2009{\em{}}) {\em {Influenza Fact sheet}}.
+\newblock Available at http://www.who.int/mediacentre/factsheets/fs211/en/.
+
+\end{thebibliography}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2015-04-08 manuscript/correspondence/cover_letter.tex
Include Bioinformatics cover letter.

\documentclass[stdletter,letterpaper,addrfromleft,orderfromdateto,dateleft,11pt,noaddrto,sigleft]{newlfm}
\topmarginskip{-0.4in}
\bottommarginskip{-1.5in}
\leftmarginsize{1in}
\rightmarginsize{1.25in}
\sigskipbefore{0.2in}
\sigskipafter{0in}
\noLines
\nolines
\noHeadline
\noheadline
\signature{Trevor Bedford and Richard Neher}

\usepackage{graphicx}
\usepackage{hyperref}

\namefrom{}
\addrfrom{
	\includegraphics[width=6.5cm]{figures/fhcrc_logo} \\ 
	Vaccine and Infectious Disease Division \\ Fred Hutchinson Cancer Research Center \\ Seattle, WA, USA \\
	\\
	\includegraphics[width=4cm]{figures/max_planck_logo} \\ 
	Max Planck Institute for Developmental Biology \\ T\"ubingen, Germany}

\greetto{Dear Editor,}
\closeline{Sincerely,}

% comments
\usepackage{color} 
\usepackage{ulem}
\definecolor{purple}{rgb}{0.459,0.109,0.538}
\def\tb#1#2{\sout{#1} \textcolor{purple}{#2}} 
\def\tbc#1{\textcolor{purple}{[#1]}}

 \newcommand{\nextflu}{\textbf{nextflu}}

\begin{document}

\begin{newlfm}

Please find, attached, our application note entitled ``nextflu: Real-time tracking of seasonal influenza virus evolution in humans,'' which we would be grateful for your consideration for publication in \textit{Bioinformatics}.
We believe that \textit{Bioinformatics} is an appropriate venue to showcase our software application, as \nextflu{} represents a valuable contribution to public health bioinformatics as it allows better understanding of circulating viruses in an extremely timely fashion.
Source code to \nextflu{} is freely available from \url{https://github.com/blab/nextflu}, while a live web application is publicly available at \url{http://nextflu.org}.
At this point, we've received positive encouragement that \nextflu{} will be useful to several National Influenza Centers and WHO Collaborating Centers in aiding surveillance and strain selection efforts.
As such, we believe that \nextflu{} will see substantial adoption by the influenza research community.

\end{newlfm}
\end{document}

### 2015-04-08 manuscript/nextflu.tex
tiny revisions

@@
 We have thus developed a processing pipeline and browser-based visualization that allows convenient exploration and analysis of the most recent influenza virus sequence data.
-This web-application displays a phylogenetic tree that can be decorated with additional information such the viral genotype at specific sites, sampling location, and derived statistics that have been shown to predictive of future virus dynamics.
+This web-application displays a phylogenetic tree that can be decorated with additional information such the viral genotype at specific sites, sampling location, and derived statistics that have been shown to be predictive of future virus dynamics.
 Additionally, mutation, genotype and clade frequency trajectories are calculated and displayed.
@@
 For this purpose, influenza sequence data for the hemagglutinin (HA) gene is downloaded from the GISAID EpiFlu database \citep{GISAID}, which contains the most up-to-date collection of seasonal influenza viruses.
-The first step in the processing pipeline is to automatically select a subset of representative H3N2 viruses.
+The first step in the processing pipeline is to automatically select a subset of representative viruses.
 Here, viruses without complete date or geographic information, viruses passaged in eggs and sequences less than 987 bases are removed.
@@
 Internal branches without mutations are collapsed into polytomies.
-The final tree is decorated with the attributes to be displayed in the browser, as described below.
+The final tree is decorated with the attributes to be displayed in the browser. %, as described below.
 
@@
 	g(\mathbf{x}) = \mathrm{exp}\left(
-		-\sum_i \frac{(\Delta x_i - \epsilon\Delta x_{i-1})^2}{2\gamma \delta t_i}
+		-\sum_i \frac{(\Delta x_i - \epsilon\Delta x_{i-1})^2}{2\gamma \Delta t_i}
 	\right)
 \end{equation}
-where $\Delta x_i = x_i-x_{i-1}$, $\Delta t_i = t_i-t_{i-1}$, $\gamma$ parameterizes the smoothing imposed on the frequency estimate and $\epsilon$ parameterizes our expectation of frequency changes in interval $i$ based on the observed change in interval $i-1$.
+where $\Delta x_i = x_i-x_{i-1}$, $\Delta t_i = t_i-t_{i-1}$, $\gamma$ parameterizes the smoothing imposed on the frequency estimate and $\epsilon$ parameterizes our expectation of frequency changes in interval $i$ based on the  change in interval $i-1$.
 $\epsilon=0$ implies that frequency changes are uncorrelated from one interval to the next, while $\epsilon=1$ implies that the most likely $\Delta x_i$ is $\Delta x_{i-1}$.
@@
 By using the most recent data and integrating phylogenies with frequency trajectories and predictors of successful clades, we hope that \nextflu{} can inform the choice of strains used in seasonal influenza vaccines.
-\nextflu{} was designed to be readily adapted to other rapidly evolving viruses.
+\nextflu{} was designed to be readily adapted to other rapidly evolving viruses and we have implemented pipelines for all four seasonal influenza viruses.
 

### 2015-04-09 manuscript/nextflu.tex
Minor revisions and swap screenshots.

@@
 The program \nextflu{} presents a near real-time display of genetic relationships among influenza viruses and allows investigation of currently available sequence data.
+Currently, \nextflu{} tracks all four circulating lineages of seasonal influenza: A/H3N2, A/H1N1pdm, B/Victoria and B/Yamagata.
 
@@
 For our standard display period of 3 years and 50 viruses per month, this typically results in $\sim$1800 viruses, which we align using MAFFT \citep{katoh_mafft_2013}.
-Once aligned, the set of virus sequences is further cleaned by removing insertions relative to the outgroup to enforce canonical HA site numbering and by removing sequences that show either too much or too little divergence relative to the expectation given sampling date.
-Additionally, known reassortant clusters are removed, including the triple-reassortant influenza viruses sporadically sampled in humans throughout North America \citep{bastien_human_2010}.
+Once aligned, the set of virus sequences is further cleaned by removing insertions relative to the outgroup to enforce canonical HA site numbering, by removing sequences that show either too much or too little divergence relative to the expectation given sampling date, and by removing known reassortant clusters.
 
@@
 Frequency trajectories are represented as a linear interpolation $x(t)$ of pivot frequencies $x_i$ at time points $t_i$ spaced one month apart. 
-The frequencies $x_i$ corresponding to a feature $\phi$ (e.g. mutation) are estimated by simultaneously maximizing the Bernoulli observation model 
+The frequencies $x_i$ corresponding to a feature $\phi$ (e.g.\ mutation) are estimated by simultaneously maximizing the Bernoulli observation model 
 \begin{equation}
@@
 
-The frequency plot below the tree (\FIG{freq}) displays the frequency trajectory of clades in the tree whenever the mouse hovers above the branch defining the clade. 
-Furthermore, trajectories of individual mutations, combinations of two mutations, and predefined clades such as 3c3.a can be plotted.
-
 \begin{figure}[t]
@@
 
+The frequency plot below the tree (\FIG{freq}) displays the frequency trajectory of clades in the tree whenever the mouse hovers above the branch defining the clade. 
+Furthermore, trajectories of individual mutations, combinations of two mutations, and predefined clades such as 3c3.a can be plotted.
+
 %\section{Conclusion}
@@
 By using the most recent data and integrating phylogenies with frequency trajectories and predictors of successful clades, we hope that \nextflu{} can inform the choice of strains used in seasonal influenza vaccines.
-\nextflu{} was designed to be readily adapted to other rapidly evolving viruses and we have implemented pipelines for all four seasonal influenza viruses.
+\nextflu{} was designed to be readily adapted to other rapidly evolving viruses and we see significant room for future developments in this area.
 
@@
 \begin{thebibliography}{}
-
-\bibitem[Bastien {\em et~al.}, 2010]{bastien_human_2010}
-Bastien,N. {\em et~al.} (2010{\em{}})
-  Human infection with a triple-reassortant swine influenza {A (H1N1)} virus
-  containing the hemagglutinin and neuraminidase genes of seasonal influenza
-  virus.
-\newblock {\em J Infect Dis, } {\bf 201} (8), 1178--1182.
 

### 2015-04-09 augur/src/make_all.py
Include S3 push/pull for FASTA files in make_all script.
Lineage FASTA are stored in the S3 bucket "nextflu-data". With the flag
--s3 the build script will pull lineage FASTAs from the bucket, amend files
with local FASTA and push back to S3. If running on a server, would only
require updating of local fasta file before running build script.

@@ -10,6 +10,40 @@ patterns = {('A / H3N2', ''):'H3N2',
 			('A / H1N1', 'seasonal'):'H1N1',
 			}
 
+def pull_fasta_from_s3(lineages = ['H3N2', 'H1N1pdm', 'Vic', 'Yam'], directory = 'data/', bucket = 'nextflu-data'):
+	"""Retrieve FASTA files from S3 bucket"""
+	"""Boto expects environmental variables AWS_ACCESS_KEY_ID and AWS_SECRET_ACCESS_KEY"""
+	directory = directory.rstrip('/')+'/'
+
+	import boto
+	conn = boto.connect_s3()
+	b = conn.get_bucket(bucket)
+	k = boto.s3.key.Key(b)
+
+	for lineage in lineages:
+		print "Retrieving FASTA for",lineage
+		fasta = lineage+'_gisaid_epiflu_sequence.fasta'
+		k.key = fasta
+		k.get_contents_to_filename(directory+fasta)
+		print fasta,"retrieved"
+		
+def push_fasta_to_s3(lineages= ['H3N2', 'H1N1pdm', 'Vic', 'Yam'], directory = 'data/', bucket = 'nextflu-data'):
+	"""Upload FASTA files to S3 bucket"""
+	"""Boto expects environmental variables AWS_ACCESS_KEY_ID and AWS_SECRET_ACCESS_KEY"""
+	directory = directory.rstrip('/')+'/'
+
+	import boto
+	conn = boto.connect_s3()
+	b = conn.get_bucket(bucket)
+	k = boto.s3.key.Key(b)
+
+	for lineage in lineages:
+		print "Uploading FASTA for",lineage
+		fasta = lineage+'_gisaid_epiflu_sequence.fasta'
+		k.key = fasta
+		k.set_contents_from_filename(directory+fasta)
+		print fasta,"uploaded"
+
 def ammend_files(fname, lineages= ['H3N2', 'H1N1pdm', 'Vic', 'Yam'], threshold = 10, directory = 'data/'):
 	directory = directory.rstrip('/')+'/'
 	updated = []
@@ -58,8 +92,12 @@ if __name__=="__main__":
 	parser.add_argument('--bin', type = str, default = "python")
 	parser.add_argument('--ATG', action = "store_true", default = False, help ="include full HA sequence starting at ATG")
 	parser.add_argument('--all', action = "store_true", default = False)
[…]

### 2015-04-10 augur/src/make_all.py
Improve lineage logic and add theshold option.
I felt like having methods only deal with one lineage at a time was
cleaner.

@@ -9,8 +9,9 @@ patterns = {('A / H3N2', ''):'H3N2',
 			('B / H0N0', 'Yamagata'):'Yam',
 			('A / H1N1', 'seasonal'):'H1N1',
 			}
+lineages = ['H3N2', 'H1N1pdm', 'Vic', 'Yam']
 
-def pull_fasta_from_s3(lineages = ['H3N2', 'H1N1pdm', 'Vic', 'Yam'], directory = 'data/', bucket = 'nextflu-data'):
+def pull_fasta_from_s3(lineage, directory = 'data/', bucket = 'nextflu-data'):
 	"""Retrieve FASTA files from S3 bucket"""
 	"""Boto expects environmental variables AWS_ACCESS_KEY_ID and AWS_SECRET_ACCESS_KEY"""
 	directory = directory.rstrip('/')+'/'
@@ -20,14 +21,13 @@ def pull_fasta_from_s3(lineages = ['H3N2', 'H1N1pdm', 'Vic', 'Yam'], directory =
 	b = conn.get_bucket(bucket)
 	k = boto.s3.key.Key(b)
 
-	for lineage in lineages:
-		print "Retrieving FASTA for",lineage
-		fasta = lineage+'_gisaid_epiflu_sequence.fasta'
-		k.key = fasta
-		k.get_contents_to_filename(directory+fasta)
-		print fasta,"retrieved"
+	print "Retrieving FASTA for",lineage
+	fasta = lineage+'_gisaid_epiflu_sequence.fasta'
+	k.key = fasta
+	k.get_contents_to_filename(directory+fasta)
+	print fasta,"retrieved"
 		
-def push_fasta_to_s3(lineages= ['H3N2', 'H1N1pdm', 'Vic', 'Yam'], directory = 'data/', bucket = 'nextflu-data'):
+def push_fasta_to_s3(lineage, directory = 'data/', bucket = 'nextflu-data'):
 	"""Upload FASTA files to S3 bucket"""
 	"""Boto expects environmental variables AWS_ACCESS_KEY_ID and AWS_SECRET_ACCESS_KEY"""
 	directory = directory.rstrip('/')+'/'
@@ -37,52 +37,50 @@ def push_fasta_to_s3(lineages= ['H3N2', 'H1N1pdm', 'Vic', 'Yam'], directory = 'd
 	b = conn.get_bucket(bucket)
 	k = boto.s3.key.Key(b)
 
-	for lineage in lineages:
[…]

### 2015-04-10 augur/src/make_all.py
Include JSON push to dev.

@@ -26,7 +26,7 @@ def pull_fasta_from_s3(lineage, directory = 'data/', bucket = 'nextflu-data'):
 	k.key = fasta
 	k.get_contents_to_filename(directory+fasta)
 	print fasta,"retrieved"
-		
+
 def push_fasta_to_s3(lineage, directory = 'data/', bucket = 'nextflu-data'):
 	"""Upload FASTA files to S3 bucket"""
 	"""Boto expects environmental variables AWS_ACCESS_KEY_ID and AWS_SECRET_ACCESS_KEY"""
@@ -42,6 +42,23 @@ def push_fasta_to_s3(lineage, directory = 'data/', bucket = 'nextflu-data'):
 	k.key = fasta
 	k.set_contents_from_filename(directory+fasta)
 	print fasta,"uploaded"
+	
+def push_json_to_s3(lineage, directory = '../auspice/data/', bucket = 'nextflu-dev'):
+	"""Upload JSON files to S3 bucket"""
+	"""Boto expects environmental variables AWS_ACCESS_KEY_ID and AWS_SECRET_ACCESS_KEY"""
+	directory = directory.rstrip('/')+'/'
+
+	import boto
+	conn = boto.connect_s3()
+	b = conn.get_bucket(bucket)
+	k = boto.s3.key.Key(b)
+
+	print "Uploading JSONs for",lineage
+	for postfix in ['_tree.json', '_sequences.json', '_frequencies.json', '_meta.json']:
+		json = lineage + postfix
+		k.key = 'data/'+json
+		k.set_contents_from_filename(directory+json)
+		print json,"uploaded"	
 
 def ammend_fasta(fname, lineage, threshold = 10, directory = 'data/'):
 	directory = directory.rstrip('/')+'/'
@@ -76,7 +93,7 @@ def ammend_fasta(fname, lineage, threshold = 10, directory = 'data/'):
 					print tmp_lineage,"not found"
 
 	print "Found", len(new_seqs), 'new for lineage', lineage
-	if len(new_seqs)>threshold:
+	if len(new_seqs)>=threshold:
 		with open(ex_fname, 'a') as outfile:
 			SeqIO.write(new_seqs, outfile, 'fasta')
 		updated = True
@@ -90,18 +107,20 @@ if __name__=="__main__":
 	parser.add_argument('--bin', type = str, default = "python")
 	parser.add_argument('--ATG', action = "store_true", default = False, help = "include full HA sequence starting at ATG")
 	parser.add_argument('--all', action = "store_true", default = False)
[…]

### 2015-05-19 manuscript/correspondence/response.tex
Include reviewer criticisms.

\documentclass[11pt,oneside,letterpaper]{article}

% graphicx package, useful for including eps and pdf graphics
\usepackage{graphicx}
\DeclareGraphicsExtensions{.pdf,.png,.jpg}

% basic packages
\usepackage{color} 
\usepackage{parskip}
\usepackage{float}
\usepackage{hyperref}

% text layout
\usepackage{geometry}
\geometry{textwidth=15.25cm} % 15.25cm for single-space, 16.25cm for double-space
\geometry{textheight=22cm} % 22cm for single-space, 22.5cm for double-space

% helps to keep figures from being orphaned on a page by themselves
\renewcommand{\topfraction}{0.85}
\renewcommand{\textfraction}{0.1}

% bold the 'Figure #' in the caption and separate it with a period
% Captions will be left justified
\usepackage[labelfont=bf,labelsep=period,font=small]{caption}

% review layout with double-spacing
%\usepackage{setspace} 
%\doublespacing
%\captionsetup{labelfont=bf,labelsep=period,font=doublespacing}

% cite package, to clean up citations in the main text. Do not remove.
\usepackage{cite}
%\renewcommand\citeleft{(}
%\renewcommand\citeright{)}
%\renewcommand\citeform[1]{\textsl{#1}}

% Remove brackets from numbering in list of References
\renewcommand\refname{\large References}
\makeatletter
\renewcommand{\@biblabel}[1]{\quad#1.}
\makeatother

\usepackage{authblk}
\renewcommand\Authands{ \& }
\renewcommand\Authfont{\normalsize \bf}
\renewcommand\Affilfont{\small \normalfont}
\makeatletter
\renewcommand\AB@affilsepx{, \protect\Affilfont}
\makeatother

\begin{document}

\newgeometry{top=4cm}

\today

\vspace{1cm}

Dear editor,

Please find attached our revised manuscript entitled ```nextflu: Real-time tracking of seasonal influenza virus evolution in humans''.  This is a resubmission of manuscript number BIOINF-2015-0560.

Sincerely,\\
Richard Neher and Trevor Bedford

\restoregeometry

\newpage

\section*{Reviewer responses}

Original reviewer criticisms are in plain text.  Our responses follow in \textbf{bold}.

%%% REVIEWER 1 %%%
\section*{Reviewer 1}

\subsection*{Comments to the Author}

Neher and Bedford report “nextflu”, an interactive near real time web interface to browse and explore the most informative approx. 2000 influenza HA gene sequences deposited in GISAID for the four most common strains. The automated pipeline “augur” fetches most recent data from GISAID, prunes the sequences which are subsequently combined to one multiple alignment with MAFFT. From this alignment, a phylogenetic tree is constructed and additional information is inferred, both of which are displayed on the interactive web site via “auspice”.

The freely available Python source code is of exemplary structure and quality, the website is well documented and clearly structured. Nevertheless, I want to raise some critical comments regarding the application note as well as the application itself:

\subsection*{Manuscript}

1. Despite the fact that I do not fully understand the computation of the ``frequency trajectories of mutations'' I suspect that is tries to express the same measure as the ``frequencies of alleles'' in the ``Allele dynamic plots'' of Steinbrueck \& McHardy (NAR 2011) with the addition of relating it to a noise model. You should include a statement of comparison to AD  plots in your manuscript.

2. The prefix ``next'' of ``nextflu'' raises hope that the application will provide some forecast or recommendation which influenza might be the next to be battled with vaccination. This hope is further fueled by the expression ``predictors of successful clades''. However, I cannot see how to obtain such predictions with nextflu. Also the frequency plot does not seem to state something about future strain frequencies.

3. The use case of the application is not clear to me. Can you give a more explicit example how information retrieval with nextflu could foster selection of the ``next successful clade'', rather than browsing the plain GISAID sequences? Of course, nextflu just visualizes existing data and thus cannot make up something that could not also be found in the database itself, but it would be nice to see how you intend to use nextflu.

4. Please give some hints why ``nextflu'' is not real-time but only near-real-time.

5. I suggest to add the word ``as'' in ``additional information such \textit{as} the viral genotype at specific sites'' in the Abstract.

6. Typo: ``by laboratories around to world'' → ``by laboratories around the world''

7. You report that HA genes from GISAID are downloaded. Is the alignment build by whole genes or just the CDS?

8. Please explain what the variable $i$ stands for.

9. To me, the presented explanation of the frequency calculation is not very comprehensible. I want to suggest to skip the maths (which better suits a research paper) and use the gained space to further convince readers of the usability of the tool. Anyways, a reference (or at least as supplemental material) to a detailed explanation of these computations should be provided.

\subsection*{Application}

1. You give information about the workflow of the augur pipeline to process data of the underlying GISAID database. What happens if GISAID is updated. Do you have to manually restart augur? Is it possible to trigger such an update by the user of nextflu? If it is not possible, is the term ``real-time'' really satisfied?

2. You describe the data pruning of the GISAID sequences, which also involves a step of pruning the alignment to various criteria. Is this done automatically? It would be nice to give the user the capability to parameterize some of these criteria prior to visualizing the tree.

3. It is rather complicated to zoom in and out or to scroll in the tree. Would it be possible to add according navigation buttons, rather then resetting the whole display when zooming out a bit?

4. It is difficult to use the date slider to actually move through time since the refresh is only accomplished when releasing the mouse button. Also slight horizontal movements can cause unwanted jumps. Could you maybe provide two buttons to jump to the neighboring date points? Or a functionality to automatically move through time like in a movie.

5. Unfortunately, observing the tree and the frequencies is impossible because of the large vertical expansion of the tree. Could you find some way to have both information more close to each other?

6. What is the X-Axis for the tree? Is it somehow related with time? If so, how comes that the sequence A/Georgia/16/2012 (2012-09-02) is farther to the right then sequence A/Georgia/3058/2012 (2012-09-26)?

%%% REVIEWER 2 %%%
\section*{Reviewer 2}

\subsection*{Comments to the Author}

This Applications Note introduces Nextflu, an original and important online tool to track the evolution of human Influenza in real time.

The paper is interesting and clearly written so I don't have any major comment.

\subsection*{Minor revisions}

p1, col1, l47: ``near real-time'', can the authors specify the updating frequency? from the website, it seems to be every 5 days.

p1, col2, l54: how is the outgroup chosen?

p2, col1, l21: typo: the sampling time point is $t_v$ not $x(t_v)$

%%% REVIEWER 3 %%%
\section*{Reviewer 3}

\subsection*{Major comments}

This may be beyond the scope of the tool, but it would be extremely useful if the tool could visualize not just the \# of epitope or receptor binding site changes, but which specific amino acid changes are involved -- at least for the branches defining the main clusters identified (e.g., 3c3.a and 3c2.a).

\subsection*{Minor comment}

Could the authors clarify what they mean by `removing known reassortant clusters'?  This is very vague and I assume naturally occurring reassortant viruses are not being removed.

\end{document}

### 2015-05-19 manuscript/correspondence/response.tex
Start on response.

@@
 
-Neher and Bedford report “nextflu”, an interactive near real time web interface to browse and explore the most informative approx.
+Neher and Bedford report ``nextflu'', an interactive near real time web interface to browse and explore the most informative approx.
 2000 influenza HA gene sequences deposited in GISAID for the four most common strains.
-The automated pipeline “augur” fetches most recent data from GISAID, prunes the sequences which are subsequently combined to one multiple alignment with MAFFT.
-From this alignment, a phylogenetic tree is constructed and additional information is inferred, both of which are displayed on the interactive web site via “auspice”.
+The automated pipeline ``augur'' fetches most recent data from GISAID, prunes the sequences which are subsequently combined to one multiple alignment with MAFFT.
+From this alignment, a phylogenetic tree is constructed and additional information is inferred, both of which are displayed on the interactive web site via ``auspice''.
 
@@
 
-6. Typo: ``by laboratories around to world'' → ``by laboratories around the world''
+\textbf{Done.}
+
+6. Typo: ``by laboratories around to world'' $\rightarrow$ ``by laboratories around the world''
+
+\textbf{Fixed.}
 
@@
 
+\textbf{Thank you for the suggestion.
+We experimented extensively with 3-column layouts and other methods to help display frequency plot alongside the phylogeny.
+However, the resulting layouts did not quite work aesthetically or in terms of usability space on a small display.
+Our suggestion here would be to `zoom' out with the web browser.
+The SVG implementation is fully responsive and the shrunk display will nicely fit both the frequency plot and the phylogeny.
+This seemed like the best solution to this particular issue.}
+
 6.
@@
 If so, how comes that the sequence A/Georgia/16/2012 (2012-09-02) is farther to the right then sequence A/Georgia/3058/2012 (2012-09-26)?
+
+\textbf{The x-axis is scaled in terms of substitutions per site, which is the native units of RAxML.
+This distance is across all sites.
+As influenza shows clock-like behavior we expect a general correlation between date of sampling and x coordinate.
+However, this correlation is not perfect and there will be cases like A/Georgia/16/2012 vs A/Georgia/3058/2012.
+We've clarified this point in the text.}
 

### 2015-05-19 manuscript/nextflu.tex
Reorganize for Bioinformatics submission.
This commit was submitted as an Application Note on April 8, 2015.
Start on response.

@@
 \usepackage{amsmath}
-\usepackage{natbib}
 
@@
 \maketitle
-\bibliographystyle{bioinformatics}
 
@@
 We have thus developed a processing pipeline and browser-based visualization that allows convenient exploration and analysis of the most recent influenza virus sequence data.
-This web-application displays a phylogenetic tree that can be decorated with additional information such the viral genotype at specific sites, sampling location, and derived statistics that have been shown to be predictive of future virus dynamics.
+This web-application displays a phylogenetic tree that can be decorated with additional information such as the viral genotype at specific sites, sampling location, and derived statistics that have been shown to be predictive of future virus dynamics.
 Additionally, mutation, genotype and clade frequency trajectories are calculated and displayed.
@@
 	\centering
-	\includegraphics[width=0.99\columnwidth]{figures/tree}
+	\includegraphics[width=0.99\columnwidth]{tree}
 	\caption[]{The \nextflu{} website with the user interface on the left and the phylogenetic tree on the right.}
@@
 
-These JSON files are then visualized by \auspice{} using D$^3$ \citep{bostock_d3_2011} and the phylogenetic tree displayed (\FIG{tree}).
+These JSON files are then visualized by \auspice{} using D$^3$ \citep{bostock_d3_2011} and the phylogenetic tree displayed with branches scaled according to evolutionary distance across all sites (\FIG{tree}).
 The user can explore the data interactively by selecting viruses from different dates or by coloring the tree by attributes such as:
@@
 	\centering
-	\includegraphics[width=0.99\columnwidth]{figures/frequencies}
+	\includegraphics[width=0.99\columnwidth]{frequencies}
 	\caption[]{The frequency diagram allows geography-specific plotting of frequencies of individual mutations, pairs of mutations and clades in the tree.}
@@
 %\section{Conclusion}
-We built \nextflu{} to facilitate the analysis and exploration of seasonal influenza sequence data collected by laboratories around to world.
+We built \nextflu{} to facilitate the analysis and exploration of seasonal influenza sequence data collected by laboratories around the world.
 By using the most recent data and integrating phylogenies with frequency trajectories and predictors of successful clades, we hope that \nextflu{} can inform the choice of strains used in seasonal influenza vaccines.

### 2015-05-22 manuscript/correspondence/response.tex
continued working on response

@@
 
+We would like to thank you and the reviewers for their supportive comments.
+We have addressed all comments in detailed poin-by-point response below.
+We also have implemented the LBI calculation that can be used as a coloring option an is dynamically updated when the date-slider is changed.
+
 Sincerely,\\
@@
 
+{\bf We would like to thank the reviewer for the supportive and constructive comments.}
+
 \subsection*{Manuscript}
@@
 You should include a statement of comparison to AD  plots in your manuscript.
+
+{\bf yes, the mutation trajectories are essentially the same, but nextflu also provides frequiencies of clades in the tree.
+The trajectories provide a useful sense of the temporal change.
+We have included a reference to the suggested paper.
+TODO}
 
@@
 
+{\bf The purpose of nextflu is to integrate as much information as possible to get an overview of the currently circulating viruses.
+As such, nextflu can help making an informed vaccine strain selection.
+We included quantities that have been suggested to be predictive for this purpose.
+However, we feel that predictors are not mature or specific enough to make an automated prediction and many practical constraints (virus/serum availability, cell-culture behavior etc) needs to be taken into account when selecting the vaccine strain.
+By providing the predictors along with explicit temporal dynamics of clades in the tree, the user can evaluate their relative merits and pick the most promising strain, given other constraints.}
+
 3.
@@
 
+{\bf Nextflu is meant as a tool to explore the most recent influenza sequences, their geographic distribution, and the dynamics of different variants.
+We have received enthusiastic feedback from the influenza community}
+
 4. Please give some hints why ``nextflu'' is not real-time but only near-real-time.
+
+{\bf As of now, nextflu requires manual update which we do on a weekly basis.
+Future integration into GISAID might make this step obsolete.
+Furthermore, nextflu can only track sequences in the data base, and as they are deposited with some lag, nextflu isn't traking influenza in real time -- strictly speaking.}
 
@@
 
+{\bf We use the entire HA segment, but include shorter sequences if necessary.}
+
 8. Please explain what the variable $i$ stands for.
+
+{\bf We have worked extensively on this section, see below.}
 
@@
 Anyways, a reference (or at least as supplemental material) to a detailed explanation of these computations should be provided.
+
+{\bf We provided a method tab in nextflu.}
 
@@
 
+{\bf as of now, we need to update manually. In any case, the deposition of sequences incurs a greater lag than our update frequency. We settled for ``near real time''.}
+
 2.
@@
 
+{\bf Since building aligning $>2000$ sequences and building a tree requires substantial computation, we can't have the user choose sequence filtering criteria and rerun the pipeline on demand.
+To address this concern, we have included multiple time resolutions.
+Users who want to customize this further can always clone the github repo and run the pipeline locally.}
+
 3. It is rather complicated to zoom in and out or to scroll in the tree. Would it be possible to add according navigation buttons, rather then resetting the whole display when zooming out a bit?
+
+{\bf With the current set-up, this is not readily achieved.}
 
@@
 Or a functionality to automatically move through time like in a movie.
+
+{\bf Dynamic update of the tree imposes a heavy computational burden.
+To address this issue, we now update a limited set of features as the slider is moved.
+We feel that the current compromise of updating tip circles without recalculating predictors and colors is a good compromise.
+Depending on the speed of the user's computer, this well be more or less smooth.}
 
@@
 
+{\bf We choose a well characterized strain predating the time interval of interest by a few years. We updated the text to clarify this.}
+
 p2, col1, l21: typo: the sampling time point is $t_v$ not $x(t_v)$
+{\bf Fixed, thank you.}
 
@@
 
+{\bf Tool tips of branches now show the amino-acid mutations associated with the branch.
+Furthermore, the we added a display of all alignment variability that allows to spot variable positions.
+Clicking on these positions in the graph will plot the frequencies of amino acids and color the tree accordingly.
+}
+
 \subsection*{Minor comment}
@@
 
+{\bf We filter sequences that are very similar to sequences from outbreaks of triple reassortants with pH1N1. ADD DETAIL.}
+
 \end{document}

### 2015-05-22 manuscript/methods.tex
continued working on response

--- 2015-04-09 manuscript/nextflu.tex
@@
-\documentclass{bioinfo}
-\copyrightyear{2015}
-\pubyear{2015}
-\usepackage[english]{babel}
-\usepackage{amssymb,amsfonts,amsmath}
-\usepackage{hyperref}
-\usepackage{graphicx}
-\usepackage{amsmath}
-\usepackage{natbib}
+\title{nextflu.org methods summary}
 
-\newcommand{\EQ}[1]{Eq.~(\ref{eq:#1})}
-\newcommand{\EQS}[2]{Eqs.~(\ref{eq:#1}) and (\ref{eq:#2})}
-\newcommand{\FIG}[1]{Fig.~\ref{fig:#1}}
-\newcommand{\TAB}[1]{Tab.~\ref{tab:#1}}
-\newcommand{\REF}[1]{ref.~\citep{#1}}
-\newcommand{\augur}{\textbf{augur}}
-\newcommand{\auspice}{\textbf{auspice}}
-\newcommand{\nextflu}{\textbf{nextflu}}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{document}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\title[Tracking of seasonal influenza virus evolution]{nextflu: Real-time tracking of seasonal influenza virus evolution in humans}
-\author{Richard~A.~Neher$^{1}$ and Trevor Bedford$^{2}$}
-\address{$^{1}$Max Planck Institute for Developmental Biology, 72076 T\"ubingen, Germany, and $^{2}$Vaccine and Infectious Disease Division, Fred Hutchinson Cancer Research Center, Seattle, WA 98109, USA}
-\history{Received on XXXXX; revised on XXXXX; accepted on XXXXX}
-
-\editor{Associate Editor: XXXXXXX}
-
-%\date{\today}
-\maketitle
-\bibliographystyle{bioinformatics}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{abstract} \section{Summary:} Seasonal influenza viruses evolve rapidly, allowing them to evade immunity in their human hosts and reinfect previously infected individuals.
-Similarly, vaccines against seasonal influenza need to be updated frequently to protect against an evolving virus population.
-We have thus developed a processing pipeline and browser-based visualization that allows convenient exploration and analysis of the most recent influenza virus sequence data.
-This web-application displays a phylogenetic tree that can be decorated with additional information such the viral genotype at specific sites, sampling location, and derived statistics that have been shown to be predictive of future virus dynamics.
-Additionally, mutation, genotype and clade frequency trajectories are calculated and displayed.
-
-\section{Availability and implementation:} Python and Javascript source code is freely available from \url{https://github.com/blab/nextflu}, while the web-application is live at \url{http://nextflu.org}.
-
-\section{Contact:} tbedford@fredhutch.org
-
-\end{abstract}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%\section*{Introduction}
-
-Every year, seasonal influenza infects between 10\% and 20\% of the global population resulting in substantial human morbidity and mortality \citep{flufactsheet}.
-Vaccination remains the most effective public health measure to combat seasonal epidemics.
-However, influenza viruses constantly evolve and thereby undergo antigenic drift, allowing drifted viruses to reinfect those in the human population with acquired immunity to previously circulating strains.
-Owing to antigenic drift, the seasonal influenza vaccine needs frequent updating to remain effective.
-In any given year, the particular choice of vaccine strain plays a major role in determining vaccine efficacy and so it is of critical importance to develop tools to analyze the ongoing evolution of the influenza virus population in order to aid vaccine strain selection.
-The program \nextflu{} presents a near real-time display of genetic relationships among influenza viruses and allows investigation of currently available sequence data.
-Currently, \nextflu{} tracks all four circulating lineages of seasonal influenza: A/H3N2, A/H1N1pdm, B/Victoria and B/Yamagata.
-
-\begin{figure}[t!]
-	\centering
-	\includegraphics[width=0.99\columnwidth]{figures/tree}
-	\caption[]{The \nextflu{} website with the user interface on the left and the phylogenetic tree on the right.}
-	\label{fig:tree}
-\end{figure}
-
-In implementation, \nextflu{} consists of a processing pipeline written in Python called \augur{} that analyzes virus sequence data and a JavaScript-based browser visualization called \auspice{} that displays this processed information.
-As input, \augur{} requires a FASTA file of sequences with FASTA header labels containing relevant information such as the strain name, the sampling date and passage history.
-For this purpose, influenza sequence data for the hemagglutinin (HA) gene is downloaded from the GISAID EpiFlu database \citep{GISAID}, which contains the most up-to-date collection of seasonal influenza viruses.
+\section{Pruning and alignment}
 The first step in the processing pipeline is to automatically select a subset of representative viruses.
@@
 Once aligned, the set of virus sequences is further cleaned by removing insertions relative to the outgroup to enforce canonical HA site numbering, by removing sequences that show either too much or too little divergence relative to the expectation given sampling date, and by removing known reassortant clusters.
+As outgroup for each viral lineage, we chose a well characterized virus without insertions relative to the canonical amino-acid numbering and a sampling date a few years before the time interval of interest.
 
+\section{Tree building}
 From the filtered and cleaned alignment, \augur{} builds a phylogenetic tree using FastTree \citep{price_fasttree_2009}, which is then further refined using RAxML \citep{stamatakis_raxml_2014}.
@@
 Internal branches without mutations are collapsed into polytomies.
-The final tree is decorated with the attributes to be displayed in the browser. %, as described below.
+The final tree is decorated with the attributes to be displayed in the browser.
 
-In addition to the phylogenetic tree, \augur{} estimates the frequency trajectories of mutations, genotypes and clades in the tree. 
+\section{Frequency calculation}
+\augur{} estimates the frequency trajectories of mutations, genotypes and clades in the tree. 
 Frequency trajectories are represented as a linear interpolation $x(t)$ of pivot frequencies $x_i$ at time points $t_i$ spaced one month apart. 
@@
 \end{equation}
-where $V_{\phi}$ is the set of viruses carring attribute $\phi$ and $x(t_v)$ is the sampling timepoint of virus $v$, and the process noise model
+where $V_{\phi}$ is the set of viruses carring attribute $\phi$ and $t_v$ is the collection date of virus $v$, and the process noise model
 \begin{equation}
@@
 Frequency estimation is done for mutations, particular genotypes, and major clades of the tree.
-At the end of the \augur{} pipeline, JSON files are exported containing the annotated phylogenetic tree, sequence data and frequency trajectories.
 
-These JSON files are then visualized by \auspice{} using D$^3$ \citep{bostock_d3_2011} and the phylogenetic tree displayed (\FIG{tree}).
-The user can explore the data interactively by selecting viruses from different dates or by coloring the tree by attributes such as:
-\begin{itemize}
-	\item \textit{epitope mutations} at sites generally associated with antibody binding have been suggested to be predictive of future clade success \citep{luksza_predictive_2014},
-%	\item non-epitope mutations: 
-%	Mutations outside epitopes tend to be damaging and anti-correlated with clade expansion \citep{luksza_predictive_2014}.
-    \item \textit{receptor binding mutations} at seven positions close to the receptor binding site have been shown to be responsible for major antigenic transitions in the past decades \citep{koel_substitutions_2013},
-    \item \textit{local branching index} is the exponentially weighted tree length surrounding a node, which is associated with rapid branching and expansion of clades \citep{neher_predicting_2014},
- %   \item geographic location
-    \item \textit{HA genotype} directly colors the tree by genotype at specific amino acid positions.
-\end{itemize}
-
-\begin{figure}[t]
-	\centering
-	\includegraphics[width=0.99\columnwidth]{figures/frequencies}
-	\caption[]{The frequency diagram allows geography-specific plotting of frequencies of individual mutations, pairs of mutations and clades in the tree.}
-	\label{fig:freq}
-\end{figure}
-
-The frequency plot below the tree (\FIG{freq}) displays the frequency trajectory of clades in the tree whenever the mouse hovers above the branch defining the clade. 
-Furthermore, trajectories of individual mutations, combinations of two mutations, and predefined clades such as 3c3.a can be plotted.
-
-%\section{Conclusion}
-We built \nextflu{} to facilitate the analysis and exploration of seasonal influenza sequence data collected by laboratories around to world.
-By using the most recent data and integrating phylogenies with frequency trajectories and predictors of successful clades, we hope that \nextflu{} can inform the choice of strains used in seasonal influenza vaccines.
-\nextflu{} was designed to be readily adapted to other rapidly evolving viruses and we see significant room for future developments in this area.
-
-\paragraph{Funding\textcolon}This work is supported by the ERC though Stg-260686 and by the NIH through U54 GM111274.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{thebibliography}{}
-
-\bibitem[Bogner {\em et~al.}, 2006]{GISAID}
-Bogner,P. {\em et~al.} (2006{\em{}}) A global
-  initiative on sharing avian flu data.
-\newblock {\em Nature, } {\bf 442} (7106), 981--981.
-
-\bibitem[Bostock {\em et~al.}, 2011]{bostock_d3_2011}
-Bostock,M. {\em et~al.} (2011{\em{}}) D$^3$: data-driven
-  documents.
-\newblock {\em IEEE Trans Vis Comput Graphics, } {\bf 17} (12), 2301--2309.
-
-\bibitem[Katoh and Standley, 2013]{katoh_mafft_2013}
-Katoh,K. and Standley,D.M. (2013{\em{}}) {MAFFT} multiple sequence alignment
-  software version 7: improvements in performance and usability.
-\newblock {\em Mol Biol Evol, } {\bf 30}, 772--780.
-
-\bibitem[Koel {\em et~al.}, 2013]{koel_substitutions_2013}
-Koel,B.F. {\em et~al.} (2013{\em{}})
-  Substitutions near the receptor binding site determine major antigenic change
-  during influenza virus evolution.
-\newblock {\em Science, } {\bf 342} (6161), 976--979.
-
-\bibitem[\L{}uksza and L\"a{}ssig, 2014]{luksza_predictive_2014}
-\L{}uksza,M. and L\"a{}ssig,M. (2014{\em{}}) A predictive fitness model for
-  influenza.
-\newblock {\em Nature, } {\bf 507} (7490), 57--61.
-
-\bibitem[Neher {\em et~al.}, 2014]{neher_predicting_2014}
-Neher,R.A. {\em et~al.} (2014{\em{}}) Predicting evolution
-  from the shape of genealogical trees.
-\newblock {\em eLife Sciences, } {\bf 3}, e03568.
-
-\bibitem[Oliphant, 2007]{oliphant_python_2007}
-Oliphant,T. (2007{\em{}}) Python for scientific computing.
-\newblock {\em Comput Sci Eng, } {\bf 9} (3), 10--20.
-
-\bibitem[Price {\em et~al.}, 2009]{price_fasttree_2009}
-Price,M.N. {\em et~al.} (2009{\em{}}) {FastTree:} computing
-  large minimum evolution trees with profiles instead of a distance matrix.
-\newblock {\em Mol Biol Evol, } {\bf 26} (7), 1641--50.
-
-\bibitem[Stamatakis, 2014]{stamatakis_raxml_2014}
-Stamatakis,A. (2014{\em{}}) {RAxML} version 8: a tool for phylogenetic analysis
-  and post-analysis of large phylogenies.
-\newblock {\em Bioinformatics, } {\bf 30} (9), 1312--1313.
-
-\bibitem[{World Health Organization}, 2009]{flufactsheet}
-{World Health Organization} (2009{\em{}}) {\em {Influenza Fact sheet}}.
-\newblock Available at http://www.who.int/mediacentre/factsheets/fs211/en/.
-
-\end{thebibliography}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\end{document}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{References}

### 2015-05-22 manuscript/nextflu.tex
continued working on response

@@
 The program \nextflu{} presents a near real-time display of genetic relationships among influenza viruses and allows investigation of currently available sequence data.
+By visualizing many different observables and predictors, we hope that \nextflu{} will help vaccine strain selection.
 Currently, \nextflu{} tracks all four circulating lineages of seasonal influenza: A/H3N2, A/H1N1pdm, B/Victoria and B/Yamagata.
@@
 Once aligned, the set of virus sequences is further cleaned by removing insertions relative to the outgroup to enforce canonical HA site numbering, by removing sequences that show either too much or too little divergence relative to the expectation given sampling date, and by removing known reassortant clusters.
+As outgroup for each viral lineage, we chose a well characterized virus without insertions relative to the canonical amino-acid numbering and a sampling date a few years before the time interval of interest.
 
@@
 Internal branches without mutations are collapsed into polytomies.
-The final tree is decorated with the attributes to be displayed in the browser. %, as described below.
+The final tree is decorated with the attributes to be displayed in the browser.
 
 In addition to the phylogenetic tree, \augur{} estimates the frequency trajectories of mutations, genotypes and clades in the tree.
-Frequency trajectories are represented as a linear interpolation $x(t)$ of pivot frequencies $x_i$ at time points $t_i$ spaced one month apart. 
-The frequencies $x_i$ corresponding to a feature $\phi$ (e.g.\ mutation) are estimated by simultaneously maximizing the Bernoulli observation model 
-\begin{equation}
-	\label{eq:obs}
-	f(\mathbf{x}) = \prod_{v\in V_\phi} x(t_v) \prod_{v\notin V_{\phi}} (1-x(t_v))
-\end{equation}
-where $V_{\phi}$ is the set of viruses carring attribute $\phi$ and $x(t_v)$ is the sampling timepoint of virus $v$, and the process noise model
-\begin{equation}
-	\label{eq:freq}
-	g(\mathbf{x}) = \mathrm{exp}\left(
-		-\sum_i \frac{(\Delta x_i - \epsilon\Delta x_{i-1})^2}{2\gamma \Delta t_i}
-	\right)
-\end{equation}
-where $\Delta x_i = x_i-x_{i-1}$, $\Delta t_i = t_i-t_{i-1}$, $\gamma$ parameterizes the smoothing imposed on the frequency estimate and $\epsilon$ parameterizes our expectation of frequency changes in interval $i$ based on the  change in interval $i-1$.
-$\epsilon=0$ implies that frequency changes are uncorrelated from one interval to the next, while $\epsilon=1$ implies that the most likely $\Delta x_i$ is $\Delta x_{i-1}$.
-We typically use $\epsilon = 0.7$.
-Here, the objective function $\mathrm{log}(f) + \mathrm{log}(g)$ is maximized using SciPy's implementation of a downhill simplex algorithm \citep{oliphant_python_2007}.
-To speed up convergence, optimization is first done using a coarse grid of pivot points which is subsequently refined to 12 pivots per year.
-Frequency estimation is done for mutations, particular genotypes, and major clades of the tree.
+Frequency are determined by maximizing the likelihood of sampling the observed set of virus sequences.
+In addition, we impose a smoothing that penalized rapid changes in frequency of the frequency derivative.
+\augur{} estimates frequency with up to one month resolution using a downhill simplex algorithm \citep{oliphant_python_2007}.
 At the end of the \augur{} pipeline, JSON files are exported containing the annotated phylogenetic tree, sequence data and frequency trajectories.
@@
 \end{itemize}
+The tree display can also be restricted to different geographic regions.
 
@@
 Furthermore, trajectories of individual mutations, combinations of two mutations, and predefined clades such as 3c3.a can be plotted.
+A second plot shows the variability of the alignment.
+On mouse-click on a variable positions in this plot, \auspice{} will color the tree by amino-acid at this position and plot the frequencies mutations at his positions.
 

### 2015-05-22 manuscript/correspondence/response.tex
more changes

@@
 
-{\bf yes, the mutation trajectories are essentially the same, but nextflu also provides frequiencies of clades in the tree.
+{\bf Yes, the mutation trajectories are essentially the same, but nextflu also provides frequiencies of clades in the tree.
 The trajectories provide a useful sense of the temporal change.
-We have included a reference to the suggested paper.
-TODO}
+We have included a reference to the suggested paper.}
 
@@
 We included quantities that have been suggested to be predictive for this purpose.
-However, we feel that predictors are not mature or specific enough to make an automated prediction and many practical constraints (virus/serum availability, cell-culture behavior etc) needs to be taken into account when selecting the vaccine strain.
+However, in addition to choosing the clade that best approximates the future virus population, vaccine strain selection faces many practical constraints (virus/serum availability, cell-culture behavior etc).
 By providing the predictors along with explicit temporal dynamics of clades in the tree, the user can evaluate their relative merits and pick the most promising strain, given other constraints.}
@@
 {\bf Nextflu is meant as a tool to explore the most recent influenza sequences, their geographic distribution, and the dynamics of different variants.
-We have received enthusiastic feedback from the influenza community}
+We have received enthusiastic feedback from the influenza community.}
 
@@
 
-{\bf We provided a method tab in nextflu.}
+{\bf We provided a method tab in nextflu. TODO}
 
@@
 
-{\bf With the current set-up, this is not readily achieved.}
+{\bf With the current set-up, this is not readily achieved. But we have added a dynamic inclusion of strain labels once the zoom level becomes high enough that labels can be discerned.}
 

### 2015-06-12 augur/src/tree_refine.py
started with proper genome annotation

@@ -41,9 +41,11 @@ class tree_refine(object):
 		from Bio.Seq import Seq
 		from Bio.SeqRecord import SeqRecord
 		if self.cds is not None:
-			tmp_aaseqs = [SeqRecord(Seq(node.aa_seq), id=node.strain, annotations = {'num_date':node.num_date, 'region':node.region}) for node in self.tree.leaf_iter()]
-			tmp_aaseqs.sort(key = lambda x:x.annotations['num_date'])
-			self.aa_aln = MultipleSeqAlignment(tmp_aaseqs)
+			self.aa_aln = {}
+			for anno in self.outgroup.cds:
+				tmp_aaseqs = [SeqRecord(Seq(node.aa_seq[anno]), id=node.strain, annotations = {'num_date':node.num_date, 'region':node.region}) for node in self.tree.leaf_iter()]
+				tmp_aaseqs.sort(key = lambda x:x.annotations['num_date'])
+				self.aa_aln[anno] = MultipleSeqAlignment(tmp_aaseqs)
 		tmp_nucseqs = [SeqRecord(Seq(node.seq), id=node.strain, annotations = {'num_date':node.num_date, 'region':node.region}) for node in self.tree.leaf_iter()]
 		tmp_nucseqs.sort(key = lambda x:x.annotations['num_date'])
 		self.nuc_aln = MultipleSeqAlignment(tmp_nucseqs)
@@ -101,15 +103,18 @@ class tree_refine(object):
 
 	def translate_all(self):
 		for node in self.tree.postorder_node_iter():
-			node.aa_seq = translate(node.seq[self.cds[0]:self.cds[1]])
+			node.aa_seq = {}
+			for anno, feature in self.outgroup.cds.iteritems():
+				self.aa_seq[anno] = translate(feature.extract(node.seq), to_stop = True)
 
 	def add_aa_mutations(self):
 		if hasattr(self.tree.seed_node, 'aa_seq'):
 			for node in self.tree.postorder_internal_node_iter():
-				for child in node.child_nodes():
-					child.aa_muts = ','.join([anc+str(pos)+der for pos,anc, der in 
-							zip(range(1,len(node.aa_seq)+1), node.aa_seq, child.aa_seq) if anc!=der])
-			self.tree.seed_node.aa_muts=""
+				for anno, parent_aa_seq in node.aa_seq.iteritems():
+					for child in node.child_nodes():
[…]

### 2015-06-12 augur/src/tree_refine.py
removed cds from refine. now defined in virus specific process

@@ -7,14 +7,12 @@ from seq_util import *
 from date_util import *
 
 class tree_refine(object):
-	def __init__(self,cds = (0,None), max_length = 0.01, dt=1, **kwargs):
+	def __init__(self, max_length = 0.01, dt=1, **kwargs):
 		'''
 		parameters:
-		cds 		-- coding region		
 		max_length  -- maximal length of external branches
 		dt 			-- time interval used to define the trunk of the tree
 		'''
-		self.cds = cds 
 		self.max_length = max_length
 		self.dt = dt
 

### 2015-06-12 augur/src/tree_refine.py
fixed cds reference

@@ -40,7 +40,7 @@ class tree_refine(object):
 		from Bio.SeqRecord import SeqRecord
 		if self.cds is not None:
 			self.aa_aln = {}
-			for anno in self.outgroup.cds:
+			for anno in self.cds:
 				tmp_aaseqs = [SeqRecord(Seq(node.aa_seq[anno]), id=node.strain, annotations = {'num_date':node.num_date, 'region':node.region}) for node in self.tree.leaf_iter()]
 				tmp_aaseqs.sort(key = lambda x:x.annotations['num_date'])
 				self.aa_aln[anno] = MultipleSeqAlignment(tmp_aaseqs)
@@ -102,7 +102,7 @@ class tree_refine(object):
 	def translate_all(self):
 		for node in self.tree.postorder_node_iter():
 			node.aa_seq = {}
-			for anno, feature in self.outgroup.cds.iteritems():
+			for anno, feature in self.cds.iteritems():
 				self.aa_seq[anno] = translate(feature.extract(node.seq), to_stop = True)
 
 	def add_aa_mutations(self):

### 2015-06-12 augur/src/tree_refine.py
fixed smth in refine

@@ -103,7 +103,7 @@ class tree_refine(object):
 		for node in self.tree.postorder_node_iter():
 			node.aa_seq = {}
 			for anno, feature in self.cds.iteritems():
-				self.aa_seq[anno] = translate(feature.extract(node.seq), to_stop = True)
+				node.aa_seq[anno] = translate(feature.extract(node.seq), to_stop = True)
 
 	def add_aa_mutations(self):
 		if hasattr(self.tree.seed_node, 'aa_seq'):

### 2015-06-12 augur/src/tree_refine.py
moved all translation stuff into single function

@@ -28,22 +28,11 @@ class tree_refine(object):
 		self.add_nuc_mutations()
 		if self.cds is not None:
 			self.translate_all()
-			self.add_aa_mutations()
 		self.add_node_attributes()
 		self.reduce()
 		self.layout()
 		self.define_trunk()
 
-		# make an amino acid aligment
-		from Bio.Align import MultipleSeqAlignment
-		from Bio.Seq import Seq
-		from Bio.SeqRecord import SeqRecord
-		if self.cds is not None:
-			self.aa_aln = {}
-			for anno in self.cds:
-				tmp_aaseqs = [SeqRecord(Seq(node.aa_seq[anno]), id=node.strain, annotations = {'num_date':node.num_date, 'region':node.region}) for node in self.tree.leaf_iter()]
-				tmp_aaseqs.sort(key = lambda x:x.annotations['num_date'])
-				self.aa_aln[anno] = MultipleSeqAlignment(tmp_aaseqs)
 		tmp_nucseqs = [SeqRecord(Seq(node.seq), id=node.strain, annotations = {'num_date':node.num_date, 'region':node.region}) for node in self.tree.leaf_iter()]
 		tmp_nucseqs.sort(key = lambda x:x.annotations['num_date'])
 		self.nuc_aln = MultipleSeqAlignment(tmp_nucseqs)
@@ -100,10 +89,21 @@ class tree_refine(object):
 				node._child_nodes.sort(key=lambda n:n.tree_length, reverse=True)
 
 	def translate_all(self):
+		# make an amino acid aligment
+		from Bio.Align import MultipleSeqAlignment
+		from Bio.Seq import Seq
+		from Bio.SeqRecord import SeqRecord
 		for node in self.tree.postorder_node_iter():
 			node.aa_seq = {}
 			for anno, feature in self.cds.iteritems():
 				node.aa_seq[anno] = translate(feature.extract(node.seq), to_stop = True)
+
+		self.add_aa_mutations()
+		self.aa_aln = {}
+		for anno in self.cds:
+			tmp_aaseqs = [SeqRecord(Seq(node.aa_seq[anno]), id=node.strain, annotations = {'num_date':node.num_date, 'region':node.region}) for node in self.tree.leaf_iter()]
+			tmp_aaseqs.sort(key = lambda x:x.annotations['num_date'])
+			self.aa_aln[anno] = MultipleSeqAlignment(tmp_aaseqs)
 
[…]

### 2015-06-12 augur/src/process.py
added genes to export

@@ -143,17 +143,16 @@ class process(virus_frequencies):
 		elems = {}
 		for node in self.tree:
 			if hasattr(node, "clade") and hasattr(node, "seq"):
-				if seq == 'nuc':
-					elems[node.clade] = {pos:state for pos, (state, ancstate) in 
+				elems[node.clade] = {}
+				elems[node.clade]['nuc'] = {pos:state for pos, (state, ancstate) in 
 								enumerate(izip(node.seq, self.tree.seed_node.seq)) if state!=ancstate}
-				else:
-					elems[node.clade] = {pos:state for pos, (state, ancstate) in 
-								enumerate(izip(node.aa_seq, self.tree.seed_node.aa_seq)) if state!=ancstate}
+				for anno, aa_seq in node.aa_seq:
+					elems[node.clade][anno] = {pos:state for pos, (state, ancstate) in 
+								enumerate(izip(aa_seq, self.tree.seed_node.aa_seq[anno])) if state!=ancstate}
 
-		if seq == 'nuc':
-			elems['root'] = self.tree.seed_node.seq
-		else:
-			elems['root'] = self.tree.seed_node.aa_seq
+		elems['root']['nuc'] = self.tree.seed_node.seq
+		for anno, aa_seq in self.tree.seed_node.aa_seq:
+			elems['root'] = aa_seq
 		write_json(elems, self.auspice_sequences_fname, indent=None)
 
 		print "Writing tree"

### 2015-06-12 augur/src/tree_refine.py

@@ -96,7 +96,7 @@ class tree_refine(object):
 		for node in self.tree.postorder_node_iter():
 			node.aa_seq = {}
 			for anno, feature in self.cds.iteritems():
-				node.aa_seq[anno] = translate(feature.extract(node.seq), to_stop = True)
+				node.aa_seq[anno] = translate(feature.extract(node.seq), to_stop = False)
 
 		self.add_aa_mutations()
 		self.aa_aln = {}

### 2015-06-12 augur/src/bernoulli_frequency.py
adjusted frequencies to multiple amino acid sequences

@@ -246,11 +246,11 @@ class virus_frequencies(object):
 			if self.verbose: print "too few observations"
 			return None, (tps, obs)
 
-	def get_sub_alignment(self, regions=None, nuc=False):
+	def get_sub_alignment(self, regions=None, gene='nuc'):
 		from Bio.Align import MultipleSeqAlignment
 		sub_aln = []
 		all_dates = []
-		for seq in self.nuc_aln if nuc else self.aa_aln:
+		for seq in self.nuc_aln if gene=='nuc' else self.aa_aln[gene]:
 			if regions is None or seq.annotations['region'] in regions:
 				seq_date = seq.annotations['num_date']
 				if seq_date>=self.time_interval[0] and seq_date < self.time_interval[1]:
@@ -258,32 +258,32 @@ class virus_frequencies(object):
 					all_dates.append(seq_date)				 
 		return MultipleSeqAlignment(sub_aln)
 
-	def determine_mutation_frequencies(self, regions=None, threshold=0.01, nuc=False):
+	def determine_mutation_frequencies(self, regions=None, threshold=0.01, gene='nuc'):
 		'''
 		determine the abundance of all single position variants  
 		'''
-		sub_aln = self.get_sub_alignment(regions, nuc=nuc)
-		if nuc:
+		sub_aln = self.get_sub_alignment(regions, gene=gene)
+		if gene=='nuc':
 			alpha, freqs = self.nuc_alphabet, self.nuc_frequencies
 		else:
-			alpha, freqs = self.aa_alphabet, self.aa_frequencies
+			alpha, freqs = self.aa_alphabet, self.aa_frequencies[gene]
 
 		mutation_frequencies = {"pivots":list(self.pivots)}
 		for pos in xrange(sub_aln.get_alignment_length()):
 			for ai, aa in enumerate(alpha):
 				if freqs[ai,pos]>threshold and freqs[ai,pos]<1.0-threshold:
-					mut = str(pos+1)+aa
+					mut = gene+': '+str(pos+1)+aa
 					print "estimating freq of ", mut, "total frequency:", freqs[ai,pos]
 					est_freq, (tps, obs) = self.estimate_genotype_frequency(sub_aln, [(pos, aa)])
 					if est_freq is not None:
 						mutation_frequencies[mut] = list(np.round(logit_inv(est_freq.y),3))
 		return mutation_frequencies
 
-	def determine_genotype_frequencies(self, regions=None, threshold=0.1, nuc=False):
[…]

### 2015-06-12 augur/src/process.py

@@ -388,23 +388,28 @@ class process(virus_frequencies):
 											for pos in xrange(self.nuc_frequencies.shape[1])]
 
 		if hasattr(self, 'aa_aln'):
-			aln_array = np.array(self.aa_aln)
-			self.aa_frequencies = np.zeros((len(self.aa_alphabet),aln_array.shape[1]))
-			for ai,aa in enumerate(self.aa_alphabet):
-				self.aa_frequencies[ai,:]=(aln_array==aa).mean(axis=0)
+			self.variable_aa = {}
+			self.consensus_aa = {}
+			self.aa_entropy = {}
+			self.variable_aa_identities = {}
+			for anno, aln in self.aa_aln:
+				aln_array = np.array(aln)
+				self.aa_frequencies = np.zeros((len(self.aa_alphabet),aln_array.shape[1]))
+				for ai,aa in enumerate(self.aa_alphabet):
+					self.aa_frequencies[ai,:]=(aln_array==aa).mean(axis=0)
 
-			self.variable_aa = np.where(np.max(self.aa_frequencies,axis=0)<1.0-self.min_mutation_frequency)[0]
-			self.consensus_aa = "".join(np.fromstring(self.aa_alphabet, 'S1')[np.argmax(self.aa_frequencies,axis=0)])
-			self.aa_entropy = -np.sum(self.aa_frequencies*np.log(np.maximum(1e-10,self.aa_frequencies)), axis=0)
-			self.variable_aa_identities = [ [self.aa_alphabet[ii] for ii in np.where(self.aa_frequencies[:,pos])[0]]
-											for pos in xrange(self.aa_frequencies.shape[1])]
+				self.variable_aa[anno] = np.where(np.max(self.aa_frequencies,axis=0)<1.0-self.min_mutation_frequency)[0]
+				self.consensus_aa[anno] = "".join(np.fromstring(self.aa_alphabet, 'S1')[np.argmax(self.aa_frequencies,axis=0)])
+				self.aa_entropy[anno] = -np.sum(self.aa_frequencies*np.log(np.maximum(1e-10,self.aa_frequencies)), axis=0)
+				self.variable_aa_identities[anno] = [ [self.aa_alphabet[ii] for ii in np.where(self.aa_frequencies[:,pos])[0]]
+												for pos in xrange(self.aa_frequencies.shape[1])]
 
 
 	def estimate_frequencies(self, tasks = ['mutations','genotypes', 'clades', 'tree']):
 		if 'mutations' in tasks:
[…]

### 2015-06-12 auspice/js/frequencies.js
Don't color tree by genotype on mouseover. Require click.
It was too easy to get the tree stuck with a genotype coloring requiring
manual deletion of text in the "Genomic position" box.

@@ -351,10 +351,6 @@ d3.json(path + file_prefix + "frequencies.json", function(error, json){
 							for (var ii=0;ii<frequencies["entropy"][d-1][2].length;ii+=1){
 								tmp.push(["global",d+frequencies["entropy"][d-1][2][ii]]);
 							}
-							console.log("tooltip");
-							colorBy = "genotype";
-							colorByGenotypePosition([d-1]);
-							d3.select("#gt-color").property("value", d);
 						}
 					}
 	            	return 'Position ' + d + frequencies["entropy"][d-1][2].join(","); 

### 2015-06-12 augur/src/process.py
added gene specific mutations to toolkit and genes to entropy export

@@ -204,9 +204,11 @@ class process(virus_frequencies):
 			if not hasattr(self, 'aa_entropy') and not hasattr(self, 'nuc_entropy'):
 				self.determine_variable_positions()
 
-			if seq=='aa' and hasattr(self, 'aa_entropy'):
-				self.frequencies["entropy"] = [ [pos, S, muts] for pos,S,muts in 
-						izip(xrange(self.aa_entropy.shape[0]), self.aa_entropy,self.variable_aa_identities) ]
+			if hasattr(self, 'aa_entropy'):
+				self.frequencies["entropy"] = {}
+				for anno, alnS in self.aa_entropy.iteritems():
+					self.frequencies["entropy"][anno] = [ [pos, S, muts] for pos,S,muts in 
+						izip(xrange(alnS.shape[0]), alnS,self.variable_aa_identities[anno]) ]
 			elif seq=='nuc' and hasattr(self, 'nuc_entropy'):
 				self.frequencies["entropy"] = [ [pos, S, muts] for pos,S,muts in 
 						izip(xrange(self.nuc_entropy.shape[0]), self.nuc_entropy,self.variable_nuc_identities) ]

### 2015-06-12 auspice/js/colors.js

@@ -66,11 +66,11 @@ function adjust_coloring_by_date() {
 	}	
 }
 
-function stateAtPosition(clade, pos){
-	if (typeof cladeToSeq[clade][pos] == "undefined"){
-		return cladeToSeq["root"][pos];
+function stateAtPosition(clade, gene, pos){
+	if (typeof cladeToSeq[clade][gene][pos] == "undefined"){
+		return cladeToSeq["root"][gene][pos];
 	}else{
-		return cladeToSeq[clade][pos];		
+		return cladeToSeq[clade][gene][pos];
 	}
 }
 
@@ -172,7 +172,7 @@ function colorByGenotypePosition (positions) {
 	var gts = nodes.map(function (d) {
 		var tmp = [];
 		for (var i=0; i<positions.length; i++){
-			tmp[tmp.length] = (positions[i]+1)+stateAtPosition(d.clade, positions[i]);
+			tmp[tmp.length] = (positions[i]+1)+stateAtPosition(d.clade, gene, positions[i]);
 		}
 		d.coloring = tmp.join(" / "); 
 		return d.coloring;});

### 2015-06-12 auspice/js/tooltips.js

@@ -70,9 +70,12 @@ var linkTooltip = d3.tip()
 		if (typeof d.frequency != "undefined") {
 			string += "Frequency: " + (100 * d.frequency).toFixed(1) + "%"
 		}
-		if ((typeof d.aa_muts !="undefined")&&(d.aa_muts.length)){
-			string+="<br>Mutations: "+d.aa_muts.replace(/,/g, ', ');
-		}else if ((typeof d.nuc_muts !="undefined")&&(d.nuc_muts.length)){
+		if (typeof d.aa_muts !="undefined"){
+			for (tmp_gene in d.aa_muts){
+				if (d.aa_muts[tmp_gene].length){
+					string+="<br>"+tmp_gene+": "+d.aa_muts[tmp_gene].replace(/,/g, ', ');
+				}
+		else if ((typeof d.nuc_muts !="undefined")&&(d.nuc_muts.length)){
 			string+="<br>Mutations: "+d.nuc_muts.replace(/,/g, ', ');
 		}
 		return string;

### 2015-06-12 augur/src/mers_process.py
changed mers_refine

@@ -48,6 +48,11 @@ class mers_filter(virus_filter):
 		virus_filter.__init__(self, **kwargs)
 		self.min_length = min_length
 		self.vaccine_strains =[]
+		genome_annotation = SeqIO.read('source-data/outgroup.gb', 'genbank').features
+		self.cds = {x.qualifiers['gene'][0]:x for x in genome_annotation
+				if 'gene' in x.qualifiers and x.type=='CDS' and 
+				x.qualifiers['gene'][0] in 
+				['ORF1ab', 'S', 'ORF3', 'ORF4a', 'ORF4b', 'E', 'N', 'M', 'ORF8b']}
 		self.outgroup = {
 			'strain': 'Camel_Egypt_NRCE-HKU270',
 			'accession':'KJ477103',
@@ -90,6 +95,7 @@ class mers_refine(tree_refine):
 		self.collapse()
 		self.add_nuc_mutations()
 		self.add_node_attributes()
+		self.translate_all()
 		self.reduce()
 		self.layout()
 		self.define_trunk()

### 2015-06-12 auspice/js/mathjax.config.js
Revert "Include methods page."
This reverts commit 07afb440c0c5864937616f43d3b301573251a0ae.

@@ -1,35 +1,0 @@
-// configure mathjax to work with markdown through script tags: `
-
-MathJax.Hub.Config({
-	tex2jax: { skipTags: ['script', 'noscript', 'style', 'textarea', 'pre'], inlineMath: [["$","$"],["\\(","\\)"]] }
-});
-	
-MathJax.Hub.Queue(function() {
-	var all = MathJax.Hub.getAllJax(), i;
-	for(i=0; i < all.length; i += 1) {
-		all[i].SourceElement().parentNode.className += ' has-jax';
-	}
-});
-
-// Switch MathJax font to sans serif
-
-MathJax.Hub.Config({
-  "HTML-CSS": {availableFonts: ["TeX"]},
-	  MMLorHTML: {prefer: "HTML"}
-});
-
-MathJax.Hub.Register.StartupHook("HTML-CSS Jax Ready",function () {
-  var VARIANT = MathJax.OutputJax["HTML-CSS"].FONTDATA.VARIANT;
-  VARIANT["normal"].fonts.unshift("MathJax_SansSerif");
-  VARIANT["bold"].fonts.unshift("MathJax_SansSerif-bold");
-  VARIANT["italic"].fonts.unshift("MathJax_SansSerif-italic");
-  VARIANT["-tex-mathit"].fonts.unshift("MathJax_SansSerif-italic");
-});
-
-MathJax.Hub.Register.StartupHook("SVG Jax Ready",function () {
-  var VARIANT = MathJax.OutputJax.SVG.FONTDATA.VARIANT;
-  VARIANT["normal"].fonts.unshift("MathJax_SansSerif");
-  VARIANT["bold"].fonts.unshift("MathJax_SansSerif-bold");
-  VARIANT["italic"].fonts.unshift("MathJax_SansSerif-italic");
-  VARIANT["-tex-mathit"].fonts.unshift("MathJax_SansSerif-italic");
-});

### 2015-06-12 manuscript/correspondence/response.tex
additions to response
Revise response and main text.

@@
 
-We would like to thank you and the reviewers for their supportive comments.
-We have addressed all comments in detailed poin-by-point response below.
-We also have implemented the LBI calculation that can be used as a coloring option an is dynamically updated when the date-slider is changed.
+We would like to thank you and the reviewers for their supportive comments. We have addressed all comments in detailed point-by-point response below.
+%We also have implemented the LBI calculation that can be used as a coloring option an is dynamically updated when the date-slider is changed.
 
@@
 {\bf Yes, the mutation trajectories are essentially the same, but nextflu also provides frequiencies of clades in the tree.
-The trajectories provide a useful sense of the temporal change.
-We have included a reference to the suggested paper.}
+As with the ``allele dynamic plots'', we believe trajectories provide a useful sense of the temporal change when combined with maximum-likelihood phylogenies.
+We have now included a reference to the suggested paper.}
 
@@
 We included quantities that have been suggested to be predictive for this purpose.
-However, in addition to choosing the clade that best approximates the future virus population, vaccine strain selection faces many practical constraints (virus/serum availability, cell-culture behavior etc).
-By providing the predictors along with explicit temporal dynamics of clades in the tree, the user can evaluate their relative merits and pick the most promising strain, given other constraints.}
+However, in addition to choosing the clade that best approximates the future virus population, vaccine strain selection faces many practical constraints (virus/serum availability, cell-culture behavior, etc...).
+By providing the predictors along with explicit temporal dynamics of clades in the tree, the user can evaluate their relative merits and pick the most promising strain, given other constraints.
+We believe that continually up-to-date displays of circulating viruses will be of the most use to WHO vaccine strain selection.
+These displays could help week-to-week in making it clear which viruses should be chosen to become test viruses in HI and other phenotypic assays.
+This said, we would eventually like to experiment with fully predictive models of virus population dynamics (like the models used by Steinbr\"{u}ck et al.
+2014 or \L{}uksza and L\"{a}ssig 2014), however, we believe that this work is beyond the scope of nextflu 1.0.}
 
@@
 
-{\bf Nextflu is meant as a tool to explore the most recent influenza sequences, their geographic distribution, and the dynamics of different variants.
-We have received enthusiastic feedback from the influenza community.}
+{\bf nextflu is meant as a tool to explore the most recent influenza sequences, their geographic distribution, and the dynamics of different variants.
+We have received enthusiastic feedback from the influenza community, including representatives from all five WHO Collaborating Centers and from the GISAID Initiative.}
 
@@
 
-{\bf As of now, nextflu requires manual update which we do on a weekly basis.
+{\bf As of now, nextflu requires manual update which we do on a weekly basis, as there is no API to allow automatic updating from the GISAID database.
 Future integration into GISAID might make this step obsolete.
-Furthermore, nextflu can only track sequences in the data base, and as they are deposited with some lag, nextflu isn't traking influenza in real time -- strictly speaking.}
+We understand that nextflu can only track sequences in the database, and as they are deposited with some lag, nextflu isn't traking influenza in real time -- strictly speaking.}
 
@@
 
-{\bf We use the entire HA segment, but include shorter sequences if necessary.}
+{\bf We use the entire HA segment, but include shorter sequences if necessary. This has been noted in the manuscript.}
 
@@
 
-{\bf We provided a method tab in nextflu. TODO}
+{\bf We agree that breaking out the mathematical details behind the frequency estimation is warranted.
+We've removed these details from the manuscript and describe them online at \href{http://nextflu.org/methods/}{\texttt{nextflu.org/methods/}}.}
 
@@
 
-{\bf as of now, we need to update manually. In any case, the deposition of sequences incurs a greater lag than our update frequency. We settled for ``near real time''.}
+{\bf As of now, we need to update manually, which involves clicking through to download all recent human sequences from GISAID and running the augur processing script.
+In any case, the deposition of sequences incurs a greater lag than our update frequency.
+During the North Hemisphere influenza season, when data deposition is frequent, GISAID and nextflu contain sequences within ~3 weeks of the present day.
+This is as close to `real-time' as present surveillance allows and we believe labeling nextflu as `real-time' is warranted.
+Also, as surveillance pipelines continue to improve, this lag will continue to decrease.}
 
@@
 {\bf Since building aligning $>2000$ sequences and building a tree requires substantial computation, we can't have the user choose sequence filtering criteria and rerun the pipeline on demand.
-To address this concern, we have included multiple time resolutions.
-Users who want to customize this further can always clone the github repo and run the pipeline locally.}
+However, to address this concern, we have invested a substantial of effort to provide multiple temporal `resolutions' displaying different sampling windows.
+For example, \href{http://nextflu.org/H3N2/1y/}{\texttt{nextflu.org/H3N2/1y/}} displays a detailed view of the last year of viral evolution and consequently deep sampling, while \href{http://nextflu.org/H3N2/12y/}{\texttt{nextflu.org/H3N2/12y/}} displays a courser view of the last 12 years of viral evolution.
+Furthermore, users who want to customize sampling can easily clone the GitHub repo and run the pipeline locally.}
 
@@
 {\bf Dynamic update of the tree imposes a heavy computational burden.
-To address this issue, we now update a limited set of features as the slider is moved.
+To address this issue, we now show/hide virus circles as the slider is moved to show sampling time, but `gray out' the virus colors while manipulating the time window.
 We feel that the current compromise of updating tip circles without recalculating predictors and colors is a good compromise.
@@
 The SVG implementation is fully responsive and the shrunk display will nicely fit both the frequency plot and the phylogeny.
-This seemed like the best solution to this particular issue.}
+After substantial experimentation, this seemed like the best solution to this particular issue.}
 
@@
 
-\textbf{The x-axis is scaled in terms of substitutions per site, which is the native units of RAxML.
+\textbf{The x-axis is scaled in terms of substitutions per site, which is the native unit of RAxML.
 This distance is across all sites.

### 2015-06-12 manuscript/nextflu.tex
more changes
Revise response and main text.

@@
 Following filtering, the viruses are subsampled to achieve a more equitable temporal and geographic distribution.
-For our standard display period of 3 years and 50 viruses per month, this typically results in $\sim$1800 viruses, which we align using MAFFT \citep{katoh_mafft_2013}.
+For our standard display period of 3 years and 50 viruses per month, this typically results in $\sim$1800 viruses, in which we align full-length HA sequences, where available and partial sequences otherwise, using MAFFT \citep{katoh_mafft_2013}.
 Once aligned, the set of virus sequences is further cleaned by removing insertions relative to the outgroup to enforce canonical HA site numbering, by removing sequences that show either too much or too little divergence relative to the expectation given sampling date, and by removing known reassortant clusters.
@@
 \augur{} estimates frequency with up to one month resolution using a downhill simplex algorithm \citep{oliphant_python_2007}.
+The result is similar to ``allele dynamics'' plots in \citep{steinbruck_allele_2011}, but provides frequencies of clades in the tree in addition to point mutations.
+
 At the end of the \augur{} pipeline, JSON files are exported containing the annotated phylogenetic tree, sequence data and frequency trajectories.
@@
 
+\bibitem[Steinbr\"{u}ck, 2011]{steinbruck_allele_2011}
+Steinbr\"{u}ck,L. and McHardy,A.C. (2011{\em{}}) Allele dynamics plots for the study of evolutionary dynamics in viral populations.
+\newblock {\em Nucleic Acids Res, } {\bf 39} (1), e4.
+
 \bibitem[Stamatakis, 2014]{stamatakis_raxml_2014}

### 2015-06-13 manuscript/correspondence/response.tex
Further revise manuscript and response.

@@
 
+{\bf At the moment, GISAID does not provide an API to automatically download sequences, and so, as of now, we need to update manually, which involves clicking through to download all recent human sequences from GISAID and running the augur processing script.
+We monitor the addition of new sequences in the database and update the site when more than 10 new sequences have appeared relative to what's present on the nextflu site.
+In practice, this means updates every 3-7 days.
+We've made a note of this in the main text.
+From what we understand, a forthcoming version of the GISAID database will make the process of updating easier and allow automated downloads through an API.
+When this occurs, nextflu could be fully automated and update daily.}
+
 p1, col2, l54: how is the outgroup chosen?
 
-{\bf We choose a well characterized strain predating the time interval of interest by a few years. We updated the text to clarify this.}
+{\bf We choose a well characterized strain predating the time interval of interest by a few years (A/Beijing/32/1992 for H3N2). We've updated the manuscript text to clarify this.}
 
 p2, col1, l21: typo: the sampling time point is $t_v$ not $x(t_v)$
+
 {\bf Fixed, thank you.}
@@
 
-{\bf Tool tips of branches now show the amino-acid mutations associated with the branch.
-Furthermore, the we added a display of all alignment variability that allows to spot variable positions.
+{\bf This is a good suggestion.
+We've revised nextflu to make it so that tool tips of branches now show the amino-acid mutations associated with the branch.
+Furthermore, we've added a display of all alignment variability that allows one to spot variable positions.
 Clicking on these positions in the graph will plot the frequencies of amino acids and color the tree accordingly.
-}
+This allows investigation of how specific amino acid changes map to the phylogeny.}
 
@@
 
-{\bf We filter sequences that are very similar to sequences from outbreaks of triple reassortants with pH1N1. ADD DETAIL.}
+{\bf We filter sequences that are very similar to sequences from outbreaks of triple-reassortant swine influenza H1N1 that contain hemagglutinin (HA) from seasonal H3N2 viruses as described in Bastien et al.
+(2010, Journal of Infectious Diseases).
+This is a specific clade of reassortant viruses that has circulated sporadically in North America since 2009.
+We just filter out this clade, rather than looking for any reassortant virus.
+This has been clarified in the text.}
 

### 2015-06-13 manuscript/nextflu.tex
Further revise manuscript and response.

--- 2015-05-22 manuscript/nextflu.tex
@@
 Vaccination remains the most effective public health measure to combat seasonal epidemics.
-However, influenza viruses constantly evolve and thereby undergo antigenic drift, allowing drifted viruses to reinfect those in the human population with acquired immunity to previously circulating strains.
+However, influenza viruses constantly evolve and thereby undergo antigenic drift, allowing drifted viruses to reinfect individuals with acquired immunity to previously circulating strains.
 Owing to antigenic drift, the seasonal influenza vaccine needs frequent updating to remain effective.
@@
 The program \nextflu{} presents a near real-time display of genetic relationships among influenza viruses and allows investigation of currently available sequence data.
-By visualizing many different observables and predictors, we hope that \nextflu{} will help vaccine strain selection.
+By visualizing many different genetic and epidemiological features, we hope that \nextflu{} will help vaccine strain selection.
 Currently, \nextflu{} tracks all four circulating lineages of seasonal influenza: A/H3N2, A/H1N1pdm, B/Victoria and B/Yamagata.
@@
 In implementation, \nextflu{} consists of a processing pipeline written in Python called \augur{} that analyzes virus sequence data and a JavaScript-based browser visualization called \auspice{} that displays this processed information.
-As input, \augur{} requires a FASTA file of sequences with FASTA header labels containing relevant information such as the strain name, the sampling date and passage history.
+As input, \augur{} requires a FASTA file of sequences with header labels containing relevant information such as strain name, sampling date and passage history.
 For this purpose, influenza sequence data for the hemagglutinin (HA) gene is downloaded from the GISAID EpiFlu database \citep{GISAID}, which contains the most up-to-date collection of seasonal influenza viruses.
@@
 In addition, local outbreaks are filtered by keeping only one instance of identical sequences sampled at the same location on the same day.
-Following filtering, the viruses are subsampled to achieve a more equitable temporal and geographic distribution.
-For our standard display period of 3 years and 50 viruses per month, this typically results in $\sim$1800 viruses, which we align using MAFFT \citep{katoh_mafft_2013}.
-Once aligned, the set of virus sequences is further cleaned by removing insertions relative to the outgroup to enforce canonical HA site numbering, by removing sequences that show either too much or too little divergence relative to the expectation given sampling date, and by removing known reassortant clusters.
+Following filtering, viruses are subsampled to achieve a more equitable temporal and geographic distribution.
+For our standard display period of 3 years and 32 viruses per month, this typically results in $\sim$1200 viruses, for which we align full-length HA sequences where available and partial sequences otherwise, using MAFFT \citep{katoh_mafft_2013}.
+Once aligned, the set of virus sequences is further cleaned by removing insertions relative to the outgroup to enforce canonical HA site numbering, by removing sequences that show either too much or too little divergence relative to the expectation given sampling date, and by removing known reassortant clusters, such as the triple-reassortant swine influenza viruses that have sporadically circulating since 2009 \citep{bastien_human_2010}.
 As outgroup for each viral lineage, we chose a well characterized virus without insertions relative to the canonical amino-acid numbering and a sampling date a few years before the time interval of interest.
@@
 In addition to the phylogenetic tree, \augur{} estimates the frequency trajectories of mutations, genotypes and clades in the tree.
-Frequency are determined by maximizing the likelihood of sampling the observed set of virus sequences.
+Frequencies are determined by maximizing the likelihood of sampling the observed set of virus sequences.
 In addition, we impose a smoothing that penalized rapid changes in frequency of the frequency derivative.
-\augur{} estimates frequency with up to one month resolution using a downhill simplex algorithm \citep{oliphant_python_2007}.
-At the end of the \augur{} pipeline, JSON files are exported containing the annotated phylogenetic tree, sequence data and frequency trajectories.
-
-These JSON files are then visualized by \auspice{} using D$^3$ \citep{bostock_d3_2011} and the phylogenetic tree displayed with branches scaled according to evolutionary distance across all sites (\FIG{tree}).
-The user can explore the data interactively by selecting viruses from different dates or by coloring the tree by attributes such as:
-\begin{itemize}
-	\item \textit{epitope mutations} at sites generally associated with antibody binding have been suggested to be predictive of future clade success \citep{luksza_predictive_2014},
-%	\item non-epitope mutations: 
-%	Mutations outside epitopes tend to be damaging and anti-correlated with clade expansion \citep{luksza_predictive_2014}.
-    \item \textit{receptor binding mutations} at seven positions close to the receptor binding site have been shown to be responsible for major antigenic transitions in the past decades \citep{koel_substitutions_2013},
-    \item \textit{local branching index} is the exponentially weighted tree length surrounding a node, which is associated with rapid branching and expansion of clades \citep{neher_predicting_2014},
- %   \item geographic location
-    \item \textit{HA genotype} directly colors the tree by genotype at specific amino acid positions.
-\end{itemize}
-The tree display can also be restricted to different geographic regions.
+\augur{} estimates frequency with up to one month resolution.
+The result is similar to ``allele dynamics'' plots in \citet{steinbruck_allele_2011}, but provides frequencies of clades in the tree in addition to point mutations.
+The \augur{} pipeline is run every 3-7 days in response to sequence updates in the GISAID database.
 
@@
 
+At the end of the \augur{} pipeline, JSON files are exported containing the annotated phylogenetic tree, sequence data and frequency trajectories.
+These JSON files are then visualized by \auspice{} using D$^3$ \citep{bostock_d3_2011} and a phylogenetic tree is displayed with branches scaled according to evolutionary distance across all sites (\FIG{tree}).
+The user can explore the data interactively by selecting viruses from different dates or by coloring the tree by attributes such as:
+\begin{itemize}
+	\item \textit{epitope mutations} at sites generally associated with antibody binding that have been suggested to be predictive of future clade success \citep{luksza_predictive_2014},
+%	\item non-epitope mutations: 
+%	Mutations outside epitopes tend to be damaging and anti-correlated with clade expansion \citep{luksza_predictive_2014}.
+    \item \textit{receptor binding mutations} at seven positions close to the receptor binding site that have been shown to be responsible for major antigenic transitions in the past decades \citep{koel_substitutions_2013},
+    \item \textit{local branching index} indicating the exponentially weighted tree length surrounding a node, which is associated with rapid branching and expansion of clades \citep{neher_predicting_2014},
+ %   \item geographic location
+    \item \textit{HA genotype}, which directly colors the tree by genotype at specific amino acid positions.
+\end{itemize}
+The display can also be restricted to different geographic regions.
+
 The frequency plot below the tree (\FIG{freq}) displays the frequency trajectory of clades in the tree whenever the mouse hovers above the branch defining the clade.
@@
 A second plot shows the variability of the alignment.
-On mouse-click on a variable positions in this plot, \auspice{} will color the tree by amino-acid at this position and plot the frequencies mutations at his positions.
+On mouse-click on a variable position in this plot, \auspice{} will color the tree by amino-acid at this position and plot its mutation frequencies.
 
@@
 \begin{thebibliography}{}
+
+\bibitem[Bastien {\em et~al.}, 2010]{bastien_human_2010}
+Bastien,N. {\em et~al.} (2010{\em{}})
+  Human infection with a triple-reassortant swine influenza {A (H1N1)} virus
+  containing the hemagglutinin and neuraminidase genes of seasonal influenza
+  virus.
+\newblock {\em J Infect Dis, } {\bf 201} (8), 1178--1182.
 
@@
 
-\bibitem[Oliphant, 2007]{oliphant_python_2007}
-Oliphant,T. (2007{\em{}}) Python for scientific computing.
-\newblock {\em Comput Sci Eng, } {\bf 9} (3), 10--20.
+%\bibitem[Oliphant, 2007]{oliphant_python_2007}
+%Oliphant,T. (2007{\em{}}) Python for scientific computing.
+%\newblock {\em Comput Sci Eng, } {\bf 9} (3), 10--20.
 
@@
 \newblock {\em Mol Biol Evol, } {\bf 26} (7), 1641--50.
+
+\bibitem[Steinbr\"{u}ck, 2011]{steinbruck_allele_2011}
+Steinbr\"{u}ck,L. and McHardy,A.C. (2011{\em{}}) Allele dynamics plots for the study of evolutionary dynamics in viral populations.
+\newblock {\em Nucleic Acids Res, } {\bf 39} (1), e4.
 

### 2015-06-15 auspice/js/frequencies.js
fixed position in tooltip title

@@ -382,7 +382,7 @@ d3.json(path + file_prefix + "frequencies.json", function(error, json){
 								d3.select("#gt-color").property("value", d);
 							}
 						}
-		            	return gene + ' codon ' + pos + frequencies["entropy"][gene][pos][2].join(",");
+		            	return gene + ' codon ' + (pos+1) + frequencies["entropy"][gene][pos][2].join(",");
 		            }else{ return d;}},
 	            value: function (value, ratio, id) {
 	                return id.substring(id.length-4)=='anno'?"start/stop":"Variability: "+value;

### 2015-06-15 auspice/js/frequencies.js
fixed position in tooltip

@@ -309,7 +309,7 @@ d3.json(path + file_prefix + "frequencies.json", function(error, json){
 	            	if (typeof posToAA[d] != "undefined"){
 		            	var gene = posToAA[d][0];
 		            	var pos = posToAA[d][1];
-		            	return 'Genomic position ' + d + ', ' + gene + ' codon ' + pos + frequencies["entropy"][gene][pos][2].join(",");
+		            	return gene + ' codon ' + (pos+1) + frequencies["entropy"][gene][pos][2].join(",");
 		            }else{ return d;}},
 	            value: function (value, ratio, id) {
 	                return id.substring(id.length-4)=='anno'?"start/stop":"Variability: "+value;

### 2015-06-15 manuscript/correspondence/response.tex
small changes in response

@@
 
-We would like to thank you and the reviewers for their supportive comments. We have addressed all comments in detailed point-by-point response below.
+We would like to thank you and the reviewers for their supportive comments.
+We have addressed all comments in detailed point-by-point response below.
+In the mean-time, we have received very positive feedback from the community, including requests to set up separate website for ebola virus and MERS-coronovirus evolution.
+Owing to the modular and flexible structure of augur and auspice, implementation of these pipelines was straightforward, underscoring the usefulness of augur and auspice.
+We hope that you and the reviewers share this impression.
 %We also have implemented the LBI calculation that can be used as a coloring option an is dynamically updated when the date-slider is changed.
@@
 {\bf nextflu is meant as a tool to explore the most recent influenza sequences, their geographic distribution, and the dynamics of different variants.
-We have received enthusiastic feedback from the influenza community, including representatives from all five WHO Collaborating Centers and from the GISAID Initiative.}
+We have received enthusiastic feedback from the influenza community, including representatives from all five WHO Collaborating Centers and from the GISAID Initiative.
+
+In response to requests from the community, we have recently extended nextflu to ebola virus and MERS coronavirus, underscoring both the need for tools like nextflu and the versatility of augur and auspice.}
 
@@
 
-{\bf As of now, we need to update manually, which involves clicking through to download all recent human sequences from GISAID and running the augur processing script.
+{\bf As of now, we need to update manually, which involves only downloading all recent human sequences from GISAID and running the augur processing script.
 In any case, the deposition of sequences incurs a greater lag than our update frequency.
@@
 However, to address this concern, we have invested a substantial of effort to provide multiple temporal `resolutions' displaying different sampling windows.
-For example, \href{http://nextflu.org/H3N2/1y/}{\texttt{nextflu.org/H3N2/1y/}} displays a detailed view of the last year of viral evolution and consequently deep sampling, while \href{http://nextflu.org/H3N2/12y/}{\texttt{nextflu.org/H3N2/12y/}} displays a courser view of the last 12 years of viral evolution.
+For example, \href{http://nextflu.org/H3N2/1y/}{\texttt{nextflu.org/H3N2/1y/}} displays a detailed view of the last year of viral evolution and consequently deep sampling, while \href{http://nextflu.org/H3N2/12y/}{\texttt{nextflu.org/H3N2/12y/}} displays a coarser view of the last 12 years of viral evolution.
 Furthermore, users who want to customize sampling can easily clone the GitHub repo and run the pipeline locally.}
@@
 We feel that the current compromise of updating tip circles without recalculating predictors and colors is a good compromise.
-Depending on the speed of the user's computer, this well be more or less smooth.}
+Depending on the speed of the user's computer, this well be more or less smooth.
+We have also added more flexibility to choose the focal time interval.
+Users can now manipulate the upper and lower time cut-off separately.}
 

### 2015-06-15 manuscript/correspondence/response.tex
fiddling

@@
 In the mean-time, we have received very positive feedback from the community, including requests to set up separate website for ebola virus and MERS-coronovirus evolution.
-Owing to the modular and flexible structure of augur and auspice, implementation of these pipelines was straightforward, underscoring the usefulness of augur and auspice.
-We hope that you and the reviewers share this impression.
+Owing to the modular and flexible structure of augur and auspice, implementation of these pipelines was straightforward.
+This shows that the nextflu framework can be readily applied to very different pathogens.
 %We also have implemented the LBI calculation that can be used as a coloring option an is dynamically updated when the date-slider is changed.

### 2015-06-15 manuscript/correspondence/response.tex
Small addition to response letter.

@@
 
+\newcommand{\augur}{\textbf{augur}}
+\newcommand{\auspice}{\textbf{auspice}}
+\newcommand{\nextflu}{\textbf{nextflu}}
+
 \begin{document}
@@
 We have addressed all comments in detailed point-by-point response below.
-In the mean-time, we have received very positive feedback from the community, including requests to set up separate website for ebola virus and MERS-coronovirus evolution.
-Owing to the modular and flexible structure of augur and auspice, implementation of these pipelines was straightforward.
+In the mean-time, we have received very positive feedback from the community, including enthusiastic responses from the many involved with the World Health Organization Global Influenza Surveillance and Response System.
+Additionally, we've received requests to set up separate websites for Ebola virus and MERS-CoV evolution.
+Owing to the modular and flexible structure of \augur{} and \auspice{}, implementation of these pipelines was straightforward.
 This shows that the nextflu framework can be readily applied to very different pathogens.
-%We also have implemented the LBI calculation that can be used as a coloring option an is dynamically updated when the date-slider is changed.
+However, the work with Ebola and MERS-CoV is still somewhat experimental, while the pipelines in \nextflu{} proper are very robust.
+We believe \nextflu{} will prove highly useful in continued influenza surveillance.
 

### 2015-06-15 manuscript/correspondence/response.tex
Include point-by-point for editor.

@@
 
-\newgeometry{top=4cm}
+\newgeometry{top=6cm}
 
@@
 We believe \nextflu{} will prove highly useful in continued influenza surveillance.
+
+Additionally, we've made the following specific changes to the program / manuscript:
+
+``1) Prediction of successful clades: you could easily implement your Local Branching Index (LBI), instead of the full likelihood calculation, for fitness calculation.''
+
+We provide LBI for each node in the tree at any point in time.
+This has previously been shown to be predictive of clade success.
+However, a full analysis of predictive clade dynamics is beyond version 1.0 of the platform.
+More work on this is forthcoming.
+We'd like to do this as properly as possible to keep trust in the platform by public health officials.
+If we provide explicit clade frequency predictions, we want them to be right.
+
+``2) Try to clarify the example case as suggested''
+
+As reviewer 1 suggested, we've clarified the example case.
+
+``3) Try to update the visualization while sliding through the data''
+
+We've clarified and improved the `date slider' behavior.
+
+``4) Move technical parts to supplementary material''
+
+We've moved technical details to an online methods page at \href{http://nextflu.org/methods/}{\texttt{nextflu.org/methods/}}.
+
+``5) It would be convenient to provide the users some control of how the data is being pruned''
+
+We've invested significant effort to making this happen.
+Users can now select a `data resolution'.
+\href{http://nextflu.org/H3N2/1y/}{\texttt{nextflu.org/H3N2/1y/}} displays a detailed view of the last year of viral evolution and consequently uses deeper sampling, while \href{http://nextflu.org/H3N2/12y/}{\texttt{nextflu.org/H3N2/12y/}} displays a coarser view of the last 12 years of viral evolution.
+
+``6) Explain how the tool accepts updateds of the database''
+
+This has been clarified.
 

### 2015-06-15 manuscript/correspondence/response.tex
Revise response letter.

@@
 
-We provide LBI for each node in the tree at any point in time.
-This has previously been shown to be predictive of clade success.
-However, a full analysis of predictive clade dynamics is beyond version 1.0 of the platform.
-More work on this is forthcoming.
-We'd like to do this as properly as possible to keep trust in the platform by public health officials.
-If we provide explicit clade frequency predictions, we want them to be right.
+We have implemented the LBI calculation that can be used as a coloring option and is dynamically updated when the date-slider is changed.
 
@@
 
-We've invested significant effort to making this happen.
-Users can now select a `data resolution'.
-\href{http://nextflu.org/H3N2/1y/}{\texttt{nextflu.org/H3N2/1y/}} displays a detailed view of the last year of viral evolution and consequently uses deeper sampling, while \href{http://nextflu.org/H3N2/12y/}{\texttt{nextflu.org/H3N2/12y/}} displays a coarser view of the last 12 years of viral evolution.
+To provide customization of data pruning, users can now select between alternative `data resolutions'.
+For example, \href{http://nextflu.org/H3N2/1y/}{\texttt{nextflu.org/H3N2/1y/}} displays a detailed view of the last year of viral evolution and consequently uses deeper sampling, while \href{http://nextflu.org/H3N2/12y/}{\texttt{nextflu.org/H3N2/12y/}} displays a coarser view of the last 12 years of viral evolution.
 

### 2015-09-18 auspice/_plugins/embed-image.rb
Include more detail on Yam and H3N2 titers.

@@ -22,7 +22,7 @@ module Jekyll
 			border = (12-width)/2
 			
 			html = ""
-			html += "<div class=\"spacer\"></div>"
+			html += "<div class=\"bigspacer\"></div>"
 			html += "<div class=\"row\">"
 			if border > 0 then
 				html += "<div class=\"col-lg-#{border}\"></div>"
@@ -36,7 +36,7 @@ module Jekyll
 				html += "<div class=\"col-lg-#{border}\"></div>"
 			end
 			html += "</div>"
-			html += "<div class=\"spacer\"></div>"			
+			html += "<div class=\"bigspacer\"></div>"			
 			html 
 			
 		end

### 2015-09-18 augur/src/process.py
Further adjustments to frequency parameters. Pick up smaller tree clades.

@@ -31,7 +31,7 @@ virus_config = {
 	# frequency estimation parameters
 	'aggregate_regions': [  ("global", None), ("NA", ["NorthAmerica"]), ("EU", ["Europe"]), 
 							("AS", ["China", "SoutheastAsia", "JapanKorea"]), ("OC", ["Oceania"]) ],
-	'frequency_stiffness':5.0,
+	'frequency_stiffness':10.0,
 	'verbose':2, 
 	'tol':2e-4, #tolerance for frequency optimization
 	'pc':1e-3, #pseudocount for frequencies 
@@ -429,4 +429,4 @@ class process(virus_frequencies):
 		if 'nuc_clades' in tasks:
 			self.all_clade_frequencies(gene='nuc') 
 		if 'tree' in tasks:
-			self.all_tree_frequencies() 
+			self.all_tree_frequencies(threshold = 5) 

### 2015-09-18 auspice/reports/tex/sep-2015.tex
Include TeX report.

\documentclass[11pt,oneside,letterpaper]{article}

% graphicx package, useful for including eps and pdf graphics
\usepackage{graphicx}
\DeclareGraphicsExtensions{.pdf,.png,.jpg}

% basic packages
\usepackage{color} 
\usepackage{parskip}
\usepackage{float}

% text layout
\usepackage{geometry}
\geometry{textwidth=16.25cm} % 15.25cm for single-space, 16.25cm for double-space
\geometry{textheight=22.5cm} % 22cm for single-space, 22.5cm for double-space

% helps to keep figures from being orphaned on a page by themselves
\renewcommand{\topfraction}{0.85}
\renewcommand{\textfraction}{0.1}

% bold the 'Figure #' in the caption and separate it with a period
% Captions will be left justified
\usepackage[labelfont=bf,labelsep=period,font=small]{caption}

% review layout with double-spacing
%\usepackage{setspace} 
%\doublespacing
%\captionsetup{labelfont=bf,labelsep=period,font=doublespacing}

% cite package, to clean up citations in the main text. Do not remove.
\usepackage{cite}
%\renewcommand\citeleft{(}
%\renewcommand\citeright{)}
%\renewcommand\citeform[1]{\textsl{#1}}

% Remove brackets from numbering in list of References
\renewcommand\refname{\large References}
\makeatletter
\renewcommand{\@biblabel}[1]{\quad#1.}
\makeatother

\usepackage{authblk}
\renewcommand\Authands{ \& }
\renewcommand\Authfont{\normalsize \bf}
\renewcommand\Affilfont{\small \normalfont}
\makeatletter
\renewcommand\AB@affilsepx{, \protect\Affilfont}
\makeatother

%%% TITLE %%%
\title{\vspace{2cm} \Large \bf 
Seasonal influenza circulation patterns and future projections
}

\author[1]{Trevor Bedford}
\author[2]{Richard A.\ Neher}

\affil[1]{Vaccine and Infectious Disease Division, Fred Hutchinson Cancer Research Center, Seattle, WA, USA.}
\affil[2]{Max Planck Institute for Developmental Biology, T\"ubingen, Germany.}

\date{September 16, 2015}

\begin{document}

\maketitle

\pagebreak

\section*{A/H3N2}

\textbf{We expect clade 3c2.a viruses to further predominate over 3c3.a and 3c3.b viruses. We do not observe the appearance of additional fast growing clades indicative of the emergence of new antigenic variants.}

There were 4 clades of H3N2 viruses circulating in 2015 (3c2.a, 3c3.a, 3c3.b and ancestral 3c3) (Fig.~\ref{H3N2_HA1-159}). 3c2.a viruses have predominated with 78\% of 2015 viruses belonging to clade 3c2.a, while 12\% of viruses belonged to clade 3c3.a and 6\% of viruses belonged to clade 3c3.b. Clades 3c2.a, 3c3.a and 3c3.b can easily be distinguished by the residue at HA1 site 159. In 3c2.a this is tyrosine (Y), in 3c3.a this is serine (S) and in 3c3.b this is phenylalanine (F).

\begin{figure}[h!]
	\centering		
	\includegraphics[width=0.8\textwidth]{../figures/sep-2015/H3N2_HA1-159.png}
	\caption{\textbf{H3N2 phylogeny colored by genotype at HA1 159.} 
	}
	\label{H3N2_HA1-159}
\end{figure}

3c2.a has been increasing in frequency, going from 50\% globally in October 2014 to 80\% of post-July viruses (Fig.~\ref{H3N2_frequencies}). 3c3.a has been steadily decreasing in frequency, going from a peak of 54\% in August 2014 to a present low of 10\% of post-July viruses. The first 3c3.b viruses appear in the database in August 2014. They have been slow to take off; just 2\% of post-July viruses belong to clade 3c3.b. In fact, 3c3.b viruses appear to have peaked in frequency in May/June at 8\%. For perspective, 3c2.a viruses first appear in the database in November 2013 and reach 10\% frequency two months later in January 2014 and 50\% eleven months later. 3c3.b viruses show no where near this level of success. Barring additional antigenic changes, it appears that 3c2.a viruses will take over the virus population and that 3c3.a and 3c3.b will soon be extinct.

\begin{figure}[h!]
	\centering		
	\includegraphics[width=1.0\textwidth]{../figures/sep-2015/H3N2_frequencies.png}
	\caption{\textbf{H3N2 frequencies across clades.} 
	}
	\label{H3N2_frequencies}
\end{figure}

These patterns of clade growth and decline extend to the regional level (Fig.~\ref{H3N2_HA1-159_frequencies}). 3c2.a viruses now dominate throughout the world. 3c3.b viruses briefly increased in frequency in Europe, but have declined in frequency in recent samples.

\begin{figure}[h!]
	\centering		
	\includegraphics[width=0.8\textwidth]{../figures/sep-2015/H3N2_HA1-159_frequencies.png}
	\caption{\textbf{H3N2 regional frequencies across clades.} 
	}
	\label{H3N2_HA1-159_frequencies}
\end{figure}

The continued dominance of 3c2.a is corroborated with a large local branching index and comparatively low mutational load as measured by the number of non-epitope substitutions since the root of the tree. Additionally, we observe no rapidly expanding subclades within 3c2.a, 3c3.a or 3c3.b. Within 3c2.a, the fastest branching patterns still appear at the root of the clade, indicative of no further adaptive events occurring (Fig.~\ref{H3N2_lbi}).

\begin{figure}[h!]
	\centering		
	\includegraphics[width=0.8\textwidth]{../figures/sep-2015/H3N2_lbi.png}
	\caption{\textbf{H3N2 phylogeny colored by local branching index.} 
	}
	\label{H3N2_lbi}
\end{figure}

Summarizing data from recent reports by the WHO CC London, we observe approximately a 2.8-fold drop in titer between serum raised against vaccine strain A/Switzerland/9715293/2013 and 3c2.a viruses. This corresponds to a 1.5 antigenic unit distance between clades. This can be seen on the left in the raw titer data (viruses are colored based on titer drop relative to A/Switzerland/9715293/2013 serum) and also on the right in a statistical model of titer drop (Fig.~\ref{H3N2_HI_titers}). The statistical model finds that 3c2.a and 3c3.a both appear homogeneous in terms of titer measurements within the clade, further supporting the absence of new adaptive variants. Note however, that this analysis is averaging across the serum NIBF13/14 raised against cell-culture virus and the serum F25/14 raised against egg-culture virus.

\begin{figure}[h!]
	\centering		
	\includegraphics[width=0.9\textwidth]{../figures/sep-2015/H3N2_HI_titers.png}
	\caption{\textbf{H3N2 phylogeny colored by antigenic distance to A/Switzerland/9715293/2013.} 
	}
	\label{H3N2_HI_titers}
\end{figure}

Here we compare titer measurements between cell-culture based serum NIBF13/14 on the left and the egg-culture based serum F25/14 on the right (Fig.~\ref{H3N2_HI_titers_ce}). Averaged across measurements, titer drop between cell-culture serum and egg-culture serum is similar with an average antigenic distance of 1.5 for cell-culture measurements and an average antigenic distance of 1.5 for egg-culture measurements. This suggests that egg-adaptation during production of serum F25/14 did not result in antigenic change relative to 3c2.a viruses.

\begin{figure}[h!]
	\centering		
	\includegraphics[width=0.9\textwidth]{../figures/sep-2015/H3N2_HI_titers_ce.png}
	\caption{\textbf{H3N2 phylogeny colored by antigenic distance to A/Switzerland/9715293/2013.} 
	}
	\label{H3N2_HI_titers_ce}
\end{figure}

Thus, 3c2.a and 3c3.a viruses, although not antigenically identical appear quite similar. Observed rapid displacement of 3c3.a viruses by 3c2.a viruses makes sense in this context, i.e. we expect strong immunological competition between clades.

\clearpage
\pagebreak

\section*{A/H1N1pdm}

\textbf{We observe the continued dominance of clade 6b viruses, although the substitution HA1:84N has recently risen to high frequency in South Asia.}

For H1N1pdm viruses, the clade 6b has been slowly increasing in frequency since 2013, so that we now find that 99.5\% of 2015 viruses belong to clade 6b (Fig.~\ref{H1N1pdm_HA2-164_HA1-84}). Within this clade two recent substitutions in particular have reached appreciable frequency, HA1:84N and HA2:164G. We find that 20\% of 2015 viruses are 84N and 11\% of 2015 viruses are 164G.

\begin{figure}[h!]
	\centering		
	\includegraphics[width=0.8\textwidth]{../figures/sep-2015/H1N1pdm_HA2-164_HA1-84.png}
	\caption{\textbf{H1N1pdm phylogeny colored by genotype at HA1 84 and HA1 164.} 
	}
	\label{H1N1pdm_HA2-164_HA1-84}
\end{figure}

The HA1:84N substitution shows substantial geographic subdivision with this clade predominating in South Asia (Fig.~\ref{H1N1pdm_geo}). Fully 66\% of 2015 viruses isolated from South Asia possess HA1:84N.

\begin{figure}[h!]
	\centering		
	\includegraphics[width=0.8\textwidth]{../figures/sep-2015/H1N1pdm_geo.png}
	\caption{\textbf{H1N1pdm phylogeny colored by geographic region.} 
	}
	\label{H1N1pdm_geo}
\end{figure}

Analyzing frequency trajectories of HA1:84N and HA2:164G shows a similar picture with a recent increase in frequency of 84N in South Asia (Fig.~\ref{H1N1pdm_HA1-84_HA1-164_frequencies}). It remains to be fully resolved whether there is an association between the HA1:84N substitution and the severe 2015 Indian H1N1pdm epidemic.

\begin{figure}[h!]
	\centering		
	\includegraphics[width=0.8\textwidth]{../figures/sep-2015/H1N1pdm_HA1-84_HA1-164_frequencies.png}
	\caption{\textbf{H1N1pdm regional frequencies at HA1 84 and HA1 164.} 
	}
	\label{H1N1pdm_HA1-84_HA1-164_frequencies}
\end{figure}

We observe no emerging clades of H1N1pdm that appear to be rapidly expanding, suggesting the absence of novel antigenic variants. Still, in light of the previous season's Indian outbreak, we suggest careful attention to the HA1:84N substitution to monitor further geographic spread and clade growth.

\clearpage
\pagebreak

\section*{B/Vic}

\textbf{Clade 1A viruses have predominated the B/Vic population with a recent emergence and spread of substitution HA1:129D.}

B/Victoria viruses isolated during 2015 were almost entirely 1A viruses, with 1B viruses almost entirely extinct (Fig.~\ref{Vic_HA1-129_HA1-209}). Within 1A viruses, there were three main groups: ancestral 1A viruses possessing 129N and 209K, derived viruses possessing 129N and 209N and derived viruses possessing 129D and 209K.

\begin{figure}[h!]
	\centering		
	\includegraphics[width=0.8\textwidth]{../figures/sep-2015/Vic_HA1-129_HA1-209.png}
	\caption{\textbf{Vic phylogeny colored by genotype at HA1 129 and HA1 209.} 
	}
	\label{Vic_HA1-129_HA1-209}
\end{figure}

The 129D/209K clade has been continually expanding after first appearing in late 2012 (Fig.~\ref{Vic_frequencies}). It comprised 77\% of 2015 viruses.

\begin{figure}[h!]
	\centering		
	\includegraphics[width=0.8\textwidth]{../figures/sep-2015/Vic_frequencies.png}
	\caption{\textbf{Vic frequencies at HA1 129.} 
	}
	\label{Vic_frequencies}
\end{figure}

\clearpage
\pagebreak

\section*{B/Yam}

\textbf{Clade 3 has predominated the B/Yamagata population with the recent emergence and spread of the substitution HA1:172Q.}

B/Yamagata viruses isolated during 2015 were predominately clade 3, with 97\% of viruses belonging to clade 3 and just 3\% belonging to clade 2 (Fig.~\ref{Yam_HA1-172}). Within clade 3 viruses, the substitution HA1:172Q rose rapidly in frequency during 2014 (Fig.~\ref{Yam_HA1-172_frequencies}).

\begin{figure}[h!]
	\centering		
	\includegraphics[width=0.8\textwidth]{../figures/sep-2015/Yam_HA1-172.png}
	\caption{\textbf{Yam phylogeny colored by genotype.} 
	}
	\label{Yam_HA1-172}
\end{figure}

\begin{figure}[h!]
	\centering		
	\includegraphics[width=1.0\textwidth]{../figures/sep-2015/Yam_HA1-172_frequencies.png}
	\caption{\textbf{Yam frequencies at HA1 172.} 
	}
	\label{Yam_HA1-172_frequencies}
\end{figure}

More recently, the substitution HA1:251V has emerged within the HA1:172Q clade of viruses (Fig.~\ref{Yam_HA1-251}). The HA1:251V substitution has steadily increased in frequency throughout 2015 (Fig.~\ref{Yam_HA1-251_frequencies}).

\begin{figure}[h!]
	\centering		
	\includegraphics[width=0.8\textwidth]{../figures/sep-2015/Yam_HA1-251.png}
	\caption{\textbf{Yam phylogeny colored by genotype at HA1 251.} 
	}
	\label{Yam_HA1-251}
\end{figure}

\begin{figure}[h!]
	\centering		
	\includegraphics[width=1.0\textwidth]{../figures/sep-2015/Yam_HA1-251_frequencies.png}
	\caption{\textbf{Yam frequencies at HA1 251.} 
	}
	\label{Yam_HA1-251_frequencies}
\end{figure}

\end{document}

### 2015-09-18 auspice/reports/tex/sep-2015.tex
Port @rneher's edit to TeX.

@@
 
-The continued dominance of 3c2.a is corroborated with a large local branching index and comparatively low mutational load as measured by the number of non-epitope substitutions since the root of the tree.
-Additionally, we observe no rapidly expanding subclades within 3c2.a, 3c3.a or 3c3.b.
-Within 3c2.a, the fastest branching patterns still appear at the root of the clade, indicative of no further adaptive events occurring (Fig.~\ref{H3N2_lbi}).
+The continued dominance of 3c2.a is corroborated with a large local branching index (Fig.~\ref{H3N2_lbi}) and comparatively low mutational load as measured by the number of non-epitope substitutions since the root of the tree.
+Within 3c2.a, one clade (the uppermost clade with in 3c2.a) is starting to dominate but this clade is not characterized by any amino acid differences within HA.
+We observe no rapidly expanding subclades within 3c2.a or 3c3.b.
 

### 2015-09-18 augur/src/process.py
Adjusting tree frequency threshold to get all lineages/resolutions to run.

@@ -429,4 +429,4 @@ class process(virus_frequencies):
 		if 'nuc_clades' in tasks:
 			self.all_clade_frequencies(gene='nuc') 
 		if 'tree' in tasks:
-			self.all_tree_frequencies(threshold = 5) 
+			self.all_tree_frequencies(threshold = 20) 

### 2015-09-22 auspice/js/tree.js
margin adjustments

@@ -8,16 +8,16 @@ var dHIScale = d3.scale.linear()
 var freqScale = d3.scale.sqrt()
 	.domain([0, 1])
 	.range([1, 10]);
-\
+
 var tipRadius = 4.0;
 var left_margin = 10;
+var right_margin = 10;
 var bottom_margin = 10;
 
 var branchLabelVisFraction = 0.05;
 var top_margin = 12;
 
-if ((typeof branch_labels != "undefined")&&(branch_labels)) {top_margin +=15;}
-var right_margin = 10;
+if ((typeof branch_labels != "undefined")&&(branch_labels)) {top_margin +=5;}
 
 function initDateColorDomain(intAttributes){
 	var numDateValues = tips.map(function(d) {return d.num_date;})
@@ -430,8 +430,8 @@ d3.json(path + file_prefix + "tree.json", function(error, root) {
 			treeplot.selectAll(".branchLabel").data(nodes)
 				.transition().duration(dt)
 				.style("font-size", branchLabelSize)
-				.attr("x", function(d) {  return d.x - 6;})
-				.attr("y", function(d) {  return d.y - 3;});
+				.attr("x", function(d) {  return d.x - 9;})
+				.attr("y", function(d) {  return d.y - 6;});
 		}
 
 		if (typeof clades !="undefined"){

### 2015-12-16 augur/src/H1N1pdm_process.py
harmonized process scripts, added supplementary figures

@@ -1,5 +1,7 @@
+import time, argparse,re,os, socket
 import matplotlib as mpl
-mpl.use('pdf')
+if socket.gethostname() not in ['olt', 'rneher-iMac']:
+	mpl.use('pdf')
 import time, re, os
 from virus_filter import flu_filter, fix_name
 from virus_clean import virus_clean
@@ -232,6 +234,12 @@ class H1N1pdm_process(process, H1N1pdm_filter, H1N1pdm_clean, H1N1pdm_refine, HI
 
 		if 'HIvalidate' in steps:
 			print "--- generating validation figures " + time.strftime("%H:%M:%S") + " ---"
+			for model in ['tree', 'mutation']:
+				try:
+					tree_additivity_symmetry(self, model)
+					for fmt in fmts: plt.savefig(self.htmlpath()+'HI_symmetry_'+model+fmt)
+				except:
+					print("Can't generate symmetry/additivity figures")
 			self.generate_validation_figures()
 
 
@@ -239,6 +247,8 @@ if __name__=="__main__":
 	all_steps = ['filter', 'align', 'clean', 'tree', 'ancestral', 'refine',
 				'frequencies', 'HI', 'export']+ ['HIvalidate']
 	from process import parser
+	import matplotlib.pyplot as plt
+	plt.ion()
 	params = parser.parse_args()
 
 	lt = time.localtime()

### 2015-12-16 augur/src/Vic_process.py

@@ -221,12 +221,21 @@ class BVic_process(process, BVic_filter, BVic_clean, BVic_refine, HI_tree):
 
 		if 'HIvalidate' in steps:
 			print "--- generating validation figures " + time.strftime("%H:%M:%S") + " ---"
+			from diagnostic_figures import tree_additivity_symmetry, fmts
+			for model in ['tree', 'mutation']:
+				try:
+					tree_additivity_symmetry(self, model)
+					for fmt in fmts: plt.savefig(self.htmlpath()+'HI_symmetry_'+model+fmt)
+				except:
+					print("Can't generate symmetry/additivity figures")
 			self.generate_validation_figures()
 
 
 if __name__=="__main__":
 	all_steps = ['filter', 'align', 'clean', 'tree', 'ancestral', 'refine', 'frequencies','HI', 'export']+ ['HIvalidate']
 	from process import parser
+	import matplotlib.pyplot as plt
+	plt.ion()
 	params = parser.parse_args()
 
 	lt = time.localtime()

### 2015-12-16 augur/src/Yam_process.py

@@ -1,5 +1,7 @@
+import time, argparse,re,os, socket
 import matplotlib as mpl
-mpl.use('pdf')
+if socket.gethostname() not in ['olt', 'rneher-iMac']:
+	mpl.use('pdf')
 import time, re, os
 from virus_filter import flu_filter, fix_name
 from virus_clean import virus_clean
@@ -248,11 +250,20 @@ class BYam_process(process, BYam_filter, BYam_clean, BYam_refine, HI_tree):
 
 		if 'HIvalidate' in steps:
 			print "--- generating validation figures " + time.strftime("%H:%M:%S") + " ---"
+			from diagnostic_figures import tree_additivity_symmetry, fmts
+			for model in ['tree', 'mutation']:
+				try:
+					tree_additivity_symmetry(self, model)
+					for fmt in fmts: plt.savefig(self.htmlpath()+'HI_symmetry_'+model+fmt)
+				except:
+					print("Can't generate symmetry/additivity figures")
 			self.generate_validation_figures()
 
 if __name__=="__main__":
 	all_steps = ['filter', 'align', 'clean', 'tree', 'ancestral', 'refine', 'frequencies','HI', 'export']+ ['HIvalidate']
 	from process import parser
+	import matplotlib.pyplot as plt
+	plt.ion()
 	params = parser.parse_args()
 
 	lt = time.localtime()

### 2015-12-16 augur/src/diagnostic_figures.py

@@ -31,7 +31,7 @@ def get_slope(freq, pivots, threshold):
 ####################################
 def make_combined_accession_number_lists():
     for flu in ['H3N2', 'H1N1pdm', 'Vic', 'Yam']:
-        flist = glob.glob('../auspice/data/'+flu+'*accession_numbers.tsv')
+        flist = glob('../auspice/data/'+flu+'*accession_numbers.tsv')
         all_accessions=set()
         for fname in flist:
             with open(fname) as infile:
@@ -41,7 +41,7 @@ def make_combined_accession_number_lists():
             for strain, acc in all_accessions:
                 ofile.write(strain+'\t'+acc+'\n')
 
-def make_table_with_virus_parameter_count():
+def make_table_with_virus_parameter_count(fmt='tex'):
     numbers = {}
     fields = {
             'total number of viruses':  '# viruses',
@@ -55,6 +55,7 @@ def make_table_with_virus_parameter_count():
     for mtype in ['tree', 'mutation']:
         for flu in ['H3N2', 'H1N1pdm', 'Vic', 'Yam']:
             flist = glob.glob('../auspice/'+flu+'/*y')
+            print(flist)
             for dirname in flist:
                 res = dirname.split('/')[-1]
                 try:
@@ -66,15 +67,17 @@ def make_table_with_virus_parameter_count():
                 except:
                     print("cant open", (flu, res, mtype))
 
-        sep='& '
-        le = '\\\\ \n'
-#        sep='\t'
-#        le = '\n'
+        if fmt=='tex':
+            sep='& '
+            le = '\\\\ \n'
+        else:
+            sep='\t'
+            le = '\n'
         table_fields = ['# viruses', '# test virus','# ref viruses','# antisera','# HI titers','# genetic parameters', '# non-zero genetic parameters']
-        with open('supplementary_table.tsv', 'w') as ofile:
+        with open('supp_data/supplementary_table_parameter_count.'+fmt, 'w') as ofile:
             ofile.write(sep.join(table_fields)+le)
             for flu in ['H3N2', 'H1N1pdm', 'Vic', 'Yam']:
-                for res in ['3y', '6y', '12y', '20y']:
+                for res in ['3y', '7y', '6y', '12y', '20y']:
                     for mtype in ['tree', 'mutation']:
                         if (flu, res, mtype) in numbers:
                             vals = [numbers[(flu, res, mtype)][x] for x in table_fields]
[…]

### 2015-12-16 augur/src/make_all.py

@@ -215,7 +215,7 @@ if __name__=="__main__":
 					n_viruses = 10
 					n_years = 12
 				prefix = lineage + '_'
-				call = map(str, [params.bin, process, '-v', n_viruses, '-y', n_years, '--start','HI',
+				call = map(str, [params.bin, process, '-v', n_viruses, '-y', n_years, '--start','filter',
 				           		 '--prefix', prefix, '--resolution', resolution] + common_args)
 				print call
 				if not params.annotate:

### 2015-12-16 augur/src/trunk_cHI.py
added parameter count files, mutations effects and missing index.htmls for ms submission

@@ -38,7 +38,8 @@ plt.ylabel(r'cumulative antigenic change $\mathrm{cHI}$', fontsize=fs)
 plt.xlabel(r'year', fontsize=fs)
 plt.legend(loc=2, fontsize = fs-2)
 plt.tight_layout()
-plt.savefig('cHI_trunk.png')
+for fmt in fmts:
+    plt.savefig('cHI_trunk'+fmt)
 
 
 ############################################

### 2015-12-16 augur/src/HI_predictability.py
adjusted figure labels

@@ -172,14 +172,14 @@ if mut_models:
             stddev_success.append(np.std(tmp))
             print(HI_lower, ind.sum())
             frac_success.append((HI_max[ind,1]>freq_thres).mean())
-        ax3.errorbar(np.arange(len(frac_success))+0.5+0.03*fi, frac_success,stddev_success, label = "max freq >"+str(freq_thres), lw=2)
+        ax3.errorbar(np.arange(len(frac_success))+0.5+0.03*fi, frac_success,stddev_success, label = r"$\mathrm{freq_{max}}>"+str(freq_thres)+"$", lw=2)
 
     ax3.set_xlabel(r'contribution to HI titer $d_i$', fontsize=fs)
     ax3.set_ylabel('fraction reaching frequency threshold', fontsize=fs)
     ax3.tick_params(labelsize=fs)
     ax3.set_xticks(np.arange(len(HI_binc))+0.5)
     ax3.set_xticklabels([str(lower)+'-'+str(upper) for lower, upper in zip(HI_threshold[:-1], HI_threshold[1:])])
-    plt.legend(loc=8, fontsize=fs)
+    plt.legend(loc=8, fontsize=fs, ncol=2)
     plt.ylim([0,1])
     plt.xlim([0,len(HI_binc)])
 
@@ -193,8 +193,8 @@ if mut_models:
     plt.figure(figsize = (1.2*figheight, figheight))
     cumHI =  HI_backbone.cumsum()
     plt.plot(HI_backbone, cumHI/cumHI[-1])
-    plt.ylabel('fraction of cHI due to effects < cutoff', fontsize = fs)
-    plt.xlabel(r'contribution to HI titer', fontsize = fs)
+    plt.ylabel(r'fraction of $\mathrm{cHI}$ due $d_i<X$', fontsize = fs)
+    plt.xlabel(r'contribution to HI titer $X$', fontsize = fs)
     plt.tick_params(labelsize=fs)
     plt.tight_layout()
     if save_figs:
@@ -294,7 +294,7 @@ sm._A = []
 cb = plt.colorbar(sm)
 cb.set_ticks(np.sqrt([0, 0.3, 1, 2,4]))
[…]

### 2016-02-13 auspice/reports/plot_clades.py
Switch y ticks to percentages.

@@ -8,8 +8,8 @@ from matplotlib.dates import YearLocator, MonthLocator, DateFormatter
 sns.set_style('darkgrid')
 plt.ion()
 
-#virus = 'H3N2'
-virus = 'H1N1pdm'
+virus = 'H3N2'
+#virus = 'H1N1pdm'
 #virus = 'Vic'
 #virus = 'Yam'
 
@@ -63,7 +63,7 @@ if len(clades):
         ax.set_ylim(0,1)
         ax.text(pivots[-1]-700, 0.9, clade)
         ax.tick_params(labelsize=fs)
-        ax.set_yticks([0, 0.25, 0.5, 0.75, 1.0])
+        ax.set_yticklabels(['{:3.0f}%'.format(x*100) for x in [0, 0.2, 0.4, 0.6, 0.8, 1.0]])        
         ax.xaxis.set_major_locator(months)
         ax.xaxis.set_major_formatter(monthsFmt)
     fig.autofmt_xdate()
@@ -86,7 +86,7 @@ for mutation, ax in zip(mutations, axs):
     ax.set_xlim([pivots[-1]-700,pivots[-1]+30])
     ax.set_ylim(0,1)
     ax.text(pivots[-1]-700, 0.9, mutation)
-    ax.set_yticks([0, 0.25, 0.5, 0.75, 1.0])
+    ax.set_yticklabels(['{:3.0f}%'.format(x*100) for x in [0, 0.2, 0.4, 0.6, 0.8, 1.0]]) 
     ax.tick_params(labelsize=fs)
     ax.xaxis.set_major_locator(months)
     ax.xaxis.set_major_formatter(monthsFmt)

### 2016-02-13 auspice/reports/plot_clades.py
Add "Frequency" as y axes label.

@@ -63,14 +63,19 @@ if len(clades):
         ax.set_ylim(0,1)
         ax.text(pivots[-1]-700, 0.9, clade)
         ax.tick_params(labelsize=fs)
-        ax.set_yticklabels(['{:3.0f}%'.format(x*100) for x in [0, 0.2, 0.4, 0.6, 0.8, 1.0]])        
+        ax.set_yticklabels(['{:3.0f}%'.format(x*100) for x in [0, 0.2, 0.4, 0.6, 0.8, 1.0]])
         ax.xaxis.set_major_locator(months)
         ax.xaxis.set_major_formatter(monthsFmt)
     fig.autofmt_xdate()
+    fax = fig.add_axes( [0., 0., 1, 1] )
+    fax.set_axis_off()
+    fax.set_xlim(0, 1)
+    fax.set_ylim(0, 1)
+    fax.text(0.02, 0.54, "Frequency", rotation='vertical', horizontalalignment='center', verticalalignment='center')
     axs[clade_legend['panel']].legend(loc=clade_legend['loc'], ncol=1, bbox_to_anchor=(1.0, 0.2))
     #plt.tight_layout(h_pad=0.01)
     bottom_margin = 0.2 - 0.03*len(clades)
-    plt.subplots_adjust(left=0.08, right=0.86, top=0.96, bottom=bottom_margin)
+    plt.subplots_adjust(left=0.12, right=0.86, top=0.96, bottom=bottom_margin)
     plt.savefig('figures/feb-2016/'+virus+'_clades.png')
 
 
@@ -90,9 +95,14 @@ for mutation, ax in zip(mutations, axs):
     ax.tick_params(labelsize=fs)
     ax.xaxis.set_major_locator(months)
     ax.xaxis.set_major_formatter(monthsFmt)
-fig.autofmt_xdate()    
+fig.autofmt_xdate()
+fax = fig.add_axes( [0., 0., 1, 1] )
+fax.set_axis_off()
+fax.set_xlim(0, 1)
+fax.set_ylim(0, 1)
+fax.text(0.02, 0.54, "Frequency", rotation='vertical', horizontalalignment='center', verticalalignment='center')
 axs[mut_legend['panel']].legend(loc=mut_legend['loc'], ncol=1, bbox_to_anchor=(1.0, 0.2))
 #plt.tight_layout(h_pad=0.01)
 bottom_margin = 0.2 - 0.03*len(mutations)
[…]

### 2016-02-13 auspice/reports/plot_clades.py
Fix region labeling.

@@ -44,6 +44,7 @@ elif virus=='Yam':
 offset = datetime(2000,1,1).toordinal()
 pivots = [offset+(x-2000)*365.25 for x in  freqs['clades']['global']['pivots']]
 regions = ['global', 'NA', 'AS', 'EU', 'OC']
+region_label = {'global': 'Global', 'NA': 'N America', 'AS': 'Asia', 'EU': 'Europe', 'OC': 'Oceania'}
 cols = sns.color_palette(n_colors=len(regions))
 fs=12
 months = MonthLocator(range(1, 13), bymonthday=1, interval=3)
@@ -56,7 +57,7 @@ if len(clades):
             try:
                 tmp_freq = freqs['clades'][region][clade]
                 if tmp_freq is not None:
-                    ax.plot_date(pivots, tmp_freq,'-o', label = region, c=c, lw=3 if region=='global' else 1)
+                    ax.plot_date(pivots, tmp_freq,'-o', label = region_label[region], c=c, lw=3 if region=='global' else 1)
             except:
                 print "skipping", clade, region
         ax.set_xlim([pivots[-1]-700,pivots[-1]+30])
@@ -75,7 +76,7 @@ if len(clades):
     axs[clade_legend['panel']].legend(loc=clade_legend['loc'], ncol=1, bbox_to_anchor=(1.0, 0.2))
     #plt.tight_layout(h_pad=0.01)
     bottom_margin = 0.2 - 0.03*len(clades)
-    plt.subplots_adjust(left=0.12, right=0.86, top=0.96, bottom=bottom_margin)
+    plt.subplots_adjust(left=0.12, right=0.84, top=0.96, bottom=bottom_margin)
     plt.savefig('figures/feb-2016/'+virus+'_clades.png')
 
 
@@ -85,7 +86,7 @@ for mutation, ax in zip(mutations, axs):
         try:
             tmp_freq = freqs['mutations'][region][mutation]
             if tmp_freq is not None:
-                ax.plot_date(pivots, tmp_freq, '-o',label = region, c=c, lw=3 if region=='global' else 1)
+                ax.plot_date(pivots, tmp_freq, '-o', label = region_label[region], c=c, lw=3 if region=='global' else 1)
         except:
             print "skipping", mutation, region
     ax.set_xlim([pivots[-1]-700,pivots[-1]+30])
[…]

### 2016-02-13 auspice/reports/plot_clades.py
Update Vic and Yam text. Swap ordering of Yam panels.

@@ -35,8 +35,8 @@ elif virus=='Vic':
     mut_legend = {'panel':0, 'loc':3}
 elif virus=='Yam':
     freqs = json.load(open('../data/Yam_2y_frequencies.json'))
-    clades = ['3', '2']
-    mutations = ['HA1:251V', 'HA1:172Q']
+    clades = ['2', '3']
+    mutations = ['HA1:172Q', 'HA1:251V']
     clade_legend = {'panel':0, 'loc':3}
     mut_legend = {'panel':0, 'loc':3}
 

### 2016-02-13 auspice/reports/tex/feb-2016.tex
Include TeX version of Feb 2016 report.

\documentclass[11pt,oneside,letterpaper]{article}

% graphicx package, useful for including eps and pdf graphics
\usepackage{graphicx}
\DeclareGraphicsExtensions{.pdf,.png,.jpg}

% basic packages
\usepackage{color} 
\usepackage{parskip}
\usepackage{float}

% text layout
\usepackage{geometry}
\geometry{textwidth=16.25cm} % 15.25cm for single-space, 16.25cm for double-space
\geometry{textheight=22.5cm} % 22cm for single-space, 22.5cm for double-space

% helps to keep figures from being orphaned on a page by themselves
\renewcommand{\topfraction}{0.85}
\renewcommand{\textfraction}{0.1}

% bold the 'Figure #' in the caption and separate it with a period
% Captions will be left justified
\usepackage[labelfont=bf,labelsep=period,font=small]{caption}

% review layout with double-spacing
%\usepackage{setspace} 
%\doublespacing
%\captionsetup{labelfont=bf,labelsep=period,font=doublespacing}

% cite package, to clean up citations in the main text. Do not remove.
\usepackage{cite}
%\renewcommand\citeleft{(}
%\renewcommand\citeright{)}
%\renewcommand\citeform[1]{\textsl{#1}}

% Remove brackets from numbering in list of References
\renewcommand\refname{\large References}
\makeatletter
\renewcommand{\@biblabel}[1]{\quad#1.}
\makeatother

\usepackage{authblk}
\renewcommand\Authands{ \& }
\renewcommand\Authfont{\normalsize \bf}
\renewcommand\Affilfont{\small \normalfont}
\makeatletter
\renewcommand\AB@affilsepx{, \protect\Affilfont}
\makeatother

%%% TOC
\usepackage{patchcmd}
\makeatletter
\patchcommand\@starttoc{\begin{quote}}{\end{quote}}
\usepackage{tocloft}
\renewcommand{\cftsecleader}{\cftdotfill{\cftdotsep}}

%%% TITLE %%%
\title{\vspace{2cm} \Large \bf 
Seasonal influenza circulation patterns and future projections
}

\author[1]{Trevor Bedford}
\author[2]{Richard A.\ Neher}

\affil[1]{Vaccine and Infectious Disease Division, Fred Hutchinson Cancer Research Center, Seattle, WA, USA.}
\affil[2]{Max Planck Institute for Developmental Biology, T\"ubingen, Germany.}

\date{February 15, 2016}

\begin{document}

\maketitle

\tableofcontents

\pagebreak

%%% A/H3N2 %%%
\section*{A/H3N2}
\addcontentsline{toc}{section}{A/H3N2}

\textbf{We expect clade 3c2.a viruses to continue to predominate over 3c3.a and 3c3.b viruses. Diversity within 3c2.a is beginning to accrue with the emergence of several clades of appreciable frequency, the most notable of which is the 171K/77V clade that has very recently risen to high frequency.}

Throughout 2015, clade 3c2.a viruses predominated over 3c3.a and 3c3.b viruses (Fig.\ \ref{H3N2_clades}). The only exception to this trend has been the recent reemergence of 3c3.a viruses in Europe. However, this clade of European 3c3.a viruses lacks recent amino acid substitutions relative to the 3c3.a parent. Thus it appears that the reemergence of 3c3.a viruses in Europe is due to epidemiological factors, rather than genetic factors. In addition, we estimate that 3c2.a viruses still predominate in Europe over 3c3.a viruses despite their recent uptick and 3c3.a isolates have been declining in recent weeks.

\begin{figure}[h!]
	\centering		
	\includegraphics[width=0.9\textwidth]{../figures/feb-2016/H3N2_clades.png}
	\caption{\textbf{Frequency trajectories of H3N2 clades.} 
	}
	\label{H3N2_clades}
\end{figure}

Within clade 3c2.a, there have emerged multiple circulating amino acid variants including HA1:114T, HA1:142K/197R and (HA1:171K, HA2:77V/155E) (Fig.\ \ref{H3N2_3c2a_tree}). The substitution 171K has emerged twice within 3c3.a. The earlier substitution did not spread far, however, the recent substitution has occurred alongside HA2:77V/155E and is present in more recent viral isolates.

\begin{figure}[h!]
	\centering		
	\includegraphics[width=0.8\textwidth]{../figures/feb-2016/H3N2_3c2a_tree.png}
	\caption{\textbf{H3N2 / 3c2.a phylogeny colored by genotype.} 
	}
	\label{H3N2_3c2a_tree}
\end{figure}

These variants show varying dynamics (Fig.\ \ref{H3N2_mutations}). The 114T variant peaked in frequency around July 2015 and similarly the 142K variant has been around for more than a year with a low but even geographic distribution. Consequently, we expect neither of these variants to be under strong positive selection. The T160K substitution is observed scattered within clade 3c2.a and overrepresented in sequences obtained from cultured virus. However, we observe a recent rapid increase in frequency of (HA1:171K, HA2:77V/155E) especially in Asia and North America. This has been the fastest growing variant within 3c2.a that we have observed. This clade also shows rapid phylogenetic branching \cite{neher2014predicting}, corroborating its rapid rise. Barring substantial changes in other clades, we predict the (HA1:171K, HA2:77V/155E) variant to dominate.

\begin{figure}[h!]
	\centering		
	\includegraphics[width=0.9\textwidth]{../figures/feb-2016/H3N2_mutations.png}
	\caption{\textbf{Frequency trajectories of 3c2.a subclades.} 
	}
	\label{H3N2_mutations}
\end{figure}

Due to poorly performing HI assays on 3c2.a samples and the lack of recent data, antigenic diversity within 3c2.a is hard to assess. The data available to us do not suggest significant antigenic evolution. Here we show recent antigenic data using sera against A/HongKong/5738/2014 from the Crick Worldwide Influenza Centre alongside model fits \cite{neher2015prediction} (Fig.\ \ref{H3N2_HI}). Cooler color indicate greater antigenic similarity. Note that we lack HI data for (HA1:171K, HA2:77V/155E) viruses.

\begin{figure}[h!]
	\centering		
	\includegraphics[width=1.0\textwidth]{../figures/feb-2016/H3N2_HI.png}
	\caption{\textbf{H3N2 phylogeny colored by antigenicity.} 
	}
	\label{H3N2_HI}
\end{figure}

\clearpage
\pagebreak

%%% A/H1N1pdm %%%
\section*{A/H1N1pdm}
\addcontentsline{toc}{section}{A/H1N1pdm}

\textbf{Within clade 6b, the clade 6b.1, comprising 84N/162N/216T, has rapidly risen to predominate the H1N1pdm population in the latter half of 2015. Although there is no obvious serological evidence for antigenic novelty of 6b.1, its extremely rapid rise in frequency suggests a selective mechanism promoting its spread.}

Within clade 6b, genetic variants have begun to emerge, most notably HA1:84N/162N/216T, referred to here as clade 6b.1, and (HA1:152T, HA2:174E), referred to here as clade 6b.2 (Fig.\ \ref{H1N1pdm_tree}).

\begin{figure}[h!]
	\centering		
	\includegraphics[width=0.8\textwidth]{../figures/feb-2016/H1N1pdm_tree.png}
	\caption{\textbf{H1N1pdm phylogeny colored by genotype.} 
	}
	\label{H1N1pdm_tree}
\end{figure}

\pagebreak

Both 6b.1 and 6b.2 viruses have recently increased in frequency (Fig.\ \ref{H1N1pdm_mutations}). With the exception of China, all regions of the world are now dominated by 6b.1 (with the caveat that most recent data comes from Europe and data on almost all other regions is sparse). The 84N substitution rose in frequency around April 2015 and 162N followed around October 2015. Additionally, China and Oceania have seen a recent rise in 6b.2 viruses. Recent samples from China have been dominated by 6b.2 viruses.

\begin{figure}[h!]
	\centering		
	\includegraphics[width=0.9\textwidth]{../figures/feb-2016/H1N1pdm_mutations.png}
	\caption{\textbf{Frequency trajectories of H1N1pdm variants.} 
	}
	\label{H1N1pdm_mutations}
\end{figure}

The 6b.1 clade comprising 84N/162N/216T rose from very low counts in early summer 2015 to almost 100\% by early 2016 in Europe and North America. The global rise of 6b.1 is extremely rapid relative to average H1N1pdm dynamics. This clade is also singled out by the local branching index \cite{neher2014predicting} as the clade most likely to persist. The rapid rise of 6b.1 viruses is likely selective in origin. Along these lines, we observe a very slight drop in titer associated with the 84N substitution of approximately 0.5 log2 HI dilutions. We lack antigenic data to catalog an effect from the 162N, 216T or 152T substitutions. Here we show recent antigenic data from the Crick Worldwide Influenza Centre alongside model fits \cite{neher2015prediction} (Fig.\ \ref{H1N1pdm_HI}).

\begin{figure}[h!]
	\centering		
	\includegraphics[width=1.0\textwidth]{../figures/feb-2016/H1N1pdm_HI.png}
	\caption{\textbf{H1N1pdm phylogeny colored by antigenicity.} 
	}
	\label{H1N1pdm_HI}
\end{figure}

\pagebreak

Generally, we would not expect such a minor antigenic effect to have a major selective advantage. It was previously suggested that the cause of the 2013-2014 H1N1pdm season was due to antigenic changes via K163Q despite a lack of an effect on naive ferret antisera \cite{linderman2014potential}. The spread of 163Q corresponds to the previous increase in frequency of 6b viruses. Here, we observe a significantly faster take off of 6b.1 viruses than observed for 6b, with 6b.1 viruses reaching high frequency in $\sim$6 months time rather than the $\sim$12 months required for clade 6b.

\clearpage
\pagebreak

%%% B/Victoria %%%
\section*{B/Vic}
\addcontentsline{toc}{section}{B/Vic}

\textbf{Within clade 1A viruses, there are few examples of selectively interesting clades. The clade 129D/146I/117V has risen to high frequency, but at a rate that suggests little effect from natural selection.}

North American and European isolates almost all contain 129D. While the majority of North American isolates also contain 117V, recent European isolates more ancestral with 117I instead of 117V. Recent isolates from the Southern hemisphere (and some Hong Kong isolates) are from the highly derived clade 124A/56K. Here we show frequency trajectories of B/Vic variants (Fig.\ \ref{Vic_mutations}).

\begin{figure}[h!]
	\centering		
	\includegraphics[width=0.9\textwidth]{../figures/feb-2016/Vic_mutations.png}
	\caption{\textbf{Frequency trajectories of B/Vic variants.} 
	}
	\label{Vic_mutations}
\end{figure}

No prominent amino acid substitutions are observed within the 117V clade, but this clade has been growing most strongly and is picked by the local branching index \cite{neher2014predicting} as the putatively most successful clade.

\clearpage
\pagebreak

%%% B/Yamagata %%%
\section*{B/Yam}
\addcontentsline{toc}{section}{B/Yam}

\textbf{Clade 3 has predominated the B/Yamagata population with the recent emergence and spread of the substitution HA1:172Q. At this point clade 2 viruses are close to extinct.}

In 2015, the vast majority of B/Yamagata isolates were of clade 3 viruses (Fig.\ \ref{Yam_clades}). Clade 2 appears to be barely hanging on and we expect it to continue to persist only if it acquires additional selective substitutions.

\begin{figure}[h!]
	\centering		
	\includegraphics[width=0.9\textwidth]{../figures/feb-2016/Yam_clades.png}
	\caption{\textbf{Frequency trajectories of B/Yam clades.} 
	}
	\label{Yam_clades}
\end{figure}

\pagebreak

Within clade 3, HA1:172Q has predominated during the whole of 2015, while the subclade HA1:172Q/251V has been slowly rising to frequencies close to 100\% in all geographic regions (Fig.\ \ref{Yam_mutations}). The HI model \cite{neher2015prediction} assigns a modest titer drop of about 0.3 units to the M251V branch. Within the clade 251V there are multiple subclades of intermediate size, one of which predominates in China. None of these exhibit a particularly striking dynamics.

\begin{figure}[h!]
	\centering		
	\includegraphics[width=0.9\textwidth]{../figures/feb-2016/Yam_mutations.png}
	\caption{\textbf{Frequency trajectories of B/Yam variants.} 
	}
	\label{Yam_mutations}
\end{figure}

%%% References %%%
\bibliographystyle{plos}
\bibliography{references}

\end{document}

### 2016-02-17 augur/src/H1N1pdm_process.py
Add numbering notes and adjust frequency estimation parameters.

@@ -14,6 +14,9 @@ from Bio.Align import MultipleSeqAlignment
 import numpy as np
 from itertools import izip
 
+# HA2 AA sites are shifted by +327 relative to HA1
+# So HA2:174E is 501E in HA1 numbering 
+
 # numbering starting at methionine including the signal peptide
 sp = 17
 epitope_mask = np.array(['1' if pos in [141,142,145,146,172,176,178,179,180,181,183,184,185, #Sa

### 2016-02-17 augur/src/H3N2_process.py

@@ -14,6 +14,9 @@ from Bio.Align import MultipleSeqAlignment
 import numpy as np
 from itertools import izip
 
+# HA2 AA sites are shifted by +329 relative to HA1
+# So HA2:77V is 406V in HA1 numbering 
+
 virus_config.update({
 	# data source and sequence parsing/cleaning/processing
 	'virus':'H3N2',
@@ -180,7 +183,7 @@ class H3N2_clean(virus_clean):
 
 	def clean(self):
 		self.clean_generic()
-		#self.clean_outbreaks()
+		self.clean_outbreaks()
 		#print "Number of viruses after outbreak filtering:",len(self.viruses)
 		self.clean_reassortants()
 		print "Number of viruses after reassortant filtering:",len(self.viruses)

### 2016-02-17 augur/src/make_all.py

@@ -197,7 +197,7 @@ if __name__=="__main__":
 					n_viruses = 100
 					n_years = 1
 				if resolution == '2y':
-					n_viruses = 60
+					n_viruses = 100
 					n_years = 2
 				if resolution == '3y':
 					n_viruses = 60

### 2016-02-17 augur/src/process.py

@@ -36,15 +36,15 @@ virus_config = {
 	# frequency estimation parameters
 	'aggregate_regions': [  ("global", None), ("NA", ["NorthAmerica"]), ("EU", ["Europe"]),
 							("AS", ["China", "SoutheastAsia", "JapanKorea"]), ("OC", ["Oceania"]) ],
-	'frequency_stiffness':10.0,
+	'frequency_stiffness':30.0,
 	'verbose':2,
 	'tol':2e-4, #tolerance for frequency optimization
 	'pc':1e-3, #pseudocount for frequencies
 
 	'extra_pivots': 12,  # number of pivot point for or after the last observations of a mutations
-	'inertia':0.5,		# fraction of frequency change carry over in the stiffness term
+	'inertia':0.8,		# fraction of frequency change carry over in the stiffness term
 	'n_iqd':3,     # standard deviations from clock
-	'min_mutation_frequency':0.02,
+	'min_mutation_frequency':0.01,
 	'min_genotype_frequency':0.1,
 }
 

### 2016-02-17 auspice/reports/plot_clades.py
Update report with new data and discussion of 168V.

@@ -17,7 +17,7 @@ virus = 'H3N2'
 if virus=='H3N2': ########## H3N2
     freqs = json.load(open('../data/H3N2_2y_frequencies.json'))
     clades = ['3c2.a', '3c3.a', '3c3.b']
-    mutations = ['HA1:114T', 'HA1:142K', 'HA1:171K']
+    mutations = ['HA1:114T', 'HA1:142K', 'HA1:168V', 'HA1:171K']
     #mutations = ['HA1:94H', 'HA1:114T', 'HA1:142K', 'HA1:171K']
     clade_legend = {'panel':0, 'loc':3}
     mut_legend = {'panel':0, 'loc':3}

### 2016-02-17 auspice/reports/tex/feb-2016.tex
Fix typo.
Include notes on HA1 numbering of HA2 substitutions.
Add numbering notes and adjust frequency estimation parameters.
Update Feb report TeX.

@@
 
-\date{February 15, 2016}
+\date{February 16, 2016}
 
@@
 \textbf{We expect clade 3c2.a viruses to continue to predominate over 3c3.a and 3c3.b viruses.
-Diversity within 3c2.a is beginning to accrue with the emergence of several clades of appreciable frequency, the most notable of which is the 171K/77V clade that has very recently risen to high frequency.}
+Diversity within 3c2.a is beginning to accrue with the emergence of several clades of appreciable frequency, the most notable of which is the new 171K clade that has very recently risen to high frequency.}
 
@@
 Within clade 3c2.a, there have emerged multiple circulating amino acid variants including HA1:114T, HA1:142K/197R and (HA1:171K, HA2:77V/155E) (Fig.\ \ref{H3N2_3c2a_tree}).
+This last clade corresponds to HA1:171K/406V/484E.
 The substitution 171K has emerged twice within 3c3.a.
 The earlier substitution did not spread far, however, the recent substitution has occurred alongside HA2:77V/155E and is present in more recent viral isolates.
+Within clade 142K/197R there is a subclade of significant size bearing 168V.
 
@@
 The T160K substitution is observed scattered within clade 3c2.a and overrepresented in sequences obtained from cultured virus.
-However, we observe a recent rapid increase in frequency of (HA1:171K, HA2:77V/155E) especially in Asia and North America.
+
+\pagebreak
+
+We observe a recent rapid increase in frequency of (HA1:171K, HA2:77V/155E) especially in Asia and North America.
 This has been the fastest growing variant within 3c2.a that we have observed.
 This clade also shows rapid phylogenetic branching \cite{neher2014predicting}, corroborating its rapid rise.
+Additionally, the subclade bearing HA1:142K/197R/168V has recently risen in frequency, although not as rapidly as the (HA1:171K, HA2:77V/155E) clade.
+We estimate clade HA1:142K/197R/168V has gone from 1\% global frequency in March 2015 to 19\% global frequency today, while clade (HA1:171K, HA2:77V/155E) has gone from 1\% global frequency in June 2015 to 29\% global frequency today.
+
 Barring substantial changes in other clades, we predict the (HA1:171K, HA2:77V/155E) variant to dominate.
@@
 
+\pagebreak
+
 Due to poorly performing HI assays on 3c2.a samples and the lack of recent data, antigenic diversity within 3c2.a is hard to assess.
@@
 Note that we lack HI data for (HA1:171K, HA2:77V/155E) viruses.
+Epitope mutation count would predict that HA1:142K/197R/168V is more drifted than (HA1:171K, HA2:77V/155E) with 2 epitope mutations vs 1 epitope mutation, although this requires confirmation through HI or other antigenic characterization.
 
@@
 Within clade 6b, genetic variants have begun to emerge, most notably HA1:84N/162N/216T, referred to here as clade 6b.1, and (HA1:152T, HA2:174E), referred to here as clade 6b.2 (Fig.\ \ref{H1N1pdm_tree}).
+The 6b.2 clade corresponds to HA1:152T/501E.
 
@@
 Within the clade 251V there are multiple subclades of intermediate size, one of which predominates in China.
-None of these exhibit a particularly striking dynamics.
+None of these exhibit particularly striking dynamics.
 

### 2016-02-18 auspice/reports/plot_clades.py
Make stacked bar plot, prettify all plots and add sampling text. Fixes #128.

@@ -1,11 +1,11 @@
 import matplotlib.pyplot as plt
 import numpy as np
 import seaborn as sns
-from datetime import datetime
+from datetime import datetime, timedelta
 import json
 from matplotlib.dates import YearLocator, MonthLocator, DateFormatter
 
-sns.set_style('darkgrid')
+sns.set_style('ticks')
 plt.ion()
 
 show_errorbars = True
@@ -64,11 +64,11 @@ l = len(freqs['clades']['global']['pivots'])
 bins = np.array([c[0] for c in counts])[-l:]
 date_bins = []
 for b in bins:
-	date_bins.append(datetime.strptime(b, "%Y-%m"))
+	date_bins.append(datetime.strptime(b, "%Y-%m") - timedelta(days=8))
 count_array = np.array([c[1:] for c in counts])[-l:,:].T
 count_by_region = {region: np.sum([count_array[region_names.index(r)] for r in region_codes[region]], axis=0)
                             for region in regions if region!='global'}
-count_by_region['global'] = count_array.sum(axis=0)                            
+count_by_region['global'] = count_array.sum(axis=0)
 smoothed_count_array = np.array([np.convolve(np.ones(n, dtype=float)/n, c, mode='same')
                         for c in count_array])
 smoothed_count_by_region = {region: np.sum([smoothed_count_array[region_names.index(r)] for r in region_codes[region]], axis=0)
@@ -76,17 +76,22 @@ smoothed_count_by_region = {region: np.sum([smoothed_count_array[region_names.in
 smoothed_count_by_region['global'] = smoothed_count_array.sum(axis=0)
 
 fig, ax = plt.subplots(figsize=(8, 3))
+tmpcounts = np.zeros(len(date_bins[2:]))
+plt.bar(date_bins[2:], count_by_region['global'][2:], width=18, linewidth=0, label="Other", color="#bbbbbb", clip_on=False)
 for c,region in zip(cols, regions):
-	plt.plot_date(date_bins[2:], count_by_region[region][2:], '-o', label = region_label[region], c=c, lw=3 if region=='global' else 1, clip_on=False)
-ax.tick_params(which = 'major', labelsize=fs, pad=20)
-ax.tick_params(which = 'minor', pad=7)
+    if region!='global':
[…]

### 2016-02-19 auspice/reports/plot_clades.py
Make sample counts x axis consistent with trajectories x axis.

@@ -76,12 +76,13 @@ smoothed_count_by_region = {region: np.sum([smoothed_count_array[region_names.in
 smoothed_count_by_region['global'] = smoothed_count_array.sum(axis=0)
 
 fig, ax = plt.subplots(figsize=(8, 3))
-tmpcounts = np.zeros(len(date_bins[2:]))
-plt.bar(date_bins[2:], count_by_region['global'][2:], width=18, linewidth=0, label="Other", color="#bbbbbb", clip_on=False)
+drop = 3
+tmpcounts = np.zeros(len(date_bins[drop:]))
+plt.bar(date_bins[drop:], count_by_region['global'][drop:], width=18, linewidth=0, label="Other", color="#bbbbbb", clip_on=False)
 for c,region in zip(cols, regions):
     if region!='global':
-        plt.bar(date_bins[2:], count_by_region[region][2:], bottom=tmpcounts, width=18, linewidth=0, label=region_label[region], color=c, clip_on=False)
-        tmpcounts += count_by_region[region][2:]
+        plt.bar(date_bins[drop:], count_by_region[region][drop:], bottom=tmpcounts, width=18, linewidth=0, label=region_label[region], color=c, clip_on=False)
+        tmpcounts += count_by_region[region][drop:]
 ax.tick_params(axis='x', which='major', labelsize=fs, pad=20)
 ax.tick_params(axis='x', which='minor', pad=7)
 ax.xaxis.set_major_locator(years)

### 2016-02-19 auspice/reports/tex/feb-2016.tex
Update TeX.

@@
 \usepackage{float}
+\usepackage{microtype}
+\usepackage{url}
 
@@
 \usepackage{geometry}
-\geometry{textwidth=16.25cm} % 15.25cm for single-space, 16.25cm for double-space
+\geometry{textwidth=16.5cm} % 15.25cm for single-space, 16.25cm for double-space
 \geometry{textheight=22.5cm} % 22cm for single-space, 22.5cm for double-space
@@
 
-\date{February 16, 2016}
+\date{February 18, 2016}
 
@@
 
+We base our primary analysis on a set of viruses collected between Feb 2014 and Jan 2016, comprising approximately 100 viruses per month where available and seeking to equilibriate sample counts geographically where possible (Fig.\ \ref{H3N2_counts}).
+This equilibration attempts to collect equal samples from Africa, China, Europe, Japan/South Korea, North America, Oceania, South America, South Asia, Southeast Asia and West Asia.
+In the following analysis we collapse samples from China, South Asia, Southeast Asia and West Asia into a single region, resulting in Asia possessing greater sample counts than North America or Europe.
+The only month that significantly departs from equitable sampling is Jan 2016 with 61 viruses, primarily from Europe and North America.
+Oct 2015 to Jan 2016 are lacking Southern Hemisphere viruses.
+
+\begin{figure}[h!]
+	\centering		
+	\includegraphics[width=0.95\textwidth]{../figures/feb-2016/H3N2_counts.png}
+	\caption{\textbf{Sample counts through time and across regions.} 
+	}
+	\label{H3N2_counts}
+\end{figure}
+
+For each virus, we know its date of collection and we know its genetic make up, allowing us count the number of viruses belonging to a particular genetic lineage through time.
+We use these timeseries of presence/absence to estimate clade frequencies through time.
+This estimation smooths month-to-month sampling variation.
+
 Throughout 2015, clade 3c2.a viruses predominated over 3c3.a and 3c3.b viruses (Fig.\ \ref{H3N2_clades}).
@@
 Thus it appears that the reemergence of 3c3.a viruses in Europe is due to epidemiological factors, rather than genetic factors.
-In addition, we estimate that 3c2.a viruses still predominate in Europe over 3c3.a viruses despite their recent uptick and 3c3.a isolates have been declining in recent weeks.
+In addition, we find that 3c2.a viruses still predominate in Europe over 3c3.a viruses despite their recent uptick and 3c3.a isolates have been declining in recent weeks.
 
@@
 	\centering		
-	\includegraphics[width=0.9\textwidth]{../figures/feb-2016/H3N2_clades.png}
+	\includegraphics[width=0.95\textwidth]{../figures/feb-2016/H3N2_clades.png}
 	\caption{\textbf{Frequency trajectories of H3N2 clades.} 
@@
 The substitution 171K has emerged twice within 3c3.a.
-The earlier substitution did not spread far, however, the recent substitution has occurred alongside HA2:77V/155E and is present in more recent viral isolates.
+The earlier substitution did not spread far.
+However, the recent 171K substitution has occurred alongside HA2:77V/155E and is present in more recent viral isolates.
 Within clade 142K/197R there is a subclade of significant size bearing 168V.
@@
 	\centering		
-	\includegraphics[width=0.8\textwidth]{../figures/feb-2016/H3N2_3c2a_tree.png}
+	\includegraphics[width=0.9\textwidth]{../figures/feb-2016/H3N2_3c2a_tree.png}
 	\caption{\textbf{H3N2 / 3c2.a phylogeny colored by genotype.} 
@@
 These variants show varying dynamics (Fig.\ \ref{H3N2_mutations}).
-The 114T variant peaked in frequency around July 2015 and similarly the 142K variant has been around for more than a year with a low but even geographic distribution.
-Consequently, we expect neither of these variants to be under strong positive selection.
+The 114T variant peaked in frequency around July 2015 and similarly the 142K variant has been around for more than a year with a low but even geographic distribution, although has seen a recent uptick in frequency since October due to 168V.
+We expect neither of these variants to be under strong positive selection.
 The T160K substitution is observed scattered within clade 3c2.a and overrepresented in sequences obtained from cultured virus.
-
-\pagebreak
 
@@
 We estimate clade HA1:142K/197R/168V has gone from 1\% global frequency in March 2015 to 19\% global frequency today, while clade (HA1:171K, HA2:77V/155E) has gone from 1\% global frequency in June 2015 to 29\% global frequency today.
+For comparison, 3c3.a viruses increased in frequency from 1\% to 31\% in a similar span of 8 months time.
 
@@
 	\centering		
-	\includegraphics[width=0.9\textwidth]{../figures/feb-2016/H3N2_mutations.png}
+	\includegraphics[width=0.95\textwidth]{../figures/feb-2016/H3N2_mutations.png}
 	\caption{\textbf{Frequency trajectories of 3c2.a subclades.} 
@@
 The data available to us do not suggest significant antigenic evolution.
-Here we show recent antigenic data using sera against A/HongKong/5738/2014 from the Crick Worldwide Influenza Centre alongside model fits \cite{neher2015prediction} (Fig.\ \ref{H3N2_HI}).
+Here we show recent antigenic data (through September 2015) using sera against A/HongKong/5738/2014 from the Crick Worldwide Influenza Centre alongside model fits \cite{neher2015prediction} (Fig.\ \ref{H3N2_HI}).
 Cooler color indicate greater antigenic similarity.
@@
 
+We would prioritize further characterization of viruses basal to the (HA1:171K, HA2:77V/155E) and HA1:142K/197R/168V clades.
+These include A/Washington/51/2015 and A/Laos/928/2015 for 171K and A/Bangladesh/3400/2015 and A/Roma/29/2015 for 168V (Fig.\ \ref{H3N2_basal_viruses}).
+
+\begin{figure}[h!]
+	\centering		
+	\includegraphics[width=1.0\textwidth]{../figures/feb-2016/H3N2_basal_viruses.png}
+	\caption{\textbf{Viruses basal to the (HA1:171K, HA2:77V/155E) clade (left panel) and to the HA1:142K/197R/168V clade (right panel).} 
+	}
+	\label{H3N2_basal_viruses}
+\end{figure}
+
 \clearpage
@@
 
+We base our primary analysis on a set viruses collected between Feb 2014 and Feb 2016, comprising approximately 100 viruses per month where available and seeking to equilibriate sample counts geographically where possible.
+As discussed above, this equilibration attempts to collect equal samples from Africa, China, Europe, Japan/South Korea, North America, Oceania, South America, South Asia, Southeast Asia and West Asia.
+Recent months through Jan 2016 have largely sufficient sample counts and sample distributions, although are lacking Southern Hemisphere viruses.
+
+\begin{figure}[h!]
+	\centering		
+	\includegraphics[width=0.95\textwidth]{../figures/feb-2016/H1N1pdm_counts.png}
+	\caption{\textbf{Sample counts through time and across regions.} 
+	}
+	\label{H1N1pdm_counts}
+\end{figure}
+
 Within clade 6b, genetic variants have begun to emerge, most notably HA1:84N/162N/216T, referred to here as clade 6b.1, and (HA1:152T, HA2:174E), referred to here as clade 6b.2 (Fig.\ \ref{H1N1pdm_tree}).
@@
 	\centering		
-	\includegraphics[width=0.8\textwidth]{../figures/feb-2016/H1N1pdm_tree.png}
+	\includegraphics[width=0.9\textwidth]{../figures/feb-2016/H1N1pdm_tree.png}
 	\caption{\textbf{H1N1pdm phylogeny colored by genotype.} 
@@
 Both 6b.1 and 6b.2 viruses have recently increased in frequency (Fig.\ \ref{H1N1pdm_mutations}).
-With the exception of China, all regions of the world are now dominated by 6b.1 (with the caveat that most recent data comes from Europe and data on almost all other regions is sparse).
+With the exception of China, all regions of the world are now dominated by 6b.1.
 The 84N substitution rose in frequency around April 2015 and 162N followed around October 2015.
 Additionally, China and Oceania have seen a recent rise in 6b.2 viruses.
-Recent samples from China have been dominated by 6b.2 viruses.
+Recent samples from China have been mixed between 6b.1 and 6b.2 viruses.
+We note that that are very few Oceania viruses on which to base estimates resulting in wide confidence intervals.
 
@@
 	\centering		
-	\includegraphics[width=0.9\textwidth]{../figures/feb-2016/H1N1pdm_mutations.png}
+	\includegraphics[width=0.95\textwidth]{../figures/feb-2016/H1N1pdm_mutations.png}
 	\caption{\textbf{Frequency trajectories of H1N1pdm variants.} 
@@
 	\label{H1N1pdm_mutations}
+\end{figure}
+
+Although recent samples from China have contained a large fraction of 6b.2 viruses, samples have been limited in geographic scope with the majority of Chinese 6b.2 isolates originating in Shanghai (Fig.\ \ref{H1N1pdm_6b2_tree}).
+Further sampling of a more diverse geographic area may revise our estimates of the frequency of 6b.2 in Asia downwards.
+
+\begin{figure}[h!]
+	\centering		
+	\includegraphics[width=1.0\textwidth]{../figures/feb-2016/H1N1pdm_6b2_tree.png}
+	\caption{\textbf{H1N1pdm phylogeny showing clade 6b.2 viruses colored by sampling location.} 
+	}
+	\label{H1N1pdm_6b2_tree}
 \end{figure}
@@
 We lack antigenic data to catalog an effect from the 162N, 216T or 152T substitutions.
-Here we show recent antigenic data from the Crick Worldwide Influenza Centre alongside model fits \cite{neher2015prediction} (Fig.\ \ref{H1N1pdm_HI}).
+Here we show recent antigenic data (through September 2015) from the Crick Worldwide Influenza Centre alongside model fits \cite{neher2015prediction} (Fig.\ \ref{H1N1pdm_HI}).
 
@@
 	\centering		
-	\includegraphics[width=0.9\textwidth]{../figures/feb-2016/Vic_mutations.png}
+	\includegraphics[width=0.95\textwidth]{../figures/feb-2016/Vic_mutations.png}
 	\caption{\textbf{Frequency trajectories of B/Vic variants.} 
@@
 	\centering		
-	\includegraphics[width=0.9\textwidth]{../figures/feb-2016/Yam_clades.png}
+	\includegraphics[width=0.95\textwidth]{../figures/feb-2016/Yam_clades.png}
 	\caption{\textbf{Frequency trajectories of B/Yam clades.} 
@@
 	\centering		
-	\includegraphics[width=0.9\textwidth]{../figures/feb-2016/Yam_mutations.png}
+	\includegraphics[width=0.95\textwidth]{../figures/feb-2016/Yam_mutations.png}
 	\caption{\textbf{Frequency trajectories of B/Yam variants.} 

### 2016-02-19 auspice/reports/plot_clades.py
Replace with 95% CIs.

@@ -59,7 +59,7 @@ yearsFmt = DateFormatter('%Y')
 monthsFmt = DateFormatter("%b")
 
 n=2
-n_std_dev=1.5
+n_std_dev=2
 l = len(freqs['clades']['global']['pivots'])
 bins = np.array([c[0] for c in counts])[-l:]
 date_bins = []

### 2016-02-19 auspice/reports/tex/feb-2016.tex
Update TeX and include figure legends. Fixes #130.

@@
 \usepackage{geometry}
-\geometry{textwidth=16.5cm} % 15.25cm for single-space, 16.25cm for double-space
+\geometry{textwidth=16.75cm} % 15.25cm for single-space, 16.25cm for double-space
 \geometry{textheight=22.5cm} % 22cm for single-space, 22.5cm for double-space
@@
 
-\begin{figure}[h!]
+\begin{figure}[H]
 	\centering		
@@
 	\caption{\textbf{Sample counts through time and across regions.}
+	This is a stacked bar plot, so that most months there are $\sim$100 total samples and $\sim$15 samples each from North America and from Europe.
 	}
@@
 
+\pagebreak
+
 Throughout 2015, clade 3c2.a viruses predominated over 3c3.a and 3c3.b viruses (Fig.\ \ref{H3N2_clades}).
@@
 	\caption{\textbf{Frequency trajectories of H3N2 clades.}
+	We estimate frequencies of different clades based on sample counts and collection dates.
+	We use a Brownian motion process prior to smooth frequencies from month-to-month.
+	Transparent bands show an estimate the 95\% confidence interval based on sample counts.
+	The final point represents our frequency estimate for Feb 1 2016.
 	}
@@
 \end{figure}
+
+\pagebreak
 
@@
 
+\pagebreak
+
 Barring substantial changes in other clades, we predict the (HA1:171K, HA2:77V/155E) variant to dominate.
 
-\begin{figure}[h!]
+\begin{figure}[H]
 	\centering		
@@
 	\caption{\textbf{Frequency trajectories of 3c2.a subclades.}
+	We estimate frequencies of different clades based on sample counts and collection dates.
+	We use a Brownian motion process prior to smooth frequencies from month-to-month.
+	Transparent bands show an estimate the 95\% confidence interval based on sample counts.
+	The final point represents our frequency estimate for Feb 1 2016.	
 	}
@@
 	\caption{\textbf{H3N2 phylogeny colored by antigenicity.}
+	The left panel shows individual HI measurements of viruses against serum from A/HongKong/5738/2014.
+	Cooler color indicates great antigenic similarity (less titer drop going from homologous to heterologous titers).
+	The right panel shows model estimates of antigenic divergence from the root of the phylogeny.
+	These estimates combine HI measurements from many different sera.
+	Data is current through September 2015.
 	}
@@
 	\caption{\textbf{Viruses basal to the (HA1:171K, HA2:77V/155E) clade (left panel) and to the HA1:142K/197R/168V clade (right panel).}
+	Viruses are colored according to sampling region.
 	}
@@
 	\caption{\textbf{Sample counts through time and across regions.}
+	This is a stacked bar plot, so that in good months there are $\sim$100 total samples and $\sim$15 samples each from North America and from Europe.	
 	}
@@
 
+\pagebreak
+
 Within clade 6b, genetic variants have begun to emerge, most notably HA1:84N/162N/216T, referred to here as clade 6b.1, and (HA1:152T, HA2:174E), referred to here as clade 6b.2 (Fig.\ \ref{H1N1pdm_tree}).
@@
 	\caption{\textbf{Frequency trajectories of H1N1pdm variants.}
+	We estimate frequencies of different clades based on sample counts and collection dates.
+	We use a Brownian motion process prior to smooth frequencies from month-to-month.
+	Transparent bands show an estimate the 95\% confidence interval based on sample counts.
+	The final point represents our frequency estimate for Feb 1 2016.	
 	}
@@
 
+\pagebreak
+
 Although recent samples from China have contained a large fraction of 6b.2 viruses, samples have been limited in geographic scope with the majority of Chinese 6b.2 isolates originating in Shanghai (Fig.\ \ref{H1N1pdm_6b2_tree}).
@@
 	\caption{\textbf{H1N1pdm phylogeny colored by antigenicity.}
+	The left panel shows individual HI measurements of viruses against serum from A/HongKong/5659/2012.
+	Cooler color indicates great antigenic similarity (less titer drop going from homologous to heterologous titers).
+	The right panel shows model estimates of antigenic divergence from the root of the phylogeny.
+	These estimates combine HI measurements from many different sera.
+	Data is current through September 2015.	
 	}
@@
 \end{figure}
-
-\pagebreak
 
@@
 	\caption{\textbf{Frequency trajectories of B/Vic variants.}
+	We estimate frequencies of different clades based on sample counts and collection dates.
+	We use a Brownian motion process prior to smooth frequencies from month-to-month.
+	Transparent bands show an estimate the 95\% confidence interval based on sample counts.
+	The final point represents our frequency estimate for Feb 1 2016.	
 	}
@@
 	\caption{\textbf{Frequency trajectories of B/Yam clades.}
+	We estimate frequencies of different clades based on sample counts and collection dates.
+	We use a Brownian motion process prior to smooth frequencies from month-to-month.
+	Transparent bands show an estimate the 95\% confidence interval based on sample counts.
+	The final point represents our frequency estimate for Feb 1 2016.	
 	}
@@
 	\caption{\textbf{Frequency trajectories of B/Yam variants.}
+	We estimate frequencies of different clades based on sample counts and collection dates.
+	We use a Brownian motion process prior to smooth frequencies from month-to-month.
+	Transparent bands show an estimate the 95\% confidence interval based on sample counts.
+	The final point represents our frequency estimate for Feb 1 2016.	
 	}

### 2016-02-19 auspice/reports/plot_clades.py
updated Asia count list

@@ -18,7 +18,7 @@ virus = 'H3N2'
 freqs = json.load(open('../data/'+virus+'_2y_frequencies.json'))
 counts = json.load(open('../data/'+virus+'_2y_meta.json'))['virus_stats']
 region_names = json.load(open('../data/'+virus+'_2y_meta.json'))['regions']
-region_codes = {'EU':['Europe'], 'AS':['China', 'JapanKorea', 'SouthAsia', 'SoutheastAsia', 'WestAsia'],
+region_codes = {'EU':['Europe'], 'AS':['China', 'JapanKorea','SoutheastAsia'],
                 'NA':["NorthAmerica"], 'OC':["Oceania"]}
 
 

### 2016-02-19 auspice/reports/tex/feb-2016.tex
Fix sampling description.

@@
 This equilibration attempts to collect equal samples from Africa, China, Europe, Japan/South Korea, North America, Oceania, South America, South Asia, Southeast Asia and West Asia.
-In the following analysis we collapse samples from China, South Asia, Southeast Asia and West Asia into a single region, resulting in Asia possessing greater sample counts than North America or Europe.
+In the following analysis we collapse samples from China, Southeast Asia, Japan and Korea into a single region that we're calling "East Asia", resulting in East Asia possessing greater sample counts than North America or Europe.
 The only month that significantly departs from equitable sampling is Jan 2016 with 61 viruses, primarily from Europe and North America.
 Oct 2015 to Jan 2016 are lacking Southern Hemisphere viruses.
+We subsample to 100 viruses per month and not more to keep sample counts as equitable as possible across space and time.
 
@@
 
-We base our primary analysis on a set viruses collected between Feb 2014 and Feb 2016, comprising approximately 100 viruses per month where available and seeking to equilibriate sample counts geographically where possible.
+We base our primary analysis on a set viruses collected between Feb 2014 and Feb 2016, comprising approximately 100 viruses per month where available and seeking to equilibriate sample counts geographically where possible (Fig.\ \ref{H1N1pdm_counts}).
 As discussed above, this equilibration attempts to collect equal samples from Africa, China, Europe, Japan/South Korea, North America, Oceania, South America, South Asia, Southeast Asia and West Asia.
+In the following analysis we collapse samples from China, Southeast Asia, Japan and Korea into a single region that we're calling "East Asia", resulting in East Asia possessing greater sample counts than North America or Europe.
 Recent months through Jan 2016 have largely sufficient sample counts and sample distributions, although are lacking Southern Hemisphere viruses.
+We subsample to 100 viruses per month and not more to keep sample counts as equitable as possible across space and time.
 

### 2016-02-19 auspice/reports/plot_clades.py

@@ -49,7 +49,7 @@ offset = datetime(2000,1,1).toordinal()
 pivots = [offset+(x-2000)*365.25 for x in  freqs['clades']['global']['pivots']]
 pivots.pop()
 regions = ['global', 'NA', 'AS', 'EU', 'OC']
-region_label = {'global': 'Global', 'NA': 'N America', 'AS': 'Asia', 'EU': 'Europe', 'OC': 'Oceania'}
+region_label = {'global': 'Global', 'NA': 'N America', 'AS': 'E Asia', 'EU': 'Europe', 'OC': 'Oceania'}
 cols = sns.color_palette(n_colors=len(regions))
 fs=12
 

### 2016-09-17 auspice/reports/tex/sep-2016.tex
Include TeX version of Sep report.

\documentclass[11pt,oneside,letterpaper]{article}

% graphicx package, useful for including eps and pdf graphics
\usepackage{graphicx}
\DeclareGraphicsExtensions{.pdf,.png,.jpg}

% basic packages
\usepackage{color}
\usepackage{parskip}
\usepackage{float}
\usepackage{microtype}
\usepackage{url}

% text layout
\usepackage{geometry}
\geometry{textwidth=16.75cm} % 15.25cm for single-space, 16.25cm for double-space
\geometry{textheight=22.5cm} % 22cm for single-space, 22.5cm for double-space

% helps to keep figures from being orphaned on a page by themselves
\renewcommand{\topfraction}{0.85}
\renewcommand{\textfraction}{0.1}

% bold the 'Figure #' in the caption and separate it with a period
% Captions will be left justified
\usepackage[labelfont=bf,labelsep=period,font=small]{caption}

% review layout with double-spacing
%\usepackage{setspace}
%\doublespacing
%\captionsetup{labelfont=bf,labelsep=period,font=doublespacing}

% cite package, to clean up citations in the main text. Do not remove.
\usepackage{cite}
%\renewcommand\citeleft{(}
%\renewcommand\citeright{)}
%\renewcommand\citeform[1]{\textsl{#1}}

% Remove brackets from numbering in list of References
\renewcommand\refname{\large References}
\makeatletter
\renewcommand{\@biblabel}[1]{\quad#1.}
\makeatother

\usepackage{authblk}
\renewcommand\Authands{ \& }
\renewcommand\Authfont{\normalsize \bf}
\renewcommand\Affilfont{\small \normalfont}
\makeatletter
\renewcommand\AB@affilsepx{, \protect\Affilfont}
\makeatother

%%% TOC
\usepackage{patchcmd}
\makeatletter
\patchcommand\@starttoc{\begin{quote}}{\end{quote}}
\usepackage{tocloft}
\renewcommand{\cftsecleader}{\cftdotfill{\cftdotsep}}

%%% TITLE %%%
\title{\vspace{2cm} \LARGE \bf
Seasonal influenza circulation patterns and projections for 2016-2017
}

\author[1]{Trevor Bedford}
\author[2]{Richard A.\ Neher}

\affil[1]{Vaccine and Infectious Disease Division, Fred Hutchinson Cancer Research Center, Seattle, WA, USA}
\affil[2]{Max Planck Institute for Developmental Biology, T\"ubingen, Germany}

\date{September 16, 2016}

\begin{document}

\maketitle

\tableofcontents

\pagebreak

%%% A/H3N2 %%%
\section*{A/H3N2}
\addcontentsline{toc}{section}{A/H3N2}

\textbf{Despite the late-season 3c3.a epidemic in the USA, we predict clade 3c2.a viruses will continue to predominate in the H3N2 population. Within clade 3c2.a, the 171K variant has spread rapidly so that the majority of recent H3N2 infections are comprised of 171K viruses. Barring the emergence of a new antigenic variant, we believe 171K will continue to predominate.}

We base our primary analysis on a set of viruses collected between Sep 2014 and Aug 2016, comprising approximately 100 viruses per month where available and seeking to equilibrate sample counts geographically where possible (Fig.\ \ref{H3N2_counts}). This equilibration attempts to collect equal samples from Africa, China, Europe, Japan/South Korea, North America, Oceania, South America, South Asia, Southeast Asia and West Asia. In the following analysis we collapse samples from China, South Asia, Southeast Asia, Japan and Korea into a single region referred to here as ``Asia'', resulting in Asia possessing greater sample counts than North America or Europe. The only month that significantly departs from equitable sampling is Aug 2016 with 38 viruses, primarily from Europe and North America. We subsample to 100 viruses per month and not more to keep sample counts as equitable as possible across space and time. Repeating mutation and clade frequency calculations with up to 200 viruses per month and region yields similar results (see below).

\begin{figure}[H]
	\centering
	\includegraphics[width=0.95\textwidth]{../figures/sep-2016/H3N2_counts.png}
	\caption{\textbf{Sample counts through time and across regions.}
	This is a stacked bar plot, so that most months there are $\sim$100 total samples and $\sim$15 samples each from North America and from Europe.
	}
	\label{H3N2_counts}
\end{figure}

\pagebreak

Viral clades 3c3.a, 3c3.b and 3c2.a emerged from the Texas/2012 background in early 2014 and rapidly spread through the viral population. Subsequently, we have observed competition among these clades, with 3c2.a viruses being globally dominant beginning in 2015 (Fig.\ \ref{H3N2_clades}). Recently, we have observed the steady decline of 3c3.b viruses. At this point, we estimate that they are either extinct or nearly extinct. 3c3.a viruses were largely replaced by 3c2.a viruses starting in 2015. However, we have observed an anomalous late-season epidemic of 3c3.a viruses within the USA from Jan to Aug 2016. Elsewhere in the world, 3c2.a viruses have remained dominant, particularly in Asia, where we estimate 3c2.a frequencies to be $>$95\% throughout 2016.

\begin{figure}[H]
	\centering
	\includegraphics[width=0.95\textwidth]{../figures/sep-2016/H3N2_clades.png}
	\caption{\textbf{Frequency trajectories of H3N2 clades.}
	We estimate frequencies of different clades based on sample counts and collection dates.
	We use a Brownian motion process prior to smooth frequencies from month-to-month.
	Transparent bands show an estimate the 95\% confidence interval based on sample counts.
	The final point represents our frequency estimate for Sep 1 2016.
	}
	\label{H3N2_clades}
\end{figure}

\pagebreak

The late-season 3c3.a epidemic in the USA is puzzling. However, we suspect that this epidemic is due to epidemiologic circumstance rather than the emergence of a selective variant. The strongest evidence for this is that there is not a single clade within 3c3.a that is spreading throughout the USA (Fig.\ \ref{H3N2_3c3a_tree}). Instead, a variety of 3c3.a viruses are spreading, each with different HA1 mutations. The emergence and spread of an adaptive variant would have appeared as a single clade bearing a characteristic epitope mutation. This is not what we see. One small clade, however, carries the S145N and F193S mutations close to the receptor binding site which resulted in antigenic evolution in previous years. The clade is restricted to viruses from US and remains at less than 1\% global frequency. We doubt that these 3c3.a viruses will spread globally. It is conceivable however, that the 2016-2017 USA season could derive from over-summering transmission chains. This would be an unusual event, but not unheard of \cite{bedford2010global}. Still, we believe that 2016-2017 USA season will likely derive from 171K viruses due to their rapid spread (see below).

\begin{figure}[H]
	\centering
	\includegraphics[width=0.9\textwidth]{../figures/sep-2016/H3N2_3c3a_tree.png}
	\caption{\textbf{H3N2 / 3c3.a phylogeny colored by geography.}
	}
	\label{H3N2_3c3a_tree}
\end{figure}

\pagebreak

The dominant 3c2.a viruses have continued to genetically diversify with the emergence of multiple subclades (Fig.\ \ref{H3N2_3c2a_tree}). We observe 3 subclades of decent frequency in 2016 viruses. These are characterized by HA1 mutations 142K/197R, HA1 mutation 197K and HA1 mutation 171K + HA2 mutations 77V/155E. The 171K variant comprises approximately 54\% of 3c2.a viruses collected in 2016.

\begin{figure}[H]
	\centering
	\includegraphics[width=0.9\textwidth]{../figures/sep-2016/H3N2_3c2a_tree.png}
	\caption{\textbf{H3N2 / 3c2.a phylogeny colored by genotype.}
	}
	\label{H3N2_3c2a_tree}
\end{figure}

The 142K/197R clade first emerged around June 2015 and rose to nearly 20\% global frequency in Nov 2015 (Fig.\ \ref{H3N2_mutations}). However, it's declined throughout 2016. We doubt this is a competitive virus. The 197K clade emerged in late 2015 and has slowly grown in frequency since. We estimate that it now comprises $\sim$14\% of H3N2 viruses. However, 197K is at extremely low frequency in Asia with an estimated present-day frequency $\sim$2\%. On the other hand, clade 171K viruses have done remarkably well throughout 2016 and now comprise an estimated 69\% of currently circulating H3N2 viruses. This steady and rapid increase is strongly suggestive of an adaptive origin. Notably, 171K emerged and spread first in Asia, reaching nearly 80\% frequency in April 2016. Higher frequency of 171K in Asia is expected to spread to the rest of the world given historic geographic observations \cite{bedford2015global}.

\pagebreak

\textit{Without strong competition from another novel H3N2 virus, we believe that 171K will continue to increase in frequency in the global population and predominate in the 2016-2017 influenza season.} The continued spread of 171K is fully in line with the predictions we made in Feb 2016 \cite{feb2016report}.

\begin{figure}[H]
	\centering
	\includegraphics[width=0.95\textwidth]{../figures/feb-2016/H3N2_mutations.png}
	\caption{\textbf{Frequency trajectories of 3c2.a subclades.}
	We estimate frequencies of different clades based on sample counts and collection dates.
	We use a Brownian motion process prior to smooth frequencies from month-to-month.
	Transparent bands show an estimate the 95\% confidence interval based on sample counts.
	The final point represents our frequency estimate for Sep 1 2016.
	}
	\label{H3N2_mutations}
\end{figure}

\pagebreak

Other indicators suggest evolutionary success of 171K viruses. Notably ``local branching index'' \cite{neher2014predicting} supports 171K as a high fitness virus (Fig.\ \ref{H3N2_LBI}). 3c3.a viruses and other clades within 3c2.a and do not show signal in the ``local branching index'' analysis.

\begin{figure}[H]
	\centering
	\includegraphics[width=0.9\textwidth]{../figures/sep-2016/H3N2_LBI.png}
	\caption{\textbf{H3N2 phylogeny colored by local branching index.}
	}
	\label{H3N2_LBI}
\end{figure}

Unfortunately, we lack sufficient recent serological data to distinguish antigenic evolution for most subclades within 3c2.a and 3c3.a. These observations derive entirely from genetic data.

\pagebreak

There are a variety of viruses at the base of the 171K clade, possessing HA1:171K along with HA2:77V/155E, but lacking further amino acid changes in HA (Fig.\ \ref{H3N2_171K_basal}).

\begin{figure}[h!]
	\centering
	\includegraphics[width=1.0\textwidth]{../figures/sep-2016/H3N2_171K_basal.png}
	\caption{\textbf{Viruses basal to the (HA1:171K, HA2:77V/155E) clade.}
	Viruses are colored according to sampling region.
	}
	\label{H3N2_171K_basal}
\end{figure}

\clearpage
\pagebreak

%%% A/H1N1pdm %%%
\section*{A/H1N1pdm}
\addcontentsline{toc}{section}{A/H1N1pdm}

\textbf{The clade 6b.1, comprising 84N/162N/216T, has continued to rise and predominate in the H1N1pdm population throughout 2016. Almost all circulating H1N1pdm viruses are now 6b.1. There is not yet obvious evolution within this clade.}

As discussed above, we base our primary analysis on a set of viruses collected between Sep 2014 and Aug 2016, comprising approximately 100 viruses per month where available and seeking to equilibrate sample counts geographically where possible (Fig.\ \ref{H1N1pdm_counts}). Recent months through June 2016 have largely sufficient sample counts and sample distributions. There are fewer samples from July to present.

\begin{figure}[H]
	\centering
	\includegraphics[width=0.95\textwidth]{../figures/sep-2016/H1N1pdm_counts.png}
	\caption{\textbf{Sample counts through time and across regions.}
	This is a stacked bar plot, so that in good months there are $\sim$100 total samples and $\sim$15 samples each from North America and from Europe.
	}
	\label{H1N1pdm_counts}
\end{figure}

\pagebreak

Within clade 6b, two major genetic variants have emerged (Fig.\ \ref{H1N1pdm_tree}). These are clade 6b.1 comprised of HA1:84N/162N/216T and clade 6b.2 comprised of HA1:152T and HA2:174E. Most recent samples have been from 6b.1 viruses, with 85\% of 2016 samples being from 6b.1 viruses.

\begin{figure}[H]
	\centering
	\includegraphics[width=0.9\textwidth]{../figures/sep-2016/H1N1pdm_tree.png}
	\caption{\textbf{H1N1pdm phylogeny colored by genotype.}
	}
	\label{H1N1pdm_tree}
\end{figure}

\pagebreak

At this point, all regions of the world are dominated by 6b.1 viruses (Fig.\ \ref{H1N1pdm_clades}). This clade rose from low frequency in Aug 2015 to reach present day global frequencies of $\sim$98\%. There remain a minority of circulating 6b.2 viruses. We estimate 2\% of H1N1pdm viruses globally to be 6b.2. These are slightly higher prevalence in Asia, but still a distinct minority. We estimate that 6b.1 is at 85\% frequency in Asia, while 6b.2 is at 15\% in Asia. Notably, the frequency of 6b.2 has remained stable for almost 12 months.

\textit{Every indication suggests the continued dominance of 6b.1 viruses, and their extremely rapid rise suggests a selective origin. We are now watching for the emergence of genetic variants within the 6b.1 clade.} Notably, the continued rise and dominance of 6b.1 viruses fits with our predictions from Feb 2016 \cite{feb2016report}.

\begin{figure}[H]
	\centering
	\includegraphics[width=0.95\textwidth]{../figures/feb-2016/H1N1pdm_clades.png}
	\caption{\textbf{Frequency trajectories of H1N1pdm clades.}
	We estimate frequencies of different clades based on sample counts and collection dates.
	We use a Brownian motion process prior to smooth frequencies from month-to-month.
	Transparent bands show an estimate the 95\% confidence interval based on sample counts.
	The final point represents our frequency estimate for Sep 1 2016.
	}
	\label{H1N1pdm_clades}
\end{figure}

\clearpage
\pagebreak

%%% B/Victoria %%%
\section*{B/Vic}
\addcontentsline{toc}{section}{B/Vic}

\textbf{Within clade 1A viruses, the clade 129D/146I/117V has risen to high frequency, but at a rate that suggests a smaller effect of natural selection. At this point, 117V viruses are dominant in the global viral population.}

As above, we base our primary analysis on a set of viruses collected between Sep 2014 and Aug 2016, comprising approximately 100 viruses per month where available and seeking to equilibrate sample counts geographically where possible (Fig.\ \ref{Vic_counts}).

\begin{figure}[H]
	\centering
	\includegraphics[width=0.95\textwidth]{../figures/sep-2016/Vic_counts.png}
	\caption{\textbf{Sample counts through time and across regions.}
	This is a stacked bar plot, so that in good months there are $\sim$100 total samples and $\sim$15 samples each from North America and from Europe.
	}
	\label{Vic_counts}
\end{figure}

\pagebreak

In the past year, B/Vic clade 129D/146I/117V has dominated the viral population, with 94\% of samples possessing the 117V mutation (Fig.\ \ref{Vic_mutations}). This dominance has increased throughout 2016 and we estimate that currently circulating Vic viruses are 94\% 117V. At this point, we haven't observed new variants of appreciable frequency within the 117V clade and there aren't decent competitors outside the 117V clade.

\begin{figure}[H]
	\centering
	\includegraphics[width=0.95\textwidth]{../figures/sep-2016/Vic_mutations.png}
	\caption{\textbf{Frequency trajectories of B/Vic variants.}
	We estimate frequencies of different clades based on sample counts and collection dates.
	We use a Brownian motion process prior to smooth frequencies from month-to-month.
	Transparent bands show an estimate the 95\% confidence interval based on sample counts.
	The final point represents our frequency estimate for Sep 1 2016.
	}
	\label{Vic_mutations}
\end{figure}

\pagebreak

Additionally, the 117V clade has been growing most quickly and is picked by the local branching index \cite{neher2014predicting} as the currently most successful clade (Fig.\ \ref{Vic_LBI}). \textit{All indicators suggest the continued success of 117V in the coming year.}

\begin{figure}[H]
	\centering
	\includegraphics[width=0.9\textwidth]{../figures/sep-2016/Vic_LBI.png}
	\caption{\textbf{B/Vic phylogeny colored by local branching index.}
	}
	\label{Vic_LBI}
\end{figure}

\clearpage
\pagebreak

%%% B/Yamagata %%%
\section*{B/Yam}
\addcontentsline{toc}{section}{B/Yam}

\textbf{Clade 3 has continued to predominate the B/Yamagata population. Within clade 3 the 172Q/251Q subclade has risen to predominate. There is limited diversity within 172Q/251Q viruses, with the largest subclade comprised of 211R viruses.}

As above, we base our primary analysis on a set of viruses collected between Sep 2014 and Jul 2016, comprising approximately 100 viruses per month where available and seeking to equilibrate sample counts geographically where possible (Fig.\ \ref{Yam_counts}).

\begin{figure}[H]
	\centering
	\includegraphics[width=0.95\textwidth]{../figures/sep-2016/Yam_counts.png}
	\caption{\textbf{Sample counts through time and across regions.}
	This is a stacked bar plot, so that in good months there are $\sim$100 total samples and $\sim$15 samples each from North America and from Europe.
	}
	\label{Yam_counts}
\end{figure}

\pagebreak

During 2016, the vast majority of B/Yamagata isolates were of clade 3 viruses (Fig.\ \ref{Yam_clades}). We estimate that the current frequency of clade 2 is now $\sim$2\%. At this rate, we expect clade 2 to go extinct in the coming year.

\begin{figure}[H]
	\centering
	\includegraphics[width=0.95\textwidth]{../figures/feb-2016/Yam_clades.png}
	\caption{\textbf{Frequency trajectories of B/Yam clades.}
	We estimate frequencies of different clades based on sample counts and collection dates.
	We use a Brownian motion process prior to smooth frequencies from month-to-month.
	Transparent bands show an estimate the 95\% confidence interval based on sample counts.
	The final point represents our frequency estimate for Sep 1 2016.
	}
	\label{Yam_clades}
\end{figure}

\pagebreak

Within clade 3, HA1:172Q has predominated throughout 2015 and 2016 (Fig.\ \ref{Yam_tree}). On this background the HA1:251V variant has emerged and on top of 251V, the HA1:211R variant has appeared.

\begin{figure}[H]
	\centering
	\includegraphics[width=0.9\textwidth]{../figures/sep-2016/Yam_tree.png}
	\caption{\textbf{H3N2 phylogeny colored by genotype.}
	}
	\label{Yam_tree}
\end{figure}

\pagebreak

The 251V clade increased from low frequency in Oct 2014 to predominate in the population (Fig.\ \ref{Yam_mutations}). We estimate that 251V is currently at 79\% globally. The 211R variant rose from low frequency in Apr 2015 to reach 34\% in currently circulating viruses. The rate of increase, however, has been rather mild.

\begin{figure}[H]
	\centering
	\includegraphics[width=0.95\textwidth]{../figures/feb-2016/Yam_mutations.png}
	\caption{\textbf{Frequency trajectories of B/Yam variants.}
	We estimate frequencies of different clades based on sample counts and collection dates.
	We use a Brownian motion process prior to smooth frequencies from month-to-month.
	Transparent bands show an estimate the 95\% confidence interval based on sample counts.
	The final point represents our frequency estimate for Sep 1 2016.
	}
	\label{Yam_mutations}
\end{figure}

\clearpage
\pagebreak

%%% A/H3N2 (all sequences) %%%
\section*{A/H3N2 (all sequences)}
\addcontentsline{toc}{section}{A/H3N2 (all sequences)}

The clade and mutation frequencies discussed above were based on a limited sequence subsample with an equitable geographical distribution. More accurate region specific trajectories can be estimated using all sequence data available in GISAID. We repeated the clade and mutation frequency estimation using up to 200 sequences per month and region (Fig.\ \ref{H3N2_counts_all}).

\begin{figure}[H]
	\centering
	\includegraphics[width=0.95\textwidth]{../figures/sep-2016/H3N2_counts_all.png}
	\caption{\textbf{Sample counts through time and across regions.}
	This is a stacked bar plot, so that in the peak of the 2014-2015 season there are $\sim$700 total samples per month.
	}
	\label{H3N2_counts_all}
\end{figure}

\pagebreak

With this more inclusive sampling, northern hemisphere winter months tend to be dominated by North America and Europe, while other months are dominated by samples from North America, Asia, and Oceania (Fig.\ \ref{H3N2_clades_all}). The global average will track the regions contributing the majority of the sequence data.

\begin{figure}[H]
	\centering
	\includegraphics[width=0.95\textwidth]{../figures/sep-2016/H3N2_clades_all.png}
	\caption{\textbf{Frequency trajectories of H3N2 clades.}
	We estimate frequencies of different clades based on sample counts and collection dates.
	We use a Brownian motion process prior to smooth frequencies from month-to-month.
	Transparent bands show an estimate the 95\% confidence interval based on sample counts.
	The final point represents our frequency estimate for Sep 1 2016.
	}
	\label{H3N2_clades_all}
\end{figure}

\pagebreak

The H3N2 clades 3c3.a (159S) and 3c2.a (159Y) show broadly similar trajectories as discussed above (Fig.\ \ref{H3N2_mutations_all}). 3c2.a came to dominate Asia completely, while viruses outside of 3c3.a and 3c2.a continue to circulate at low levels in Oceania. Within 3c2.a, the mutation 171K dominates across all regions.

\begin{figure}[H]
	\centering
	\includegraphics[width=0.95\textwidth]{../figures/sep-2016/H3N2_mutations_all.png}
	\caption{\textbf{Frequency trajectories of H3N2 mutations.}
	We estimate frequencies of different clades based on sample counts and collection dates.
	We use a Brownian motion process prior to smooth frequencies from month-to-month.
	Transparent bands show an estimate the 95\% confidence interval based on sample counts.
	The final point represents our frequency estimate for Sep 1 2016.
	}
	\label{H3N2_mutations_all}
\end{figure}

%%% References %%%
\bibliographystyle{plos}
\bibliography{references}

\end{document}

### 2016-09-17 augur/src/Vic_process.py
Update Vic outliers.

@@ -134,8 +134,8 @@ class BVic_clean(virus_clean):
 		new_viruses = []
 		outlier_strains = ["B/Kol/2024/2008", "B/Kolkata/2024/2008", "B/Brisbine/33/2008", "B/Kolkata/1373/2008",
 			"B/Kolkata/372/2010", "B/Stockholm/7/2011", "B/Bangkok/SI17/2012", "B/Bangkok/SI58/2012",
-			"B/Togo/LNG/419/2013", "B/Netherlands/76/2014", "B/NewCaledonia/119/2015", "B/Brisbane/14/2016",
-			"B/Sydney/6/2016"]
+			"B/Togo/LNG/419/2013", "B/Netherlands/76/2014", "B/NewCaledonia/119/2015", "B/Thailand/CU-B11637/2015",
+			"B/Brisbane/14/2016", "B/Sydney/6/2016", "A/Malaysia/438/2016"]
 		for v in self.viruses:
 			if v.strain in outlier_strains:
 				if self.verbose > 1:

### 2016-09-17 auspice/reports/tex/sep-2016.tex
Fix typos in Sep report.

@@
 	\centering
-	\includegraphics[width=0.95\textwidth]{../figures/feb-2016/H3N2_mutations.png}
+	\includegraphics[width=0.95\textwidth]{../figures/sep-2016/H3N2_mutations.png}
 	\caption{\textbf{Frequency trajectories of 3c2.a subclades.}
@@
 	\centering
-	\includegraphics[width=0.95\textwidth]{../figures/feb-2016/H1N1pdm_clades.png}
+	\includegraphics[width=0.95\textwidth]{../figures/sep-2016/H1N1pdm_clades.png}
 	\caption{\textbf{Frequency trajectories of H1N1pdm clades.}
@@
 	\centering
-	\includegraphics[width=0.95\textwidth]{../figures/feb-2016/Yam_clades.png}
+	\includegraphics[width=0.95\textwidth]{../figures/sep-2016/Yam_clades.png}
 	\caption{\textbf{Frequency trajectories of B/Yam clades.}
@@
 	\includegraphics[width=0.9\textwidth]{../figures/sep-2016/Yam_tree.png}
-	\caption{\textbf{H3N2 phylogeny colored by genotype.}
+	\caption{\textbf{B/Yam phylogeny colored by genotype.}
 	}
@@
 	\centering
-	\includegraphics[width=0.95\textwidth]{../figures/feb-2016/Yam_mutations.png}
+	\includegraphics[width=0.95\textwidth]{../figures/sep-2016/Yam_mutations.png}
 	\caption{\textbf{Frequency trajectories of B/Yam variants.}

### 2016-09-20 auspice/reports/plot_clades.py
Include new H3 and H1 figures. Still need to update report text.

@@ -27,7 +27,7 @@ region_codes = {'EU':['europe'], 'AS':['china', 'south_asia', 'japan_korea','sou
 
 if virus=='H3N2': ########## H3N2
     clades = ['3c2.a', '3c3.a', '3c3.b']
-    mutations = ['HA1:142K', 'HA1:197K', 'HA1:171K']
+    mutations = ['HA1:142K', 'HA1:197K', 'HA1:171K', 'HA1:121K']
     clade_legend = {'panel':0, 'loc':3}
     mut_legend = {'panel':0, 'loc':3}
 elif virus=='H1N1pdm': ########## H1N1pdm

### 2016-09-20 auspice/reports/tex/sep-2016.tex
Update TeX for Sep report.

@@
 
-\date{September 16, 2016}
+\date{September 20, 2016}
 
@@
 One small clade, however, carries the S145N and F193S mutations close to the receptor binding site which resulted in antigenic evolution in previous years.
-The clade is restricted to viruses from US and remains at less than 1\% global frequency.
-We doubt that these 3c3.a viruses will spread globally.
-It is conceivable however, that the 2016-2017 USA season could derive from over-summering transmission chains.
-This would be an unusual event, but not unheard of \cite{bedford2010global}.
-Still, we believe that 2016-2017 USA season will likely derive from 171K viruses due to their rapid spread (see below).
+This is a small clade restricted to viruses from US and remains at less than 1\% global frequency.
 
@@
 
+We doubt that these 3c3.a viruses will spread globally.
+It is conceivable however, that the 2016-2017 USA season could derive from over-summering transmission chains.
+This would be an unusual event, but not unheard of \cite{bedford2010global,bedford2015global}.
+In one recent season (2008-2009), USA viruses derived primarily from the previous USA season.
+In other years (at least back to 2000), USA seasons were primarily reseeded from elsewhere (Fig.\ \ref{H3N2_historical_tree_na}).
+We believe that the 2016-2017 USA season will likely derive from 171K viruses due to their rapid spread (see below).
+
+\begin{figure}[H]
+	\centering
+	\includegraphics[width=0.9\textwidth]{../figures/sep-2016/H3N2_historical_tree_na.png}
+	\caption{\textbf{Historical phylogeny of H3N2 viruses from North America from Bedford et al.\ (2015) \cite{bedford2015global}.}
+	}
+	\label{H3N2_historical_tree_na}
+\end{figure}
+
+\pagebreak
+
 The dominant 3c2.a viruses have continued to genetically diversify with the emergence of multiple subclades (Fig.\ \ref{H3N2_3c2a_tree}).
@@
 These are characterized by HA1 mutations 142K/197R, HA1 mutation 197K and HA1 mutation 171K + HA2 mutations 77V/155E.
-The 171K variant comprises approximately 54\% of 3c2.a viruses collected in 2016.
+The 171K variant comprises approximately 58\% of 3c2.a viruses collected in 2016.
+Within 171K, the variant 121K has emerged and comprises approximately 23\% of 3c2.a viruses collected in 2016.
 
@@
 Higher frequency of 171K in Asia is expected to spread to the rest of the world given historic geographic observations \cite{bedford2015global}.
+Within 171K, the variant 121K has spread.
+However, it doesn't appear to be spreading more rapidly than its parent 171K clade.
+Phylogenetic patterns suggest 171K as driver rather than 121K.
+Still, the rate of increase of 121K suggests that a sizable fraction of 2016-2017 viruses will be comprised of 121K.
 
@@
 \textit{Without strong competition from another novel H3N2 virus, we believe that 171K will continue to increase in frequency in the global population and predominate in the 2016-2017 influenza season.} The continued spread of 171K is fully in line with the predictions we made in Feb 2016 \cite{feb2016report}.
+At this point, we can't say whether the 171K mutation in HA1 or the HA2 mutations 77V/155E (or some combination) is driving the selective spread of this clade.
 
@@
 
+\pagebreak
+
+Analysis of local branching index \cite{neher2014predicting} also flags the clade 6b.1 as rapidly expanding and the most successful lineages within H1N1pdm (Fig.\ \ref{H1N1pdm_LBI}).
+
+\begin{figure}[H]
+	\centering
+	\includegraphics[width=0.9\textwidth]{../figures/sep-2016/H1N1pdm_LBI.png}
+	\caption{\textbf{H1N1pdm phylogeny colored by local branching index.}
+	}
+	\label{H1N1pdm_LBI}
+\end{figure}
+
+\pagebreak
+
+Analysis of antigenic evolution via the method of Neher et al.
+(2015) \cite{neher2015prediction} and using recent HI data from the Feb 2016 VCM report \cite{crickfeb2016report} from the Crick Institute Collaborating Center suggests a small effect on HI titer in both the 6b.1 and 6b.2 clades (Fig.\ \ref{H1N1pdm_HI}).
+In this case, 6b.1 shows an antigenic impact of 0.65 log2 units and 6b.2 shows an impact of 0.66 log2 units.
+Interesting, this analysis places the impact on clade 6b.1 from the 84N mutation rather than the later 162N or 216T mutations.
+
+\begin{figure}[h!]
+	\centering
+	\includegraphics[width=1.0\textwidth]{../figures/sep-2016/H1N1pdm_HI.png}
+	\caption{\textbf{H1N1pdm phylogeny colored by antigenicity.}
+	The left panel shows individual HI measurements of viruses against serum from A/HongKong/5659/2012.
+	Cooler color indicates great antigenic similarity (less titer drop going from homologous to heterologous titers).
+	The right panel shows model estimates of antigenic divergence from A/HongKong/5659/2012.
+	These estimates combine HI measurements from many different sera.
+	Data is current through Feb 2016.
+	}
+	\label{H1N1pdm_HI}
+\end{figure}
+
 \clearpage
@@
 \textbf{Clade 3 has continued to predominate the B/Yamagata population.
-Within clade 3 the 172Q/251Q subclade has risen to predominate.
-There is limited diversity within 172Q/251Q viruses, with the largest subclade comprised of 211R viruses.}
+Within clade 3 the 172Q/251V subclade has risen to predominate.
+There is limited diversity within 172Q/251V viruses, with the largest subclade comprised of 211R viruses.}
 

### 2016-09-21 auspice/reports/tex/sep-2016.tex

@@
 
-\date{September 20, 2016}
+\date{September 21, 2016}
 
@@
 We subsample to 100 viruses per month and not more to keep sample counts as equitable as possible across space and time.
-Repeating mutation and clade frequency calculations with up to 200 viruses per month and region yields similar results (see below).
+Repeating mutation and clade frequency calculations with up to 1200 viruses per month yields similar results (see below).
 
@@
 However, it's declined throughout 2016.
-We doubt this is a competitive virus.
+We doubt this is a competitive virus based on current clade success, but site 142 mutated several times within 3c2a to 142G or 142K suggesting that mutating this position might enable the virus to escape preexisting immunity at a fitness cost; it warrants continued observation.
 The 197K clade emerged in late 2015 and has slowly grown in frequency since.
@@
 More accurate region specific trajectories can be estimated using all sequence data available in GISAID.
-We repeated the clade and mutation frequency estimation using up to 200 sequences per month and region (Fig.\ \ref{H3N2_counts_all}).
+We repeated the clade and mutation frequency estimation using up to 500 sequences per month and region (Fig.\ \ref{H3N2_counts_all}).
 

### 2016-09-21 auspice/reports/tex/sep-2016.tex
Update references and add disclaimer to tex report.

@@
 \usepackage{url}
+\usepackage{hyperref}
 
@@
 \maketitle
+
+\textit{This is not meant as a comprehensive report, but is instead intended as particular observations that may be of relevance.
+Please also note that observed patterns reflect the GISAID database and may not be entirely representative of underlying dynamics.
+All analyses are based on the nextflu pipeline \cite{neher2015nextflu} with continual updates posted to \href{http://nextflu.org}{nextflu.org}.}
 

### 2017-02-17 auspice/reports/plot_clades.py
Update H1N1pdm section

@@ -10,8 +10,8 @@ plt.ion()
 
 show_errorbars = True
 
-virus = 'H3N2'
-#virus = 'H1N1pdm'
+#virus = 'H3N2'
+virus = 'H1N1pdm'
 #virus = 'Vic'
 #virus = 'Yam'
 
@@ -33,7 +33,7 @@ if virus=='H3N2': ########## H3N2
     mut_legend = {'panel':0, 'loc':3}
 elif virus=='H1N1pdm': ########## H1N1pdm
     clades = ['6b.1', '6b.2']
-    mutations = ['HA1:84N','HA1:162N','HA1:152T'] #these don't add up to one in Asia, probably due to sketchy sampling.
+    mutations = ['HA1:205K', 'HA1:183P', 'HA1:166V']
     clade_legend = {'panel':0, 'loc':3}
     mut_legend = {'panel':0, 'loc':3}
 elif virus=='Vic':
@@ -100,6 +100,7 @@ sns.despine()
 plt.savefig('figures/' + report + '/'+virus+'_counts.png')
 
 if len(clades):
+    print "Plotting clade frequencies"
     fig, axs = plt.subplots(len(clades), 1, sharex=True, figsize=(8, len(clades)*2))
     for clade, ax in zip(clades, axs):
         for c,region in zip(cols, regions):
@@ -135,7 +136,7 @@ if len(clades):
     sns.despine()
     plt.savefig('figures/' + report + '/'+virus+'_clades.png')
 
-
+print "Plotting mutation frequencies"
 fig, axs = plt.subplots(len(mutations), 1, sharex=True, figsize=(8, len(mutations)*2))
 for mutation, ax in zip(mutations, axs):
     for c,region in zip(cols, regions):

### 2017-02-17 auspice/reports/plot_clades.py
Update Vic in Feb report

@@ -11,8 +11,8 @@ plt.ion()
 show_errorbars = True
 
 #virus = 'H3N2'
-virus = 'H1N1pdm'
-#virus = 'Vic'
+#virus = 'H1N1pdm'
+virus = 'Vic'
 #virus = 'Yam'
 
 resolution = '2y'

### 2017-02-17 auspice/reports/plot_clades.py
Update Yam section of Feb report

@@ -10,9 +10,9 @@ plt.ion()
 
 show_errorbars = True
 
-#virus = 'H3N2'
+virus = 'H3N2'
 #virus = 'H1N1pdm'
-virus = 'Vic'
+#virus = 'Vic'
 #virus = 'Yam'
 
 resolution = '2y'

### 2017-02-17 auspice/reports/tex/feb-2017.tex
Include TeX version of Feb report

\documentclass[11pt,oneside,letterpaper]{article}

% graphicx package, useful for including eps and pdf graphics
\usepackage{graphicx}
\DeclareGraphicsExtensions{.pdf,.png,.jpg}

% basic packages
\usepackage{color}
\usepackage{parskip}
\usepackage{float}
\usepackage{microtype}
\usepackage{url}
\usepackage{hyperref}

% text layout
\usepackage{geometry}
\geometry{textwidth=16.75cm} % 15.25cm for single-space, 16.25cm for double-space
\geometry{textheight=22.5cm} % 22cm for single-space, 22.5cm for double-space

% helps to keep figures from being orphaned on a page by themselves
\renewcommand{\topfraction}{0.85}
\renewcommand{\textfraction}{0.1}

% bold the 'Figure #' in the caption and separate it with a period
% Captions will be left justified
\usepackage[labelfont=bf,labelsep=period,font=small]{caption}

% review layout with double-spacing
%\usepackage{setspace}
%\doublespacing
%\captionsetup{labelfont=bf,labelsep=period,font=doublespacing}

% cite package, to clean up citations in the main text. Do not remove.
\usepackage{cite}
%\renewcommand\citeleft{(}
%\renewcommand\citeright{)}
%\renewcommand\citeform[1]{\textsl{#1}}

% Remove brackets from numbering in list of References
\renewcommand\refname{\large References}
\makeatletter
\renewcommand{\@biblabel}[1]{\quad#1.}
\makeatother

\usepackage{authblk}
\renewcommand\Authands{ \& }
\renewcommand\Authfont{\normalsize \bf}
\renewcommand\Affilfont{\small \normalfont}
\makeatletter
\renewcommand\AB@affilsepx{, \protect\Affilfont}
\makeatother

%%% TOC
\usepackage{patchcmd}
\makeatletter
\patchcommand\@starttoc{\begin{quote}}{\end{quote}}
\usepackage{tocloft}
\renewcommand{\cftsecleader}{\cftdotfill{\cftdotsep}}

%%% TITLE %%%
\title{\vspace{2cm} \LARGE \bf
Seasonal influenza circulation patterns and projections for 2017-2018
}

\author[1]{Trevor Bedford}
\author[2]{Richard A.\ Neher}

\affil[1]{Vaccine and Infectious Disease Division, Fred Hutchinson Cancer Research Center, Seattle, WA, USA}
\affil[2]{Biozentrum, University of Basel, Basel, Switzerland}

\date{February 16, 2017}

\begin{document}

\maketitle

\textit{This is not meant as a comprehensive report, but is instead intended as particular observations that may be of relevance. Please also note that observed patterns reflect the GISAID database and may not be entirely representative of underlying dynamics. All analyses are based on the nextflu pipeline \cite{neher2015nextflu} with continual updates posted to \href{http://nextflu.org}{nextflu.org}.}

\tableofcontents

\pagebreak

%%% A/H3N2 %%%
\section*{A/H3N2}
\addcontentsline{toc}{section}{A/H3N2}

\textbf{In H3N2, clade 3c2.a has continued to diversify genetically with complicated and rapid dynamics of different subclades.
This diversification is not reflected in serological data that shows only minor to moderate antigenic evolution.
Nevertheless, the highly parallel mutation patterns and the rapid rise and fall of clades suggests competitive dynamics of phenotypically distinct viruses.}

We base our primary analysis on a set of viruses collected between Jan 2015 and Feb 2017 (Fig.\ \ref{H3N2_counts}).
We analyze mutation frequency trajectories separately in different geographic regions using all available sequence data.
Our phylogenetic analysis is based on a subsample of approximately 100 viruses per month where available and seeking to equilibrate sample counts geographically where possible.
This equilibration attempts to collect equal samples from Africa, China, Europe, Japan/South Korea, North America, Oceania, South America, South Asia, Southeast Asia and West Asia.
In the following analysis we collapse samples from China, South Asia, Southeast Asia, Japan and Korea into a single region referred to here as ``Asia'', resulting in Asia possessing greater sample counts than North America or Europe.
There is reasonably broad geographic sampling throughout, except Dec 2016 and Jan 2017, which have more samples from Europe and North America.

\begin{figure}[H]
	\centering
	\includegraphics[width=0.95\textwidth]{../figures/feb-2017/H3N2_counts.png}
	\caption{\textbf{Sample counts through time and across regions.}
	This is a stacked bar plot, so that most months there are $\sim$100 total samples and $\sim$15 samples each from North America and from Europe.
	}
	\label{H3N2_counts}
\end{figure}

\pagebreak

Of the clades that emerged from the Texas/2012 background in early 2014, only 3c2.a currently remains at globally high prevalence (Fig.\ \ref{H3N2_clades}).
3c3.b viruses have scarcly been seen since the end of 2015.
While the uptick of 3c3.a viruses in mid-2016 in the US has failed to carry forward to the 2016-2017 US winter season.
Oceania still harbors 3c3.a viruses, but we fully expect these to be replaced by another clade in the coming year as Oceania is reseeded from the global reservoir.

\begin{figure}[H]
	\centering
	\includegraphics[width=0.95\textwidth]{../figures/feb-2017/H3N2_clades.png}
	\caption{\textbf{Frequency trajectories of H3N2 clades.}
	We estimate frequencies of different clades based on sample counts and collection dates.
	We use a Brownian motion process prior to smooth frequencies from month-to-month.
	Transparent bands show an estimate the 95\% confidence interval based on sample counts.
	The final point represents our frequency estimate for Feb 1 2017.
	}
	\label{H3N2_clades}
\end{figure}

\pagebreak

Within clade 3c2.a, a number of subclades with signature mutations at N171K, T131K/R142K, R142K, N121K/R144K, Q197K have emerged (Fig.\ \ref{H3N2_tree}).
Clades with R142K, N121K/R144K, Q197K arose in late 2014 or early 2015 but have not come to dominate.
However, the 171K clade that was first observed in mid-2015 has continued to increase in frequency at a moderate rate and is now at a global frequency of approximately 60\%.
Within the 171K clade, a subclade bearing the 121K mutation emerged in at the beginning of 2016 and has steadily increased in frequency now making of the majority of the parent 171K clade.

\begin{figure}[H]
	\centering
	\includegraphics[width=0.9\textwidth]{../figures/feb-2017/H3N2_tree.png}
	\caption{\textbf{H3N2 / 3c2.a phylogeny colored by genotype.}
	}
	\label{H3N2_tree}
\end{figure}

Currently, the only noteworthy sister clade of 171K is the clade with mutations T131K/R142K (Fig.\ \ref{H3N2_mutations}).
This clade was first observed mid-2016 in China and dominated in China during last 3 month according to sequences available in GISAID, making up ~80\% of Chinese isolate since October.
It has been rising rapidly and has spread to Europe and North America.
In the branch leading up to this clade, we observe a rapid succession of mutation in codon 131 from threonine to asparagine and then to lysine and the mutation R142K.
Rapid succession of multiple mutations in the same codon suggest that these changes are adaptive.
Interestingly, we observe a large number of mutations to positively charged residues and particularly many parallel mutations to lysine at positions 121, 131, 142, 144, 171, and 197.

\pagebreak

\begin{figure}[H]
	\centering
	\includegraphics[width=0.95\textwidth]{../figures/feb-2017/H3N2_mutations.png}
	\caption{\textbf{Frequency trajectories of 3c2.a subclades.}
	We estimate frequencies of different clades based on sample counts and collection dates.
	We use a Brownian motion process prior to smooth frequencies from month-to-month.
	Transparent bands show an estimate the 95\% confidence interval based on sample counts.
	The final point represents our frequency estimate for Feb 1 2017.
	}
	\label{H3N2_mutations}
\end{figure}

\pagebreak

Despite the large genetic diversity at sites in the vicinity of the receptor binding site, HI or FRA assays show no evidence of substantial antigenic change (Fig.\ \ref{H3N2_tree_titer_model}).
HI assays by the WHO collaborating center at the Crick Institute in London indicate a moderate two-fold titer drop relative to A/HongKong/4801/2014 for several subclades, including the 171K subclade.
However, there is substantial variability in those data.
More dense and recent data by the CDC in Atlanta, Georgia, do not support a homogeneous titer drop of the 171K clade relative to the current vaccine strain Hong Kong/4801/2014.
The figure below shows trees colored by predicted titer drop relative to Hong Kong/4801/2014 using the tree model \cite{neher2015prediction} trained on HI data from Crick and CDC as well as FRA data from CDC.
Taken together, these data provide little evidence for substantial antigenic evolution within 3c2.a.

\begin{figure}[h!]
	\centering
	\includegraphics[width=1.0\textwidth]{../figures/feb-2017/H3N2_tree_titer_model.png}
	\caption{\textbf{H3N2 phylogeny colored by antigenicity.}
	The left panel shows individual HI measurements of viruses against serum from A/HongKong/5738/2014.
	Cooler color indicates great antigenic similarity (less titer drop going from homologous to heterologous titers).
	The right panel shows model estimates of antigenic divergence from the root of the phylogeny.
	These estimates combine HI measurements from many different sera.
	Crick data is current through Sep 2016 and CDC data is current through Feb 2017.
	}
	\label{H3N2_tree_titer_model}
\end{figure}

The local branching index (LBI), a phylogenetic indicator of clade growth \cite{neher2014predicting}, corroborates the observations made on the basis of clade frequencies (Fig.\ \ref{H3N2_lbi}).
Over the last 6 monthes, the 131K clade has highest LBI, while the clade 171K has highest LBI when considering sequences in all of 2016.
Interpretation of the sequence based indices such as number of epitope or non-epitope mutations is difficult due the extensive genetic heterogeneity.
The clade 131K has accumulated the same number of epitope mutations as 171K but has fewer non-epitope mutations relative to A/Texas/50/2012.

\begin{figure}[H]
	\centering
	\includegraphics[width=0.9\textwidth]{../figures/feb-2017/H3N2_lbi.png}
	\caption{\textbf{H3N2 phylogeny colored by local branching index.}
	}
	\label{H3N2_lbi}
\end{figure}

\clearpage
\pagebreak

%%% A/H1N1pdm %%%
\section*{A/H1N1pdm}
\addcontentsline{toc}{section}{A/H1N1pdm}

\textbf{Very few H1N1pdm viruses have been observed in recent monthes.
The dominant clade continues to be 6b.1 and there is little amino acid sequence variation within HA.
The only notable subclade that has been growing recently is the clade bearing HA1:R205K/S183P.
This clade is dominated by North American viruses and we see no evidence that this clade has a particular competitive advantage.}

As above, we base our primary analysis on a set of viruses collected between Jan 2015 and Feb 2017, comprising approximately 100 viruses per month where available and seeking to equilibrate sample counts geographically where possible (Fig.\ \ref{H1N1pdm_counts}). However, in the case of H1N1pdm we were not able to collect a full 100 samples per month from the database in many months due to low absolute prevalence.

\begin{figure}[H]
	\centering
	\includegraphics[width=0.95\textwidth]{../figures/feb-2017/H1N1pdm_counts.png}
	\caption{\textbf{Sample counts through time and across regions.}
	This is a stacked bar plot, so that in good months there are $\sim$100 total samples and $\sim$15 samples each from North America and from Europe.
	}
	\label{H1N1pdm_counts}
\end{figure}

\pagebreak

There has been continued dominance of clade 6b.1 (characterized by mutation 162N) over clade 6b.2 (characterized by mutation 152T) (Fig.\ \ref{H1N1pdm_clades}).

\begin{figure}[H]
	\centering
	\includegraphics[width=0.95\textwidth]{../figures/feb-2017/H1N1pdm_clades.png}
	\caption{\textbf{Frequency trajectories of H1N1pdm clades.}
	We estimate frequencies of different clades based on sample counts and collection dates.
	We use a Brownian motion process prior to smooth frequencies from month-to-month.
	Transparent bands show an estimate the 95\% confidence interval based on sample counts.
	The final point represents our frequency estimate for Feb 1 2017.
	}
	\label{H1N1pdm_clades}
\end{figure}

\pagebreak

Within clade 6b.1 there is very little amino acid sequence variation (Fig.\ \ref{H1N1pdm_tree}). The only mutations of any size are I166V, S183P and R205K. Interestingly, these three amino acid mutations occur in rapid succession along a single lineage, resulting in a series of nested clades.

\begin{figure}[H]
	\centering
	\includegraphics[width=0.9\textwidth]{../figures/feb-2017/H1N1pdm_tree.png}
	\caption{\textbf{H1N1pdm / 6b.1 phylogeny colored by genotype.}
	}
	\label{H1N1pdm_tree}
\end{figure}

\pagebreak

Mutations 183P and 205K has been rising in Europe and North America starting in early-mid 2016, but this rise corresponds to a small number of cases and might turn out insignificant during the next H1N1pdm dominated season (Fig.\ \ref{H1N1pdm_mutations}). The pace of increase is only moderate, suggesting a lack of strong selective pressure. Additionally, no consistent antigenic variation is observed among recently circulating H1N1pdm viruses (according to primary infection in naive animals).

\begin{figure}[H]
	\centering
	\includegraphics[width=0.95\textwidth]{../figures/feb-2017/H1N1pdm_mutations.png}
	\caption{\textbf{Frequency trajectories of 6b.1 subclades.}
	We estimate frequencies of different clades based on sample counts and collection dates.
	We use a Brownian motion process prior to smooth frequencies from month-to-month.
	Transparent bands show an estimate the 95\% confidence interval based on sample counts.
	The final point represents our frequency estimate for Feb 1 2017.
	}
	\label{H1N1pdm_mutations}
\end{figure}

\clearpage
\pagebreak

%%% B/Victoria %%%
\section*{B/Vic}
\addcontentsline{toc}{section}{B/Vic}

\textbf{Clade 1A has continued to dominate and mutation 117V has all but taken over the global population.
The rise of this mutation was fairly gradual and we have no evidence that it is associated with antigenic change or other benefit to the virus.}

As above, we base our primary analysis on a set of viruses collected between Jan 2015 and Feb 2017, comprising approximately 100 viruses per month where available and seeking to equilibrate sample counts geographically where possible (Fig.\ \ref{Vic_counts}).

\begin{figure}[H]
	\centering
	\includegraphics[width=0.95\textwidth]{../figures/feb-2017/Vic_counts.png}
	\caption{\textbf{Sample counts through time and across regions.}
	This is a stacked bar plot, so that in good months there are $\sim$100 total samples and $\sim$15 samples each from North America and from Europe.
	}
	\label{Vic_counts}
\end{figure}

\pagebreak

We observe the continued dominance of the clade bearing mutations N129D, V146I and I117V (Fig.\ \ref{Vic_mutations}).
There is very little HA amino acid variation within the 117V viruses with no subclades rising to appreciable frequency.

\begin{figure}[H]
	\centering
	\includegraphics[width=0.95\textwidth]{../figures/feb-2017/Vic_mutations.png}
	\caption{\textbf{Frequency trajectories of B/Vic variants.}
	We estimate frequencies of different clades based on sample counts and collection dates.
	We use a Brownian motion process prior to smooth frequencies from month-to-month.
	Transparent bands show an estimate the 95\% confidence interval based on sample counts.
	The final point represents our frequency estimate for Feb 1 2017.
	}
	\label{Vic_mutations}
\end{figure}

\pagebreak

Neither the local branching index nor clade frequencies single out a particular variant within the 1A/I117V clade (Fig.\ \ref{Vic_lbi}).

\begin{figure}[H]
	\centering
	\includegraphics[width=0.9\textwidth]{../figures/feb-2017/Vic_lbi.png}
	\caption{\textbf{B/Vic phylogeny colored by local branching index.}
	}
	\label{Vic_lbi}
\end{figure}

\clearpage
\pagebreak

%%% B/Yamagata %%%
\section*{B/Yam}
\addcontentsline{toc}{section}{B/Yam}

\textbf{Clade 3 has continued to dominate. Within clade 3, a clade with mutation HA1:251V is globally at frequency of about 80\% throughout 2016. Within this clade, mutation 211R is at 25\% frequency. In addition, a clade without prominent amino acid mutations has been rising throughout 2016.}

As above, we base our primary analysis on a set of viruses collected between Jan 2015 and Feb 2017, comprising approximately 100 viruses per month where available and seeking to equilibrate sample counts geographically where possible (Fig.\ \ref{Yam_counts}).

\begin{figure}[H]
	\centering
	\includegraphics[width=0.95\textwidth]{../figures/feb-2017/Yam_counts.png}
	\caption{\textbf{Sample counts through time and across regions.}
	This is a stacked bar plot, so that in good months there are $\sim$100 total samples and $\sim$15 samples each from North America and from Europe.
	}
	\label{Yam_counts}
\end{figure}

\pagebreak

Clade 3 viruses have continued to dominate with clade 2 viruses scarcely observed throughout 2016.
Within clade 3, mutations L172Q and M251V remain dominant and mutation K211R continues to persist at low frequency (Fig.\ \ref{Yam_mutations}, \ref{Yam_tree}). There is little evidence of emerging clades of strong selective effect.

\begin{figure}[H]
	\centering
	\includegraphics[width=0.95\textwidth]{../figures/feb-2017/Yam_mutations.png}
	\caption{\textbf{Frequency trajectories of B/Yam variants.}
	We estimate frequencies of different clades based on sample counts and collection dates.
	We use a Brownian motion process prior to smooth frequencies from month-to-month.
	Transparent bands show an estimate the 95\% confidence interval based on sample counts.
	The final point represents our frequency estimate for Feb 1 2017.
	}
	\label{Yam_mutations}
\end{figure}

\begin{figure}[H]
	\centering
	\includegraphics[width=0.8\textwidth]{../figures/feb-2017/Yam_tree.png}
	\caption{\textbf{B/Yam phylogeny colored by genotype.}
	}
	\label{Yam_tree}
\end{figure}

\pagebreak

The LBI highlights the central clade in the tree below (Fig.\ \ref{Yam_lbi}). This clade has no amino acid mutations in HA along its backbone relative to its parent clade 251V.

\begin{figure}[H]
	\centering
	\includegraphics[width=0.9\textwidth]{../figures/feb-2017/Yam_lbi.png}
	\caption{\textbf{B/Vic phylogeny colored by local branching index.}
	}
	\label{Yam_lbi}
\end{figure}

\clearpage
\pagebreak

%%% References %%%
\bibliographystyle{plos}
\bibliography{references}

\end{document}

### 2017-02-17 auspice/reports/tex/feb-2017.tex
revert some of the frequency figures, fix mismatches in capitalization, add L&L

@@
 We base our primary analysis on a set of viruses collected between Jan 2015 and Feb 2017 (Fig.\ \ref{H3N2_counts}).
-We analyze mutation frequency trajectories separately in different geographic regions using all available sequence data.
+We analyze mutation frequency trajectories in different geographic regions using all available sequence data for each region.
 Our phylogenetic analysis is based on a subsample of approximately 100 viruses per month where available and seeking to equilibrate sample counts geographically where possible.
@@
 	\centering
-	\includegraphics[width=0.9\textwidth]{../figures/feb-2017/H3N2_tree.png}
+	\includegraphics[width=0.9\textwidth]{../figures/feb-2017/h3n2_tree.png}
 	\caption{\textbf{H3N2 / 3c2.a phylogeny colored by genotype.}
@@
 	\centering
-	\includegraphics[width=0.95\textwidth]{../figures/feb-2017/H3N2_mutations.png}
+	\includegraphics[width=0.95\textwidth]{../figures/feb-2017/h3n2_frequencies.pdf}
 	\caption{\textbf{Frequency trajectories of 3c2.a subclades.}
@@
 Over the last 6 monthes, the 131K clade has highest LBI, while the clade 171K has highest LBI when considering sequences in all of 2016.
-Interpretation of the sequence based indices such as number of epitope or non-epitope mutations is difficult due the extensive genetic heterogeneity.
+Interpretation of the sequence based indices such as number of epitope or non-epitope mutations \cite{luksza2014predictive} is difficult due the extensive genetic heterogeneity.
 The clade 131K has accumulated the same number of epitope mutations as 171K but has fewer non-epitope mutations relative to A/Texas/50/2012.
@@
 	\centering
-	\includegraphics[width=0.9\textwidth]{../figures/feb-2017/H3N2_lbi.png}
+	\includegraphics[width=0.9\textwidth]{../figures/feb-2017/h3n2_lbi.png}
 	\caption{\textbf{H3N2 phylogeny colored by local branching index.}
@@
 	\centering
-	\includegraphics[width=0.9\textwidth]{../figures/feb-2017/H1N1pdm_tree.png}
+	\includegraphics[width=0.9\textwidth]{../figures/feb-2017/h1n1pdm_tree.png}
 	\caption{\textbf{H1N1pdm / 6b.1 phylogeny colored by genotype.}
@@
 	\centering
-	\includegraphics[width=0.95\textwidth]{../figures/feb-2017/H1N1pdm_mutations.png}
+	\includegraphics[width=0.95\textwidth]{../figures/feb-2017/h1n1pdm_frequencies.pdf}
 	\caption{\textbf{Frequency trajectories of 6b.1 subclades.}
@@
 	\centering
-	\includegraphics[width=0.95\textwidth]{../figures/feb-2017/Vic_mutations.png}
+	\includegraphics[width=0.95\textwidth]{../figures/feb-2017/vic_frequencies.pdf}
 	\caption{\textbf{Frequency trajectories of B/Vic variants.}
@@
 	\centering
-	\includegraphics[width=0.95\textwidth]{../figures/feb-2017/Yam_mutations.png}
+	\includegraphics[width=0.95\textwidth]{../figures/feb-2017/yam_frequencies.pdf}
 	\caption{\textbf{Frequency trajectories of B/Yam variants.}
@@
 	\centering
-	\includegraphics[width=0.8\textwidth]{../figures/feb-2017/Yam_tree.png}
+	\includegraphics[width=0.8\textwidth]{../figures/feb-2017/yam_tree.png}
 	\caption{\textbf{B/Yam phylogeny colored by genotype.}
@@
 	\centering
-	\includegraphics[width=0.9\textwidth]{../figures/feb-2017/Yam_lbi.png}
+	\includegraphics[width=0.9\textwidth]{../figures/feb-2017/yam_lbi.png}
 	\caption{\textbf{B/Vic phylogeny colored by local branching index.}

### 2017-02-23 auspice/_includes/3y_meta.js
added index.htmls for vic and yam

@@ -1,4 +1,4 @@
-var time_window = 1.0;
+var time_window = 2.5;
 var full_data_time_window = 3.0;
 var time_ticks=[2013.0, 2013.5, 2014.0, 2014.5, 2015.0, 2015.5, 2016.0, 2016.5];
 var dfreq_dn = 2;

### 2017-02-24 auspice/reports/tex/feb-2017.tex
Update report text

@@
 \usepackage{geometry}
-\geometry{textwidth=16.75cm} % 15.25cm for single-space, 16.25cm for double-space
+\geometry{textwidth=17cm} % 15.25cm for single-space, 16.25cm for double-space
 \geometry{textheight=22.5cm} % 22cm for single-space, 22.5cm for double-space
@@
 
-\date{February 16, 2017}
+\date{February 23, 2017}
 
@@
 3c3.b viruses have scarcly been seen since the end of 2015.
-While the uptick of 3c3.a viruses in mid-2016 in the US has failed to carry forward to the 2016-2017 US winter season.
+The uptick of 3c3.a viruses in mid-2016 in the US has failed to carry forward to the 2016-2017 US winter season.
 Oceania still harbors 3c3.a viruses, but we fully expect these to be replaced by another clade in the coming year as Oceania is reseeded from the global reservoir.
@@
 
-Within clade 3c2.a, a number of subclades with signature mutations at N171K, T131K/R142K, R142K, N121K/R144K, Q197K have emerged (Fig.\ \ref{H3N2_tree}).
+Within clade 3c2.a, a number of subclades have emerged (Fig.\ \ref{H3N2_tree}).
+Notable clades include N171K (clade 3c2a.1), N171K/N121K, N121K/S144K, 131K/142K and R142K/Q197K.
 Clades with R142K, N121K/R144K, Q197K arose in late 2014 or early 2015 but have not come to dominate.
-However, the 171K clade that was first observed in mid-2015 has continued to increase in frequency at a moderate rate and is now at a global frequency of approximately 60\%.
+However, the 171K clade (3c2a.1) that was first observed in mid-2015 has continued to increase in frequency at a moderate rate and is now at a global frequency of approximately 60\%.
 Within the 171K clade, a subclade bearing the 121K mutation emerged in at the beginning of 2016 and has steadily increased in frequency now making of the majority of the parent 171K clade.
@@
 Interestingly, we observe a large number of mutations to positively charged residues and particularly many parallel mutations to lysine at positions 121, 131, 142, 144, 171, and 197.
+Clades N171K/N121K and T131K/R142K are on opposing genetic backgrounds so that there is competition between these clades; both cannot succeed.
 
@@
 	\centering
-	\includegraphics[width=0.95\textwidth]{../figures/feb-2017/h3n2_frequencies.pdf}
+	\includegraphics[width=0.75\textwidth]{../figures/feb-2017/h3n2_frequencies.pdf}
 	\caption{\textbf{Frequency trajectories of 3c2.a subclades.}
@@
 The clade 131K has accumulated the same number of epitope mutations as 171K but has fewer non-epitope mutations relative to A/Texas/50/2012.
+
+ \textit{Given current patterns of clade growth and decline, we predict clades 171K/121K and 131K/142K to be the most successful of currently circulating clades.
+We expect both to increase in the coming months.
+Longer term projections are difficult.
+171K/121K may continue to dominate, 131K/142K may displace 171K/121K or another mutation may appear that determines the eventual outcome.}
 
@@
 
-\pagebreak
-
 Mutations 183P and 205K has been rising in Europe and North America starting in early-mid 2016, but this rise corresponds to a small number of cases and might turn out insignificant during the next H1N1pdm dominated season (Fig.\ \ref{H1N1pdm_mutations}).
@@
 
+\textit{Given the lack of competition, we predict the continued dominance of 6b.1 viruses.}
+
+\pagebreak
+
 \begin{figure}[H]
 	\centering
-	\includegraphics[width=0.95\textwidth]{../figures/feb-2017/h1n1pdm_frequencies.pdf}
+	\includegraphics[width=0.9\textwidth]{../figures/feb-2017/h1n1pdm_frequencies.pdf}
 	\caption{\textbf{Frequency trajectories of 6b.1 subclades.}
@@
 There is very little HA amino acid variation within the 117V viruses with no subclades rising to appreciable frequency.
+There is deletion at AA site 162 that first appeared in mid-2016, but remains at low global frequency.
 
@@
 Neither the local branching index nor clade frequencies single out a particular variant within the 1A/I117V clade (Fig.\ \ref{Vic_lbi}).
+
+\textit{Given the lack of competition, we predict the continued dominance of 117V viruses.}
 
@@
 
+\textit{Genetic variation within 172Q/251V viruses is beginning to develop, but given lack of credible competition, we expect 172Q/251V to continual to dominate in the coming months.}
+
 \begin{figure}[H]

### 2017-02-24 auspice/reports/tex/feb-2017.tex
Fix figure legend.

@@
 	\caption{\textbf{H3N2 phylogeny colored by antigenicity.}
-	The left panel shows individual HI measurements of viruses against serum from A/HongKong/5738/2014.
-	Cooler color indicates great antigenic similarity (less titer drop going from homologous to heterologous titers).
-	The right panel shows model estimates of antigenic divergence from the root of the phylogeny.
+	Each panel shows model estimates of antigenic divergence relative to serum from A/HongKong/5738/2014.
+	Cooler color indicates greater antigenic similarity (less titer drop going from homologous to heterologous titers).
 	These estimates combine HI measurements from many different sera.
+	The left panel uses HI data from the Crick Institute in London, the center panel uses HI data from the CDC in Atlanta and the right panel uses FRA data from the CDC.
 	Crick data is current through Sep 2016 and CDC data is current through Feb 2017.

### 2017-09-15 auspice/reports/plot_clades.py
Update 171K to 3c2.a1. Fix H3N2 clades plot.

@@ -19,7 +19,7 @@ drop = 1
 resolution = '2y'
 
 if virus=='h3n2':
-    clades = ['3c2.a', '3c3.a', '171K']
+    clades = ['3c2.a', '3c3.a', '3c2.a1']
     # mutations = ['HA1:121K', 'HA1:92R', 'HA1:131K', 'HA1:31S','HA1:198P', 'HA1:193S']
     # Large subclades
     # 1. 31S, 53N, 144R, 171K, 192T, 197H, characteristic: 197H

### 2017-09-15 auspice/reports/tex/sep-2017.tex
tex version of the report

\documentclass[11pt,oneside,letterpaper]{article}

% graphicx package, useful for including eps and pdf graphics
\usepackage{graphicx}
\DeclareGraphicsExtensions{.png,.png,.jpg}

% basic packages
\usepackage{color}
\usepackage{parskip}
\usepackage{float}
\usepackage{microtype}
\usepackage{url}
\usepackage{hyperref}

% text layout
\usepackage{geometry}
\geometry{textwidth=17cm} % 15.25cm for single-space, 16.25cm for double-space
\geometry{textheight=22.5cm} % 22cm for single-space, 22.5cm for double-space

% helps to keep figures from being orphaned on a page by themselves
\renewcommand{\topfraction}{0.85}
\renewcommand{\textfraction}{0.1}

% bold the 'Figure #' in the caption and separate it with a period
% Captions will be left justified
\usepackage[labelfont=bf,labelsep=period,font=small]{caption}

% review layout with double-spacing
%\usepackage{setspace}
%\doublespacing
%\captionsetup{labelfont=bf,labelsep=period,font=doublespacing}

% cite package, to clean up citations in the main text. Do not remove.
\usepackage{cite}
%\renewcommand\citeleft{(}
%\renewcommand\citeright{)}
%\renewcommand\citeform[1]{\textsl{#1}}

% Remove brackets from numbering in list of References
\renewcommand\refname{\large References}
\makeatletter
\renewcommand{\@biblabel}[1]{\quad#1.}
\newcommand{\FIG}[1]{Fig.~\ref{#1}}
\makeatother

\usepackage{authblk}
\renewcommand\Authands{ \& }
\renewcommand\Authfont{\normalsize \bf}
\renewcommand\Affilfont{\small \normalfont}
\makeatletter
\renewcommand\AB@affilsepx{, \protect\Affilfont}
\makeatother

%%% TOC
\usepackage{patchcmd}
\makeatletter
\patchcommand\@starttoc{\begin{quote}}{\end{quote}}
\usepackage{tocloft}
\renewcommand{\cftsecleader}{\cftdotfill{\cftdotsep}}

%%% TITLE %%%
\title{\vspace{2cm} \LARGE \bf
Seasonal influenza circulation patterns and projections for 2017-2018
}

\author[1]{Trevor Bedford}
\author[2]{Richard A.\ Neher}

\affil[1]{Vaccine and Infectious Disease Division, Fred Hutchinson Cancer Research Center, Seattle, WA, USA}
\affil[2]{Biozentrum, University of Basel, Basel, Switzerland}

\date{September 15, 2017}
\begin{document}

\maketitle

\emph{This is not meant as a comprehensive report, but is instead
intended as particular observations that we've made that may be of
relevance. Please also note that observed patterns reflect the GISAID
database and may not be entirely representative of underlying dynamics.}

\tableofcontents

\pagebreak
\section*{A/H3N2}
\addcontentsline{toc}{section}{A/H3N2}

\textbf{H3N2 continues to diversify with many coexisting clades, all of
which carry several amino acid mutations at previously characterized
epitopes. The majority of viruses fall into the 3c2.a clade which has
been dominating globally for \textgreater{}3 years, but 3c3.a viruses
continue to persist. The common ancestor of circulating H3N2 viruses is
now more than 5 years old, which is rare for H3N2. Despite extensive
genetic diversity, serological assays suggest limited, but non-zero,
antigenic evolution. We expect multiple competing clades within 3c2.a to
persist into the future with no clear immediate winner.}

We base our primary analysis on a set of viruses collected between Oct
2015 and Aug 2017, comprising \textgreater{}800 viruses per month in Dec
2016 to Mar 2017 and 100-200 viruses per month more recently, see \FIG{H3N2_counts}. We use all
available data when estimating frequencies of mutations and weight
samples appropriately by regional population size and relative sampling
intensity to arrive at a putatively unbiased global frequency estimate.
Phylogenetic analyses are based on a representative sample of about 2000
viruses.

\begin{figure}[H]
  \centering
  \includegraphics[width=0.9\textwidth]{../figures/sep-2017/h3n2_counts.png}
  \caption{\textbf{Sample counts through time and across regions.}
  This is a stacked bar plot, so that in all months there are at least 100 samples, in some month $>500$ sequences.
  }
  \label{H3N2_counts}
\end{figure}

\clearpage
\begin{figure}[H]
  \centering
  \includegraphics[width=0.9\textwidth]{../figures/sep-2017/h3n2_clades.png}
  \caption{\textbf{Dynamics of major H3N2 clades.} Clade 3c3.a continues to circulate at low frequency, while 3c3.b has not been observed for several month (not shown).}
  \label{H3N2_clades}
\end{figure}
At the level of major clades, 3c2.a has continued to dominate while
3c3.a persisted at about 20\% frequency in Oceania and North America in
the 2016-2017 season, see \FIG{H3N2_clades}. Within 3c2.a, the clade 3c2.a1 comprising HA1
mutation N171K and HA2 mutations I77V and G155E had risen rapidly in
early 2016 but later stabilized at about 50\% with somewhat higher
frequencies in Europe.

\clearpage
\begin{figure}[H]
  \centering
  \includegraphics[width=0.8\textwidth]{../figures/sep-2017/h3n2_tree.png}
  \caption{\textbf{H3N2 / 3c2.a phylogeny colored by genotype.} Colors are chosen to highlight the clades 1--5. Clades are in reverse order, that is clade 5 is on top, clade 1 is at the bottom.
  }
  \label{H3N2_tree}
\end{figure}

We observe substantial diversification within the 3c2.a clade over the
past year. There are 5 major subclades within 3c2.a that have reached
appreciable frequency. These clades are characterized by the following
mutations and highlighed in the phylogenetic tree in \FIG{H3N2_tree}:

\begin{enumerate}
\item
  53N, 144R, 171K, 192T, 197H
\item
  121K, 144K
\item
  131K, 142K, 261Q
\item
  135K, HA2:150E
\item
  92R, 311Q
\end{enumerate}

All of these clades possess mutations at previously characterized
epitope sites. We will refer to these clades as clade 1--5.

All of these clades have risen in frequency and are now competing with
one another, see \FIG{H3N2_mutations}. Clade 1 (53N, 144R, 171K, 192T, 197H) is highly derived in
terms of sequence, but is at low frequency globally (currently less than
1\%) despite making up the a substantial fraction of 2017 Oceania
viruses. Clade 2 (121K, 144K) is beginning to decline in frequency and
is now at \textasciitilde{}21\% global frequency. Clade 3 (131K, 142K,
261Q) and clade 4 (135K, HA2:150E) have each slowly climbed to
approximately 20\% frequency. However, clade 5 (92R, 311Q) has recently
increased to 39\% in global frequency.

\begin{figure}[h!]
  \centering
  \includegraphics[width=0.95\textwidth]{../figures/sep-2017/h3n2_mutations.png}
  \caption{\textbf{Frequency trajectories of 3c2.a subclades.}
  We estimate frequencies of different clades based on sample counts and collection dates.
  We use a Brownian motion process prior to smooth frequencies from month-to-month.
  Transparent bands show an estimate the 95\% confidence interval based on sample counts.
  }
  \label{H3N2_mutations}
\end{figure}

\clearpage
\begin{figure}[h!]
  \centering
  \includegraphics[width=1.0\textwidth]{../figures/sep-2017/h3n2_tree_titer_model.png}
  \caption{\textbf{H3N2 phylogeny colored by antigenic advance.}
  Each panel shows model estimates of antigenic divergence relative to the root of the tree, which is similar to the titer drop relative to A/HongKong/5738/2014.
  Cooler color indicates greater antigenic similarity (less titer drop going from homologous to heterologous titers).
  }
  \label{H3N2_tree_titer_model}
\end{figure}

All 5 of these clades appear somewhat drifted according to antigenic
analysis Neher et al \cite{neher2015prediction} of HI measurements provided by the Influenza Division at the US CDC, see \FIG{H3N2_tree_titer_model}.
With antigenic differences estimated between 0.5 units and 2.0 units.

The mutation T135K has come up in clade 2, 4, and 5 and is now globally
at 40\% frequency. The most rapidly growing clade carries mutations
N121K/K92R/H311Q/E62G/R142G/T135K. This is a subclade of clade 5 above
with addition of 62G, 142G and 135K. The model fit to the antigenic data
estimates that this subclade of clade 5 has drifted by 2.2 antigenic
units, 1 of which is associated with the E62G/R142G mutations and 1.2
with the T135K mutation (on this background), see \FIG{H3N2_135K}.
Note, however, that there
is no evidence of antigenic evolution in the FRA data (one data point
only). A small sister clade with mutation T135N (also part of clade 5)
recently developed. Taken together, changes at position 135 likely have
selective origin.

\begin{figure}[h!]
  \centering
  \includegraphics[width=0.8\textwidth]{../figures/sep-2017/h3n2_tree_135k.png}
  \caption{\textbf{Zoom into clade 5, colored by antigenic advance.} A subclade with a T135K mutation has recently risen in frequency and has drifted $>$2 antigenic units.
  }
  \label{H3N2_135K}
\end{figure}

Clades 1,2, and 3 are part of 3c2.a but outside of the 3c2.a1 subclade.
Clade 3 with T131K/R142K mutations continues to be common in China (but
not in South Asia) with an embedded cluster of South and North American
isolates from last summer. Subclusters within this clade have been
expanding recently, but there is little evidence for antigenic evolution
in the HI or FRA data. The 131K substitution is not seen outside this
clade.

Clade 1 with substitutions N31S/D53N/S144R/N171K/I192T/Q197H was
observed globally over the past 9 month but is now decreasing everywhere
except Oceania HI data and the model fit suggest a moderate (4-fold,
about 2 antigenic units) titer drop relative to A/HongKong/4801/2014.
Despite the titer drop, this calde seems to be outcompeted almost
everywhere.

The most notable subclade of 3c3.a is a clade carrying the mutation
F193S. However, this clade along with its sister clades has been
decreasing in recent month and it doesn't show a consistent signal of
antigenic change. The F193D mutations is observed sporadically within
the 3c2.a clade as well.

% \begin{figure}[h!]
%   \centering
%   \includegraphics[width=1.0\textwidth]{../figures/sep-2017/h3n2_3c2a.png}
%   \caption{\textbf{Recent frequency changes.} The tree, zoomed into clade 3c2.a, is colored by change in clade frequency over the past 4 month.
%   }
%   \label{H3N2_135K}
% \end{figure}

In summary, we think that the clade most likely to increase further in
frequency is the E62G/R142G/T135K clade. However, two other clades
(T131K/R142K and 144K) have similar potential. The former has very
rapidly growing subclade, the latter rose rapidly last year and has a
slightly higher LBI than the other two.

\clearpage
\section*{A/H1N1pdm}
\addcontentsline{toc}{section}{A/H1N1pdm}

\textbf{A clade comprising mutations S74R and I295V has recently risen
to \textgreater{}90\% global frequency. Although it shows no antigenic
distinction by ferret HI data, the rapidity of its rise suggests a
selective origin.}

We base our primary analysis on a set of viruses collected between Oct
2015 and Aug 2017, comprising approximately 100 viruses per month during
relevant months of 2017, see \FIG{H1N1pdm_counts}. We use all available data when estimating
frequencies of mutations and weight samples appropriately by regional
population size and relative sampling intensity to arrive at a
putatively unbiased global frequency estimate. Phylogenetic analyses are
based on a representative sample of about 2000 viruses.

\begin{figure}[H]
  \centering
  \includegraphics[width=0.95\textwidth]{../figures/sep-2017/h1n1pdm_counts.png}
  \caption{\textbf{A/H1N1pdm sample counts through time and across regions.}}
  \label{H1N1pdm_counts}
\end{figure}

H1N1pdm recently underwent a selective sweep of the 6b.1 clade
comprising mutations S162N and I216T. Nearly all currently circulating
H1N1pdm viruses are clade 6b.1. The H1N1pdm vaccine strain was updated
in Sep 2016 to match 6b.1 viruses.

There has been little diversification within clade 6b.1 with a couple
notable exceptions. There is a clade comprising mutations S74R and I295V
that has been very successful. Within this clade, there is a more recent
subclade with the additional mutation S164T.

\begin{figure}[H]
  \centering
  \includegraphics[width=0.95\textwidth]{../figures/sep-2017/h1n1pdm_mutations.png}
  \caption{\textbf{Frequency trajectories of recent mutations in A/H1N1pdm viruses.}  }
  \label{H1N1pdm_clades}
\end{figure}

The S74R and I295V clade has risen rapidly in frequency starting around
Jan 2017. This clade has been very successful and is now at
approximately 92\% global frequency. Within this clade the subclade
comprising S164T has risen in frequency from April 2017 and is now at
approximately 35\% global frequency, see \FIG{H1N1pdm_clades}

Other mutations have appeared within 6b.1 viruses including S183P, R205K
and I324V, but none of these have been very successful and are now
declining in response to the rise of S74R and I295V.

\clearpage
\begin{figure}[H]
  \centering
  \includegraphics[width=0.9\textwidth]{../figures/sep-2017/h1n1pdm_tree_titer_model.png}
  \caption{\textbf{Antigenic distance from A/Michigan/45/2015.} No antigenic variation is picked up by the model.
  }
  \label{H1N1pdm_mutations}
\end{figure}
Antigenic analysis using HI measurements provided by the Influenza
Division at the US CDC fails to show any distinction within recently
circulating H1N1pdm viruses. Like the clades 6b and 6b.1 clade before
it, the new S74R and I295V clade does not show any titer drop despite
there being ample measurements.

\begin{figure}[H]
  \centering
  \includegraphics[width=0.9\textwidth]{../figures/sep-2017/h1n1pdm_mutations_6b_6b1.png}
  \caption{\textbf{Sequential rapid replacement of clades in A/H1N1pdm.} The latest rapid rise of 74R took place at similar or greater speed as the spread of 162N.
  }
  \label{H1N1pdm_mutations}
\end{figure}
It was previously observed that clade 6b viruses with mutation K163Q
showed decreased titers in a subset of adult human sera due to
immunodominance effects, despite continued high titers in ferret sera
(\href{http://www.pnas.org/content/111/44/15798}{Linderman et al 2014}).
Importantly, the new S74R and I295V clade has spread at a similarly
rapid rate as did clades 6b and 6b.1. This rate of spread and
displacement of existing viral diversity is consistent with a selective
origin.

\emph{Given current patterns of growth and decline it is almost certain
that the S74R and I295V clade will dominant in a years time. Although
there is no antigenic effect visible via HI from ferret antisera, the
rapidity of the rise of S74R and I295V suggests a selective origin. We
would suggest a detailed examination of human serological data as it's
already been established that human serology may differ from ferret
serology in H1N1pdm viruses.}

\clearpage
%%% B/Victoria %%%
\section*{B/Vic}
\addcontentsline{toc}{section}{B/Vic}

\textbf{A clade with a two amino acid deletion 162-/163- has altered
serological properties and is increasing in frequency, albeit slowly.
Two other clades (carrying mutations K209N and V87A/I175V) have
increased in frequency moderately.}

We base our primary analysis on a set of viruses collected between Oct
2015 and Aug 2017, comprising between 50 and 250 viruses per month
during relevant months of 2017, see \FIG{Vic_counts}. We use all available data when
estimating frequencies of mutations and weight samples appropriately by
regional population size and relative sampling intensity to arrive at a
putatively unbiased global frequency estimate. Phylogenetic analyses are
based on a representative sample of about 2000 viruses.

\begin{figure}[H]
  \centering
  \includegraphics[width=0.95\textwidth]{../figures/sep-2017/vic_counts.png}
  \caption{\textbf{Sample counts through time and across regions.}
  This is a stacked bar plot, so that in good months there are $\sim$200 total samples.
  }
  \label{Vic_counts}
\end{figure}
\clearpage
\begin{figure}[H]
  \centering
  \includegraphics[width=0.95\textwidth]{../figures/sep-2017/vic_tree.png}
  \caption{\textbf{B/VIC -- Phylogenetic tree colored by genotype as positions 162, 175, 209.}
  }
  \label{Vic_tree}
\end{figure}
The mutation I117V has continued its slow ascent in frequency as is now
\textgreater{}99\% in the global population. Within this clade there is
relatively little circulating genetic diversity. The only clades of note
are a small clade with mutations I180V, D129G and deletions K162- and
K163-, another small clade with mutations V87A and I175V and a third
small clade with mutation K209N.

\clearpage
\begin{figure}[H]
  \centering
  \includegraphics[width=0.95\textwidth]{../figures/sep-2017/vic_mutations.png}
  \caption{\textbf{Frequency trajectories of B/Vic variants.}
  We estimate frequencies of different clades based on sample counts and collection dates.
  We use a Brownian motion process prior to smooth frequencies from month-to-month.
  Transparent bands show an estimate the 95\% confidence interval based on sample counts.
  }
  \label{Vic_mutations}
\end{figure}
The clade bearing 129G, 162--, 163-- and 180V has remained at low
frequency globally with current global estimates at
\textasciitilde{}4\%, see \FIG{Vic_mutations}. The clade bearing 87A and 175V has increased in
frequency substantially in North America but remains low in the rest of
the world. The clade bearing 209N has been more successful, approaching
15\% globally and circulating in Asia.

\clearpage
\begin{figure}[H]
  \centering
  \includegraphics[width=0.95\textwidth]{../figures/sep-2017/vic_tree_titer_model.png}
  \caption{\textbf{Antigenic evolution of the B/Vic 162-/163- clade.}
  The tree model estimates a drop in HI titers by 3 units.
  }
  \label{Vic_titers}
\end{figure}

The clade bearing 129G, 162--, 163-- and 180V appears clearly drifted
according to HI measurements provided by the Influenza Division at the
US CDC, see \FIG{Vic_titers}. We estimate an approximately 8-fold titer drop (3 antigenic
units) of drift from vaccine strain B/Brisbane/60/2008. Separately, a
small number of recent viruses sampled in Hong Kong have a 9bp deletion
overlapping codons 162, 163, 164, and 165 along with a mutation I180T.
While based on limited data, the tree model estimates an 1.2 unit titer
drop.

\emph{Although the clade bearing 129G, 162--, 163-- and 180V is almost
certainly antigenically drifted, it's slow spread up to the this point
suggests that it is unlikely to predominate in a year's time. We would
advise careful monitoring as further mutations may enhance spread or
similar mutations might emerge on fitter backgrounds.}

\clearpage
\section*{B/Yam}
\addcontentsline{toc}{section}{B/Yam}

\textbf{A clade comprising M251V with clade 3 viruses continues to
dominate. The is little genetic differentiation within this clade and no
evidence of antigenic evolution.}

We base our primary analysis on a set of viruses collected between Oct
2015 and Aug 2017, comprising between 50 and 250 viruses per month
during relevant months of 2017, see \FIG{Yam_counts}. We use all available data when
estimating frequencies of mutations and weight samples appropriately by
regional population size and relative sampling intensity to arrive at a
putatively unbiased global frequency estimate. Phylogenetic analyses are
based on a representative sample of about 2000 viruses.

\begin{figure}[H]
  \centering
  \includegraphics[width=0.95\textwidth]{../figures/sep-2017/yam_counts.png}
  \caption{\textbf{Sample counts through time and across regions.}
  This is a stacked bar plot, so that in good months there are $\sim$200 total samples.
  }
  \label{Yam_counts}
\end{figure}

\clearpage
\begin{figure}[H]
  \centering
  \includegraphics[width=0.95\textwidth]{../figures/sep-2017/yam_mutations.png}
  \caption{\textbf{Frequency trajectories of B/Yam variants.}
  }
  \label{Yam_mutations}
\end{figure}
Clade 2 viruses appear to be extinct from the world. Clade 3 viruses
have continued to evolve and genetic diversity relative to vaccine
strain B/Phuket/3073/2013 has emerged. Of note the clade comprising
M251V dominates the population with over 99\% frequency, see \FIG{Yam_mutations}. The subclade
within M251V comprising K211R has decreased in frequency and is now
extinct or nearly extinct. The only notable subclade with the M251V
clade is that bearing T76I that has circulated at low frequency.

\begin{figure}[H]
  \centering
  \includegraphics[width=0.8\textwidth]{../figures/sep-2017/yam_tree_titer_model.png}
  \caption{\textbf{B/Yam phylogeny colored by antigenic distance from B/Phuket/3073/2013.} The tree model of antigenic distances infers no antigenic evolution.
  }
  \label{Yam_tree}
\end{figure}

Based on HI measurements provided by the Influenza Division at the US
CDC, we find no evidence for antigenic evolution within clade 3 viruses
and the vaccine B/Phuket/3073/2013 remains well matched to circulating
viruses, see \FIG{Yam_tree}.

\emph{Barring the emergence of a new selective variant we anticipate
relative stasis of B/Yam viruses in the coming year.}

\clearpage
\pagebreak

%%% References %%%
\bibliographystyle{plos}
\bibliography{references}

\end{document}

### 2017-09-18 auspice/reports/tex/sep-2017.tex
added clade labels to tex
Revision pass through H3N2

@@
 relevance. Please also note that observed patterns reflect the GISAID
-database and may not be entirely representative of underlying dynamics.}
+database and may not be entirely representative of underlying dynamics.
+All analyses are based on the nextflu pipeline \cite{neher2015nextflu}
+with continual updates posted to \href{http://nextflu.org}{nextflu.org}.}
 
@@
 2015 and Aug 2017, comprising \textgreater{}800 viruses per month in Dec
-2016 to Mar 2017 and 100-200 viruses per month more recently, see \FIG{H3N2_counts}. We use all
+2016 to Mar 2017 and 100-200 viruses per month more recently (\FIG{H3N2_counts}). We use all
 available data when estimating frequencies of mutations and weight
@@
   \caption{\textbf{Sample counts through time and across regions.}
-  This is a stacked bar plot, so that in all months there are at least 100 samples, in some month $>500$ sequences.
+  This is a stacked bar plot, so that in all months there are at least 100 samples and in some months there are $>500$ sequences.
   }
@@
   \includegraphics[width=0.9\textwidth]{../figures/sep-2017/h3n2_clades.png}
-  \caption{\textbf{Dynamics of major H3N2 clades.} Clade 3c3.a continues to circulate at low frequency, while 3c3.b has not been observed for several month (not shown).}
+  \caption{\textbf{Dynamics of major H3N2 clades.} Clade 3c2.a continues to dominate, clade 3c3.a continues to circulate at low frequency, and 3c3.b has not been observed for several months (not shown).}
   \label{H3N2_clades}
@@
 3c3.a persisted at about 20\% frequency in Oceania and North America in
-the 2016-2017 season, see \FIG{H3N2_clades}. Within 3c2.a, the clade 3c2.a1 comprising HA1
+the 2016-2017 season (\FIG{H3N2_clades}). Within 3c2.a, the clade 3c2.a1 comprising HA1
 mutation N171K and HA2 mutations I77V and G155E had risen rapidly in
@@
   \includegraphics[width=0.8\textwidth]{../figures/sep-2017/h3n2_tree.png}
-  \caption{\textbf{H3N2 / 3c2.a phylogeny colored by genotype.} Colors are chosen to highlight the clades 1--5. Clades are in reverse order, that is clade 5 is on top, clade 1 is at the bottom.
+  \caption{\textbf{H3N2 / 3c2.a phylogeny colored by genotype.} Colors are chosen to highlight clades 1--5.
+Clades are in reverse order, that is clade 5 is on top, clade 1 is at the bottom.
+Each of these clades is non-nested and so all compete against each other.
   }
@@
 past year. There are 5 major subclades within 3c2.a that have reached
-appreciable frequency. These clades are characterized by the following
+appreciable frequency (\FIG{H3N2_tree}). These clades are characterized by the following
 mutations and highlighed in the phylogenetic tree in \FIG{H3N2_tree}:
@@
 \begin{enumerate}
-\item
-  53N, 144R, 171K, 192T, 197H
-\item
-  121K, 144K
-\item
-  131K, 142K, 261Q
-\item
-  135K, HA2:150E
-\item
-  92R, 311Q
+\item  Clade 1: 53N, 144R, 171K, 192T, 197H
+\item  Clade 2: 121K, 144K
+\item  Clade 3: 131K, 142K, 261Q
+\item  Clade 4: 135K, HA2:150E
+\item  Clade 5: 92R, 311Q
 \end{enumerate}
@@
 All of these clades possess mutations at previously characterized
-epitope sites. We will refer to these clades as clade 1--5.
+epitope sites.
 
 All of these clades have risen in frequency and are now competing with
-one another, see \FIG{H3N2_mutations}. Clade 1 (53N, 144R, 171K, 192T, 197H) is highly derived in
+one another (\FIG{H3N2_mutations}). Clade 1 (53N, 144R, 171K, 192T, 197H) is highly derived in
 terms of sequence, but is at low frequency globally (currently less than
@@
   We estimate frequencies of different clades based on sample counts and collection dates.
+  Each clade is estimated according to a characteristic mutation unique to this clade.
+  These estimates are based on all available data and global frequencies are weighted according to regional population size and relative sampling intensity .
   We use a Brownian motion process prior to smooth frequencies from month-to-month.
@@
   \includegraphics[width=1.0\textwidth]{../figures/sep-2017/h3n2_tree_titer_model.png}
-  \caption{\textbf{H3N2 phylogeny colored by antigenic advance.}
-  Each panel shows model estimates of antigenic divergence relative to the root of the tree, which is similar to the titer drop relative to A/HongKong/5738/2014.
+  \caption{\textbf{H3N2 phylogeny colored by antigenic advancement.}
+  Model estimates of antigenic divergence relative to the root of the tree, which is similar to the titer drop relative to A/HongKong/4801/2014 or A/Michigan/15/2014.
   Cooler color indicates greater antigenic similarity (less titer drop going from homologous to heterologous titers).
@@
 All 5 of these clades appear somewhat drifted according to antigenic
-analysis Neher et al \cite{neher2015prediction} of HI measurements provided by the Influenza Division at the US CDC, see \FIG{H3N2_tree_titer_model}.
-With antigenic differences estimated between 0.5 units and 2.0 units.
+analysis \cite{neher2015prediction} of HI measurements provided by the Influenza Division at the US CDC (\FIG{H3N2_tree_titer_model}).
+Antigenic differences from vaccine strain A/HongKong/4801/2014 are estimated to be between 0.5 and 2.0 antigenic units (two-fold HI dilutions).
 
-The mutation T135K has come up in clade 2, 4, and 5 and is now globally
-at 40\% frequency. The most rapidly growing clade carries mutations
+The mutation T135K has appeared independently in clades 2, 4 and 5 and is now globally at 40\% frequency.
+The most rapidly growing clade carries mutations
 N121K/K92R/H311Q/E62G/R142G/T135K. This is a subclade of clade 5 above
@@
 estimates that this subclade of clade 5 has drifted by 2.2 antigenic
-units, 1 of which is associated with the E62G/R142G mutations and 1.2
-with the T135K mutation (on this background), see \FIG{H3N2_135K}.
+units, 1 of which is derived from the E62G/R142G mutations and 1.2 of which is derived from
+the T135K mutation (on this background), see \FIG{H3N2_135K}.
 Note, however, that there
@@
 
-Clades 1,2, and 3 are part of 3c2.a but outside of the 3c2.a1 subclade.
+Clades 1, 2 and 3 are part of 3c2.a but outside of the 3c2.a1 subclade.
+
+Clade 1 with substitutions N31S/D53N/S144R/N171K/I192T/Q197H was observed globally over the past 9 months but is now decreasing everywhere except Oceania.
+HI data and the model fit suggest a moderate (4-fold, about 2 antigenic units) titer drop relative to A/HongKong/4801/2014.
+Despite the titer drop, this clade seems to be outcompeted almost everywhere (with the exception of Oceania).
+It seems likely that this clade will die back with the ending of the Oceania influenza season.
+
 Clade 3 with T131K/R142K mutations continues to be common in China (but
 not in South Asia) with an embedded cluster of South and North American
-isolates from last summer. Subclusters within this clade have been
+isolates from last summer. Subclades within this clade have been
 expanding recently, but there is little evidence for antigenic evolution
@@
 
-Clade 1 with substitutions N31S/D53N/S144R/N171K/I192T/Q197H was
-observed globally over the past 9 month but is now decreasing everywhere
-except Oceania HI data and the model fit suggest a moderate (4-fold,
-about 2 antigenic units) titer drop relative to A/HongKong/4801/2014.
-Despite the titer drop, this calde seems to be outcompeted almost
-everywhere.
-
 The most notable subclade of 3c3.a is a clade carrying the mutation
@@
 decreasing in recent month and it doesn't show a consistent signal of
-antigenic change. The F193D mutations is observed sporadically within
+antigenic change. The F193D mutation is observed sporadically within
 the 3c2.a clade as well.
@@
 
-In summary, we think that the clade most likely to increase further in
-frequency is the E62G/R142G/T135K clade. However, two other clades
-(T131K/R142K and 144K) have similar potential. The former has very
-rapidly growing subclade, the latter rose rapidly last year and has a
-slightly higher LBI than the other two.
+\textit{In summary, we think that the clade most likely to increase further in frequency is the E62G/R142G/T135K subclade within clade 5.
+However, clade 2 viruses (characterized by N121K/R144K) and clade 3 viruses (characterized by T131K/R142K/R261Q) also appear competitive, although less so than E62G/R142G/T135K viruses.
+We expect all three of these clades to contribute to the future H3N2 population.}
 

### 2017-09-18 auspice/reports/tex/sep-2017.tex
Minor revisions

@@
 2015 and Aug 2017, comprising \textgreater{}800 viruses per month in Dec
-2016 to Mar 2017 and 100-200 viruses per month more recently (\FIG{H3N2_counts}). We use all
+2016 to Mar 2017 and 100-200 viruses per month more recently (\FIG{h3n2_counts}). We use all
 available data when estimating frequencies of mutations and weight
@@
   }
-  \label{H3N2_counts}
+  \label{h3n2_counts}
 \end{figure}
@@
   \caption{\textbf{Dynamics of major H3N2 clades.} Clade 3c2.a continues to dominate, clade 3c3.a continues to circulate at low frequency, and 3c3.b has not been observed for several months (not shown).}
-  \label{H3N2_clades}
+  \label{h3n2_clades}
 \end{figure}
@@
 3c3.a persisted at about 20\% frequency in Oceania and North America in
-the 2016-2017 season (\FIG{H3N2_clades}). Within 3c2.a, the clade 3c2.a1 comprising HA1
+the 2016-2017 season (\FIG{h3n2_clades}). Within 3c2.a, the clade 3c2.a1 comprising HA1
 mutation N171K and HA2 mutations I77V and G155E had risen rapidly in
@@
   }
-  \label{H3N2_tree}
+  \label{h3n2_tree}
 \end{figure}
@@
 past year. There are 5 major subclades within 3c2.a that have reached
-appreciable frequency (\FIG{H3N2_tree}). These clades are characterized by the following
-mutations and highlighed in the phylogenetic tree in \FIG{H3N2_tree}:
+appreciable frequency (\FIG{h3n2_tree}). These clades are characterized by the following
+mutations and highlighed in the phylogenetic tree in \FIG{h3n2_tree}:
 
@@
 All of these clades have risen in frequency and are now competing with
-one another (\FIG{H3N2_mutations}). Clade 1 (53N, 144R, 171K, 192T, 197H) is highly derived in
+one another (\FIG{h3n2_mutations}). Clade 1 (53N, 144R, 171K, 192T, 197H) is highly derived in
 terms of sequence, but is at low frequency globally (currently less than
@@
   }
-  \label{H3N2_mutations}
+  \label{h3n2_mutations}
 \end{figure}
@@
   }
-  \label{H3N2_tree_titer_model}
+  \label{h3n2_tree_titer_model}
 \end{figure}
@@
 All 5 of these clades appear somewhat drifted according to antigenic
-analysis \cite{neher2015prediction} of HI measurements provided by the Influenza Division at the US CDC (\FIG{H3N2_tree_titer_model}).
+analysis \cite{neher2015prediction} of HI measurements provided by the Influenza Division at the US CDC (\FIG{h3n2_tree_titer_model}).
 Antigenic differences from vaccine strain A/HongKong/4801/2014 are estimated to be between 0.5 and 2.0 antigenic units (two-fold HI dilutions).
@@
 units, 1 of which is derived from the E62G/R142G mutations and 1.2 of which is derived from
-the T135K mutation (on this background), see \FIG{H3N2_135K}.
+the T135K mutation (on this background), see \FIG{h3n2_tree_135k}.
 Note, however, that there
@@
   }
-  \label{H3N2_135K}
+  \label{h3n2_tree_135k}
 \end{figure}
@@
 2015 and Aug 2017, comprising approximately 100 viruses per month during
-relevant months of 2017, see \FIG{H1N1pdm_counts}. We use all available data when estimating
+relevant months of 2017 (\FIG{h1n1pdm_counts}). We use all available data when estimating
 frequencies of mutations and weight samples appropriately by regional
@@
   \includegraphics[width=0.95\textwidth]{../figures/sep-2017/h1n1pdm_counts.png}
-  \caption{\textbf{A/H1N1pdm sample counts through time and across regions.}}
-  \label{H1N1pdm_counts}
+  \caption{\textbf{A/H1N1pdm sample counts through time and across regions.}
+  This is a stacked bar plot, so that in many months there are more than 100 samples.
+  }
+  \label{h1n1pdm_counts}
 \end{figure}
@@
   \includegraphics[width=0.95\textwidth]{../figures/sep-2017/h1n1pdm_mutations.png}
-  \caption{\textbf{Frequency trajectories of recent mutations in A/H1N1pdm viruses.}  }
-  \label{H1N1pdm_clades}
+  \caption{\textbf{Frequency trajectories of recent mutations in A/H1N1pdm viruses.}
+  We estimate frequencies of different mutations based on sample counts and collection dates.
+  These estimates are based on all available data and global frequencies are weighted according to regional population size and relative sampling intensity .
+  We use a Brownian motion process prior to smooth frequencies from month-to-month.
+  Transparent bands show an estimate the 95\% confidence interval based on sample counts.
+  }
+  \label{h1n1pdm_mutations}
 \end{figure}
@@
 The S74R and I295V clade has risen rapidly in frequency starting around
-Jan 2017. This clade has been very successful and is now at
+Jan 2017 (\FIG{h1n1pdm_mutations}). This clade has been very successful and is now at
 approximately 92\% global frequency. Within this clade the subclade
 comprising S164T has risen in frequency from April 2017 and is now at
-approximately 35\% global frequency, see \FIG{H1N1pdm_clades}
+approximately 35\% global frequency.
 
@@
   \includegraphics[width=0.9\textwidth]{../figures/sep-2017/h1n1pdm_tree_titer_model.png}
-  \caption{\textbf{Antigenic distance from A/Michigan/45/2015.} No antigenic variation is picked up by the model.
+  \caption{\textbf{Antigenic distance from A/Michigan/45/2015.}
+  Model estimates of antigenic divergence relative to A/Michigan/45/2015.
+  No antigenic variation is picked up by the model.
   }
-  \label{H1N1pdm_mutations}
+  \label{h1n1pdm_tree_titer_model}
 \end{figure}
@@
 Division at the US CDC fails to show any distinction within recently
-circulating H1N1pdm viruses. Like the clades 6b and 6b.1 clade before
+circulating H1N1pdm viruses (\FIG{h1n1pdm_tree_titer_model}). Like the clades 6b and 6b.1 clade before
 it, the new S74R and I295V clade does not show any titer drop despite
@@
   }
-  \label{H1N1pdm_mutations}
+  \label{h1n1pdm_mutations_6b_6b1}
 \end{figure}
@@
 immunodominance effects, despite continued high titers in ferret sera
-(\href{http://www.pnas.org/content/111/44/15798}{Linderman et al 2014}).
+\cite{linderman2014potential}.
 Importantly, the new S74R and I295V clade has spread at a similarly
-rapid rate as did clades 6b and 6b.1. This rate of spread and
+rapid rate as previous clades 6b and 6b.1 (\FIG{h1n1pdm_mutations_6b_6b1}). This rate of spread and
 displacement of existing viral diversity is consistent with a selective
@@
 2015 and Aug 2017, comprising between 50 and 250 viruses per month
-during relevant months of 2017, see \FIG{Vic_counts}. We use all available data when
+during relevant months of 2017, see \FIG{vic_counts}. We use all available data when
 estimating frequencies of mutations and weight samples appropriately by
@@
   }
-  \label{Vic_counts}
+  \label{vic_counts}
 \end{figure}
@@
   \includegraphics[width=0.95\textwidth]{../figures/sep-2017/vic_tree.png}
-  \caption{\textbf{B/VIC -- Phylogenetic tree colored by genotype as positions 162, 175, 209.}
+  \caption{\textbf{B/Vic phylogeny colored by genotype at positions 162, 175 and 209.}
   }
-  \label{Vic_tree}
+  \label{vic_tree}
 \end{figure}
@@
 \textgreater{}99\% in the global population. Within this clade there is
-relatively little circulating genetic diversity. The only clades of note
+relatively little circulating genetic diversity (\FIG{vic_tree}). The only clades of note
 are a small clade with mutations I180V, D129G and deletions K162- and
@@
   }
-  \label{Vic_mutations}
+  \label{vic_mutations}
 \end{figure}
@@
 frequency globally with current global estimates at
-\textasciitilde{}4\%, see \FIG{Vic_mutations}. The clade bearing 87A and 175V has increased in
+$\sim$4\%, see \FIG{vic_mutations}. The clade bearing 87A and 175V has increased in
 frequency substantially in North America but remains low in the rest of
@@
   }
-  \label{Vic_titers}
+  \label{vic_tree_titer_model}
 \end{figure}
@@
 according to HI measurements provided by the Influenza Division at the
-US CDC, see \FIG{Vic_titers}. We estimate an approximately 8-fold titer drop (3 antigenic
+US CDC (\FIG{vic_tree_titer_model}). We estimate an approximately 8-fold titer drop (3 antigenic
 units) of drift from vaccine strain B/Brisbane/60/2008. Separately, a
@@
 \emph{Although the clade bearing 129G, 162--, 163-- and 180V is almost
-certainly antigenically drifted, it's slow spread up to the this point
+certainly antigenically drifted, its slow spread up to the this point
 suggests that it is unlikely to predominate in a year's time. We would
@@
 
-\textbf{A clade comprising M251V with clade 3 viruses continues to
+\textbf{A clade comprising M251V within clade 3 viruses continues to
 dominate. The is little genetic differentiation within this clade and no
@@
 2015 and Aug 2017, comprising between 50 and 250 viruses per month
-during relevant months of 2017, see \FIG{Yam_counts}. We use all available data when
+during relevant months of 2017 (\FIG{yam_counts}). We use all available data when
 estimating frequencies of mutations and weight samples appropriately by
@@
   }
-  \label{Yam_counts}
+  \label{yam_counts}
 \end{figure}
@@
   \caption{\textbf{Frequency trajectories of B/Yam variants.}
+  We estimate frequencies of different clades based on sample counts and collection dates.
+  We use a Brownian motion process prior to smooth frequencies from month-to-month.
+  Transparent bands show an estimate the 95\% confidence interval based on sample counts.
   }
-  \label{Yam_mutations}
+  \label{yam_mutations}
 \end{figure}
@@
 strain B/Phuket/3073/2013 has emerged. Of note the clade comprising
-M251V dominates the population with over 99\% frequency, see \FIG{Yam_mutations}. The subclade
+M251V dominates the population with over 99\% frequency (\FIG{yam_mutations}). The subclade
 within M251V comprising K211R has decreased in frequency and is now
-extinct or nearly extinct. The only notable subclade with the M251V
-clade is that bearing T76I that has circulated at low frequency.
+extinct or nearly extinct. The only notable subclade within the M251V
+clade is that bearing T76I which is circulating at low frequency.
 
@@
   }
-  \label{Yam_tree}
+  \label{yam_tree_titer_model}
 \end{figure}
@@
 and the vaccine B/Phuket/3073/2013 remains well matched to circulating
-viruses, see \FIG{Yam_tree}.
+viruses (\FIG{yam_tree_titer_model}).
 

### 2017-09-18 auspice/reports/plot_clades.py
added by region plot

@@ -66,7 +66,7 @@ offset = datetime(2000,1,1).toordinal()
 pivots = [offset+(x-2000)*365.25 for x in freqs['pivots']]
 meta = json.load(open(file_prefix+'_meta.json'))
 region_names = meta['geo']['region'].keys()
-region_codes = {'EU':['europe'], 'AS':['china', 'south_asia', 'japan_korea','southeast_asia'],
+region_codes = {'EU':['europe'], 'EUAS':['china', 'south_asia', 'japan_korea','southeast_asia'],
                 'NA':["north_america"], 'OC':["oceania"]}
 
 offset = datetime(2000,1,1).toordinal()
@@ -125,20 +125,21 @@ plt.subplots_adjust(left=0.1, right=0.82, top=0.94, bottom=0.22)
 sns.despine()
 plt.savefig('figures/' + report + '/'+virus+'_counts.png')
 
+##########################################################################################
 if len(clades):
     print "Plotting clade frequencies"
     fig, axs = plt.subplots(len(clades), 1, sharex=True, figsize=(8, len(clades)*2))
     for clade, ax in zip(clades, axs):
         for c,(region, r1) in zip(cols, [('north_america', 'NA'), ('china', 'AS'), ('europe','EU'), ('oceania','OC'), ('global', 'global')]):
-            if True: #try:
+            try:
                 tmp_freq = np.array(freqs['%s_%s'%(region, clade)])
                 if tmp_freq is not None:
                     std_dev = np.sqrt(tmp_freq*(1-tmp_freq)/(smoothed_count_by_region[r1]+1))
                     ax.plot(pivots, tmp_freq,'-o', label = region_label[r1], c=c, lw=3 if region=='global' else 1)
                     if show_errorbars:
                         ax.fill_between(pivots, tmp_freq-n_std_dev*std_dev, tmp_freq+n_std_dev*std_dev, facecolor=c, linewidth=0, alpha=0.1)
-            # except:
-            #     print "skipping", clade, region
+            except:
+                print "skipping", clade, region
         ax.set_xlim([pivots[0], pivots[-1]])
         ax.set_ylim(0,1)
         ax.text(pivots[0]+5, 0.88, clade)
[…]

### 2017-09-18 auspice/js/auspice/tooltips.js
Add nuc mutations to tooltip

@@ -143,7 +143,7 @@ var linkTooltip = d3.tip()
 		string += "<div class=\"smallnote\">";
 		var ncount = d.muts.length;
 		if (ncount) {string += "<b>Mutations:</b><ul>";}
-		if ((typeof d.aa_muts !="undefined")&&(mutType=='aa'||mutType=='both')){
+		if ((typeof d.aa_muts !="undefined")){
 			for (tmp_gene in d.aa_muts) {ncount+=d.aa_muts[tmp_gene].length;}
 			for (tmp_gene in d.aa_muts){
 				if (d.aa_muts[tmp_gene].length){
@@ -151,7 +151,7 @@ var linkTooltip = d3.tip()
 				}
 			}
 		}
-		if ((typeof d.muts !="undefined")&&(mutType=='nuc'||mutType=='both')&&(d.muts.length)){
+		if ((typeof d.muts !="undefined") && (d.muts.length)){
 			var nmuts = d.muts.length;
 			var tmp_muts = d.muts.slice(0,Math.min(10, nmuts))
 			string += "<li>nuc: "+tmp_muts.join(', ');

### 2017-09-19 auspice/reports/tex/sep-2017.tex
Add reference to Koel et al
Swap in new frequency figures for final GISAID data

@@
   \includegraphics[width=0.9\textwidth]{../figures/sep-2017/h3n2_clades.png}
-  \caption{\textbf{Dynamics of major H3N2 clades.} Clade 3c2.a continues to dominate, clade 3c3.a continues to circulate at low frequency, and 3c3.b has not been observed for several months (not shown).}
+  \caption{\textbf{Dynamics of major H3N2 clades.} Clade 3c2.a continues to dominate, clade 3c3.a continues to circulate at low frequency, and 3c3.b has not been observed for several months (not shown).
+Within 3c3.a the subclade 3c2.a1 has persisted at $\sim$50\% for the previous year.
+Viruses belonging to 3c2.a but not 3c2.a1 are also at $\sim$50\% frequency.}
   \label{h3n2_clades}
@@
 viruses. Clade 2 (121K, 144K) is beginning to decline in frequency and
-is now at \textasciitilde{}21\% global frequency. Clade 3 (131K, 142K,
+is now at $\sim$21\% global frequency. Clade 3 (131K, 142K,
 261Q) and clade 4 (135K, HA2:150E) have each slowly climbed to
@@
 selective origin.
+Previously, the mutation K135T was associated with the Sidney/97 antigenic transition as an ``accessory'' mutation \cite{koel2013substitutions}.
 
@@
   \includegraphics[width=0.8\textwidth]{../figures/sep-2017/h3n2_tree_135k.png}
-  \caption{\textbf{Zoom into clade 5, colored by antigenic advance.} A subclade with a T135K mutation has recently risen in frequency and has drifted $>$2 antigenic units.
+  \caption{\textbf{Zoom into subclade of 3c2.a1 characterized by mutation 121K comprising clades 4 and 5, colored by antigenic advance.} A subclade within clade 5 bearing a T135K mutation has recently risen in frequency and has drifted $>$2 antigenic units.
+  Parallel T135K mutations in clades 4 and 5 combine together to show an rapid recent increase in frequency.
   }

### 2017-09-19 auspice/reports/plot_clades.py

@@ -23,11 +23,11 @@ if virus=='h3n2':
     # mutations = ['HA1:121K', 'HA1:92R', 'HA1:131K', 'HA1:31S','HA1:198P', 'HA1:193S']
     # Large subclades
     # 1. 31S, 53N, 144R, 171K, 192T, 197H, characteristic: 197H
-    # 2. 121K, 144K, characteristic: 144K
+    # 2. 121K, 144K, characteristic: nuc:1320T
     # 3. 131K, 142K, 261Q, characteristic: 131K
     # 4. 142R, HA2:150E, characteristic: HA2:150E ---> additional 135K
-    # 5. 92R, 311Q, characteristic: 92R ---> additional 135K
-    mutations = ['HA1:197H', 'HA1:144K', 'HA1:131K', 'HA2:150E', 'HA1:92R']
+    # 5. 92R, 311Q, characteristic: nuc:538C ---> additional 135K
+    mutations = ['HA1:197H', 'nuc:1320T', 'HA1:131K', 'HA2:150E', 'nuc:538C']
     clade_legend = {'panel':0, 'loc':3}
     mut_legend = {'panel':0, 'loc':3}
     ymax = 1000
@@ -182,7 +182,7 @@ for ci, mutation, ax in zip(range(len(mutations)), mutations, axs):
                 ax.fill_between(pivots[drop:], (tmp_freq-n_std_dev*std_dev)[drop:], (tmp_freq+n_std_dev*std_dev)[drop:], facecolor=c, linewidth=0, alpha=0.1)
     ax.set_xlim([pivots[drop-1], pivots[-1]])
     ax.set_ylim(0,1)
-    ax.text(pivots[drop-1]+10, 0.88, ("clade %d"%(ci+1) if virus=='h3n2' else mutation), fontsize=fs*1.3)
+    ax.text(pivots[drop-1]+10, 0.88, ("clade %d"%(ci+1) if virus=='h3n2' else mutation), fontsize=fs*1.2)
     ax.set_yticklabels(['{:3.0f}%'.format(x*100) for x in [0, 0.2, 0.4, 0.6, 0.8, 1.0]])
     ax.tick_params(axis='x', which='major', labelsize=fs, pad=20)
     ax.tick_params(axis='x', which='minor', pad=7)

### 2017-09-19 auspice/reports/tex/sep-2017.tex
More revisions

@@
 1\%) despite making up the a substantial fraction of 2017 Oceania
-viruses. Clade 2 (121K, 144K) is beginning to decline in frequency and
-is now at $\sim$21\% global frequency. Clade 3 (131K, 142K,
+viruses. Clade 2 (121K, 144K) had a large upswing around Dec 2016, but has since declined slightly to $\sim$21\% global frequency. Clade 3 (131K, 142K,
 261Q) and clade 4 (135K, HA2:150E) have each slowly climbed to
@@
 decreasing in recent month and it doesn't show a consistent signal of
-antigenic change. The F193D mutation is observed sporadically within
+antigenic change. The F193S mutation is observed sporadically within
 the 3c2.a clade as well.
 
-% \begin{figure}[h!]
-%   \centering
-%   \includegraphics[width=1.0\textwidth]{../figures/sep-2017/h3n2_3c2a.png}
-%   \caption{\textbf{Recent frequency changes.} The tree, zoomed into clade 3c2.a, is colored by change in clade frequency over the past 4 month.
-%   }
-%   \label{H3N2_135K}
-% \end{figure}
+Measuring clade frequency growth and decline over the previous 6 months shows that subclades within clades 3 and 5 are growing the most rapidly (\FIG{h3n2_tree_dfreq}).
+The most rapidly growing subclade is the E62G/R142G/T135K subclade within clade 5.
+This analysis is borne out in "local branching index" \cite{neher2014predicting}, which also supports this subclade as a high fitness clade.
+
+\begin{figure}[h!]
+  \centering
+  \includegraphics[width=0.8\textwidth]{../figures/sep-2017/h3n2_tree_dfreq.png}
+	\caption{\textbf{H3N2 / 3c2.a phylogeny colored by clade growth.}
+  The tree, zoomed into clade 3c2.a, is colored by change in clade frequency over the past 6 months.
+  }
+  \label{h3n2_tree_dfreq}
+\end{figure}
 
@@
 
-The S74R and I295V clade has risen rapidly in frequency starting around
-Jan 2017 (\FIG{h1n1pdm_mutations}). This clade has been very successful and is now at
-approximately 92\% global frequency. Within this clade the subclade
-comprising S164T has risen in frequency from April 2017 and is now at
-approximately 35\% global frequency.
+The S74R and I295V clade has risen rapidly in frequency starting around Jan 2017 (\FIG{h1n1pdm_mutations}).
+This clade has been very successful and is now at approximately 66\% global frequency.
+Notably, this clade is at \textgreater{}90\% frequency in North America, Europe and Oceania but is at lower frequency in Asia.
+Within this clade the subclade comprising S164T has risen in frequency from April 2017 and is now at approximately 32\% global frequency.
 
@@
 \cite{linderman2014potential}.
-Importantly, the new S74R and I295V clade has spread at a similarly
-rapid rate as previous clades 6b and 6b.1 (\FIG{h1n1pdm_mutations_6b_6b1}). This rate of spread and
+Importantly, the new S74R and I295V clade has spread at a similar
+rate as previous clades 6b and 6b.1 (\FIG{h1n1pdm_mutations_6b_6b1}). This rate of spread and
 displacement of existing viral diversity is consistent with a selective
@@
 
-\emph{Given current patterns of growth and decline it is almost certain
+\emph{Given current patterns of growth and decline it is highly likely
 that the S74R and I295V clade will dominant in a years time. Although
 there is no antigenic effect visible via HI from ferret antisera, the
-rapidity of the rise of S74R and I295V suggests a selective origin. We
+rapidity of the rise of S74R and I295V suggests a selective origin. The S164T subclade also appears to be selectively driven. We
 would suggest a detailed examination of human serological data as it's

### 2017-09-19 auspice/_includes/2y_meta.js
tweaked LBI params

@@ -2,5 +2,5 @@ var time_window = 2.0;
 var full_data_time_window = 2.0;
 var time_ticks=[2014.0, 2014.5, 2015.0, 2015.5, 2016.0, 2016.5, 2017.0, 2017.5];
 var dfreq_dn = 6;
-var LBItime_window = 0.75;
-var LBItau = 0.25;
+var LBItime_window = 0.5;
+var LBItau = 0.3;

### 2017-09-19 auspice/_includes/3y_meta.js

@@ -2,5 +2,5 @@ var time_window = 3.0;
 var full_data_time_window = 3.0;
 var time_ticks=[2014.0, 2014.5, 2015.0, 2015.5, 2016.0, 2016.5, 2017.0, 2017.5];
 var dfreq_dn = 6;
-var LBItime_window = 0.75;
-var LBItau = 0.25;
+var LBItime_window = 0.6;
+var LBItau = 0.4;

### 2017-09-19 auspice/js/auspice/colors.js

@@ -30,7 +30,8 @@ var receptorBindingColorScale = d3.scale.linear().clamp([true])
 	.range(colors[4]);
 
 var lbiColorScale = d3.scale.linear()
-	.domain([0.0, 0.02, 0.04, 0.07, 0.1, 0.2, 0.4, 0.7, 0.9, 1.0])
+//	.domain([0.0, 0.02, 0.04, 0.07, 0.1, 0.2, 0.4, 0.7, 0.9, 1.0])
+	.domain([0.0, 0.03, 0.06, 0.12, 0.20, 0.3, 0.4, 0.5, 0.6, 0.7])
 	.range(colors[10]);
 
 var dfreqColorScale = d3.scale.linear()

### 2017-09-19 auspice/reports/tex/sep-2017.tex
Revise report

@@
   \centering
-  \includegraphics[width=0.8\textwidth]{../figures/sep-2017/h3n2_tree.png}
+  \includegraphics[width=0.8\textwidth]{../figures/sep-2017/h3n2_tree_with_timetree.png}
   \caption{\textbf{H3N2 / 3c2.a phylogeny colored by genotype.} Colors are chosen to highlight clades 1--5.
@@
 Each of these clades is non-nested and so all compete against each other.
+Also shown is a time calibrated phylogeny with equivalent coloring.
   }
-  \label{h3n2_tree}
+  \label{h3n2_tree_with_timetree}
 \end{figure}
@@
 past year. There are 5 major subclades within 3c2.a that have reached
-appreciable frequency (\FIG{h3n2_tree}). These clades are characterized by the following
+appreciable frequency (\FIG{h3n2_tree_with_timetree}). These clades are characterized by the following
 mutations and highlighed in the phylogenetic tree in \FIG{h3n2_tree}:
@@
 All of these clades have risen in frequency and are now competing with
-one another (\FIG{h3n2_mutations}). Clade 1 (53N, 144R, 171K, 192T, 197H) is highly derived in
+one another (\FIG{h3n2_mutations}, \FIG{h3n2_mutations_by_region}). Clade 1 (53N, 144R, 171K, 192T, 197H) is highly derived in
 terms of sequence, but is at low frequency globally (currently less than
@@
   \includegraphics[width=0.95\textwidth]{../figures/sep-2017/h3n2_mutations.png}
-  \caption{\textbf{Frequency trajectories of 3c2.a subclades.}
+  \caption{\textbf{Frequency trajectories of 3c2.a subclades partitioned by clade then by region.}
   We estimate frequencies of different clades based on sample counts and collection dates.
@@
   \label{h3n2_mutations}
+\end{figure}
+
+\begin{figure}[h!]
+  \centering
+  \includegraphics[width=0.95\textwidth]{../figures/sep-2017/h3n2_mutations_by_region.png}
+  \caption{\textbf{Frequency trajectories of 3c2.a subclades partitioned by region then by clade.}
+  We estimate frequencies of different clades based on sample counts and collection dates.
+  Each clade is estimated according to a characteristic mutation unique to this clade.
+  These estimates are based on all available data and global frequencies are weighted according to regional population size and relative sampling intensity .
+  We use a Brownian motion process prior to smooth frequencies from month-to-month.
+  Transparent bands show an estimate the 95\% confidence interval based on sample counts.
+  }
+  \label{h3n2_mutations_by_region}
 \end{figure}
@@
 
-Measuring clade frequency growth and decline over the previous 6 months shows that subclades within clades 3 and 5 are growing the most rapidly (\FIG{h3n2_tree_dfreq}).
-The most rapidly growing subclade is the E62G/R142G/T135K subclade within clade 5.
-This analysis is borne out in "local branching index" \cite{neher2014predicting}, which also supports this subclade as a high fitness clade.
+Additionally, an analysis of ``local branching index'' \cite{neher2014predicting} suggests that the E62G/R142G/T135K subclade within clade 5 is the most rapidly growing of circulating viruses and is thus estimated to have the highest fitness.
+With the exception of background 3c2.a1 viruses, other circulating 3c2.a are seen to have similar local branching index values.
 
@@
   \centering
-  \includegraphics[width=0.8\textwidth]{../figures/sep-2017/h3n2_tree_dfreq.png}
-	\caption{\textbf{H3N2 / 3c2.a phylogeny colored by clade growth.}
-  The tree, zoomed into clade 3c2.a, is colored by change in clade frequency over the past 6 months.
+  \includegraphics[width=0.8\textwidth]{../figures/sep-2017/h3n2_tree_lbi.png}
+	\caption{\textbf{H3N2 / 3c2.a phylogeny colored by local branching index.}
+  The tree, zoomed into clade 3c2.a, is colored by local branching index.
   }
@@
 \textit{In summary, we think that the clade most likely to increase further in frequency is the E62G/R142G/T135K subclade within clade 5.
-However, clade 2 viruses (characterized by N121K/R144K) and clade 3 viruses (characterized by T131K/R142K/R261Q) also appear competitive, although less so than E62G/R142G/T135K viruses.
+However, clade 2 viruses (characterized by N121K/R144K) and clade 3 viruses (characterized by T131K/R142K/R261Q) also appear competitive (they are present at moderate frequency, have somewhat diverged antigenic phenotypes and are increasing in frequency), although less so than E62G/R142G/T135K viruses (which are growing faster and appear more diverged in antigenic phenotype).
 We expect all three of these clades to contribute to the future H3N2 population.}

### 2017-09-19 auspice/reports/tex/sep-2017.tex
Textual changes

@@
 which carry several amino acid mutations at previously characterized
-epitopes. The majority of viruses fall into the 3c2.a clade which has
+epitope sites. The majority of viruses fall into the 3c2.a clade which has
 been dominating globally for \textgreater{}3 years, but 3c3.a viruses
@@
 past year. There are 5 major subclades within 3c2.a that have reached
-appreciable frequency (\FIG{h3n2_tree_with_timetree}). These clades are characterized by the following
-mutations and highlighed in the phylogenetic tree in \FIG{h3n2_tree}:
+appreciable frequency (\FIG{h3n2_tree_with_timetree}). All of these clades possess mutations at previously characterized
+epitope sites. These clades are characterized by the following
+mutations and highlighed in the phylogenetic tree in \FIG{h3n2_tree_with_timetree}:
 
@@
 
-All of these clades possess mutations at previously characterized
-epitope sites.
-
 All of these clades have risen in frequency and are now competing with
@@
 1\%) despite making up the a substantial fraction of 2017 Oceania
-viruses. Clade 2 (121K, 144K) had a large upswing around Dec 2016, but has since declined slightly to $\sim$21\% global frequency. Clade 3 (131K, 142K,
-261Q) and clade 4 (135K, HA2:150E) have each slowly climbed to
-approximately 20\% frequency. However, clade 5 (92R, 311Q) has recently
-increased to 39\% in global frequency.
+viruses.
+Clade 2 (121K, 144K) had a large upswing around Dec 2016, but has since declined slightly to $\sim$27\% global frequency, but is now increasing again.
+Clade 3 (131K, 142K, 261Q) has steadily climbed to $\sim$39\% global frequency and importantly is at relatively high frequency in Asia.
+Clade 4 (135K, HA2:150E) has persisted at low frequency and is currently at $\sim$8\% globally.
+Clade 5 (92R, 311Q) has recently increased to $\sim$25\% in global frequency, also with relatively high frequency in Asia.
 
@@
 analysis \cite{neher2015prediction} of HI measurements provided by the Influenza Division at the US CDC (\FIG{h3n2_tree_titer_model}).
-Antigenic differences from vaccine strain A/HongKong/4801/2014 are estimated to be between 0.5 and 2.0 antigenic units (two-fold HI dilutions).
+Antigenic differences from vaccine strain A/HongKong/4801/2014 are estimated to be between 0.5 and 2.2 antigenic units (or up to a four-fold HI dilution).
 
-The mutation T135K has appeared independently in clades 2, 4 and 5 and is now globally at 40\% frequency.
+The mutation T135K has appeared independently in clades 2, 4 and 5 and is now globally at 31\% frequency.
 The most rapidly growing clade carries mutations
@@
 isolates from last summer. Subclades within this clade have been
-expanding recently, but there is little evidence for antigenic evolution
-in the HI or FRA data. The 131K substitution is not seen outside this
+expanding recently, and there is evidence for a small amount of antigenic evolution
+in the HI (0.9 antigenic units) and FRA data (0.1 antigenic units). The 131K substitution is not seen outside this
 clade.
@@
 Additionally, an analysis of ``local branching index'' \cite{neher2014predicting} suggests that the E62G/R142G/T135K subclade within clade 5 is the most rapidly growing of circulating viruses and is thus estimated to have the highest fitness.
-With the exception of background 3c2.a1 viruses, other circulating 3c2.a are seen to have similar local branching index values.
+With the exception of background 3c2.a1 viruses, other circulating 3c2.a viruses are seen to have similar local branching index values.
 
@@
 \textit{In summary, we think that the clade most likely to increase further in frequency is the E62G/R142G/T135K subclade within clade 5.
-However, clade 2 viruses (characterized by N121K/R144K) and clade 3 viruses (characterized by T131K/R142K/R261Q) also appear competitive (they are present at moderate frequency, have somewhat diverged antigenic phenotypes and are increasing in frequency), although less so than E62G/R142G/T135K viruses (which are growing faster and appear more diverged in antigenic phenotype).
+However, clade 2 viruses (characterized by N121K/R144K) and clade 3 viruses (characterized by T131K/R142K/R261Q) also appear competitive (they are present at moderate frequency, have slightly diverged antigenic phenotypes and are increasing in frequency), although less so than E62G/R142G/T135K viruses (which are growing faster and appear more diverged in antigenic phenotype).
 We expect all three of these clades to contribute to the future H3N2 population.}

### 2017-09-20 auspice/js/auspice/tree.js
Improve layout of tree control stack

@@ -704,7 +704,7 @@ d3.json(path + file_prefix + "tree.json", function(error, root) {
 	var mc = autocomplete(document.getElementById('straininput'))
 		.keys(tips)
 		.dataField("strain")
-		.placeHolder("search strains...")
+		.placeHolder("strain name...")
 		.width(800)
 		.height(500)
 		.onSelected(highlightStrainSearch)

### 2017-09-20 auspice/reports/tex/feb-2017.tex
Fix figure references in Feb 2017 report

@@
 	\centering
-	\includegraphics[width=0.75\textwidth]{../figures/feb-2017/h3n2_frequencies.pdf}
+	\includegraphics[width=0.75\textwidth]{../figures/feb-2017/h3n2_frequencies.png}
 	\caption{\textbf{Frequency trajectories of 3c2.a subclades.}
@@
 	\centering
-	\includegraphics[width=0.9\textwidth]{../figures/feb-2017/h1n1pdm_frequencies.pdf}
+	\includegraphics[width=0.9\textwidth]{../figures/feb-2017/h1n1pdm_frequencies.png}
 	\caption{\textbf{Frequency trajectories of 6b.1 subclades.}
@@
 	\centering
-	\includegraphics[width=0.95\textwidth]{../figures/feb-2017/vic_frequencies.pdf}
+	\includegraphics[width=0.95\textwidth]{../figures/feb-2017/vic_frequencies.png}
 	\caption{\textbf{Frequency trajectories of B/Vic variants.}
@@
 	\centering
-	\includegraphics[width=0.95\textwidth]{../figures/feb-2017/yam_frequencies.pdf}
+	\includegraphics[width=0.95\textwidth]{../figures/feb-2017/yam_frequencies.png}
 	\caption{\textbf{Frequency trajectories of B/Yam variants.}
